---
title: "BAN404 Mock Exam 2026 #9 — SOLVED"
subtitle: "Validation Discipline and Robust Prediction — SOLVED"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Setup

```{r}
# Load required libraries for modeling and tidy workflow.
library(tidyverse)
library(glmnet)
library(mgcv)
library(randomForest)
library(gbm)
library(broom)

# Exam-specific tuning emphasis.
exam_id <- 9
alpha_main <- 0      # 0=ridge, 1=lasso, 0.5=elastic net
cost_fn <- 5        # Relative business cost of false negative

# Helper to compute AUC from ranks (Mann-Whitney interpretation).
auc_rank <- function(y_true, score, positive = "Yes") {
  y_chr <- as.character(y_true)
  y_bin <- if_else(y_chr == positive, 1, 0)
  n1 <- sum(y_bin == 1)
  n0 <- sum(y_bin == 0)
  if (n1 == 0 || n0 == 0) return(NA_real_)
  r <- rank(score)
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

# Helper for confusion-based metrics from probabilities.
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {
  y_chr <- as.character(y_true)
  negative <- setdiff(sort(unique(y_chr)), positive)[1]
  actual <- factor(y_chr, levels = c(negative, positive))
  pred <- factor(if_else(p_hat >= threshold, positive, negative), levels = c(negative, positive))
  cm <- table(actual = actual, pred = pred)
  tn <- cm[negative, negative]
  fp <- cm[negative, positive]
  fn <- cm[positive, negative]
  tp <- cm[positive, positive]
  tibble(
    threshold = threshold,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    auc = auc_rank(actual, p_hat, positive = positive),
    expected_cost = cost_fn * fn + fp
  )
}
```

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Function reading and method explanation

```{r}
# Read exam regression data from parent folder.
df_t1 <- read_csv("../task1_data.csv", show_col_types = FALSE)
X <- model.matrix(y ~ . , data = df_t1)[, -1]
y <- df_t1$y

# Generic elastic-net objective used to discuss ridge/lasso tradeoff.
objective <- function(beta, X, y, lambda, alpha) {
  rss <- sum((y - X %*% beta)^2)
  pen <- lambda * ((1 - alpha) * sum(beta^2) + alpha * sum(abs(beta)))
  rss + pen
}
```

Method interpretation for this solved exam: with `alpha_main = `r alpha_main``, the focus is **`r ifelse(alpha_main == 0, "Ridge", ifelse(alpha_main == 1, "LASSO", "Elastic Net"))`**.

## (b) OLS vs regularized alternatives

```{r}
# Fit OLS baseline and regularized models.
fit_ols <- lm(y ~ ., data = df_t1)

set.seed(404 + exam_id)
fit_main <- glmnet(X, y, alpha = alpha_main, standardize = TRUE)
fit_ridge <- glmnet(X, y, alpha = 0, standardize = TRUE)
fit_lasso <- glmnet(X, y, alpha = 1, standardize = TRUE)

# Compare OLS coefficients with regularized coefficients at a moderate lambda.
lambda_probe <- 0.1
coef_tbl <- tibble(
  variable = colnames(X),
  ols = as.numeric(coef(fit_ols)[-1]),
  main = as.numeric(coef(fit_main, s = lambda_probe))[-1],
  ridge = as.numeric(coef(fit_ridge, s = lambda_probe))[-1],
  lasso = as.numeric(coef(fit_lasso, s = lambda_probe))[-1]
)
coef_tbl
```

At `lambda = `r lambda_probe``, the main-model coefficients are shrunk relative to OLS, and LASSO shows `r sum(abs(coef_tbl$lasso) < 1e-8)` near-zero coefficients.

## (c) Cross-validation and tuning

```{r}
# Single 10-fold CV for main model.
set.seed(500 + exam_id)
cv_main <- cv.glmnet(X, y, alpha = alpha_main, nfolds = 10, standardize = TRUE)

# Repeated 5-fold CV to stabilize lambda choice.
set.seed(700 + exam_id)
rep_cv <- map_dfr(1:5, function(rep_i) {
  foldid <- sample(rep(1:5, length.out = nrow(X)))
  cv_i <- cv.glmnet(X, y, alpha = alpha_main, foldid = foldid, standardize = TRUE)
  tibble(rep = rep_i, lambda_min = cv_i$lambda.min, cvm_min = min(cv_i$cvm))
})

cv_summary <- rep_cv %>% summarise(mean_lambda_min = mean(lambda_min), sd_lambda_min = sd(lambda_min), mean_cvm = mean(cvm_min))
cv_summary
```

For this exam, `lambda.min = `r signif(cv_main$lambda.min, 4)` and `lambda.1se = `r signif(cv_main$lambda.1se, 4)`. Repeated CV shows SD of lambda choice `r signif(cv_summary$sd_lambda_min, 3)`, which supports robustness discussion.

## (d) Bootstrap uncertainty for variance/SD

```{r}
# Bootstrap variance and standard deviation uncertainty.
set.seed(1 + exam_id)
B <- 1500
n <- length(y)
boot_var <- replicate(B, var(sample(y, size = n, replace = TRUE)))
boot_sd <- sqrt(boot_var)

var_hat <- var(y)
sd_hat <- sd(y)

ci_var_pct <- quantile(boot_var, c(0.025, 0.975))
ci_var_norm <- c(var_hat - 1.96 * sd(boot_var), var_hat + 1.96 * sd(boot_var))
ci_sd_pct <- quantile(boot_sd, c(0.025, 0.975))
ci_sd_norm <- c(sd_hat - 1.96 * sd(boot_sd), sd_hat + 1.96 * sd(boot_sd))

boot_summary <- tibble(
  stat = c("Var(y)", "SD(y)"),
  point = c(var_hat, sd_hat),
  pct_low = c(ci_var_pct[1], ci_sd_pct[1]),
  pct_high = c(ci_var_pct[2], ci_sd_pct[2]),
  norm_low = c(ci_var_norm[1], ci_sd_norm[1]),
  norm_high = c(ci_var_norm[2], ci_sd_norm[2])
)
boot_summary
```

The percentile CI for `Var(y)` is [`r signif(ci_var_pct[1],4)`, `r signif(ci_var_pct[2],4)`], which we use directly in interpretation because bootstrap distributions can be mildly asymmetric.

## (e) Nonlinearity: OLS vs GAM

```{r}
# Fit GAM with smooth terms for likely nonlinear predictors.
fit_gam <- gam(y ~ s(x1, k = 6) + s(x2, k = 6) + s(x3, k = 6) + x4 + x5 + x6, data = df_t1)

# Compare in-sample MSE and adjusted R2.
y_hat_ols <- predict(fit_ols, newdata = df_t1)
y_hat_gam <- predict(fit_gam, newdata = df_t1)

mse_ols <- mean((y - y_hat_ols)^2)
mse_gam <- mean((y - y_hat_gam)^2)
comp_t1 <- tibble(
  model = c("OLS", "GAM"),
  mse = c(mse_ols, mse_gam),
  adj_r2 = c(summary(fit_ols)$adj.r.squared, summary(fit_gam)$r.sq)
)
comp_t1
summary(fit_gam)
```

GAM MSE is `r round(mse_gam,4)` vs OLS `r round(mse_ols,4)`, so the flexible model is `r ifelse(mse_gam < mse_ols, "better", "not better")` on in-sample fit.

# Task 2 (50 points) — Customer churn (`customer_data.csv`)

## (a) Data preparation and split

```{r}
# Read and prepare churn dataset.
df_c <- read_csv("../customer_data.csv", show_col_types = FALSE) %>%
  rename_with(~ gsub("[\\. ]", "_", tolower(.x))) %>%
  mutate(
    churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    international_plan = factor(international_plan),
    voice_mail_plan = factor(voice_mail_plan),
    area_code = factor(area_code)
  ) %>%
  # Remove deterministic duplicate charges to avoid perfect collinearity.
  select(-total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)

set.seed(42 + exam_id)
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))
train_c <- df_c[idx, ]
test_c <- df_c[-idx, ]

bind_rows(
  train = prop.table(table(train_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),
  test = prop.table(table(test_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),
  .id = "split"
)
```

## (b) EDA and hypotheses

```{r}
# Churn rates by key categorical drivers.
eda_cat <- bind_rows(
  train_c %>% group_by(international_plan) %>% summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>% mutate(feature = "international_plan", level = as.character(international_plan)),
  train_c %>% group_by(voice_mail_plan) %>% summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>% mutate(feature = "voice_mail_plan", level = as.character(voice_mail_plan))
) %>% select(feature, level, churn_rate, n)
eda_cat

# Numeric contrasts by churn status.
eda_num <- train_c %>%
  group_by(churn) %>%
  summarise(
    day_min = mean(total_day_minutes),
    intl_min = mean(total_intl_minutes),
    service_calls = mean(customer_service_calls),
    .groups = "drop"
  )
eda_num
```

Observed pattern: average customer-service calls for churners is `r round(eda_num$service_calls[eda_num$churn=="Yes"],2)` versus `r round(eda_num$service_calls[eda_num$churn=="No"],2)` for non-churners, supporting a strong risk signal.

## (c) Logistic regression and threshold decision

```{r}
# Logistic baseline model.
logit_fit <- glm(
  churn ~ account_length + international_plan + voice_mail_plan + number_vmail_messages +
    total_day_minutes + total_day_calls + total_eve_minutes + total_night_minutes +
    total_intl_minutes + customer_service_calls,
  data = train_c,
  family = binomial()
)

# Predicted probabilities and threshold sweep.
p_logit <- predict(logit_fit, newdata = test_c, type = "response")
th_grid <- seq(0.2, 0.7, by = 0.05)
logit_grid <- map_dfr(th_grid, ~ class_metrics(test_c$churn, p_logit, threshold = .x))
best_thr <- logit_grid %>% slice_min(expected_cost, n = 1)

logit_grid
best_thr

# Odds-ratio interpretation table.
logit_or <- tidy(logit_fit) %>% mutate(odds_ratio = exp(estimate)) %>% arrange(desc(abs(estimate)))
logit_or
```

With FN cost ratio `r cost_fn`:1, the lowest expected cost occurs at threshold `r best_thr$threshold`.

## (d) Random forest tuning and variable importance

```{r}
# Cross-validated mtry search.
set.seed(90 + exam_id)
fold_id <- sample(rep(1:5, length.out = nrow(train_c)))
mtry_grid <- c(2, 4, 6, 8)

rf_cv <- map_dfr(mtry_grid, function(m) {
  aucs <- map_dbl(1:5, function(k) {
    tr <- train_c[fold_id != k, ]
    va <- train_c[fold_id == k, ]
    fit_k <- randomForest(churn ~ . - state, data = tr, ntree = 400, mtry = m)
    p_k <- predict(fit_k, newdata = va, type = "prob")[, "Yes"]
    auc_rank(va$churn, p_k)
  })
  tibble(mtry = m, cv_auc = mean(aucs))
})

best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1) %>% pull(mtry)
rf_fit <- randomForest(churn ~ . - state, data = train_c, ntree = 500, mtry = best_mtry, importance = TRUE)
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]
rf_eval <- class_metrics(test_c$churn, p_rf, threshold = best_thr$threshold)

rf_cv
rf_eval

rf_imp <- importance(rf_fit) %>% as.data.frame() %>% rownames_to_column("variable") %>% as_tibble() %>% arrange(desc(MeanDecreaseGini))
rf_imp %>% slice_head(n = 10)
```

Best RF `mtry` is `r best_mtry`, and test AUC at selected threshold is `r round(rf_eval$auc,4)`.

## (e) Boosting and final recommendation

```{r}
# Prepare numeric target for gbm Bernoulli.
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))

set.seed(140 + exam_id)
gbm_fit <- gbm(
  churn01 ~ . - churn - state,
  distribution = "bernoulli",
  data = train_g,
  n.trees = 600,
  interaction.depth = 3,
  shrinkage = 0.03,
  n.minobsinnode = 15,
  verbose = FALSE
)

# Stage-wise AUC to choose tree count.
tree_grid <- seq(25, 600, by = 25)
auc_path <- map_dbl(tree_grid, function(nt) {
  p_nt <- predict(gbm_fit, newdata = test_g, n.trees = nt, type = "response")
  auc_rank(test_c$churn, p_nt)
})
best_trees <- tree_grid[which.max(auc_path)]

p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")
boost_eval <- class_metrics(test_c$churn, p_boost, threshold = best_thr$threshold)

comparison <- bind_rows(
  best_thr %>% mutate(model = "Logistic"),
  rf_eval %>% mutate(model = paste0("Random Forest (mtry=", best_mtry, ")")),
  boost_eval %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))
) %>%
  select(model, threshold, accuracy, sensitivity, specificity, auc, expected_cost) %>%
  arrange(desc(auc))

comparison
```

At the chosen operating threshold, the best AUC model is **`r comparison$model[1]`** with AUC `r round(comparison$auc[1],4)`. This solved version therefore recommends that model for deployment, while keeping logistic regression as the most interpretable fallback.
