---
title: "BAN404 – Predicted Exam 2026 (R Version)"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

**Date:** 19 May 2026  
**Duration:** 6 hours  
**Support materials:** Calculator. All written material (Category III).  
**Attachments:** `task1_data.csv` · `customer_data.csv`

---

# Task 1 (50 points) — Methodological Topics

## (a) — Code reading (10 points)

Consider the following two R functions:

```{r}
library(tidyverse)

f <- function(beta, X, y, lambda) {
  rss <- sum((y - X %*% beta)^2)
  penalty <- lambda * sum(abs(beta))  # (*)
  rss + penalty
}

g <- function(X, y, lambda) {
  p <- ncol(X)
  b0 <- mean(y)                        # (**)
  y_c <- y - b0
  X_c <- sweep(X, 2, colMeans(X), "-") # (***)

  fit <- optim(
    par = rep(0, p),
    fn = f,
    X = X_c,
    y = y_c,
    lambda = lambda,
    method = "BFGS"
  )

  list(intercept = b0, beta = fit$par)
}
```

**i.** Explain what `f` and `g` do, step by step. Which BAN404 method is this?  
*Your answer here.*

**ii.** Explain lines `(*)`, `(**)`, and `(***)`.  
*Your answer here.*

**iii.** What happens to `beta` when `lambda` increases? What is the key difference from Ridge?  
*Your answer here.*

---

## (b) — Applying the method (10 points)

Use `task1_data.csv` and fit:

- OLS (`lm`)
- `g` with `lambda = 1`
- `g` with `lambda = 100`

Present coefficients in one table and comment on shrinkage and variable selection.

```{r}
xy <- readr::read_csv("task1_data.csv", show_col_types = FALSE)
X <- as.matrix(dplyr::select(xy, x1:x6))
y <- xy$y

# Your code here
```

*Your answer here.*

---

## (c) — Cross-validation (10 points)

Explain and use this function to pick an optimal `lambda` by 5-fold CV.

```{r}
cv_lasso_manual <- function(X, y, lambda, K = 5, seed = 42) {
  set.seed(seed)
  n <- length(y)
  fold_id <- sample(rep(1:K, length.out = n))
  mse <- numeric(K)

  for (k in 1:K) {
    test_idx <- which(fold_id == k)
    train_idx <- which(fold_id != k)

    fit <- g(X[train_idx, , drop = FALSE], y[train_idx], lambda)

    X_test_centered <- sweep(
      X[test_idx, , drop = FALSE],
      2,
      colMeans(X[train_idx, , drop = FALSE]),
      "-"
    )

    y_hat <- fit$intercept + X_test_centered %*% fit$beta
    mse[k] <- mean((y[test_idx] - as.numeric(y_hat))^2)
  }

  mean(mse)
}

lambda_grid <- 10^seq(-1, 3, length.out = 40)
# Your code here
```

Why use training means when centering test folds? Why prefer K-fold to LOOCV here?  
*Your answer here.*

---

## (d) — Bootstrap (10 points)

Explain and apply this bootstrap variance estimator.

```{r}
bootstrap_var <- function(y, B = 1000, seed = 42) {
  set.seed(seed)
  n <- length(y)
  boot <- numeric(B)

  for (b in 1:B) {
    idx <- sample(1:n, size = n, replace = TRUE)  # (*)
    y_b <- y[idx]                                   # (**)
    boot[b] <- var(y_b)                             # (***)
  }

  se <- sd(boot)
  ci <- quantile(boot, probs = c(0.025, 0.975))
  list(boot = boot, se = se, ci = ci)
}

# Your code here
```

*Your answer here.*

---

## (e) — Non-linearity and GAMs (10 points)

Identify possible non-linear predictors and compare OLS with at least one GAM.

```{r}
library(gam)

# Your code here
```

*Your answer here.*

---

# Task 2 (50 points) — Applied Analysis: Customer Churn

Use `customer_data.csv` to answer both:

1. Why do some customers churn?
2. How can we predict high-risk new customers?

## (a) Data preparation (6 points)

- Recode variables to correct formats.
- Handle perfect collinearity.
- Split into 80/20 train/test using seed 42.
- Explain where and why you evaluate models.

```{r}
# Your code here
```

## (b) EDA (8 points)

Use descriptive statistics and plots to identify likely important predictors.

```{r}
# Your code here
```

## (c) Logistic regression (10 points)

Fit and evaluate using test accuracy, confusion matrix, sensitivity, specificity, ROC/AUC.

```{r}
# Your code here
```

## (d) Random forest (12 points)

Fit RF with 500 trees and tune `mtry` using 5-fold CV on training data.

```{r}
# Your code here
```

## (e) Boosting (10 points)

Fit boosted trees, inspect test AUC vs number of trees, refit at optimal size, and evaluate.

```{r}
# Your code here
```

## (f) Comparison and recommendation (4 points)

Provide a summary table and final recommendation.

*Your answer here.*
