---
title: "BAN404 Mock Exam 2026 #8"
subtitle: "Resampling-heavy Methods and Explainable Churn"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Explain CV pseudocode (10 points)

Given pseudocode for K-fold CV, explain where leakage can happen and how to avoid it.

## (b) Regularized model comparison (10 points)

Compare OLS, Ridge, and Elastic Net using the same standardized inputs.

```{r}
# Your code here
```

## (c) Bootstrap coefficient stability (10 points)

Bootstrap OLS coefficients and report the most unstable predictor.

```{r}
# Your code here
```

## (d) Variance CI (10 points)

Estimate CI for `Var(y)` and interpret practical uncertainty for expected prediction error scale.

```{r}
# Your code here
```

## (e) GAM diagnostics (10 points)

Fit GAM with smooth terms and interpret EDF values.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data prep with collinearity handling (8 points)

Remove deterministic duplicates, encode factors, and split data.

```{r}
# Your code here
```

## (b) EDA and interpretable segments (8 points)

Build churn summaries by segment and identify a high-risk customer profile.

```{r}
# Your code here
```

## (c) Logistic baseline + interpretation (12 points)

Fit logistic model and interpret three key effects in odds terms.

```{r}
# Your code here
```

## (d) Random forest nonlinear interactions (12 points)

Evaluate RF and explain where it can improve over logistic.

```{r}
# Your code here
```

## (e) Boosting + deployment recommendation (10 points)

Tune boosting and recommend model under requirement: high recall while keeping acceptable specificity.

```{r}
# Your code here
```
