---
title: "BAN404 Mock Exam 2026 #9 — SOLVED"
subtitle: "Validation Discipline and Robust Prediction — SOLVED"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# How to read this solved mock exam

This solved version is written as a **lookup tool**, not merely as a short answer key. That means every code line is commented with purpose, theory, or modeling logic, and every chunk is followed by at least three explanatory sentences that can be reused in a BAN404 answer. The goal is that you should be able to return to this file later and understand both **what to run** and **why the workflow is methodologically defensible**.

# Setup

```{r}
# Load tidyverse because BAN404 workflows are easiest to follow when wrangling, summaries, iteration, and plotting use one coherent grammar.  
library(tidyverse)  # Data cleaning, grouped summaries, tibbles, purrr iteration, and ggplot graphics used throughout the solution.  
# Load glmnet because Ridge, LASSO, and Elastic Net are fit through penalized regression paths in this package.  
library(glmnet)  # Penalized regression with built-in cross-validation support.  
# Load mgcv because GAMs are the standard BAN404 tool for structured nonlinear regression.  
library(mgcv)  # Generalized Additive Models with smooth terms and EDF summaries.  
# Load randomForest because tree ensembles can capture nonlinear split structure and interactions.  
library(randomForest)  # Random Forest classification for churn prediction.  
# Load gbm because staged boosting is another high-probability BAN404 ensemble family.  
library(gbm)  # Gradient boosting machine for Bernoulli classification.  
# Load broom because tidy model summaries are easier to interpret in exam language.  
library(broom)  # Tidy coefficient tables and odds-ratio-ready output.  
# Load splines because spline-style nonlinear benchmarks help compare structured curvature against linear fits.  
library(splines)  # Natural spline basis functions for regression benchmarks.  
# Store the exam number so seeds remain systematic and easy to trace.  
exam_id <- 9  # This solved file belongs to mock exam number 9.  
# Store the main alpha choices because Task 1 compares shrinkage styles under the same validation design.  
alpha_ridge <- 0  # alpha = 0 means pure Ridge shrinkage.  
alpha_lasso <- 1  # alpha = 1 means pure LASSO shrinkage with possible exact zeros.  
alpha_elnet <- 0.45  # This mixes L1 and L2 penalties to balance sparsity and stability.  
# Store the operational constants because Task 2 uses a recall-first rule with a specificity guardrail.  
recall_floor <- 0.78  # This is the minimum sensitivity target for the final churn policy.  
specificity_floor <- 0.60  # This prevents the retention team from contacting too many safe customers.  
cost_fn <- 4  # Missing a churner is treated as four times worse than contacting a non-churner.  
# Keep a compact theme label because later interpretation should echo the main storyline of the mock exam.  
focus_label <- "Validation discipline + robust deployment"  # Short narrative label for mock exam 9.  
# Create a rank-based AUC helper because ranking quality should be separated from threshold choice.  
auc_rank <- function(y_true, score, positive = "Yes") {  # Function takes observed classes and predicted scores.  
  y_chr <- as.character(y_true)  # Convert labels to characters so the positive class is handled explicitly.  
  y_bin <- if_else(y_chr == positive, 1, 0)  # Encode the positive class as 1 for the Mann-Whitney AUC identity.  
  n1 <- sum(y_bin == 1)  # Count positives because AUC compares positive and negative rankings.  
  n0 <- sum(y_bin == 0)  # Count negatives for the same reason.  
  if (n1 == 0 || n0 == 0) return(NA_real_)  # Return missing value if one class is absent and AUC is undefined.  
  r <- rank(score)  # Rank the scores because AUC is fundamentally a ranking measure.  
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)  # Apply the standard rank-based formula for AUC.  
}  
# Create a threshold-metric helper because operational policy needs more than a single ranking statistic.  
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {  # Function maps probabilities into decision metrics.  
  y_chr <- as.character(y_true)  # Convert classes to characters so the positive class is explicit.  
  negative <- setdiff(sort(unique(y_chr)), positive)[1]  # Identify the non-positive class as the reference level.  
  actual <- factor(y_chr, levels = c(negative, positive))  # Lock class order so confusion-matrix indexing remains stable.  
  pred <- factor(if_else(p_hat >= threshold, positive, negative), levels = c(negative, positive))  # Turn probabilities into class predictions.  
  cm <- table(actual = actual, pred = pred)  # Build the confusion matrix because threshold metrics come from these counts.  
  tn <- cm[negative, negative]  # Extract true negatives because they determine specificity.  
  fp <- cm[negative, positive]  # Extract false positives because they reflect unnecessary outreach.  
  fn <- cm[positive, negative]  # Extract false negatives because they are costly in this exam.  
  tp <- cm[positive, positive]  # Extract true positives because they are correctly identified churners.  
  tibble(  # Return a one-row tibble so many thresholds can be compared cleanly.  
    threshold = threshold,  # Store the evaluated cutoff.  
    accuracy = (tp + tn) / sum(cm),  # Compute accuracy for completeness even though it is not the main deployment objective.  
    sensitivity = tp / (tp + fn),  # Compute sensitivity because the policy is recall-first.  
    specificity = tn / (tn + fp),  # Compute specificity because the company still needs a workable contact volume.  
    auc = auc_rank(actual, p_hat, positive = positive),  # Add AUC to keep ranking quality distinct from threshold choice.  
    expected_cost = cost_fn * fn + fp  # Translate mistakes into a simple asymmetric-loss score.  
  )  # End tidy threshold-metric output.  
}  
# Create a helper that applies the recall-first guardrail because all Task 2 models should be judged under the same policy logic.  
choose_guardrail_rule <- function(metric_tbl, recall_floor, specificity_floor) {  # Function selects the best operating point under the stated constraints.  
  feasible_tbl <- metric_tbl %>% filter(sensitivity >= recall_floor, specificity >= specificity_floor)  # Keep only thresholds meeting both operational requirements.  
  if (nrow(feasible_tbl) > 0) {  # Use the feasible set when at least one threshold satisfies the rule.  
    feasible_tbl %>% arrange(expected_cost, desc(sensitivity), desc(auc)) %>% slice(1)  # Choose the cheapest feasible threshold while rewarding stronger recall and ranking.  
  } else {  # Fall back gracefully if the requested operating region is too strict.  
    metric_tbl %>% arrange(desc(sensitivity), expected_cost, desc(auc)) %>% slice(1)  # Choose the strongest recall point and then use cost and AUC to break ties.  
  }  # End guardrail-rule branch.  
}  
```

The setup chunk establishes a full BAN404 workflow rather than a few isolated commands. That matters because strong exam answers separate **ranking quality**, **threshold policy**, and **business interpretation** instead of treating them as one vague conclusion. The helper functions also reinforce a key course distinction: **AUC tells us how well the model ranks customers**, while **threshold rules determine what action policy the company actually deploys**.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Explain which validation strategy is easiest to defend when preprocessing matters

```{r}
# Read the Task 1 dataset because the validation discussion should still be tied to the actual exam data.  
df_t1 <- read_csv("../task1_data.csv", show_col_types = FALSE)  # Import quietly so the rendered document remains clean.  
# Build the predictor matrix because penalized regression functions typically work with X and y separately.  
X <- model.matrix(y ~ ., data = df_t1)[, -1]  # Remove the intercept column because glmnet handles the intercept internally.  
# Extract the response because later validation and bootstrap steps use it explicitly.  
y <- df_t1$y  # Continuous response variable for Task 1.  
# Create a theory summary because this first subquestion is conceptual rather than primarily computational.  
validation_summary <- tibble(  # Start validation-summary table.  
  strategy = c("LOOCV", "K-fold CV", "Repeated K-fold CV", "Unsafe preprocessing"),  # Name the core workflow ideas the examiner expects.  
  main_strength = c("Very low bias because almost all rows are used for training in each fit", "Good compromise between computational cost and validation realism", "More stable summary because performance is averaged over many plausible partitions", "Can look deceptively strong because validation rows indirectly influence the workflow"),  # Describe the main strength or apparent attraction of each option.  
  main_caution = c("Can have high variance and large computation cost on bigger datasets", "One unlucky split can still influence the result", "Costs more computation than one ordinary K-fold run", "Creates optimistic validation because future information leaks backward into training decisions"),  # Describe the main caution for each option.  
  ban404_takeaway = c("Not automatically best just because it uses more data per fold", "Usually defensible when preprocessing is kept inside each fold", "Often easiest to defend when tuning and preprocessing both matter", "Should be rejected because it violates fair out-of-sample logic")  # Translate each idea into BAN404-style exam language.  
)  # End validation-summary table.  
# Print the summary table because the subquestion asks for defended reasoning rather than only raw calculations.  
validation_summary  # Display the validation-discipline summary.  
```

This chunk frames validation as a **workflow design problem** rather than just a choice of function call. In BAN404 language, repeated K-fold CV is often easiest to defend because it reduces the influence of one lucky split while still remaining computationally manageable. The crucial point is that even a sophisticated fold design becomes invalid if preprocessing is performed before the resampling structure is respected.

## (b) Demonstrate why preprocessing outside the folds is optimistic

```{r}
# Set a seed because fold-based comparisons should be reproducible and not depend on one accidental partition.  
set.seed(909)  # Reproducible fold assignment for the leakage demonstration.  
# Create one shared fold assignment because the naive and safe workflows should differ only in preprocessing placement.  
fold_id <- sample(rep(1:10, length.out = nrow(X)))  # Shared 10-fold structure for both workflows.  
# Choose one common Ridge penalty because the goal is to isolate preprocessing placement rather than tuning strategy.  
lambda_probe <- 0.10  # Shared penalty strength for the leakage demonstration.  
# Standardize the full predictor matrix because this intentionally creates the naive workflow.  
X_global <- scale(X)  # Full-sample scaling leaks information from validation rows into the preprocessing step.  
# Evaluate the naive workflow because the prompt asks for a concrete optimistic-validation example.  
naive_mse <- map_dbl(1:10, function(k) {  # Loop over the shared folds for the naive workflow.  
  tr_idx <- fold_id != k  # Training rows for fold k.  
  va_idx <- fold_id == k  # Validation rows for fold k.  
  fit_k <- glmnet(X_global[tr_idx, ], y[tr_idx], alpha = alpha_ridge, lambda = lambda_probe, standardize = FALSE)  # Fit Ridge on globally scaled training data.  
  pred_k <- predict(fit_k, newx = X_global[va_idx, ], s = lambda_probe)  # Predict the held-out rows using the same leaked scaling.  
  mean((y[va_idx] - as.numeric(pred_k))^2)  # Return fold-level MSE for the naive workflow.  
})  # End naive-workflow loop.  
# Evaluate the leakage-safe workflow because scaling should be learned inside each training fold only.  
safe_mse <- map_dbl(1:10, function(k) {  # Loop over the shared folds for the safe workflow.  
  tr_idx <- fold_id != k  # Training rows for fold k.  
  va_idx <- fold_id == k  # Validation rows for fold k.  
  x_tr <- X[tr_idx, ]  # Extract training predictors for fold k.  
  x_va <- X[va_idx, ]  # Extract validation predictors for fold k.  
  tr_means <- colMeans(x_tr)  # Estimate the training-fold means because scaling should come only from the training data.  
  tr_sds <- apply(x_tr, 2, sd)  # Estimate the training-fold standard deviations for the same reason.  
  tr_sds[tr_sds == 0] <- 1  # Guard against zero standard deviations so scaling remains numerically valid.  
  x_tr_scaled <- sweep(sweep(x_tr, 2, tr_means, "-"), 2, tr_sds, "/")  # Standardize the training rows using only training-fold information.  
  x_va_scaled <- sweep(sweep(x_va, 2, tr_means, "-"), 2, tr_sds, "/")  # Apply the training-fold scaling to the held-out validation rows.  
  fit_k <- glmnet(x_tr_scaled, y[tr_idx], alpha = alpha_ridge, lambda = lambda_probe, standardize = FALSE)  # Fit Ridge on properly scaled training data.  
  pred_k <- predict(fit_k, newx = x_va_scaled, s = lambda_probe)  # Predict the held-out rows using the leakage-safe scaling.  
  mean((y[va_idx] - as.numeric(pred_k))^2)  # Return fold-level MSE for the safe workflow.  
})  # End safe-workflow loop.  
# Build a fold-wise comparison table because the question asks for an explicit contrast between the two workflows.  
preprocess_compare <- tibble(  # Start preprocessing-comparison table.  
  fold = 1:10,  # Keep the fold index visible.  
  naive_mse = naive_mse,  # Store the fold-level MSE from the globally scaled workflow.  
  safe_mse = safe_mse,  # Store the fold-level MSE from the leakage-safe workflow.  
  optimism_gap = naive_mse - safe_mse  # Measure how much more optimistic the naive workflow looks.  
)  # End preprocessing-comparison table.  
# Summarize the overall gap because BAN404 answers should usually report both detail and a compact conclusion.  
preprocess_summary <- preprocess_compare %>% summarise(mean_naive_mse = mean(naive_mse), mean_safe_mse = mean(safe_mse), mean_gap = mean(optimism_gap))  # Compute the average optimistic gap.  
preprocess_compare  # Print the fold-wise leakage comparison.  
preprocess_summary  # Print the overall optimistic-gap summary.  
```

This chunk gives a concrete demonstration of why preprocessing placement matters. The globally scaled workflow can look slightly better because information from the validation rows influenced the centering and scaling step before the model was fit. In BAN404 language, the safer workflow is the one to trust because it preserves the logic that validation rows should behave like unseen future data.

## (c) Build a same-fold penalty comparison and decide how conservative to be

```{r}
# Fit OLS because BAN404 comparisons usually start from the simplest interpretable benchmark.  
fit_ols <- lm(y ~ ., data = df_t1)  # Unpenalized linear-regression baseline.  
# Compute same-fold OLS validation error because the baseline should be judged under the same fold design.  
ols_cv <- map_dbl(1:10, function(k) {  # Loop over the shared folds for OLS.  
  tr <- df_t1[fold_id != k, ]  # Training rows for fold k.  
  va <- df_t1[fold_id == k, ]  # Validation rows for fold k.  
  fit_k <- lm(y ~ ., data = tr)  # Fit OLS only on the current training fold.  
  mean((va$y - predict(fit_k, newdata = va))^2)  # Return fold-level OLS MSE.  
})  # End same-fold OLS loop.  
# Tune Ridge on the shared folds because Ridge is the pure shrinkage benchmark in this mock exam.  
cv_ridge <- cv.glmnet(X, y, alpha = alpha_ridge, foldid = fold_id, standardize = TRUE)  # Shared-fold Ridge tuning.  
# Tune LASSO on the shared folds because sparsity is a natural comparator to pure shrinkage.  
cv_lasso <- cv.glmnet(X, y, alpha = alpha_lasso, foldid = fold_id, standardize = TRUE)  # Shared-fold LASSO tuning.  
# Tune Elastic Net on the same folds so differences reflect method behavior rather than different validation partitions.  
cv_elnet <- cv.glmnet(X, y, alpha = alpha_elnet, foldid = fold_id, standardize = TRUE)  # Shared-fold Elastic Net tuning.  
# Extract conservative coefficients because lambda.1se is often easier to defend when stability matters.  
coef_ridge <- as.numeric(coef(cv_ridge, s = "lambda.1se"))[-1]  # Ridge slopes at lambda.1se.  
coef_lasso <- as.numeric(coef(cv_lasso, s = "lambda.1se"))[-1]  # LASSO slopes at lambda.1se.  
coef_elnet <- as.numeric(coef(cv_elnet, s = "lambda.1se"))[-1]  # Elastic Net slopes at lambda.1se.  
# Build one tuning table because the examiner wants performance and penalty behavior to be interpreted together.  
penalty_compare <- tibble(  # Start penalty-comparison table.  
  model = c("OLS", "Ridge", "LASSO", "Elastic Net"),  # Name the competing models.  
  cv_mse = c(mean(ols_cv), min(cv_ridge$cvm), min(cv_lasso$cvm), min(cv_elnet$cvm)),  # Store the main validation-error summaries.  
  lambda_min = c(NA_real_, cv_ridge$lambda.min, cv_lasso$lambda.min, cv_elnet$lambda.min),  # Store the prediction-oriented penalty choice.  
  lambda_1se = c(NA_real_, cv_ridge$lambda.1se, cv_lasso$lambda.1se, cv_elnet$lambda.1se),  # Store the more conservative penalty choice.  
  active_predictors = c(ncol(X), sum(abs(coef_ridge) > 1e-8), sum(abs(coef_lasso) > 1e-8), sum(abs(coef_elnet) > 1e-8))  # Count active slopes to describe shrinkage versus sparsity.  
)  # End penalty-comparison table.  
# Build one coefficient table because aligned output is easier to interpret than many separate model summaries.  
coef_compare <- tibble(  # Start coefficient-comparison table.  
  variable = colnames(X),  # Keep the predictor names visible.  
  ols = as.numeric(coef(fit_ols)[-1]),  # OLS slopes excluding the intercept.  
  ridge_1se = coef_ridge,  # Ridge slopes at the conservative penalty choice.  
  lasso_1se = coef_lasso,  # LASSO slopes at the conservative penalty choice.  
  elastic_net_1se = coef_elnet  # Elastic Net slopes at the conservative penalty choice.  
)  # End coefficient-comparison table.  
penalty_compare  # Print the shared-fold penalty summary.  
coef_compare  # Print the side-by-side coefficient comparison.  
```

Using the same folds matters because otherwise part of the model difference could come from easier or harder validation partitions rather than from the penalties themselves. Ridge usually gives the most stable shrinkage story, LASSO usually gives the sparsest story, and Elastic Net often sits between them when correlated predictors exist. In BAN404 language, `lambda.1se` is often easier to defend because it favors a slightly more conservative answer that may be simpler and more stable even if `lambda.min` is marginally more prediction-oriented.

## (d) Use the bootstrap to discuss uncertainty in the response scale

```{r}
# Set a seed because bootstrap evidence should be reproducible and reviewable.  
set.seed(1009)  # Reproducible seed for SD(y) uncertainty.  
# Set the bootstrap size because interval estimation should be based on a large empirical sampling distribution.  
B_sd <- 1500  # Number of bootstrap replications for SD(y).  
# Generate bootstrap standard-deviation estimates because the prompt asks for uncertainty in the response scale.  
boot_sd <- replicate(B_sd, sd(sample(y, size = length(y), replace = TRUE)))  # Bootstrap sampling distribution for SD(y).  
# Compute the observed standard deviation because the intervals should be tied back to the sample estimate.  
sd_hat <- sd(y)  # Observed response standard deviation.  
# Compute the percentile interval because it follows the empirical bootstrap distribution directly.  
ci_sd_pct <- quantile(boot_sd, c(0.025, 0.975))  # Percentile CI for SD(y).  
# Compute the normal approximation because the prompt asks for an explicit comparison.  
ci_sd_norm <- c(sd_hat - 1.96 * sd(boot_sd), sd_hat + 1.96 * sd(boot_sd))  # Symmetric normal-based CI for SD(y).  
# Build one summary table because the answer should compare both intervals clearly.  
sd_summary <- tibble(  # Start SD-uncertainty summary.  
  quantity = "SD(y)",  # Name the reported quantity.  
  point_estimate = sd_hat,  # Store the observed sample standard deviation.  
  pct_low = ci_sd_pct[1],  # Store the lower percentile bound.  
  pct_high = ci_sd_pct[2],  # Store the upper percentile bound.  
  norm_low = ci_sd_norm[1],  # Store the lower normal-based bound.  
  norm_high = ci_sd_norm[2]  # Store the upper normal-based bound.  
)  # End SD-uncertainty summary.  
# Plot the bootstrap distribution because asymmetry is easier to discuss when it is visible.  
ggplot(tibble(boot_sd = boot_sd), aes(x = boot_sd)) +  # Start bootstrap-SD plot.  
  geom_histogram(bins = 35, fill = "darkorange", color = "white") +  # Show the empirical distribution of SD estimates.  
  geom_vline(xintercept = sd_hat, color = "darkgreen", linetype = "dashed") +  # Mark the observed sample standard deviation.  
  labs(title = "Bootstrap distribution of SD(y)", x = "Bootstrap SD estimate", y = "Count")  # Add clear plot labels.  
sd_summary  # Print the interval comparison table.  
```

This chunk shows that uncertainty about the response scale is itself an important modeling fact. If the bootstrap distribution is visibly asymmetric, the percentile interval is often easier to defend because it respects the empirical shape rather than forcing symmetry. In BAN404 language, uncertainty in `SD(y)` is a reminder that even a well-tuned model is predicting a noisy outcome, so final claims about fit quality should sound careful rather than overconfident.

## (e) Decide whether extra flexibility improves the answer enough to report

```{r}
# Fit the linear baseline because more flexible models should always be judged relative to a simpler benchmark.  
fit_linear <- lm(y ~ ., data = df_t1)  # Additive linear-regression baseline.  
# Fit a spline benchmark because structured curvature should be tested before jumping straight to the most flexible option.  
fit_spline <- lm(y ~ ns(x1, df = 4) + ns(x2, df = 4) + ns(x3, df = 4) + x4 + x5 + x6, data = df_t1)  # Natural-spline regression benchmark.  
# Fit a GAM because smooth terms are the BAN404 default nonlinear benchmark.  
fit_gam <- gam(y ~ s(x1, k = 6) + s(x2, k = 6) + s(x3, k = 6) + x4 + x5 + x6, data = df_t1)  # GAM with smooth terms for plausible nonlinear predictors.  
# Generate fitted values because all models should be compared on the same sample and response scale.  
y_hat_linear <- predict(fit_linear, newdata = df_t1)  # Linear fitted values.  
y_hat_spline <- predict(fit_spline, newdata = df_t1)  # Spline fitted values.  
y_hat_gam <- predict(fit_gam, newdata = df_t1)  # GAM fitted values.  
# Build a fit table because the prompt asks whether the extra flexibility is worth reporting.  
fit_compare <- tibble(  # Start nonlinear-comparison table.  
  model = c("Linear", "Spline", "GAM"),  # Name the competing models.  
  mse = c(mean((y - y_hat_linear)^2), mean((y - y_hat_spline)^2), mean((y - y_hat_gam)^2)),  # Compare mean squared error.  
  fit_measure = c(summary(fit_linear)$adj.r.squared, summary(fit_spline)$adj.r.squared, summary(fit_gam)$r.sq)  # Compare R2-style fit measures.  
)  # End nonlinear-comparison table.  
# Extract the GAM smooth summary because EDF values help decide whether the smooths are doing meaningful work.  
edf_table <- as_tibble(summary(fit_gam)$s.table, rownames = "smooth_term") %>%  # Convert the GAM smooth summary into a tidy table.  
  mutate(effect_type = case_when(edf < 1.2 ~ "approximately linear", edf < 2 ~ "mild curvature", TRUE ~ "clear nonlinearity"))  # Translate EDF values into exam-friendly interpretation labels.  
# Write a short examiner-ready Task 1 conclusion because the prompt asks for a final recommendation statement.  
final_task1_memo <- tibble(  # Start Task 1 memo table.  
  recommendation = paste0("Main Task 1 answer: ", fit_compare %>% arrange(mse) %>% slice(1) %>% pull(model), ". I would describe the nonlinear upgrade as ", if_else(min(fit_compare$mse[fit_compare$model != "Linear"]) < fit_compare$mse[fit_compare$model == "Linear"], "worth reporting if the fit improvement and EDF evidence jointly suggest real curvature", "too modest to replace the simpler linear explanation"), ".")  # Build a compact BAN404-style recommendation sentence.  
)  # End Task 1 memo table.  
fit_compare  # Print the linear-versus-nonlinear fit summary.  
edf_table  # Print the GAM EDF interpretation table.  
final_task1_memo  # Print the Task 1 recommendation.  
```

This chunk shows that extra flexibility should be defended, not assumed. A GAM is easiest to report when it improves fit and the EDF values suggest that the smooth terms are doing real work rather than behaving almost linearly. In BAN404 language, the correct conclusion is not “GAM is more advanced,” but “GAM is worth reporting only if the added flexibility produces substantively useful evidence.”

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Build a churn table that is realistic for deployment and honest for testing

```{r}
# Read the churn dataset because Task 2 is an end-to-end binary decision problem.  
df_c_raw <- read_csv("../customer_data.csv", show_col_types = FALSE)  # Import the raw churn data.  
# Clean names and recode types because deployment-ready workflows should make data preparation explicit.  
df_c <- df_c_raw %>%  # Start churn-preparation pipeline.  
  rename_with(~ gsub("[\\. ]", "_", tolower(.x))) %>%  # Convert names to lower-case snake_case.  
  mutate(churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")), international_plan = factor(international_plan), voice_mail_plan = factor(voice_mail_plan), area_code = factor(area_code)) %>%  # Recode the key categorical variables.  
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)  # Remove geography and deterministic charge duplicates for a more realistic deployment table.  
# Set a seed because the train/test split should be reproducible and reviewable.  
set.seed(1109)  # Reproducible hold-out split for Task 2.  
# Draw the training indices because all modeling work must happen before final testing.  
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))  # Sample 80 percent of the rows for training.  
# Create the training set because tuning and EDA should be restricted to this subset.  
train_c <- df_c[idx, ]  # Training data.  
# Create the test set because it should remain untouched until the final comparison stage.  
test_c <- df_c[-idx, ]  # Held-out test data.  
# Summarize class balance because churn problems often involve imbalance that affects threshold choice.  
class_split <- bind_rows(train = prop.table(table(train_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"), test = prop.table(table(test_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"), .id = "split")  # Create a tidy train-versus-test class-balance table.  
class_split  # Print the class-balance summary.  
```

This preparation step is designed to be realistic rather than purely convenient. Removing deterministic charge duplicates avoids redundant information, and removing `state` keeps the later model closer to a deployment story based on customer behaviour rather than a weak location proxy. In BAN404 language, leaving the test set untouched preserves the credibility of the final comparison because it remains a genuine out-of-sample check.

## (b) Use EDA to form risk stories before multivariate modeling

```{r}
# Build interpretable customer groupings because BAN404 EDA is strongest when it maps directly to risk stories and actions.  
train_persona <- train_c %>%  # Start persona-construction pipeline.  
  mutate(service_band = case_when(customer_service_calls >= 4 ~ "Repeated service friction", customer_service_calls >= 2 ~ "Moderate service friction", TRUE ~ "Low service friction"), day_band = case_when(total_day_minutes >= quantile(total_day_minutes, 0.75) ~ "Heavy day use", total_day_minutes <= quantile(total_day_minutes, 0.25) ~ "Light day use", TRUE ~ "Mid day use"), intl_band = case_when(total_intl_minutes >= quantile(total_intl_minutes, 0.75) ~ "Heavy international use", TRUE ~ "Lower international use"))  # Create grouped usage and friction personas.  
# Summarize persona-level churn because the prompt asks for meaningful pre-model risk stories.  
persona_summary <- train_persona %>%  # Start persona summary table.  
  group_by(service_band, day_band, intl_band, international_plan) %>%  # Group by interpretable customer-profile dimensions.  
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>%  # Compute churn rate and segment size.  
  filter(n >= 30) %>%  # Keep only segment sizes large enough to discuss credibly.  
  arrange(desc(churn_rate))  # Put the highest-risk personas first.  
# Build a simple action table because BAN404 rewards linking EDA patterns to intervention ideas.  
intervention_table <- tibble(observed_pattern = c("Repeated service friction aligns with elevated churn", "Heavy international use is often associated with higher risk", "Heavy day-use customers may be valuable enough to prioritize in retention campaigns"), suggested_action = c("Trigger service-recovery outreach after repeated complaints", "Review plan fit and price clarity for internationally active customers", "Offer proactive retention support to high-usage customers with emerging risk signals"))  # Convert descriptive patterns into possible actions.  
# Plot churn share by service-friction band because complaint history is often the clearest operational signal.  
ggplot(train_persona, aes(x = service_band, fill = churn)) +  # Start first persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions rather than raw counts.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express the vertical axis as percentages.  
  labs(title = "Churn share by service-friction band", x = "Service-friction band", y = "Proportion")  # Add clear plot labels.  
# Plot churn share by day-use band and international-plan status because joint descriptive patterns can motivate later model terms.  
ggplot(train_persona, aes(x = day_band, fill = churn)) +  # Start second persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions across usage bands.  
  facet_wrap(~ international_plan) +  # Split the plot by international-plan status.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express the vertical axis as percentages.  
  labs(title = "Churn share by day-use band and international plan", x = "Day-use band", y = "Proportion")  # Add clear plot labels.  
persona_summary  # Print the persona-level churn summary.  
intervention_table  # Print the descriptive-action mapping.  
```

This is the kind of EDA BAN404 rewards because it turns grouped summaries into **explainable risk stories** and **plausible interventions**. The plots help identify where churn risk appears to concentrate, while the persona table turns those patterns into language that can be reused in a written exam answer. The limitation is that persona summaries remain descriptive, so they may reflect confounding patterns that only multivariate models can separate properly.

## (c) Build a logistic model under a recall-first operating rule

```{r}
# Fit logistic regression because it is the standard interpretable churn baseline in BAN404.  
logit_fit <- glm(churn ~ account_length + area_code + international_plan + voice_mail_plan + number_vmail_messages + total_day_minutes + total_day_calls + total_eve_minutes + total_eve_calls + total_night_minutes + total_intl_minutes + customer_service_calls, data = train_c, family = binomial())  # Logistic model on the cleaned training data.  
# Predict test-set probabilities because ranking and policy should be assessed out of sample.  
p_logit <- predict(logit_fit, newdata = test_c, type = "response")  # Test-set churn probabilities from logistic regression.  
# Build a threshold grid because the operational cutoff is a business choice rather than a fixed model property.  
threshold_grid <- seq(0.10, 0.70, by = 0.05)  # Candidate deployment thresholds.  
# Evaluate the threshold grid because the final policy should be explicit and data-backed.  
logit_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_logit, threshold = .x))  # Threshold-by-threshold logistic performance table.  
# Apply the recall-first guardrail because the company wants high sensitivity without sacrificing too much specificity.  
best_logit <- choose_guardrail_rule(logit_grid, recall_floor = recall_floor, specificity_floor = specificity_floor)  # Best logistic operating point under the stated rule.  
# Convert coefficients to odds ratios because logistic interpretation is usually written in multiplicative-odds language.  
logit_or <- tidy(logit_fit) %>% mutate(odds_ratio = exp(estimate)) %>% arrange(desc(abs(estimate)))  # Tidy odds-ratio table sorted by effect magnitude.  
# Select exam-friendly terms because the prompt asks for interpretation rather than an undigested full coefficient dump.  
key_terms <- logit_or %>% filter(term %in% c("international_planYes", "voice_mail_planYes", "customer_service_calls"))  # Keep interpretable terms for BAN404-style discussion.  
logit_grid  # Print the threshold-by-threshold logistic performance table.  
best_logit  # Print the chosen operating point under the guardrail rule.  
key_terms  # Print the selected odds-ratio interpretations.  
```

This chunk separates three ideas that BAN404 students should keep distinct: **the model**, **the threshold**, and **the business rule**. Logistic regression provides an interpretable probability model, but the threshold is chosen later because operational priorities determine how those probabilities should be turned into action. Under a recall-first policy with a specificity guardrail, the final threshold is often below `0.5`, because catching churners matters more than maximizing default accuracy.

## (d) Test whether flexible tree models improve the churn solution

```{r}
# Set a seed because Random Forest uses random sampling and should therefore be reproducible.  
set.seed(1209)  # Reproducible Random Forest tuning seed.  
# Create inner folds because tuning must happen only inside the training data.  
rf_fold_id <- sample(rep(1:5, length.out = nrow(train_c)))  # Five-fold CV inside the training set.  
# Define the mtry grid because this parameter controls how much feature competition happens at each split.  
mtry_grid <- c(2, 4, 6, 8)  # Compact but informative Random Forest tuning grid.  
# Evaluate each mtry value because model flexibility should be chosen by validation rather than by guesswork.  
rf_cv <- map_dfr(mtry_grid, function(m) {  # Loop over candidate mtry values.  
  aucs <- map_dbl(1:5, function(k) {  # Loop over the inner CV folds.  
    tr <- train_c[rf_fold_id != k, ]  # Training portion for fold k.  
    va <- train_c[rf_fold_id == k, ]  # Validation portion for fold k.  
    rf_k <- randomForest(churn ~ ., data = tr, ntree = 500, mtry = m)  # Fit the forest for this fold and mtry.  
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]  # Predict validation probabilities.  
    auc_rank(va$churn, p_k)  # Return fold-level AUC because ranking quality is the first comparison layer.  
  })  # End inner fold loop.  
  tibble(mtry = m, cv_auc = mean(aucs))  # Store the mean validation AUC for this mtry value.  
})  # End Random Forest tuning loop.  
# Choose the best mtry because the final forest should use the strongest validated setting.  
best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1, with_ties = FALSE) %>% pull(mtry)  # Extract the best Random Forest tuning value.  
# Fit the final Random Forest because the test set should only see the chosen model once tuning is finished.  
rf_fit <- randomForest(churn ~ ., data = train_c, ntree = 650, mtry = best_mtry, importance = TRUE)  # Final tuned Random Forest.  
# Predict test-set probabilities because final comparison should use unseen data.  
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]  # Random Forest test-set probabilities.  
# Evaluate the same threshold grid because all models should face the same operating-rule logic.  
rf_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_rf, threshold = .x))  # Threshold-by-threshold Random Forest performance.  
# Apply the same guardrail rule because model comparison should not change the deployment objective.  
best_rf <- choose_guardrail_rule(rf_grid, recall_floor = recall_floor, specificity_floor = specificity_floor)  # Best Random Forest operating point.  
# Extract variable importance because tree models are interpreted through relevance rankings rather than coefficients.  
rf_imp <- importance(rf_fit) %>% as.data.frame() %>% rownames_to_column("variable") %>% as_tibble() %>% arrange(desc(MeanDecreaseGini))  # Tidy Random Forest importance table.  
# Prepare gbm-ready data because Bernoulli boosting expects a numeric target.  
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Training data with a numeric churn target.  
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Test data with a numeric churn target.  
# Create an inner training-validation split because staged boosting should choose tree count without using the final test set.  
set.seed(1309)  # Reproducible seed for the boosting early-stopping split.  
idx_gbm <- sample(seq_len(nrow(train_g)), size = floor(0.8 * nrow(train_g)))  # Draw rows for the boosting fit subset.  
fit_g <- train_g[idx_gbm, ]  # Inner training subset for gbm estimation.  
val_g <- train_g[-idx_gbm, ]  # Inner validation subset for early stopping.  
# Fit the boosting model because staged tree additions can capture complex churn structure gradually.  
gbm_fit <- gbm(churn01 ~ . - churn, distribution = "bernoulli", data = fit_g, n.trees = 850, interaction.depth = 3, shrinkage = 0.03, n.minobsinnode = 15, verbose = FALSE)  # Fit a long boosting path for evidence-based early stopping.  
# Create the staged tree grid because the final tree count should be chosen by validation rather than default settings.  
tree_grid <- seq(25, 850, by = 25)  # Candidate numbers of trees along the boosting path.  
# Evaluate validation AUC across the boosting path because early stopping should be evidence-based.  
val_auc_path <- map_dbl(tree_grid, function(nt) {  # Loop over candidate tree counts.  
  p_nt <- predict(gbm_fit, newdata = val_g, n.trees = nt, type = "response")  # Predict validation probabilities at this stage of boosting.  
  auc_rank(val_g$churn, p_nt)  # Return validation AUC for the current number of trees.  
})  # End staged boosting loop.  
# Choose the best tree count because the final boosting model should stop where validation quality peaks.  
best_trees <- tree_grid[which.max(val_auc_path)]  # Extract the best early-stopping point.  
# Predict the untouched test set because final boosting comparison should remain genuinely out of sample.  
p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")  # Final boosting probabilities on the test set.  
# Evaluate the same threshold grid because all competing models should face the same policy logic.  
boost_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_boost, threshold = .x))  # Threshold-by-threshold boosting performance.  
# Apply the same guardrail rule because policy consistency matters in model comparison.  
best_boost <- choose_guardrail_rule(boost_grid, recall_floor = recall_floor, specificity_floor = specificity_floor)  # Best boosting operating point.  
rf_cv  # Print the Random Forest tuning summary.  
best_rf  # Print the chosen Random Forest operating point.  
rf_imp %>% slice_head(n = 10)  # Print the top Random Forest importance values.  
tibble(trees = tree_grid, validation_auc = val_auc_path)  # Print the staged boosting validation path.  
best_boost  # Print the chosen boosting operating point.  
```

This chunk tests whether extra flexibility adds practically useful signal rather than only extra complexity. Random Forest and boosting may improve the churn solution when the true risk structure contains thresholds, nonlinearities, or interactions that logistic regression cannot capture naturally. In BAN404 language, the trade-off is that the explanation becomes less coefficient-based, so any performance gain must be large enough to justify the weaker interpretability.

## (e) Write the final deployment recommendation memo

```{r}
# Build one joint comparison table because the final recommendation should be based on a common operating-rule view.  
model_compare <- bind_rows(best_logit %>% mutate(model = "Logistic"), best_rf %>% mutate(model = paste0("Random Forest (mtry=", best_mtry, ")")), best_boost %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))) %>% select(model, threshold, accuracy, sensitivity, specificity, auc, expected_cost) %>% arrange(expected_cost, desc(sensitivity), desc(auc))  # Joint scorecard for the final decision.  
# Identify the recommended row because the final memo should make one explicit deployment choice.  
recommended_row <- model_compare %>% slice(1)  # Best model under the shared guardrail rule.  
# Build the final management memo because the prompt asks for a compact deployment-ready conclusion.  
final_task2_memo <- tibble(memo = paste0("Recommended deployment model: ", recommended_row$model, ". Recommended threshold: ", round(recommended_row$threshold, 2), ". This choice prioritizes recall while still enforcing a minimum specificity guardrail, so the company accepts some extra false positives in order to catch more true churners. After deployment, I would monitor class balance, threshold performance, calibration drift, and whether the main risk signals remain stable over time."))  # Compact BAN404-style management note.  
# Plot the staged validation path because early stopping is easier to defend visually than from one number alone.  
ggplot(tibble(trees = tree_grid, validation_auc = val_auc_path), aes(x = trees, y = validation_auc)) +  # Start boosting-path plot.  
  geom_line(color = "purple") +  # Show validation AUC across the boosting path.  
  geom_vline(xintercept = best_trees, color = "darkgreen", linetype = "dashed") +  # Mark the selected number of trees.  
  labs(title = "Boosting validation AUC path", x = "Number of trees", y = "Validation AUC")  # Add clear plot labels.  
model_compare  # Print the combined model-comparison scorecard.  
final_task2_memo  # Print the final deployment memo.  
```

This final chunk shows how a BAN404 answer moves from model comparison to a concrete operating recommendation. The strongest answer does not stop at “best AUC”; it states which threshold will be used, what trade-off that implies, and what should be monitored after deployment. In exam language, the best recommendation is the one that matches the actual business rule rather than the one that merely looks strongest on a single generic metric.
