---
title: "BAN404 Mock Exam 2026 #2"
subtitle: "Theory + Applied Churn Modelling"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `task1_data.csv`, `customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (task1_data.csv)

## (a) Function reading and method identification (10 points)

You are given this BAN404-style R function pair.

```{r}
f <- function(beta, X, y, lambda) {
  rss <- sum((y - X %*% beta)^2)
  penalty <- lambda * sum(abs(beta))
  rss + penalty
}

g <- function(X, y, lambda) {
  y_bar <- mean(y)
  X_c <- sweep(X, 2, colMeans(X), "-")
  y_c <- y - y_bar
  fit <- optim(par = rep(0, ncol(X)), fn = f, X = X_c, y = y_c, lambda = lambda)
  list(intercept = y_bar, beta = fit$par)
}
```

Explain method, penalty role, and difference from Ridge.

## (b) OLS vs regularization (10 points)

- Read `task1_data.csv`.
- Fit OLS and LASSO/Ridge-style alternatives.
- Compare coefficient tables.
- Explain whether variable selection appears.

```{r}
# Your code here
```

## (c) Cross-validation and tuning (10 points)

- Use K-fold CV (not LOOCV) to tune regularisation strength.
- Plot CV-MSE against penalty on log scale.
- Explain data leakage risk in fold preprocessing.

```{r}
# Your code here
```

## (d) Bootstrap confidence interval (10 points)

- Bootstrap the sampling distribution of `Var(y)` with B=1000 and seed 1.
- Show histogram.
- Compute normal-approx CI and percentile CI.

```{r}
# Your code here
```

## (e) Non-linearity and GAM (10 points)

- Diagnose non-linear predictors.
- Fit one OLS and one GAM.
- Compare in-sample MSE and R².

```{r}
# Your code here
```

# Task 2 (50 points) — Customer churn (customer_data.csv)

## (a) Data prep and design assumptions (8 points)

- Recode to correct variable formats.
- Remove variables not suitable at prediction time and motivate choices.
- Handle collinearity in telecom minute/charge pairs.
- Split to train/test with fixed seed.

```{r}
# Your code here
```

## (b) EDA and hypothesis building (8 points)

- Use descriptive stats and plots.
- Identify strongest candidate churn predictors.
- Comment on one drawback of pairwise analysis.

```{r}
# Your code here
```

## (c) Logistic regression (12 points)

- Fit logistic model.
- Evaluate with confusion matrix, sensitivity/specificity, and AUC.
- Justify threshold in business terms.

```{r}
# Your code here
```

## (d) Random forest and variable importance (12 points)

- Tune `mtry` using CV on training data.
- Evaluate on test data.
- Interpret variable importance and compare with EDA hypotheses.

```{r}
# Your code here
```

## (e) Boosting and model recommendation (10 points)

- Fit boosted trees.
- Find optimal number of trees using staged evaluation.
- Compare all models and recommend one for deployment.

```{r}
# Your code here
```
