---
title: "BAN404 Mock Exam 2026 #4"
subtitle: "Validation Discipline and Method Justification"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Explain K-fold CV pseudocode and leakage points (10 points)

Describe each step in a K-fold loop and identify where accidental test-information leakage can happen.

## (b) Leakage-safe scaling inside folds (10 points)

Implement fold-specific scaling and compare with naive global scaling for ridge tuning.

```{r}
# Your code here
```

## (c) Compare tuned Ridge and LASSO under common folds (10 points)

Use identical folds for fairness and compare best CV-MSE.

```{r}
# Your code here
```

## (d) Bootstrap CI interpretation quality check (10 points)

Estimate CI for `Var(y)` and discuss when percentile CI is preferred to normal approximation.

```{r}
# Your code here
```

## (e) Backfitting intuition and GAM implementation (10 points)

Explain backfitting intuition and compare GAM against OLS on fit quality.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data pipeline integrity (8 points)

Create a clean train/test process with no peeking and no post-outcome leakage variables.

```{r}
# Your code here
```

## (b) EDA to model-hypothesis bridge (8 points)

Produce compact EDA outputs and map each to a modeling hypothesis.

```{r}
# Your code here
```

## (c) Logistic model under threshold sweep (12 points)

Evaluate logistic across thresholds and explain business trade-offs.

```{r}
# Your code here
```

## (d) Random forest robust tuning (12 points)

Tune `mtry`, evaluate test AUC, and explain variable-importance reliability limits.

```{r}
# Your code here
```

## (e) Boosting selection with early stopping logic (10 points)

Find optimal trees via staged evaluation and justify final deployment choice.

```{r}
# Your code here
```
