---
title: "BAN404 Mock Exam 2026 #6"
subtitle: "Ridge-first Validation and Churn Decisioning"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Read method from objective function (10 points)

Given

```{r}
f <- function(beta, X, y, lambda) {
  rss <- sum((y - X %*% beta)^2)
  penalty <- lambda * sum(beta^2)
  rss + penalty
}
```

Explain method, geometry of shrinkage, and why coefficients rarely become exactly zero.

## (b) OLS vs Ridge vs LASSO (10 points)

Fit OLS, Ridge, and LASSO. Compare coefficients and discuss stability vs sparsity.

```{r}
# Your code here
```

## (c) K-fold CV + one-standard-error rule (10 points)

Tune lambda with 10-fold CV. Report `lambda.min` and `lambda.1se` and justify which you would pick for exam robustness.

```{r}
# Your code here
```

## (d) Bootstrap variance uncertainty (10 points)

Bootstrap `Var(y)` with B = 1500. Report percentile and normal CIs and explain any asymmetry.

```{r}
# Your code here
```

## (e) Linear vs smooth effects (10 points)

Diagnose non-linearity and compare OLS with GAM. Evaluate using MSE and adjusted R².

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Prediction-time data design (8 points)

Prepare data, remove deterministic duplicates, and create train/test split with fixed seed.

```{r}
# Your code here
```

## (b) EDA with actionable hypotheses (8 points)

Create churn-rate summaries and at least two visuals. State three hypotheses to test in models.

```{r}
# Your code here
```

## (c) Logistic regression + threshold strategy (12 points)

Fit logistic regression, evaluate multiple thresholds, and choose threshold using business cost of false negatives.

```{r}
# Your code here
```

## (d) Random forest tuning and interpretation (12 points)

Tune `mtry` using CV on training set. Evaluate on test set and compare importance with EDA hypotheses.

```{r}
# Your code here
```

## (e) Boosting and final recommendation (10 points)

Tune tree count by staged test AUC. Compare logistic, RF, and boosting and recommend one deployment model.

```{r}
# Your code here
```
