---
title: "BAN404 Mock Exam 2026 #3"
subtitle: "Elastic Net Trade-offs and Segment-aware Churn"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)
- **Important exam expectation:** In every subquestion, combine **method choice + runnable code + interpretation + theory/business reasoning**. A strong BAN404 answer explains why a model is appropriate, what assumptions or trade-offs matter, and what the numbers imply for a decision.

# How to use this mock exam

This mock exam is written as both a **practice exam** and a **lookup tool**. That means each subquestion tells you not only what to compute, but also what the examiner is likely to reward in a strong written answer. When you solve it, aim to write short theory-grounded paragraphs after your code rather than stopping at tables or plots.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## Context for Task 1

This task is centered on **Elastic Net**, which sits between **Ridge** and **LASSO**. That means the examiner is likely to care about whether you can explain **shrinkage**, **variable selection**, **multicollinearity handling**, and how `alpha` changes model behavior. The task also asks you to reason carefully about validation discipline, coefficient stability, and smooth terms, so method explanation is just as important as final output.

## (a) Identify Elastic Net objective mechanics (10 points)

You are given the following function:

```{r}
f <- function(beta, X, y, lambda, alpha) {
  sum((y - X %*% beta)^2) + lambda * ((1 - alpha) * sum(beta^2) + alpha * sum(abs(beta)))
}
```

Write a theory-grounded answer that explains all of the following:

1. **Which model family this objective belongs to**.
2. **What the RSS part does** and what the penalty part does.
3. **How `alpha = 0`, `alpha = 1`, and `0 < alpha < 1` should be interpreted**.
4. **Why Elastic Net can be useful when predictors are correlated**.
5. **Why Elastic Net may keep some sparsity while still stabilizing coefficients**.

Your answer should read like a BAN404 theory paragraph, not just a method label.

## (b) Ridge vs LASSO vs Elastic Net comparison (10 points)

Read `task1_data.csv` and compare **Ridge**, **LASSO**, and **Elastic Net**.

What to do:

- Fit all three methods using `glmnet`.
- Tune or compare them using cross-validated MSE.
- Report `lambda.min` for each method.
- Compare coefficient sparsity at the selected lambda values.
- Explain how the comparison reflects the theoretical difference between shrinkage-only and shrinkage-plus-selection methods.

What the examiner wants to see:

- correct separation of the three methods,
- correct interpretation of `alpha`,
- awareness that Ridge emphasizes stability, LASSO emphasizes sparsity, and Elastic Net mixes the two,
- clear link between CV performance and model complexity.

```{r}
# Your code here
```

## (c) Validation discipline under preprocessing risk (10 points)

Show how preprocessing can create leakage if it is done incorrectly.

What to do:

- Implement a validation comparison where scaling is done **incorrectly on the full dataset** and **correctly within each fold**.
- Use Elastic Net as the model.
- Compare validation MSE from the leakage-prone approach and the leakage-safe approach.
- Explain why fold-specific preprocessing is the fairer procedure.
- Briefly state what kind of exam answer would lose points here.

A strong answer should make it clear that leakage does not make the model “better”; it makes the evaluation **too optimistic**.

```{r}
# Your code here
```

## (d) Bootstrap coefficient stability (10 points)

Use the bootstrap to study **OLS coefficient stability**.

What to do:

- Set a seed.
- Draw many bootstrap samples.
- Refit the OLS model in every bootstrap sample.
- Store coefficient estimates for all predictors.
- Identify the **least stable predictor**.
- Explain what “least stable” means statistically.

The best answers connect coefficient instability to sample variation, multicollinearity, or weak signal.

```{r}
# Your code here
```

## (e) GAM smoothness interpretation (10 points)

Fit a **GAM** and interpret the smooth terms using their EDF values.

What to do:

- Fit a linear baseline.
- Fit a GAM with smooths for the most plausible nonlinear predictors.
- Report model fit using MSE and an R²-style measure.
- Extract and interpret the smooth-term EDF values.
- Decide which predictors are effectively nonlinear and which are close to linear.

A strong answer should explain that **EDF close to 1 suggests a near-linear effect**, while higher EDF values suggest more flexible curvature.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## Context for Task 2

This task focuses on **segment-aware churn prediction**. You are expected to move between **EDA**, **interpretable modeling**, and **predictive modeling** without losing sight of business usefulness. In this mock exam, assume the company treats a **false negative as 3 times as costly as a false positive**, so the final threshold policy should reflect that cost asymmetry.

## (a) Data preparation and deployment assumptions (8 points)

Prepare the churn data so it is realistic for prediction.

What to do:

- Read and clean the data.
- Recode variables into suitable types.
- Remove deterministic or duplicated predictors.
- Exclude variables you cannot defend from a deployment perspective.
- Perform an 80/20 train/test split with a fixed seed.
- Briefly justify each exclusion in terms of real-time availability or interpretation quality.

What the examiner wants:

- sensible factor handling,
- awareness of duplicated information,
- awareness of what can and cannot be known at prediction time,
- a clean split before later modeling.

```{r}
# Your code here
```

## (b) Segment-centric exploratory analysis (8 points)

Use the training data to create **risk segments** and describe a high-risk churn persona.

What to do:

- Create interpretable segments using customer behavior and/or service variables.
- Summarize churn rates across segments.
- Produce at least two informative plots.
- Identify one **high-risk persona** worth targeting.
- Explain why the EDA supports that persona description.
- Mention one limitation of segment-based descriptive analysis.

This is not only a plotting task. It is a reasoning task where you translate patterns into a story that later models can test.

```{r}
# Your code here
```

## (c) Logistic baseline with odds interpretation (12 points)

Fit a logistic regression model and interpret it carefully.

What to do:

- Fit the model on the training data.
- Predict churn probabilities on the test set.
- Report at least one global performance measure such as AUC.
- Interpret at least **three coefficients** in odds-ratio language.
- Explain which effects look practically meaningful.
- Comment on why logistic regression remains useful even if more flexible models later perform better.

A strong answer should clearly separate **direction of effect**, **size of effect**, and **business meaning**.

```{r}
# Your code here
```

## (d) Random forest interaction gains (12 points)

Fit and tune a Random Forest, then compare it with the logistic baseline.

What to do:

- Tune `mtry` using CV on the training data only.
- Fit the final forest.
- Evaluate it on the test set.
- Report variable importance.
- Compare the Random Forest story to the logistic-regression story.
- Explain where nonlinearities or interactions may help the forest outperform the logistic model.

The best answers explain that tree ensembles can discover structure that a default additive logistic model does not capture automatically.

```{r}
# Your code here
```

## (e) Boosting for final scorecard (10 points)

Fit a boosting model and recommend a final targeting strategy.

What to do:

- Fit a boosting model on the training data.
- Tune or choose the number of trees using staged validation.
- Compare logistic regression, Random Forest, and boosting.
- Choose a threshold policy under the **3:1 false-negative cost structure**.
- Give a final recommendation for campaign targeting.

Your final paragraph should explicitly state:

1. which model you recommend,
2. which metric or objective matters most,
3. why the threshold should not simply be `0.5`,
4. what trade-off the company accepts when it targets more customers to avoid missing churners.

```{r}
# Your code here
```
