---
title: "BAN404 Mock Exam 2026 #4 — SOLVED"
subtitle: "Validation Discipline and Method Justification — SOLVED"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# How to read this solved mock exam

This solved version is written as a **lookup tool**, not only as a short answer key. That means every code line is commented with purpose, theory, or modeling logic, and every code chunk is followed by at least three sentences of explanation that you can reuse in an exam answer. The goal is that you should be able to return to this file later and understand both **what to run** and **how to justify why it was the correct BAN404 step**.

# Setup

```{r}
# Load tidyverse because BAN404 workflows are easier to read when wrangling, plotting, and summaries use one coherent grammar.  
library(tidyverse)  # Data cleaning, grouped summaries, tibbles, and ggplot graphics used throughout the solution.  
# Load glmnet because Ridge and LASSO are fit efficiently through penalized regression paths in this package.  
library(glmnet)  # Penalized regression with lambda tuning and shared-fold cross-validation support.  
# Load mgcv because GAMs are the standard BAN404 tool for structured nonlinear regression.  
library(mgcv)  # Generalized Additive Models with smooth terms and EDF summaries.  
# Load randomForest because tree ensembles are useful when nonlinearities and interactions may matter.  
library(randomForest)  # Random Forest classification for churn prediction.  
# Load gbm because boosting is another high-probability BAN404 model family.  
library(gbm)  # Gradient boosting machine for staged tree-based learning.  
# Load broom because tidy coefficient tables are easier to interpret in exam answers.  
library(broom)  # Tidy model summaries and coefficient extraction.  
# Store the exam number so seeds remain systematic and easy to trace.  
exam_id <- 4  # This solved file belongs to mock exam number 4.  
# Set the main alpha because Task 1 is centered on leakage-safe Ridge tuning.  
alpha_main <- 0  # alpha = 0 corresponds to Ridge regression with pure L2 shrinkage.  
# Store the cost ratio because Task 2 assumes false negatives cost five times false positives.  
cost_fn <- 5  # Missing a churner is much costlier than unnecessarily targeting a safe customer.  
# Keep a readable label for use in prose and output summaries.  
focus_label <- "Leakage-safe validation and method justification"  # This helps keep the narrative aligned with the task.  
# Create a rank-based AUC helper because discrimination quality should be measured separately from threshold policy.  
auc_rank <- function(y_true, score, positive = "Yes") {  # Function takes observed classes and predicted scores.  
  y_chr <- as.character(y_true)  # Convert labels to characters so class handling is explicit.  
  y_bin <- if_else(y_chr == positive, 1, 0)  # Encode the positive class as 1 for the Mann-Whitney AUC formula.  
  n1 <- sum(y_bin == 1)  # Count positives because AUC compares positive and negative rankings.  
  n0 <- sum(y_bin == 0)  # Count negatives for the same reason.  
  if (n1 == 0 || n0 == 0) return(NA_real_)  # Return missing value if AUC is undefined because one class is absent.  
  r <- rank(score)  # Rank the scores because AUC is fundamentally a ranking metric.  
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)  # Apply the standard rank-based AUC identity.  
}  
# Create a threshold-metric helper because operational decisions require more than a single AUC value.  
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {  # Function maps probabilities into decision metrics.  
  y_chr <- as.character(y_true)  # Convert classes to characters so the positive class is handled explicitly.  
  negative <- setdiff(sort(unique(y_chr)), positive)[1]  # Identify the non-positive class as the reference category.  
  actual <- factor(y_chr, levels = c(negative, positive))  # Lock class order so confusion-matrix indexing stays stable.  
  pred <- factor(if_else(p_hat >= threshold, positive, negative), levels = c(negative, positive))  # Convert probabilities into class labels at a chosen threshold.  
  cm <- table(actual = actual, pred = pred)  # Build the confusion matrix because threshold metrics come from these counts.  
  tn <- cm[negative, negative]  # Extract true negatives because they determine specificity.  
  fp <- cm[negative, positive]  # Extract false positives because they represent unnecessary intervention cost.  
  fn <- cm[positive, negative]  # Extract false negatives because they are especially costly in this mock exam.  
  tp <- cm[positive, positive]  # Extract true positives because they are correctly identified churners.  
  tibble(  # Return a one-row tibble so many thresholds can be compared easily.  
    threshold = threshold,  # Store the evaluated cutoff.  
    accuracy = (tp + tn) / sum(cm),  # Compute accuracy for completeness even though business cost matters more here.  
    sensitivity = tp / (tp + fn),  # Compute sensitivity because catching churners matters when false negatives are costly.  
    specificity = tn / (tn + fp),  # Compute specificity to track how often safe customers are left alone.  
    auc = auc_rank(actual, p_hat, positive = positive),  # Add AUC to separate ranking quality from threshold choice.  
    expected_cost = cost_fn * fn + fp  # Convert mistakes into the exam's business loss structure.  
  )  # End tidy threshold-metric output.  
}  
```

The setup chunk establishes the full workflow for the rest of the exam. In BAN404, that matters because strong answers are not only about fitting models; they are about showing a **reproducible analytical pipeline** with transparent evaluation and interpretation. The helper functions also reinforce an important course distinction: **AUC evaluates ranking quality**, while **threshold metrics convert that ranking into a business action rule**.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Explain K-fold CV pseudocode and leakage points

```{r}
# Read the Task 1 dataset because all methodology examples should be grounded in the actual exam data.  
df_t1 <- read_csv("../task1_data.csv", show_col_types = FALSE)  # Import quietly so the rendered output remains tidy.  
# Build a numeric model matrix because penalized regression routines typically work with x-matrices.  
X <- model.matrix(y ~ ., data = df_t1)[, -1]  # Remove the intercept column because glmnet handles the intercept internally.  
# Extract the response because resampling loops often work with y separately from X.  
y <- df_t1$y  # Continuous response used throughout Task 1.  
# Create a compact pseudocode-style table because the prompt asks for a structured explanation of K-fold CV.  
cv_steps <- tibble(  # Start cross-validation step summary.  
  step = 1:6,  # Number the conceptual steps to mirror pseudocode logic.  
  action = c("Split data into K folds", "Hold out one fold as validation", "Train model on remaining folds", "Apply training-only preprocessing", "Predict the held-out fold", "Average validation error across folds"),  # Describe each stage of the loop.  
  leakage_risk = c("Low if split is fixed before tuning", "Low if the hold-out fold stays untouched", "Low if only training rows are used", "High if scaling/imputation uses all rows", "High if thresholds are tuned on the test fold", "High if the final test set influences model choice")  # Highlight where leakage can occur.  
)  # End pseudocode summary table.  
# Display the K-fold logic table because it turns methodology language into an exam-ready structure.  
cv_steps  # Print the pseudocode-style cross-validation summary.  
```

This table captures the core logic of K-fold cross-validation: each fold gets a turn as validation data while the remaining folds play the role of training data. The most important methodological point is that preprocessing and model selection decisions must be learned **inside the training portion of each fold**, not from the full dataset. In BAN404 language, leakage does not mean the model has genuinely improved; it means the evaluation has become **too optimistic**, which weakens the credibility of the reported performance.

## (b) Leakage-safe scaling inside folds

```{r}
# Set a seed because any fold-based comparison should be reproducible and not driven by one lucky random split.  
set.seed(104)  # Reproducible fold assignment for the leakage comparison.  
# Create one common set of five folds because the point is to isolate preprocessing effects rather than split differences.  
fold_id <- sample(rep(1:5, length.out = nrow(X)))  # Shared fold assignment for all later validation comparisons.  
# Define a lambda grid because Ridge tuning should compare many penalty strengths rather than a single arbitrary value.  
lambda_grid <- exp(seq(log(1e-3), log(10), length.out = 60))  # Wide but smooth Ridge penalty grid.  
# Create a globally scaled version of X to represent the wrong leakage-prone workflow.  
X_global_scaled <- scale(X)  # This is wrong for validation because it uses information from all observations at once.  
# Evaluate each lambda under both naive and leakage-safe scaling.  
scale_compare <- map_dfr(lambda_grid, function(lambda_i) {  # Loop over the Ridge penalty grid.  
  fold_results <- map_dfr(1:5, function(k) {  # Loop over the shared folds.  
    train_idx <- fold_id != k  # Training rows for fold k.  
    valid_idx <- fold_id == k  # Validation rows for fold k.  
    x_train <- X[train_idx, ]  # Unscaled training predictors for the safe workflow.  
    x_valid <- X[valid_idx, ]  # Unscaled validation predictors for the safe workflow.  
    y_train <- y[train_idx]  # Training outcomes for fold k.  
    y_valid <- y[valid_idx]  # Validation outcomes for fold k.  
    fit_naive <- glmnet(X_global_scaled[train_idx, ], y_train, alpha = 0, lambda = lambda_i, standardize = FALSE)  # Fit Ridge after leakage-prone global scaling.  
    pred_naive <- as.numeric(predict(fit_naive, newx = X_global_scaled[valid_idx, ], s = lambda_i))  # Predict the validation fold using the wrong globally scaled data.  
    train_means <- colMeans(x_train)  # Learn fold-specific centering values from the training fold only.  
    train_sds <- apply(x_train, 2, sd)  # Learn fold-specific scales from the training fold only.  
    train_sds[train_sds == 0] <- 1  # Prevent division by zero if a predictor has no within-fold variation.  
    x_train_safe <- scale(x_train, center = train_means, scale = train_sds)  # Scale the training fold with training-fold statistics only.  
    x_valid_safe <- scale(x_valid, center = train_means, scale = train_sds)  # Apply the same training-fold scaling to the validation fold.  
    fit_safe <- glmnet(x_train_safe, y_train, alpha = 0, lambda = lambda_i, standardize = FALSE)  # Fit leakage-safe Ridge on the fold-specific scaled data.  
    pred_safe <- as.numeric(predict(fit_safe, newx = x_valid_safe, s = lambda_i))  # Predict the validation fold using only allowed preprocessing information.  
    tibble(  # Return fold-level validation errors for this lambda.  
      fold = k,  # Fold identifier.  
      lambda = lambda_i,  # Penalty strength being evaluated.  
      mse_naive = mean((y_valid - pred_naive)^2),  # Validation MSE from the wrong globally scaled workflow.  
      mse_safe = mean((y_valid - pred_safe)^2)  # Validation MSE from the correct leakage-safe workflow.  
    )  # End fold-level summary.  
  })  # End fold loop.  
  fold_results %>% summarise(lambda = first(lambda), mean_mse_naive = mean(mse_naive), mean_mse_safe = mean(mse_safe))  # Collapse fold results into one row per lambda.  
})  # End lambda-grid loop.  
# Extract the best lambda under each workflow because the prompt asks for tuned performance comparison.  
best_naive <- scale_compare %>% slice_min(mean_mse_naive, n = 1)  # Best apparent result under leakage-prone scaling.  
best_safe <- scale_compare %>% slice_min(mean_mse_safe, n = 1)  # Best honest result under leakage-safe scaling.  
# Plot the two tuning curves because optimistic leakage is easier to see when both error profiles are on the same figure.  
ggplot(scale_compare, aes(x = lambda)) +  # Start tuning-curve plot.  
  geom_line(aes(y = mean_mse_naive, color = "Naive global scaling")) +  # Draw the leakage-prone tuning curve.  
  geom_line(aes(y = mean_mse_safe, color = "Leakage-safe fold scaling")) +  # Draw the leakage-safe tuning curve.  
  scale_x_log10() +  # Use a log scale because Ridge penalties are typically compared on multiplicative scales.  
  labs(title = "Ridge tuning: global scaling versus fold-specific scaling", x = "Lambda", y = "Mean validation MSE", color = "Workflow")  # Add exam-ready labels.  
best_naive  # Print the best naive tuning result.  
best_safe  # Print the best leakage-safe tuning result.  
```

This chunk shows why preprocessing is part of the validation pipeline rather than a harmless preliminary step. When scaling is done globally, the validation fold contributes to the centering and scaling values, which makes the reported error look slightly better than it should. In BAN404 terms, the safe workflow is the one you can defend in an exam because it preserves the idea that the validation fold must behave like **future unseen data**.

## (c) Compare tuned Ridge and LASSO under common folds

```{r}
# Reuse the common fold assignment because a fair method comparison should hold the data split constant.  
cv_ridge_common <- cv.glmnet(X, y, alpha = 0, foldid = fold_id, standardize = TRUE)  # Tune Ridge on the shared folds.  
# Tune LASSO on the exact same folds so the comparison is method-driven rather than split-driven.  
cv_lasso_common <- cv.glmnet(X, y, alpha = 1, foldid = fold_id, standardize = TRUE)  # Tune LASSO on the shared folds.  
# Extract coefficient vectors at lambda.min because this is the prediction-oriented tuning choice.  
coef_ridge <- as.numeric(coef(cv_ridge_common, s = "lambda.min"))[-1]  # Ridge coefficients at its best shared-fold lambda.  
coef_lasso <- as.numeric(coef(cv_lasso_common, s = "lambda.min"))[-1]  # LASSO coefficients at its best shared-fold lambda.  
# Summarize the shared-fold comparison because the question asks for fairness-aware method comparison.  
common_fold_compare <- tibble(  # Start shared-fold method table.  
  model = c("Ridge", "LASSO"),  # Name the compared methods.  
  lambda_min = c(cv_ridge_common$lambda.min, cv_lasso_common$lambda.min),  # Prediction-oriented tuned lambda.  
  lambda_1se = c(cv_ridge_common$lambda.1se, cv_lasso_common$lambda.1se),  # More conservative tuned lambda.  
  cv_mse = c(min(cv_ridge_common$cvm), min(cv_lasso_common$cvm)),  # Best cross-validated MSE under the shared folds.  
  near_zero = c(sum(abs(coef_ridge) < 1e-8), sum(abs(coef_lasso) < 1e-8))  # Compare sparsity at lambda.min.  
)  # End method comparison table.  
# Build a coefficient table because the methods should also be compared in terms of shrinkage pattern.  
coef_compare <- tibble(  # Start coefficient comparison.  
  variable = colnames(X),  # Keep predictor names visible.  
  ridge = coef_ridge,  # Ridge slopes at lambda.min.  
  lasso = coef_lasso  # LASSO slopes at lambda.min.  
)  # End coefficient comparison.  
common_fold_compare  # Print the shared-fold Ridge versus LASSO summary.  
coef_compare  # Print the coefficient-level comparison.  
```

Using the same folds matters because otherwise part of the observed performance difference may come from an easier or harder split rather than from the method itself. In general, Ridge is expected to keep all predictors and stabilize coefficients, while LASSO is more willing to set some coefficients exactly to zero. In BAN404 language, the fair conclusion is not just “which MSE is lower,” but whether the lower MSE is large enough to matter given the trade-off between **stability**, **sparsity**, and **interpretability**.

## (d) Bootstrap CI interpretation quality check

```{r}
# Set a seed because bootstrap results should be reproducible and reviewable.  
set.seed(204)  # Reproducible bootstrap for Var(y).  
# Set the number of bootstrap samples because a large B stabilizes the empirical uncertainty estimate.  
B <- 1200  # Use many resamples so the bootstrap distribution is smooth enough to interpret.  
# Store the sample size because each bootstrap sample should contain n rows with replacement.  
n <- length(y)  # Original sample size for the Task 1 response.  
# Preallocate storage because we want one variance estimate from each bootstrap world.  
boot_var <- numeric(B)  # Container for bootstrap estimates of Var(y).  
# Run the bootstrap because the goal is to approximate the sampling distribution empirically.  
for (b in seq_len(B)) {  # Loop over bootstrap replications.  
  idx <- sample(seq_len(n), size = n, replace = TRUE)  # Draw bootstrap indices with replacement.  
  y_b <- y[idx]  # Build the bootstrap resample of the response.  
  boot_var[b] <- var(y_b)  # Compute and store Var(y) for this bootstrap sample.  
}  
# Compute the original sample variance because intervals should be centered around the observed statistic.  
var_hat <- var(y)  # Plug-in variance estimate from the observed data.  
# Compute bootstrap standard error because the normal approximation interval uses it directly.  
se_boot <- sd(boot_var)  # Empirical standard error from the bootstrap distribution.  
# Compute the percentile interval because it respects the empirical shape of the bootstrap distribution.  
ci_pct <- quantile(boot_var, c(0.025, 0.975))  # Percentile 95% interval.  
# Compute the normal approximation interval because the prompt asks for a comparison.  
ci_norm <- c(var_hat - 1.96 * se_boot, var_hat + 1.96 * se_boot)  # Symmetric normal-based 95% interval.  
# Summarize the bootstrap evidence in one tidy table.  
boot_summary <- tibble(  # Start bootstrap summary table.  
  point_estimate = var_hat,  # Observed variance estimate.  
  se_boot = se_boot,  # Bootstrap standard error.  
  pct_low = ci_pct[1],  # Lower percentile bound.  
  pct_high = ci_pct[2],  # Upper percentile bound.  
  norm_low = ci_norm[1],  # Lower normal-approximation bound.  
  norm_high = ci_norm[2]  # Upper normal-approximation bound.  
)  # End bootstrap summary.  
# Plot the bootstrap distribution because interval choice should be justified from its shape.  
ggplot(tibble(boot_var = boot_var), aes(x = boot_var)) +  # Start histogram of bootstrap variance estimates.  
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +  # Visualize the empirical sampling distribution.  
  geom_vline(xintercept = var_hat, color = "darkgreen", linetype = "dashed") +  # Mark the observed sample variance.  
  labs(title = "Bootstrap distribution of Var(y)", x = "Bootstrap variance estimate", y = "Count")  # Add clear labels for interpretation.  
boot_summary  # Print the interval comparison table.  
```

The bootstrap is useful here because it estimates uncertainty directly from repeated resampling instead of relying only on a closed-form formula. If the histogram is noticeably skewed, the percentile interval is usually easier to defend because it follows the empirical shape of the bootstrap distribution rather than forcing symmetry. In BAN404 terms, the percentile interval is often the safer interpretation when the resampled statistic does not look approximately normal.

## (e) Backfitting intuition and GAM implementation

```{r}
# Create a small conceptual table because the prompt asks for an explanation of backfitting intuition.  
backfitting_steps <- tibble(  # Start conceptual backfitting summary.  
  stage = c("Hold other effects fixed", "Update one smooth term", "Cycle through the smooth terms", "Repeat until changes are small"),  # Describe the iterative logic of backfitting.  
  intuition = c("Treat the remaining structure as temporary residuals", "Fit the current nonlinear component to what is left", "Let each smooth adjust while the others stay provisionally fixed", "Stop when the additive components stabilize")  # Translate the algorithm into plain-language interpretation.  
)  # End conceptual table.  
# Fit the linear baseline because GAM performance should be judged relative to a simpler model.  
fit_ols <- lm(y ~ ., data = df_t1)  # OLS benchmark with additive linear effects.  
# Fit a GAM because smooth terms allow the model to learn structured nonlinearity.  
fit_gam <- gam(y ~ s(x1, k = 6) + s(x2, k = 6) + s(x3, k = 6) + x4 + x5 + x6, data = df_t1)  # Smooth the predictors most likely to be nonlinear.  
# Generate OLS fitted values so fit quality can be compared on the same dataset.  
y_hat_ols <- predict(fit_ols, newdata = df_t1)  # OLS fitted values.  
# Generate GAM fitted values for the same comparison.  
y_hat_gam <- predict(fit_gam, newdata = df_t1)  # GAM fitted values.  
# Compute OLS MSE because one key question is whether the GAM improves fit materially.  
mse_ols <- mean((y - y_hat_ols)^2)  # In-sample OLS mean squared error.  
# Compute GAM MSE as the flexible-model comparator.  
mse_gam <- mean((y - y_hat_gam)^2)  # In-sample GAM mean squared error.  
# Build a tidy model-fit table because the decision should combine error and explained-variation summaries.  
fit_compare <- tibble(  # Start OLS versus GAM comparison.  
  model = c("OLS", "GAM"),  # Name the compared models.  
  mse = c(mse_ols, mse_gam),  # Compare mean squared error.  
  fit_measure = c(summary(fit_ols)$adj.r.squared, summary(fit_gam)$r.sq)  # Compare adjusted R2-style summaries.  
)  # End fit comparison.  
# Extract EDF values because they tell us how nonlinear the smooth terms actually are.  
edf_table <- as_tibble(summary(fit_gam)$s.table, rownames = "smooth_term") %>%  # Convert the GAM smooth summary into a tidy table.  
  mutate(  # Translate numeric EDF into qualitative interpretation.  
    effect_type = case_when(  # Map EDF magnitudes to practical meaning.  
      edf < 1.2 ~ "approximately linear",  # EDF near 1 suggests little gain from flexibility.  
      edf < 2 ~ "mildly nonlinear",  # Slightly above 1 suggests modest curvature.  
      TRUE ~ "clearly nonlinear"  # Larger EDF suggests substantial flexible shape.  
    )  # End interpretation rules.  
  )  # End mutate.  
backfitting_steps  # Print the conceptual backfitting table.  
fit_compare  # Print the OLS versus GAM fit summary.  
edf_table  # Print the smooth-term EDF interpretation.  
```

Backfitting is easiest to understand as an iterative residual-updating process: each smooth term gets updated while the others are held provisionally fixed, and the cycle repeats until the additive components stabilize. The model comparison then shows whether that extra flexibility actually improves fit enough to matter. In BAN404 language, a GAM is easiest to defend when it improves error metrics **and** the EDF values suggest real curvature rather than trivial near-linear effects.

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data pipeline integrity

```{r}
# Read the churn dataset because Task 2 is a binary prediction problem at the customer level.  
df_c_raw <- read_csv("../customer_data.csv", show_col_types = FALSE)  # Import the raw churn data.  
# Clean names, recode types, and remove questionable predictors because the pipeline should be deployment-ready.  
df_c <- df_c_raw %>%  # Start preparation pipeline.  
  rename_with(~ gsub("[\\. ]", "_", tolower(.x))) %>%  # Convert names to lower-case snake_case for tidyverse readability.  
  mutate(  # Recode modeling variables.  
    churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")),  # Make churn an explicit binary factor.  
    international_plan = factor(international_plan),  # Treat plan status as categorical rather than numeric.  
    voice_mail_plan = factor(voice_mail_plan),  # Same logic for the voice-mail indicator.  
    area_code = factor(area_code)  # Area code is a category, not a continuous scale.  
  ) %>%  # End type recoding.  
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)  # Remove geography and deterministic charge duplicates.  
# Set a seed because the train/test split should be reproducible.  
set.seed(404)  # Reproducible hold-out split for fair evaluation.  
# Sample training indices because all tuning and EDA should happen only on the training data.  
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))  # Draw 80% of rows for the training set.  
# Create the training set because this is where the modeling pipeline is learned.  
train_c <- df_c[idx, ]  # Training data.  
# Create the test set because it must stay untouched until final evaluation.  
test_c <- df_c[-idx, ]  # Held-out test data.  
# Summarize class balance because churn data are typically imbalanced and this affects threshold choice.  
class_split <- bind_rows(  # Start class-balance summary.  
  train = prop.table(table(train_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),  # Training-set class proportions.  
  test = prop.table(table(test_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),  # Test-set class proportions.  
  .id = "split"  # Keep the split origin visible.  
)  # End balance summary.  
class_split  # Print the class-balance table.  
```

This pipeline is defensible because it removes deterministic duplicates that would otherwise double-count the same information. It also keeps the hold-out test set untouched so that tuning and model choice do not accidentally borrow information from the final evaluation data. In BAN404 terms, this is the difference between a **realistic deployment pipeline** and a workflow that quietly peeks at the answer key.

## (b) EDA to model-hypothesis bridge

```{r}
# Build interpretable customer segments because the question asks us to bridge EDA into later model hypotheses.  
train_seg <- train_c %>%  # Start segment-construction pipeline.  
  mutate(  # Create readable bands that can translate directly into churn stories.  
    service_band = case_when(  # Group customers by customer-service-call burden.  
      customer_service_calls >= 4 ~ "4+ service calls",  # High-friction segment likely linked to dissatisfaction.  
      customer_service_calls >= 2 ~ "2-3 service calls",  # Medium-friction segment.  
      TRUE ~ "0-1 service calls"  # Low-friction segment.  
    ),  # End service-band definition.  
    day_band = case_when(  # Group customers by daytime usage intensity.  
      total_day_minutes >= quantile(total_day_minutes, 0.75) ~ "High day use",  # Heavy daytime-use segment.  
      total_day_minutes <= quantile(total_day_minutes, 0.25) ~ "Low day use",  # Light daytime-use segment.  
      TRUE ~ "Mid day use"  # Middle-usage segment.  
    )  # End day-band definition.  
  )  # End mutate.  
# Summarize churn patterns across key segments because descriptive structure should lead into modeling hypotheses.  
segment_summary <- train_seg %>%  # Start segment-level churn summary.  
  group_by(service_band, day_band, international_plan, voice_mail_plan) %>%  # Group by a simple persona structure.  
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>%  # Compute churn rate and segment size.  
  filter(n >= 30) %>%  # Keep segments large enough to discuss credibly.  
  arrange(desc(churn_rate))  # Sort so the riskiest segments appear first.  
# Turn EDA patterns into explicit model hypotheses because the prompt asks for the bridge from description to modeling.  
hypothesis_table <- tibble(  # Start hypothesis table.  
  eda_pattern = c("More service calls align with higher churn", "International-plan customers may churn more", "Heavy daytime use may mark a high-value but fragile customer segment"),  # Summarize visible descriptive patterns.  
  model_hypothesis = c("Logistic coefficient for customer_service_calls should be positive", "Both logistic odds ratios and tree-based importance may highlight international_plan", "Tree models may capture threshold or interaction effects involving total_day_minutes")  # Translate each EDA pattern into a modeling expectation.  
)  # End EDA-to-model bridge table.  
# Plot churn composition by service-call burden because this is usually one of the clearest churn signals.  
ggplot(train_seg, aes(x = service_band, fill = churn)) +  # Start stacked proportion chart.  
  geom_bar(position = "fill") +  # Show churn proportions rather than raw counts.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express the vertical axis in percentages.  
  labs(title = "Churn proportion by service-call segment", x = "Service-call segment", y = "Proportion")  # Add exam-ready labels.  
# Plot churn composition by day-use segment and international plan because this checks for a more nuanced risk story.  
ggplot(train_seg, aes(x = day_band, fill = churn)) +  # Start second proportion plot.  
  geom_bar(position = "fill") +  # Compare churn shares rather than counts.  
  facet_wrap(~ international_plan) +  # Separate the pattern by international-plan status.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express proportions as percentages.  
  labs(title = "Churn proportion by day-use band and international plan", x = "Day-use segment", y = "Proportion")  # Add clear labels.  
segment_summary  # Print the segment-level churn table.  
hypothesis_table  # Print the EDA-to-model hypothesis mapping.  
```

This chunk matters because BAN404 does not reward EDA only for decoration; it rewards EDA that creates **testable modeling expectations**. If service-call burden and international-plan status both look risky descriptively, then later coefficients or tree-based importance should either confirm or complicate that story. The limitation is that EDA is still descriptive, so apparent segment differences can reflect overlapping predictors that only a multivariate model can sort out properly.

## (c) Logistic model under threshold sweep

```{r}
# Fit logistic regression because it is the standard interpretable churn baseline in BAN404.  
logit_fit <- glm(  # Start logistic model specification.  
  churn ~ account_length + area_code + international_plan + voice_mail_plan + number_vmail_messages + total_day_minutes + total_day_calls + total_eve_minutes + total_night_minutes + total_intl_minutes + customer_service_calls,  # Use the cleaned deployment-ready predictors.  
  data = train_c,  # Fit only on the training data to preserve honest final evaluation.  
  family = binomial()  # Choose logistic regression for the binary outcome.  
)  # End logistic fit.  
# Predict test-set probabilities because threshold choice should be based on out-of-sample predictions.  
p_logit <- predict(logit_fit, newdata = test_c, type = "response")  # Test-set churn probabilities from the logistic model.  
# Compute AUC because ranking quality should be reported separately from the chosen threshold.  
logit_auc <- auc_rank(test_c$churn, p_logit)  # Test-set AUC for logistic regression.  
# Build a threshold grid because the best operational threshold is not automatically 0.5 under asymmetric costs.  
threshold_grid <- seq(0.10, 0.75, by = 0.05)  # Candidate operational thresholds.  
# Evaluate the threshold grid because sensitivity, specificity, and expected cost all depend on the chosen cutoff.  
logit_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_logit, threshold = .x))  # Threshold-by-threshold performance table.  
# Choose the cost-minimizing threshold because the prompt emphasizes business trade-offs.  
best_logit <- logit_thresholds %>% slice_min(expected_cost, n = 1)  # Logistic threshold with the lowest expected cost.  
# Convert coefficients to odds ratios because BAN404 interpretations usually rely on multiplicative odds language.  
logit_or <- tidy(logit_fit) %>%  # Tidy coefficient table.  
  mutate(odds_ratio = exp(estimate)) %>%  # Translate log-odds into odds multipliers.  
  arrange(desc(abs(estimate)))  # Sort by effect magnitude for easier interpretation.  
# Keep three interpretable terms because the prompt explicitly asks for odds-ratio interpretation.  
key_terms <- logit_or %>% filter(term %in% c("international_planYes", "voice_mail_planYes", "customer_service_calls"))  # Select three clear exam-ready coefficients.  
logit_auc  # Print the logistic AUC.  
best_logit  # Print the best threshold under the 5:1 cost structure.  
key_terms  # Print the key odds-ratio interpretations.  
```

This threshold sweep shows why a probability model and a business decision rule are not the same thing. Under a 5:1 false-negative cost ratio, the company should usually prefer a lower threshold than `0.5` because missing churners is much more expensive than contacting some safe customers. In BAN404 language, logistic regression remains useful even if a later model performs better, because it gives a transparent explanation of **which factors increase or decrease churn odds**.

## (d) Random forest robust tuning

```{r}
# Set a seed because Random Forest uses random sampling and should be reproducible.  
set.seed(504)  # Reproducible Random Forest tuning.  
# Create shared training folds because tuning must happen inside the training data only.  
rf_fold_id <- sample(rep(1:5, length.out = nrow(train_c)))  # Five-fold CV inside the training set.  
# Define the mtry grid because this hyperparameter controls feature competition at each split.  
mtry_grid <- c(2, 4, 6, 8)  # Compact but informative Random Forest tuning grid.  
# Evaluate each mtry with cross-validated AUC because ranking quality matters before thresholding.  
rf_cv <- map_dfr(mtry_grid, function(m) {  # Loop over candidate mtry values.  
  aucs <- map_dbl(1:5, function(k) {  # Loop over the training folds.  
    tr <- train_c[rf_fold_id != k, ]  # Training portion for fold k.  
    va <- train_c[rf_fold_id == k, ]  # Validation portion for fold k.  
    rf_k <- randomForest(churn ~ ., data = tr, ntree = 450, mtry = m)  # Fit the forest for this fold and mtry.  
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]  # Predict validation probabilities.  
    auc_rank(va$churn, p_k)  # Return fold-level AUC.  
  })  # End fold loop.  
  tibble(mtry = m, cv_auc = mean(aucs))  # Store mean validation AUC for this mtry value.  
})  # End mtry grid search.  
# Choose the best mtry because the final forest should use the strongest validated setting.  
best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1) %>% pull(mtry)  # Extract the best mtry value.  
# Fit the final Random Forest on all training data.  
rf_fit <- randomForest(churn ~ ., data = train_c, ntree = 600, mtry = best_mtry, importance = TRUE)  # Final tuned forest with importance tracking.  
# Score the test set because final comparison should use unseen data.  
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]  # Random Forest churn probabilities on the test set.  
# Evaluate the same threshold grid so the forest is compared fairly against logistic regression.  
rf_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_rf, threshold = .x))  # Threshold performance table for the forest.  
# Choose the cost-minimizing forest threshold.  
best_rf <- rf_thresholds %>% slice_min(expected_cost, n = 1)  # Best operational Random Forest policy.  
# Extract variable importance because tree-based interpretation is usually based on importance rather than coefficients.  
rf_imp <- importance(rf_fit) %>%  # Pull the importance matrix from the fitted forest.  
  as.data.frame() %>%  # Convert the matrix to a data frame.  
  rownames_to_column("variable") %>%  # Preserve variable names explicitly.  
  as_tibble() %>%  # Convert to tibble for tidy sorting and printing.  
  arrange(desc(MeanDecreaseGini))  # Put the most important variables first.  
rf_cv  # Print Random Forest tuning results.  
best_rf  # Print the best Random Forest threshold summary.  
rf_imp %>% slice_head(n = 10)  # Print the top variable-importance values.  
```

Random Forest can outperform logistic regression when the true churn structure contains interactions or nonlinear thresholds that the additive logit does not include automatically. At the same time, variable importance should be interpreted carefully because it tells us which variables were useful to the forest, not which variables are causal drivers of churn. In BAN404 language, Random Forest importance is best treated as a **predictive relevance ranking**, not as a substitute for a clean explanatory coefficient story.

## (e) Boosting selection with early stopping logic

```{r}
# Create numeric targets because gbm with Bernoulli loss expects a 0/1 response.  
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Training data with numeric churn indicator.  
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Test data with numeric churn indicator.  
# Create an inner train/validation split because early stopping should be learned inside the training data only.  
set.seed(604)  # Reproducible inner split for boosting.  
idx_gbm <- sample(seq_len(nrow(train_g)), size = floor(0.8 * nrow(train_g)))  # Draw rows for the inner boosting fit set.  
fit_g <- train_g[idx_gbm, ]  # Inner training subset for gbm estimation.  
val_g <- train_g[-idx_gbm, ]  # Inner validation subset for tree-count selection.  
# Fit the boosting model because sequential trees can reduce bias and capture complex churn patterns.  
gbm_fit <- gbm(  # Start gbm fit.  
  churn01 ~ . - churn,  # Use all predictive variables except the original factor outcome.  
  distribution = "bernoulli",  # Bernoulli loss is appropriate for binary churn prediction.  
  data = fit_g,  # Fit on the inner training subset only.  
  n.trees = 800,  # Fit a long boosting path so staged selection is meaningful.  
  interaction.depth = 3,  # Allow moderate interaction structure.  
  shrinkage = 0.03,  # Use a small learning rate for gradual fitting.  
  n.minobsinnode = 15,  # Avoid tiny terminal nodes that encourage overfitting.  
  verbose = FALSE  # Keep rendered output clean.  
)  # End gbm fit.  
# Evaluate validation AUC across the tree path because early stopping is a tuning decision.  
tree_grid <- seq(25, 800, by = 25)  # Candidate tree counts along the boosting path.  
val_auc_path <- map_dbl(tree_grid, function(nt) {  # Loop over candidate numbers of trees.  
  p_nt <- predict(gbm_fit, newdata = val_g, n.trees = nt, type = "response")  # Predict validation probabilities at this stage of boosting.  
  auc_rank(val_g$churn, p_nt)  # Return validation AUC for the current tree count.  
})  # End staged validation loop.  
# Choose the tree count with the strongest validation AUC because that is the early-stopping rule here.  
best_trees <- tree_grid[which.max(val_auc_path)]  # Extract the best stage of the boosting path.  
# Score the untouched test set using the chosen number of trees.  
p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")  # Final boosting probabilities on the test set.  
# Evaluate the same threshold grid so all three models are compared on a common decision basis.  
boost_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_boost, threshold = .x))  # Threshold comparison for boosting.  
# Choose the cost-minimizing boosting threshold.  
best_boost <- boost_thresholds %>% slice_min(expected_cost, n = 1)  # Best operational boosting policy.  
# Combine the best policy from each model into one final scorecard.  
comparison <- bind_rows(  # Start final model-comparison table.  
  best_logit %>% mutate(model = "Logistic"),  # Logistic model at its best threshold.  
  best_rf %>% mutate(model = paste0("Random Forest (mtry=", best_mtry, ")")),  # Random Forest at its best threshold.  
  best_boost %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))  # Boosting at its best threshold.  
) %>%  # End row binding.  
  select(model, threshold, accuracy, sensitivity, specificity, auc, expected_cost) %>%  # Keep the decision-relevant metrics.  
  arrange(expected_cost, desc(auc))  # Rank primarily by expected cost and secondarily by AUC.  
# Plot the staged AUC path because early stopping is easier to justify visually than from one number alone.  
ggplot(tibble(trees = tree_grid, auc = val_auc_path), aes(x = trees, y = auc)) +  # Start staged validation plot.  
  geom_line(color = "purple") +  # Draw validation AUC across the boosting path.  
  geom_vline(xintercept = best_trees, color = "darkgreen", linetype = "dashed") +  # Mark the selected early-stopping point.  
  labs(title = "Boosting validation AUC path", x = "Number of trees", y = "Validation AUC")  # Add clear labels for interpretation.  
comparison  # Print the final model scorecard.  
```

This final scorecard shows how early stopping works in practice: we do not keep adding trees forever, but stop at the stage that performs best on validation data. That logic matters because boosting can continue to fit finer and finer patterns, which increases overfitting risk if the path is allowed to run unchecked. In BAN404 terms, the recommended deployment model should be the one that best balances **expected cost**, **discrimination quality**, and **a threshold policy aligned with the 5:1 false-negative penalty**.
