---
title: "BAN404 Mock Exam 2026 #5"
subtitle: "Comprehensive Final Rehearsal"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Identify methods from function snippets (10 points)

You are given multiple objective snippets from curriculum (`Task1_comp.qmd`). Match each snippet to method and state one strength/weakness.

## (b) Full regularization workflow (10 points)

Fit OLS, Ridge, LASSO, and Elastic Net. Choose one using CV performance and model simplicity.

```{r}
# Your code here
```

## (c) Combine tuning and uncertainty evidence (10 points)

Use CV for model tuning and bootstrap for uncertainty to justify a robust final Task 1 model.

```{r}
# Your code here
```

## (d) GAM enhancement + residual diagnostics (10 points)

Assess whether nonlinearity handling with GAM meaningfully improves over linear baselines.

```{r}
# Your code here
```

## (e) Exam-style interpretation summary (10 points)

Write a concise conclusion that links method choice, assumptions, and expected generalization behavior.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Build a production-ready churn dataset (8 points)

Prepare and split the data with clear justification of removals and factor handling.

```{r}
# Your code here
```

## (b) Risk-pattern EDA and intervention ideas (8 points)

Use EDA to propose two concrete retention actions and how they map to predictors.

```{r}
# Your code here
```

## (c) Logistic baseline and threshold selection (12 points)

Train logistic regression, evaluate threshold grid, and select an operating point.

```{r}
# Your code here
```

## (d) Ensemble comparison: RF vs boosting (12 points)

Tune both models, compare hold-out metrics, and explain why one may dominate under class imbalance.

```{r}
# Your code here
```

## (e) Final exam recommendation memo (10 points)

Provide a final model recommendation with metric evidence and business-constraint reasoning.

```{r}
# Your code here
```
