---
title: "BAN404 Mock Exam 2026 #7"
subtitle: "LASSO Sparsity and Cost-Sensitive Churn"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)
- **Important exam expectation:** In every subquestion, combine **method choice + runnable code + interpretation + theory/business reasoning**. A strong BAN404 answer explains why a sparse model may or may not be preferable, how validation evidence supports the choice, and what business loss structure the final churn decision is built around.

# How to use this mock exam

This mock exam is written to function both as a **practice exam** and as a **lookup tool for decision-ready exam writing**. That means each subquestion tells you what to calculate, what kind of theoretical justification the examiner is likely to reward, and what trade-off you are expected to comment on. If you solve it properly, you should practice not only **running the analysis**, but also **writing short BAN404-style explanations** about selection, uncertainty, model fairness, and deployment logic.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## Context for Task 1

This task is centered on **sparsity, defensible simplification, and model-selection evidence**. In BAN404, a strong LASSO answer is not only about obtaining zero coefficients, but about explaining **why variable selection happened**, whether the chosen penalty is **stable enough to trust**, and whether the simpler model is genuinely easier to defend than richer alternatives. You should therefore connect **penalty mechanics**, **cross-validation**, **bootstrap evidence**, and **nonlinear benchmarks** into one coherent answer.

## (a) Explain why LASSO can simplify a regression answer (10 points)

You are told that a penalized least-squares criterion adds a penalty proportional to the **sum of the absolute values of the coefficients**.

What to do:

- Identify the method and explain how you can recognize it.
- Explain why the absolute-value penalty can set some coefficients exactly to zero.
- Explain why standardization matters before penalization.
- Discuss when sparsity can improve an exam answer and when it can become misleading.
- Explain how correlated predictors can complicate interpretation of the selected model.

What the examiner wants to see:

- correct recognition of **LASSO regression**,
- awareness that LASSO changes the **bias-variance trade-off**,
- awareness that variable selection is not the same as proving a variable is truly unimportant,
- theory-grounded discussion rather than only naming the method.

```{r}
# Your code here
```

## (b) Compare OLS, LASSO, and Elastic Net under a shared workflow (10 points)

Build a fair comparison between **OLS**, **LASSO**, and **Elastic Net**.

What to do:

- Fit OLS as the unpenalized baseline.
- Fit LASSO and Elastic Net using the same fold structure.
- Report `lambda.min` and `lambda.1se` for the penalized models.
- Compare coefficient behavior at the selected penalty choices.
- Report how many predictors remain active under each penalized fit.
- Explain whether the more selective model seems preferable or whether the extra stability of Elastic Net is easier to defend.

What the examiner wants to see:

- a fair comparison built on shared validation logic,
- awareness that LASSO emphasizes sparsity while Elastic Net can behave more stably with correlated predictors,
- interpretation of both performance and coefficient behavior,
- a justified model choice rather than a number-only answer.

```{r}
# Your code here
```

## (c) Study sparse-model stability rather than only one chosen lambda (10 points)

Use validation and resampling to study whether the selected sparse model is stable enough to trust.

What to do:

- Use repeated K-fold CV to study variation in the selected `lambda.min`.
- Summarize how much the chosen penalty moves across repetitions.
- Use a bootstrap procedure to study how often predictors remain active in the LASSO solution.
- Identify which predictors look highly stable and which look fragile.
- Explain why selection stability matters for interpretation.
- State what kind of overly confident exam wording would be risky here.

A strong answer should make clear that a sparse model may look neat, while still being unstable across plausible samples or splits.

```{r}
# Your code here
```

## (d) Test whether simple nonlinearity can beat a sparse linear answer (10 points)

Compare a sparse linear answer to a more flexible nonlinear alternative.

What to do:

- Fit the chosen sparse linear model.
- Fit a polynomial or basis-expansion benchmark.
- Fit a GAM with smooth terms for plausible nonlinear predictors.
- Compare MSE and an R²-style fit measure.
- Explain whether the extra flexibility seems substantively useful or mainly cosmetic.
- State which version would be easiest to defend in a time-pressured exam answer.

What the examiner wants to see:

- awareness that sparsity is not automatically best if the functional form is wrong,
- correct interpretation of smooth/nonlinear model evidence,
- discussion of complexity versus defendability,
- a final recommendation based on fit **and** clarity.

```{r}
# Your code here
```

## (e) Write the final Task 1 recommendation paragraph (10 points)

End Task 1 with a short **exam-ready recommendation paragraph**.

What to include:

- which model you would report as the main answer,
- whether you would emphasize `lambda.min` or `lambda.1se`,
- what the stability evidence says about your confidence,
- whether nonlinearity changes your conclusion,
- how you would defend the final answer in one polished BAN404 paragraph.

The best answers sound like the final paragraph of a strong submission, not a list of outputs.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## Context for Task 2

This task focuses on **retention targeting under asymmetric decision costs**. You should keep separate the ideas of **descriptive churn patterns**, **probability ranking quality**, and **final intervention policy**. In this mock exam, assume that a **false negative costs 5 times as much as a false positive**, so the final threshold choice should be designed to catch risky customers rather than simply maximize default accuracy.

## (a) Build a deployment-aware churn dataset (8 points)

Prepare the churn data so later models can be defended as realistic and leakage-aware.

What to do:

- Read and clean the data.
- Recode categorical variables appropriately.
- Remove deterministic or duplicated predictors.
- Exclude or justify fields that are weak from an intervention or deployment perspective.
- Perform an 80/20 train/test split with a fixed seed.
- Explain why the test data must remain untouched until final model comparison.

What the examiner wants to see:

- sensible factor handling,
- awareness of duplicated information,
- awareness of deployment realism,
- correct train/test discipline.

```{r}
# Your code here
```

## (b) Use EDA to build retention personas and hypotheses (8 points)

Use the training data to create **retention personas** and translate them into model hypotheses.

What to do:

- Quantify class imbalance.
- Create grouped churn summaries for meaningful customer profiles.
- Produce at least two informative plots.
- Describe one high-risk persona and one moderate-risk persona.
- Propose at least two intervention ideas linked to the descriptive patterns.
- Mention one limitation of persona-based EDA.

This is not only a plotting task. It is a reasoning task where descriptive churn patterns become **candidate model stories** and **candidate business actions**.

```{r}
# Your code here
```

## (c) Build an interpretable logistic decision rule (12 points)

Fit a logistic regression model and choose an operating threshold under the **5:1 false-negative cost ratio**.

What to do:

- Fit logistic regression on the training data.
- Predict probabilities on the test data.
- Report a ranking metric such as AUC.
- Evaluate a grid of thresholds.
- Report sensitivity, specificity, and expected cost.
- Interpret at least three coefficients in odds-ratio language.
- Explain why the business-optimal threshold is not automatically `0.5`.

A strong answer should clearly separate **ranking quality** from **action policy**.

```{r}
# Your code here
```

## (d) Compare flexible ensembles against the logistic baseline (12 points)

Use **Random Forest** and **boosting** to test whether more flexible models improve the churn solution.

What to do:

- Tune Random Forest using the training data only.
- Fit a boosting model and choose the number of trees using staged validation logic.
- Compare all models on the hold-out test set.
- Report AUC and cost-sensitive threshold performance.
- Compare tree-based importance patterns with the earlier personas.
- Explain what the flexible models may capture that logistic regression may miss.

The best answers explain that improved prediction may come at the cost of simpler coefficient-based interpretation.

```{r}
# Your code here
```

## (e) Write the final retention-policy memo (10 points)

Provide a final recommendation memo for management.

Your memo should explicitly state:

1. which model you recommend,
2. which threshold you would deploy,
3. which metric or objective mattered most,
4. what trade-off the company is accepting,
5. what you would monitor after deployment.

A strong answer should sound like the closing paragraph of a polished BAN404 submission.

```{r}
# Your code here
```
