---
title: "BAN404 Mock Exam 2026 #4"
subtitle: "Validation Discipline and Method Justification"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)
- **Important exam expectation:** In every subquestion, combine **method choice + runnable code + interpretation + theory/business reasoning**. A BAN404 answer earns more points when it explains *why* a method is appropriate, *how* the validation is kept fair, and *what* the output means for decision-making.

# How to use this mock exam

This mock exam is written as both a **practice exam** and a **lookup tool**. Each subquestion therefore tells you not only what to compute, but also what the examiner is likely to reward in a strong answer. When you solve it, aim to write in a way that would still help you two minutes before the real exam: short method justification, clear code, and direct interpretation.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## Context for Task 1

This task is centered on **validation discipline**, **leakage prevention**, and **method justification**. That means the examiner is not only asking whether you can fit Ridge, LASSO, and GAM, but whether you understand **why a validation procedure is fair**, **why preprocessing must happen inside folds**, and **how uncertainty summaries should be defended**. A strong BAN404 answer therefore connects code to the bias-variance trade-off, honest model evaluation, and interpretation quality.

## (a) Explain K-fold CV pseudocode and leakage points (10 points)

Describe a standard **K-fold cross-validation loop** and identify where test-information leakage can occur.

What to explain:

1. **How the data are split into folds**.
2. **What happens inside each repetition of the loop**.
3. **Which object plays the role of training data and which plays the role of validation data**.
4. **Where preprocessing, feature engineering, or tuning can accidentally leak information**.
5. **Why leakage produces overly optimistic performance estimates rather than genuinely better models**.

A strong answer should sound like a BAN404 methodology paragraph, not just a list of steps.

## (b) Leakage-safe scaling inside folds (10 points)

Implement **fold-specific scaling** for Ridge regression and compare it with a naive global-scaling approach.

What to do:

- Create one common fold assignment.
- Compare a **wrong** workflow where scaling is done on the full dataset before cross-validation.
- Compare a **correct** workflow where scaling parameters are learned inside each training fold and then applied to its validation fold.
- Tune Ridge over a lambda grid.
- Compare the best validation MSE from the two procedures.
- Explain why the globally scaled workflow is not trustworthy as an evaluation strategy.

What the examiner wants to see:

- correct identification of leakage,
- fold-by-fold thinking,
- awareness that preprocessing is part of the modeling pipeline,
- explicit explanation that optimistic validation is not the same as real predictive improvement.

```{r}
# Your code here
```

## (c) Compare tuned Ridge and LASSO under common folds (10 points)

Use the **same fold assignment** to compare Ridge and LASSO fairly.

What to do:

- Reuse one common set of folds.
- Tune Ridge and LASSO with those identical folds.
- Compare `lambda.min`, `lambda.1se`, and best CV-MSE.
- Compare coefficient sparsity at the chosen lambda values.
- Explain why “same folds” matters when making a fair method comparison.

A strong answer should make it clear that if the folds differ, some of the performance difference may come from the split rather than from the method itself.

```{r}
# Your code here
```

## (d) Bootstrap CI interpretation quality check (10 points)

Use the bootstrap to estimate uncertainty for **`Var(y)`**.

What to do:

- Set a seed.
- Use many bootstrap resamples.
- Store `Var(y)` for each resample.
- Visualize the bootstrap distribution.
- Compute a **percentile confidence interval** and a **normal approximation confidence interval**.
- Explain when the percentile interval is easier to defend than the normal approximation.

The best answers explicitly link the interval choice to the **shape of the bootstrap distribution**.

```{r}
# Your code here
```

## (e) Backfitting intuition and GAM implementation (10 points)

Explain the intuition behind **backfitting** and assess whether a GAM improves on a linear baseline.

What to do:

- Give a short intuition for what backfitting is doing conceptually.
- Fit a linear baseline.
- Fit a GAM with smooth terms for plausible nonlinear predictors.
- Compare MSE and an R²-style fit measure.
- Interpret the estimated smoothness of the terms.
- Decide whether the added flexibility seems justified.

A strong answer should explain not only whether GAM fits better, but also what that says about **nonlinearity** and **model complexity**.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## Context for Task 2

This task focuses on **pipeline integrity**, **EDA-to-model reasoning**, and **cost-sensitive decision-making**. You are expected to keep separate the ideas of **descriptive insight**, **predictive discrimination**, and **operational threshold choice**. In this mock exam, assume that a **false negative is 5 times as costly as a false positive**, so the final threshold policy should reflect that business asymmetry.

## (a) Data pipeline integrity (8 points)

Create a clean churn-modeling pipeline with no peeking and no clearly post-outcome leakage variables.

What to do:

- Read and clean the data.
- Recode variables into suitable types.
- Remove duplicated or deterministic predictors.
- Exclude variables you cannot defend from a realistic deployment perspective.
- Perform an 80/20 train/test split with a fixed seed.
- Explain why the split must happen before tuning and final evaluation.

What the examiner wants:

- sensible factor handling,
- awareness of duplicated information,
- awareness of deployment realism,
- a clean separation between training work and final testing.

```{r}
# Your code here
```

## (b) EDA to model-hypothesis bridge (8 points)

Use the training data to produce **compact EDA outputs** and map each one to a modeling hypothesis.

What to do:

- Quantify class imbalance.
- Produce grouped summaries for meaningful churn segments.
- Create at least two informative plots.
- State at least three model hypotheses suggested by the EDA.
- Explain how those hypotheses might appear later in logistic regression or tree ensembles.
- Mention one limitation of the descriptive stage.

This is not only an EDA task. It is a reasoning task where descriptive patterns become **candidate modeling stories**.

```{r}
# Your code here
```

## (c) Logistic model under threshold sweep (12 points)

Fit a logistic regression model and evaluate it across a **threshold grid**.

What to do:

- Fit the model on the training set.
- Predict probabilities on the test set.
- Evaluate a grid of thresholds.
- Report sensitivity, specificity, AUC, and expected cost.
- Choose one threshold under the **5:1 false-negative cost ratio**.
- Interpret at least three coefficients in odds-ratio language.

A strong answer should make it explicit that a threshold of `0.5` is not automatically the best business choice.

```{r}
# Your code here
```

## (d) Random forest robust tuning (12 points)

Tune and evaluate a Random Forest, then discuss the limits of variable importance.

What to do:

- Tune `mtry` using CV on the training data only.
- Fit the final Random Forest.
- Evaluate the test-set AUC and threshold-based performance.
- Show the most important variables.
- Compare the Random Forest story with the logistic story.
- Explain why variable importance is useful but not the same as a causal explanation.

The best answers note that tree-based importance can be informative while still being affected by correlated predictors, split structure, and model instability.

```{r}
# Your code here
```

## (e) Boosting selection with early stopping logic (10 points)

Fit a boosting model and choose the final number of trees using staged evaluation.

What to do:

- Fit a boosting model on the training data.
- Use a validation strategy inside the training set to choose the number of trees.
- Compare logistic regression, Random Forest, and boosting.
- Choose a threshold policy under the **5:1 false-negative cost structure**.
- Give a final deployment recommendation.

Your final paragraph should explicitly state:

1. which model you recommend,
2. which metric or objective matters most,
3. how early stopping reduces overfitting risk,
4. what trade-off the company accepts when it lowers the threshold to catch more churners.

```{r}
# Your code here
```
