---
title: "BAN404 Mock Exam 2026 #1"
subtitle: "Ridge-first Generalization and Churn Operations"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)
- **Important exam expectation:** In every subquestion, combine **code + explanation + interpretation**. A good BAN404 answer does not stop after fitting a model; it also explains why the method is relevant, what assumptions matter, and what the results imply.

# How to use this mock exam

This mock exam is written as both a **practice exam** and a **revision lookup tool**. That means each task tells you not only what to compute, but also what the examiner is usually looking for in a strong written answer. When you solve it, aim to produce short theory-grounded paragraphs after the code, not only output tables.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## Context for Task 1

The dataset contains a continuous response `y` and several predictors. The structure makes this a good environment for discussing **OLS**, **regularization**, **cross-validation**, **bootstrap uncertainty**, and **GAM-based nonlinearity handling**. Your answers should therefore show both technical execution and methodological judgment.

## (a) Decode the objective: Ridge variant (10 points)

You are given the following function:

```{r}
f <- function(beta, X, y, lambda) {
  sum((y - X %*% beta)^2) + lambda * sum(beta^2)
}
```

Write a short but precise answer that explains all of the following:

1. **Which method this represents** and how you know.
2. **What the two parts of the objective function do**.
3. **What happens when `lambda` increases**, especially to coefficient size, variance, and bias.
4. **Why this can help under multicollinearity**, and how it differs from LASSO.

Your answer should read like a BAN404 theory answer, not just a one-line identification.

## (b) Baseline vs shrinkage fit comparison (10 points)

Read `task1_data.csv` and compare an **OLS model** with a **Ridge model**.

What to do:

- Fit OLS using all available predictors.
- Fit Ridge using `glmnet`.
- Compare coefficient magnitudes in a clear table.
- Comment on whether the Ridge coefficients look more stable / shrunk than OLS.
- Explain what this tells you about the possible presence of correlated predictors.

What the examiner wants to see:

- correct data setup,
- correct distinction between OLS and penalized regression,
- interpretation of shrinkage,
- clear statement of *why* Ridge is useful even if it does not perform variable selection.

```{r}
# Your code here
```

## (c) Ten-fold CV tuning with robust choice rule (10 points)

Tune `lambda` using **10-fold cross-validation**.

What to do:

- Use `cv.glmnet` on the same data.
- Report both `lambda.min` and `lambda.1se`.
- Explain the practical meaning of the two rules.
- State which one you would choose in an exam/business setting where robustness matters.
- Briefly explain why CV is more trustworthy than one random train/test split for this tuning decision.

You should also mention what kind of **data leakage mistake** would make CV unfair.

```{r}
# Your code here
```

## (d) Bootstrap uncertainty for variance (10 points)

Use the bootstrap to study uncertainty in `Var(y)`.

What to do:

- Set a seed.
- Use `B = 1500` bootstrap resamples.
- Store the bootstrap variance estimate in each repetition.
- Show the bootstrap distribution in a histogram.
- Compute both a **percentile CI** and a **normal approximation CI**.
- Explain when the percentile CI is often preferred.

Your interpretation should explicitly connect the shape of the bootstrap distribution to your choice of confidence interval language.

```{r}
# Your code here
```

## (e) Linear vs additive model improvement (10 points)

Fit both an **OLS model** and a **GAM**.

What to do:

- Fit a full linear model.
- Fit a GAM where the most plausible nonlinear predictors are modeled with smooth terms.
- Compare the models using MSE and adjusted / GAM-style R².
- Explain whether the smooth terms seem justified.
- Comment on the trade-off between flexibility and interpretability.

A strong answer does not only say which model fits better; it also explains *why* BAN404 might reward choosing the more flexible model or sticking with the simpler one.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## Context for Task 2

This is a classic BAN404 binary prediction task. A strong answer should move through the full workflow: **data design → EDA → interpretable baseline → more flexible model → recommendation**. You should keep separating *prediction usefulness* from *business usefulness* throughout the task.

## (a) Prediction-ready data design (8 points)

Prepare the churn data so it is suitable for realistic prediction.

What to do:

- Read and clean the data.
- Recode variables to appropriate types.
- Remove duplicated/deterministic predictors that can create instability.
- Explain which variables are safe to use at prediction time.
- Split the data into training and test sets using a fixed seed.
- Comment briefly on class balance in the two splits.

What the examiner wants:

- sensible factor handling,
- awareness of leakage / deployment realism,
- awareness of multicollinearity,
- a clean split before later modeling.

```{r}
# Your code here
```

## (b) EDA for churn risk framing (8 points)

Use the training data only to build an initial explanation of churn.

What to do:

- Create grouped churn-rate summaries.
- Make at least two informative plots.
- Identify three plausible churn drivers.
- Explain how EDA supports these hypotheses.
- Mention one weakness of using only pairwise summaries before multivariate modeling.

This is not only a plotting task. It is a **thinking task** where you show how descriptive patterns become modeling hypotheses.

```{r}
# Your code here
```

## (c) Logistic model with FN-cost thresholding (12 points)

Fit a logistic regression model and choose a threshold when a **false negative costs 3 times as much as a false positive**.

What to do:

- Fit the model on the training set.
- Predict probabilities on the test set.
- Evaluate several thresholds.
- Report confusion-based metrics and AUC.
- Choose one threshold and defend it in business language.
- Interpret at least a few coefficients using odds / odds ratios.

A strong answer should explain why the best threshold is not necessarily `0.5`.

```{r}
# Your code here
```

## (d) Random forest hyperparameter tuning (12 points)

Fit and tune a Random Forest model.

What to do:

- Tune `mtry` using CV on the training set only.
- Fit the final forest with the chosen `mtry`.
- Evaluate performance on the test set.
- Show the most important predictors.
- Compare the importance ranking with the hypotheses from your EDA and the story from logistic regression.

Your written answer should explain why Random Forest can outperform logistic regression when nonlinearities or interactions matter.

```{r}
# Your code here
```

## (e) Boosting selection and deployment recommendation (10 points)

Fit a boosted-tree model and compare it with the earlier models.

What to do:

- Fit a boosting model on the training data.
- Evaluate staged performance over different numbers of trees.
- Pick an appropriate number of trees.
- Compare logistic regression, Random Forest, and boosting.
- Give one final recommendation for deployment in a retention context.

The final paragraph should clearly state:

1. which model you recommend,
2. why that model is preferred,
3. which metric matters most for this business problem,
4. what trade-off the company is accepting.

```{r}
# Your code here
```
