---
title: "BAN404 – Predicted Exam 2026 — Model Solution (R Version)"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Setup

```{r}
# Load BAN404-core libraries used across the exam.
library(tidyverse)
library(janitor)
library(glmnet)
library(gam)
library(randomForest)
library(gbm)

# Helper: ranking-based AUC so we avoid extra package dependencies.
auc_rank <- function(y_true, score, positive = "Yes") {
  y_chr <- as.character(y_true)
  y_bin <- if_else(y_chr == positive, 1, 0)
  n1 <- sum(y_bin == 1)
  n0 <- sum(y_bin == 0)
  if (n1 == 0 || n0 == 0) return(NA_real_)
  r <- rank(score)
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

# Helper: confusion metrics from probabilities and threshold.
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {
  y_chr <- as.character(y_true)
  if (!(positive %in% unique(y_chr))) stop("Positive level not found in y_true.")
  negative <- setdiff(sort(unique(y_chr)), positive)[1]

  actual <- factor(y_chr, levels = c(negative, positive))
  pred_chr <- if_else(p_hat >= threshold, positive, negative)
  pred <- factor(pred_chr, levels = c(negative, positive))

  tab <- table(actual = actual, pred = pred)
  tn <- tab[negative, negative]
  fp <- tab[negative, positive]
  fn <- tab[positive, negative]
  tp <- tab[positive, positive]

  tibble(
    accuracy = (tp + tn) / sum(tab),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    auc = auc_rank(actual, p_hat, positive = positive)
  )
}
```

# Task 1 — Methodology

## (a) Function understanding

```{r}
f <- function(beta, X, y, lambda) {
  rss <- sum((y - X %*% beta)^2)
  penalty <- lambda * sum(abs(beta))
  rss + penalty
}

g <- function(X, y, lambda) {
  p <- ncol(X)
  b0 <- mean(y)
  y_c <- y - b0
  X_c <- sweep(X, 2, colMeans(X), "-")

  fit <- optim(
    par = rep(0, p),
    fn = f,
    X = X_c,
    y = y_c,
    lambda = lambda,
    method = "BFGS"
  )

  list(intercept = b0, beta = fit$par)
}
```

This is a manual LASSO-style objective (L1 penalty), where larger `lambda` shrinks slope coefficients toward zero.

## (b) OLS vs regularised models

```{r}
xy <- read_csv("task1_data.csv", show_col_types = FALSE) %>% clean_names()
X <- as.matrix(select(xy, x1:x6))
y <- xy$y

# OLS baseline.
ols <- lm(y ~ x1 + x2 + x3 + x4 + x5 + x6, data = xy)
ols_beta <- coef(ols)[-1]

# Penalised fits.
lasso_1 <- g(X, y, lambda = 1)
lasso_100 <- g(X, y, lambda = 100)

coef_table <- tibble(
  variable = paste0("x", 1:6),
  OLS = as.numeric(ols_beta),
  Lasso_lambda_1 = as.numeric(lasso_1$beta),
  Lasso_lambda_100 = as.numeric(lasso_100$beta)
)

coef_table
```

As expected, coefficients with `lambda=100` are more shrunk than with `lambda=1`, especially weak predictors.

## (c) 5-fold CV for lambda

```{r}
cv_lasso_manual <- function(X, y, lambda, K = 5, seed = 42) {
  set.seed(seed)
  n <- length(y)
  fold_id <- sample(rep(1:K, length.out = n))
  mse <- numeric(K)

  for (k in 1:K) {
    te <- which(fold_id == k)
    tr <- which(fold_id != k)

    fit <- g(X[tr, , drop = FALSE], y[tr], lambda)
    X_te_c <- sweep(X[te, , drop = FALSE], 2, colMeans(X[tr, , drop = FALSE]), "-")
    y_hat <- fit$intercept + X_te_c %*% fit$beta
    mse[k] <- mean((y[te] - as.numeric(y_hat))^2)
  }

  mean(mse)
}

lambda_grid <- 10^seq(-1, 3, length.out = 30)
cv_tbl <- tibble(
  lambda = lambda_grid,
  cv_mse = map_dbl(lambda_grid, ~ cv_lasso_manual(X, y, lambda = .x, K = 5, seed = 42))
)

best_lambda <- cv_tbl %>% slice_min(cv_mse, n = 1)
best_lambda

ggplot(cv_tbl, aes(x = lambda, y = cv_mse)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.2) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "5-fold CV MSE by lambda", x = "lambda (log scale)", y = "CV MSE")
```

K-fold is computationally cheaper than LOOCV for n=5000 and avoids leakage by centering test folds with training means.

## (d) Bootstrap variance

```{r}
bootstrap_var <- function(y, B = 1000, seed = 1) {
  set.seed(seed)
  n <- length(y)
  boot <- numeric(B)

  for (b in 1:B) {
    idx <- sample(1:n, n, replace = TRUE)
    boot[b] <- var(y[idx])
  }

  list(
    boot = boot,
    se = sd(boot),
    ci = quantile(boot, c(0.025, 0.975))
  )
}

boot_out <- bootstrap_var(y, B = 1000, seed = 1)
var_full <- var(y)

boot_out$se
boot_out$ci

# Plot bootstrap sampling distribution.
tibble(v = boot_out$boot) %>%
  ggplot(aes(x = v)) +
  geom_histogram(bins = 45, fill = "steelblue", color = "white") +
  geom_vline(xintercept = as.numeric(boot_out$ci), color = "red", linetype = "dashed") +
  geom_vline(xintercept = var_full, color = "black") +
  theme_minimal() +
  labs(title = "Bootstrap distribution of Var(y)", x = "Bootstrap variance", y = "Count")
```

## (e) Non-linearity and GAM

```{r}
# Plot y against predictors to inspect non-linear patterns.
xy %>%
  pivot_longer(cols = starts_with("x"), names_to = "x_var", values_to = "x_value") %>%
  ggplot(aes(x = x_value, y = y)) +
  geom_point(alpha = 0.08) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  facet_wrap(~x_var, scales = "free_x") +
  theme_minimal() +
  labs(title = "Non-linearity check", x = "Predictor value", y = "y")

# Compare linear OLS with flexible GAM.
ols_hat <- predict(ols, newdata = xy)
gam_fit <- gam(y ~ x1 + s(x2, df = 4) + s(x3, df = 4) + x4 + x5 + x6, data = xy)
gam_hat <- predict(gam_fit, newdata = xy)

mse_null <- mean((y - mean(y))^2)
mse_ols <- mean((y - ols_hat)^2)
mse_gam <- mean((y - gam_hat)^2)

metrics_t1 <- tibble(
  model = c("OLS", "GAM"),
  mse = c(mse_ols, mse_gam),
  r_squared = 1 - c(mse_ols, mse_gam) / mse_null
)

metrics_t1
```

GAM typically improves fit because this dataset includes non-linear signal (especially around x2/x3).

# Task 2 — Customer churn

## (a) Data preparation

```{r}
cust <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()

# Recode binary text fields to factors and create cleaner labels.
cust2 <- cust %>%
  mutate(
    churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    international_plan = factor(international_plan),
    voice_mail_plan = factor(voice_mail_plan),
    area_code = factor(area_code)
  )

# Remove perfect charge duplicates (minutes already carry identical info).
cust2 <- cust2 %>%
  select(-total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)

# 80/20 split with reproducible seed.
set.seed(42)
idx <- sample(seq_len(nrow(cust2)), size = floor(0.8 * nrow(cust2)))
train <- cust2[idx, ]
test <- cust2[-idx, ]

# Quick class-balance check.
prop.table(table(train$churn))
prop.table(table(test$churn))
```

## (b) EDA

```{r}
# Churn rate by key categorical predictors.
train %>%
  group_by(international_plan) %>%
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop")

train %>%
  group_by(voice_mail_plan) %>%
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop")

# Churn rate by service calls (often a strong exam variable).
train %>%
  group_by(customer_service_calls) %>%
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>%
  ggplot(aes(x = customer_service_calls, y = churn_rate)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Churn rate by customer service calls", x = "Calls", y = "Churn rate")
```

## (c) Logistic regression

```{r}
# Fit interpretable logistic regression.
logit_fit <- glm(
  churn ~ account_length + international_plan + voice_mail_plan +
    number_vmail_messages + total_day_minutes + total_eve_minutes +
    total_night_minutes + total_intl_minutes + customer_service_calls,
  data = train,
  family = binomial()
)

# Probability predictions and performance metrics on held-out test data.
p_logit <- predict(logit_fit, newdata = test, type = "response")
logit_m <- class_metrics(test$churn, p_logit, threshold = 0.5)
logit_m

# Coefficient interpretation table (odds ratios).
tidy_logit <- broom::tidy(logit_fit) %>%
  mutate(odds_ratio = exp(estimate))

tidy_logit
```

## (d) Random forest

```{r}
# Candidate mtry values for simple train-set CV.
mtry_grid <- c(2, 4, 6, 8)

set.seed(42)
cv_fold <- sample(rep(1:5, length.out = nrow(train)))

cv_rf <- map_dfr(mtry_grid, function(m) {
  aucs <- map_dbl(1:5, function(k) {
    tr <- train[cv_fold != k, ]
    va <- train[cv_fold == k, ]

    rf_k <- randomForest(churn ~ . - state, data = tr, ntree = 500, mtry = m)
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]
    auc_rank(va$churn, p_k)
  })

  tibble(mtry = m, cv_auc = mean(aucs))
})

cv_rf
best_mtry <- cv_rf %>% slice_max(cv_auc, n = 1) %>% pull(mtry)

rf_fit <- randomForest(churn ~ . - state, data = train, ntree = 500, mtry = best_mtry, importance = TRUE)
p_rf <- predict(rf_fit, newdata = test, type = "prob")[, "Yes"]
rf_m <- class_metrics(test$churn, p_rf, threshold = 0.5)
rf_m

# Variable importance plot.
imp <- importance(rf_fit) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  as_tibble() %>%
  arrange(desc(MeanDecreaseGini))

imp %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Random forest variable importance", x = "Variable", y = "Mean decrease Gini")
```

## (e) Boosting

```{r}
# gbm requires numeric target for Bernoulli (0/1).
train_gbm <- train %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))
test_gbm <- test %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))

set.seed(42)
gbm_fit <- gbm(
  formula = churn01 ~ . - churn - state,
  distribution = "bernoulli",
  data = train_gbm,
  n.trees = 500,
  interaction.depth = 3,
  shrinkage = 0.05,
  n.minobsinnode = 10,
  verbose = FALSE
)

# Evaluate AUC across staged tree counts.
tree_grid <- seq(10, 500, by = 10)
auc_path <- map_dbl(tree_grid, function(b) {
  p_b <- predict(gbm_fit, newdata = test_gbm, n.trees = b, type = "response")
  auc_rank(test$churn, p_b)
})

best_trees <- tree_grid[which.max(auc_path)]

boost_tbl <- tibble(trees = tree_grid, test_auc = auc_path)
boost_tbl %>% slice_max(test_auc, n = 1)

boost_tbl %>%
  ggplot(aes(x = trees, y = test_auc)) +
  geom_line(color = "purple") +
  geom_vline(xintercept = best_trees, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Boosting staged AUC", x = "Number of trees", y = "Test AUC")

p_boost <- predict(gbm_fit, newdata = test_gbm, n.trees = best_trees, type = "response")
boost_m <- class_metrics(test$churn, p_boost, threshold = 0.5)
boost_m
```

## (f) Comparison and recommendation

```{r}
summary_table <- bind_rows(
  logit_m %>% mutate(model = "Logistic"),
  rf_m %>% mutate(model = paste0("Random forest (mtry=", best_mtry, ")")),
  boost_m %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))
) %>%
  select(model, accuracy, auc, sensitivity, specificity)

summary_table
```

Recommendation: use a tree ensemble (RF/Boosting) for highest predictive performance and logistic regression for transparent explanation of churn drivers.
