---
title: "BAN404 Mock Exam 2026 #2"
subtitle: "LASSO Sparsity and Cost-Sensitive Churn"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Decode LASSO optimization code (10 points)

Explain why this objective can perform variable selection:

```{r}
f <- function(beta, X, y, lambda) {
  sum((y - X %*% beta)^2) + lambda * sum(abs(beta))
}
```

## (b) OLS vs LASSO coefficient path analysis (10 points)

Fit OLS and LASSO over a lambda grid and identify predictors that shrink to zero earliest.

```{r}
# Your code here
```

## (c) Repeated K-fold tuning and LOOCV discussion (10 points)

Use repeated 5-fold CV for lambda tuning and explain why this can be more stable than a single fold split.

```{r}
# Your code here
```

## (d) Bootstrap uncertainty for SD(y) (10 points)

Bootstrap SD(y) with `B = 1200` and compare normal and percentile CIs.

```{r}
# Your code here
```

## (e) Polynomial vs GAM flexibility check (10 points)

Compare a polynomial regression with a GAM and discuss bias-variance tradeoff.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Leakage-safe preprocessing and split (8 points)

Prepare factors, remove duplicated and non-actionable fields, and perform an 80/20 split.

```{r}
# Your code here
```

## (b) Class imbalance and churn profile EDA (8 points)

Quantify imbalance and describe high-risk segments using both grouped tables and plots.

```{r}
# Your code here
```

## (c) Logistic threshold by expected cost (12 points)

Choose threshold minimizing expected cost where FN cost is 4x FP cost. Report sensitivity/specificity/AUC.

```{r}
# Your code here
```

## (d) Random forest vs logistic interpretation (12 points)

Tune RF, evaluate test performance, and compare variable-importance ranking to logistic odds-ratio signals.

```{r}
# Your code here
```

## (e) Gradient boosting for campaign targeting (10 points)

Tune boosting rounds and recommend a final targeting model with business-metric justification.

```{r}
# Your code here
```
