---
title: "BAN404 Mock Exam 2026 #7 — SOLVED"
subtitle: "LASSO Sparsity and Cost-Sensitive Churn — SOLVED"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# How to read this solved mock exam

This solved version is written as a **lookup tool**, not only as a short answer key. That means every code line is commented with purpose, theory, or modeling logic, and every code chunk is followed by at least three explanatory sentences that can be reused in a BAN404 answer. The goal is that you should be able to revisit this file later and understand both **what to run** and **why the statistical step is defensible**.

# Setup

```{r}
# Load tidyverse because BAN404 workflows are easiest to follow when wrangling, summaries, and plotting use one coherent grammar.  
library(tidyverse)  # Data cleaning, grouped summaries, tibbles, purrr iteration, and ggplot graphics used throughout the solution.  
# Load glmnet because LASSO and Elastic Net are fit through penalized regression paths in this package.  
library(glmnet)  # Penalized regression with built-in cross-validation support.  
# Load mgcv because GAMs are the standard BAN404 tool for structured nonlinear regression.  
library(mgcv)  # Generalized Additive Models with smooth terms and EDF summaries.  
# Load randomForest because tree ensembles can capture interactions and nonlinear split behavior.  
library(randomForest)  # Random Forest classification for churn prediction.  
# Load gbm because boosting is another common BAN404 ensemble family.  
library(gbm)  # Gradient boosting machine for staged tree-based learning.  
# Load broom because tidy coefficient tables are easier to interpret in exam language.  
library(broom)  # Tidy model summaries and odds-ratio-ready output.  
# Load splines because polynomial or spline-style benchmarks help compare simple and flexible nonlinearity.  
library(splines)  # Natural spline basis functions for structured nonlinear regression.  
# Store the exam number so seeds remain systematic and easy to trace.  
exam_id <- 7  # This solved file belongs to mock exam number 7.  
# Set the main alpha because Task 1 is deliberately LASSO-centered in this mock exam.  
alpha_main <- 1  # alpha = 1 means pure LASSO with L1 shrinkage and possible exact zeros.  
# Store a secondary alpha because Elastic Net is used as the main comparator for stability.  
alpha_alt <- 0.4  # This keeps some L1 sparsity but also adds L2 stabilization.  
# Store the business cost ratio because Task 2 assumes false negatives cost five times false positives.  
cost_fn <- 5  # Missing a churner is much more expensive than contacting a safe customer.  
# Keep a readable focus label because later interpretation should echo the main Task 1 theme.  
focus_label <- "Sparse selection + stability evidence"  # Short narrative label for this mock exam.  
# Create a rank-based AUC helper because ranking quality should be separated from threshold policy.  
auc_rank <- function(y_true, score, positive = "Yes") {  # Function takes observed classes and predicted scores.  
  y_chr <- as.character(y_true)  # Convert labels to characters so class handling is explicit.  
  y_bin <- if_else(y_chr == positive, 1, 0)  # Encode the positive class as 1 for the Mann-Whitney AUC identity.  
  n1 <- sum(y_bin == 1)  # Count positives because AUC compares positive and negative score rankings.  
  n0 <- sum(y_bin == 0)  # Count negatives for the same reason.  
  if (n1 == 0 || n0 == 0) return(NA_real_)  # Return missing value if AUC is undefined because one class is absent.  
  r <- rank(score)  # Rank the scores because AUC is fundamentally a ranking metric.  
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)  # Apply the standard rank-based formula for AUC.  
}  
# Create a threshold-metric helper because a churn policy needs more than a ranking summary.  
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {  # Function maps probabilities into decision metrics.  
  y_chr <- as.character(y_true)  # Convert classes to characters so the positive class is handled explicitly.  
  negative <- setdiff(sort(unique(y_chr)), positive)[1]  # Identify the non-positive class as the reference level.  
  actual <- factor(y_chr, levels = c(negative, positive))  # Lock class order so confusion-matrix indexing stays stable.  
  pred <- factor(if_else(p_hat >= threshold, positive, negative), levels = c(negative, positive))  # Turn probabilities into class decisions.  
  cm <- table(actual = actual, pred = pred)  # Build the confusion matrix because threshold metrics come from these counts.  
  tn <- cm[negative, negative]  # Extract true negatives because they determine specificity.  
  fp <- cm[negative, positive]  # Extract false positives because they reflect unnecessary targeting cost.  
  fn <- cm[positive, negative]  # Extract false negatives because they are especially costly in this exam.  
  tp <- cm[positive, positive]  # Extract true positives because they are correctly identified churners.  
  tibble(  # Return a one-row tibble so many thresholds can be compared cleanly.  
    threshold = threshold,  # Store the evaluated cutoff.  
    accuracy = (tp + tn) / sum(cm),  # Compute accuracy for completeness even though cost matters more here.  
    sensitivity = tp / (tp + fn),  # Compute sensitivity because catching churners is the main business priority.  
    specificity = tn / (tn + fp),  # Compute specificity to track how often safe customers are left alone.  
    auc = auc_rank(actual, p_hat, positive = positive),  # Add AUC to separate ranking quality from threshold choice.  
    expected_cost = cost_fn * fn + fp  # Translate mistakes into the business loss structure given in the prompt.  
  )  # End tidy threshold-metric output.  
}  
```

The setup chunk establishes a full BAN404 workflow rather than a few isolated model calls. That matters because strong exam answers show a **repeatable analytical pipeline** in which tuning, uncertainty, and decision policy are clearly separated. The helper functions also reinforce a central course distinction: **AUC evaluates ranking quality**, while **threshold metrics determine the actual operational policy**.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Explain why LASSO can simplify a regression answer

```{r}
# Read the Task 1 dataset because the methodology discussion should be tied to the actual exam data.  
df_t1 <- read_csv("../task1_data.csv", show_col_types = FALSE)  # Import quietly so rendered output remains clean.  
# Build a numeric model matrix because penalized regression functions typically work with predictor matrices.  
X <- model.matrix(y ~ ., data = df_t1)[, -1]  # Remove the intercept column because glmnet handles the intercept internally.  
# Extract the response because resampling and model fitting often use y separately from X.  
y <- df_t1$y  # Continuous response variable for Task 1.  
# Create a compact theory table because the first subquestion is conceptual rather than prediction-heavy.  
lasso_explain <- tibble(  # Start LASSO explanation table.  
  concept = c("Method", "Why standardize", "Why zeros appear", "Benefit of sparsity", "Interpretation risk"),  # Name the core ideas the examiner expects.  
  interpretation = c(  # Provide BAN404-style explanations for each concept.  
    "LASSO regression because the penalty is proportional to the sum of absolute coefficient values",  # Identify the method from the L1 penalty.  
    "Standardization makes one unit of penalty comparable across predictors with different scales",  # Explain why fair shrinkage needs scaling.  
    "The L1 penalty has corners in the constraint geometry, so some coefficients can land exactly at zero",  # Explain the selection mechanism.  
    "Sparsity can create a shorter and easier-to-defend regression story",  # Summarize why LASSO can help in an exam setting.  
    "A coefficient set to zero is not proof that the underlying variable is universally irrelevant, especially with correlated predictors"  # Explain the main interpretation caution.  
  )  # End interpretation vector.  
)  # End LASSO explanation table.  
# Print the explanation table because the subquestion asks for theory-grounded reasoning.  
lasso_explain  # Display the compact LASSO reasoning summary.  
```

This chunk answers the method-identification part in the way BAN404 usually rewards: by linking the shape of the penalty to a practical modeling consequence. LASSO is not only “the one with absolute values”; it is the method that can simplify a regression answer by shrinking some coefficients all the way to zero. The important caution is that sparsity can reflect the **sample, the correlation structure, and the chosen penalty**, so a neat sparse solution should still be interpreted carefully.

## (b) Compare OLS, LASSO, and Elastic Net under a shared workflow

```{r}
# Set a seed because fair method comparison should be reproducible and not driven by one lucky split.  
set.seed(707)  # Reproducible fold assignment for the shared comparison.  
# Create one shared fold assignment because using the same folds keeps the comparison method-driven rather than split-driven.  
fold_id <- sample(rep(1:10, length.out = nrow(X)))  # Shared 10-fold structure for all penalized models.  
# Fit OLS because BAN404 comparisons should usually start from the simplest interpretable baseline.  
fit_ols <- lm(y ~ ., data = df_t1)  # Unpenalized linear regression benchmark.  
# Tune LASSO on the common folds because LASSO is the headline regularized model in this mock exam.  
cv_lasso <- cv.glmnet(X, y, alpha = alpha_main, foldid = fold_id, standardize = TRUE)  # Shared-fold LASSO tuning.  
# Tune Elastic Net on the exact same folds so differences reflect method behavior rather than different validation splits.  
cv_elnet <- cv.glmnet(X, y, alpha = alpha_alt, foldid = fold_id, standardize = TRUE)  # Shared-fold Elastic Net tuning.  
# Extract LASSO coefficients at lambda.1se because the one-standard-error rule emphasizes a simpler sparse answer.  
coef_lasso <- as.numeric(coef(cv_lasso, s = "lambda.1se"))[-1]  # LASSO slopes at the conservative penalty choice.  
# Extract Elastic Net coefficients at lambda.1se for the same conservative comparison logic.  
coef_elnet <- as.numeric(coef(cv_elnet, s = "lambda.1se"))[-1]  # Elastic Net slopes at its conservative penalty choice.  
# Build one aligned coefficient table because side-by-side output is easier to interpret than separate printouts.  
coef_compare <- tibble(  # Start coefficient-comparison table.  
  variable = colnames(X),  # Keep predictor names visible.  
  ols = as.numeric(coef(fit_ols)[-1]),  # OLS slopes excluding the intercept.  
  lasso_1se = coef_lasso,  # LASSO slopes at lambda.1se.  
  elastic_net_1se = coef_elnet  # Elastic Net slopes at lambda.1se.  
)  # End coefficient-comparison table.  
# Summarize tuning and active-set size because the question asks about both predictive evidence and sparsity behavior.  
cv_table <- tibble(  # Start shared-fold tuning summary.  
  model = c("LASSO", "Elastic Net"),  # Name the penalized competitors.  
  lambda_min = c(cv_lasso$lambda.min, cv_elnet$lambda.min),  # Prediction-oriented tuned lambda.  
  lambda_1se = c(cv_lasso$lambda.1se, cv_elnet$lambda.1se),  # More conservative tuned lambda.  
  cv_mse = c(min(cv_lasso$cvm), min(cv_elnet$cvm)),  # Best cross-validated MSE for each model.  
  active_predictors = c(sum(abs(coef_lasso) > 1e-8), sum(abs(coef_elnet) > 1e-8))  # Count how many predictors remain active.  
)  # End tuning summary.  
# Plot the LASSO coefficient path because path plots show how variables are removed as the penalty grows.  
plot(cv_lasso$glmnet.fit, xvar = "lambda", label = TRUE)  # Visualize the LASSO coefficient path across the penalty range.  
coef_compare  # Print the OLS-versus-penalized coefficient comparison.  
cv_table  # Print the shared-fold tuning summary.  
```

Using the same folds matters because otherwise some of the observed difference between LASSO and Elastic Net may come from different validation splits rather than from the methods themselves. The comparison typically shows LASSO as the more aggressively selective option and Elastic Net as the option that retains more stability when predictors overlap. In BAN404 language, the best conclusion is not only which CV-MSE is slightly lower, but whether the extra sparsity creates a cleaner and still believable final explanation.

## (c) Study sparse-model stability rather than only one chosen lambda

```{r}
# Set a seed because repeated resampling should be reproducible and reviewable.  
set.seed(807)  # Reproducible seed for repeated CV and bootstrap selection stability.  
# Run repeated 5-fold CV because one chosen lambda can move noticeably across plausible fold assignments.  
rep_cv <- map_dfr(1:8, function(rep_i) {  # Loop over eight repeated fold assignments.  
  foldid_i <- sample(rep(1:5, length.out = nrow(X)))  # Create one 5-fold split for this repetition.  
  cv_i <- cv.glmnet(X, y, alpha = alpha_main, foldid = foldid_i, standardize = TRUE)  # Tune LASSO on the repeated folds.  
  tibble(rep = rep_i, lambda_min = cv_i$lambda.min, lambda_1se = cv_i$lambda.1se, cvm_min = min(cv_i$cvm))  # Store the key tuning outputs.  
})  # End repeated CV loop.  
# Summarize repeated-CV stability because the prompt asks whether the selected sparse model is stable.  
rep_summary <- rep_cv %>% summarise(  # Collapse the repeated tuning evidence.  
  mean_lambda_min = mean(lambda_min),  # Average selected lambda.min across repetitions.  
  sd_lambda_min = sd(lambda_min),  # Variation in lambda.min across repetitions.  
  mean_lambda_1se = mean(lambda_1se),  # Average selected lambda.1se across repetitions.  
  sd_lambda_1se = sd(lambda_1se),  # Variation in lambda.1se across repetitions.  
  mean_cvm = mean(cvm_min)  # Average best cross-validated MSE across repetitions.  
)  # End repeated-CV summary.  
# Set the bootstrap size because selection stability should also be studied across resampled datasets.  
B <- 300  # Number of bootstrap replications for active-set frequency.  
# Preallocate a matrix because we want one active-set indicator per predictor and bootstrap sample.  
boot_active <- matrix(0, nrow = B, ncol = ncol(X))  # Container for bootstrap inclusion indicators.  
# Run the bootstrap because sparse-model interpretation is stronger when predictor inclusion is stable.  
for (b in seq_len(B)) {  # Loop over bootstrap replications.  
  idx_b <- sample(seq_len(nrow(X)), size = nrow(X), replace = TRUE)  # Draw bootstrap row indices with replacement.  
  cv_b <- cv.glmnet(X[idx_b, ], y[idx_b], alpha = alpha_main, nfolds = 5, standardize = TRUE)  # Tune LASSO inside the bootstrap sample.  
  coef_b <- as.numeric(coef(cv_b, s = "lambda.1se"))[-1]  # Extract bootstrap LASSO coefficients at lambda.1se.  
  boot_active[b, ] <- abs(coef_b) > 1e-8  # Record whether each predictor remains active in this bootstrap world.  
}  
# Build a stability table because the prompt asks which predictors look stable and which look fragile.  
selection_stability <- tibble(  # Start bootstrap stability table.  
  variable = colnames(X),  # Keep predictor names visible.  
  inclusion_rate = colMeans(boot_active)  # Estimate the proportion of bootstrap fits where each predictor stays active.  
) %>%  # End tibble creation.  
  arrange(desc(inclusion_rate))  # Put the most stable predictors first.  
# Plot repeated lambda choices because tuning stability is easier to read visually than from one standard deviation alone.  
ggplot(rep_cv, aes(x = factor(rep), y = lambda_1se)) +  # Start repeated-CV lambda plot.  
  geom_point(color = "steelblue", size = 2) +  # Show the selected lambda.1se from each repetition.  
  geom_hline(yintercept = mean(rep_cv$lambda_1se), color = "darkgreen", linetype = "dashed") +  # Mark the average conservative penalty.  
  labs(title = "Repeated 5-fold CV: selected lambda.1se across repetitions", x = "Repetition", y = "Selected lambda.1se")  # Add clear labels.  
rep_summary  # Print the repeated-CV stability summary.  
selection_stability  # Print the bootstrap inclusion-frequency table.  
```

This chunk shows why a sparse model can be attractive and still deserve caution. Repeated CV tells us whether the chosen penalty is stable across plausible fold assignments, while the bootstrap inclusion frequencies tell us whether the same predictors repeatedly survive the penalty. In BAN404 language, the safest interpretation is to describe high-inclusion predictors as **consistently selected signals**, while weaker inclusion rates should be discussed as **fragile or sample-dependent** rather than as final truth.

## (d) Test whether simple nonlinearity can beat a sparse linear answer

```{r}
# Create a compact LASSO formula because later comparisons should use the predictors chosen by the sparse model.  
active_vars <- coef_compare %>% filter(abs(lasso_1se) > 1e-8) %>% pull(variable)  # Identify the predictors kept active by LASSO at lambda.1se.  
# Build a fallback in case no variable were selected, because the comparison should remain runnable.  
active_vars <- if (length(active_vars) == 0) c("x1") else active_vars  # Guarantee at least one predictor for the follow-up models.  
# Fit the sparse linear model because this is the main simple benchmark to defend.  
fit_sparse <- lm(reformulate(active_vars, response = "y"), data = df_t1)  # OLS refit on the active LASSO predictors.  
# Fit a structured nonlinear benchmark because mild nonlinearity may outperform a sparse linear answer.  
fit_ns <- lm(y ~ ns(x1, df = 4) + ns(x2, df = 4) + x3 + x4 + x5 + x6, data = df_t1)  # Natural-spline model for plausible nonlinear effects.  
# Fit a GAM because smooth terms are the BAN404 default nonlinear benchmark.  
fit_gam <- gam(y ~ s(x1, k = 6) + s(x2, k = 6) + x3 + x4 + x5 + x6, data = df_t1)  # GAM with smooth terms for the most plausible nonlinear predictors.  
# Generate fitted values because all competing models should be compared on the same data and scale.  
y_hat_sparse <- predict(fit_sparse, newdata = df_t1)  # Fitted values from the sparse linear benchmark.  
y_hat_ns <- predict(fit_ns, newdata = df_t1)  # Fitted values from the spline benchmark.  
y_hat_gam <- predict(fit_gam, newdata = df_t1)  # Fitted values from the GAM.  
# Build a fit table because the question asks for both performance and defendability judgment.  
fit_compare <- tibble(  # Start nonlinear comparison table.  
  model = c("Sparse linear", "Natural spline", "GAM"),  # Name the competing models.  
  mse = c(mean((y - y_hat_sparse)^2), mean((y - y_hat_ns)^2), mean((y - y_hat_gam)^2)),  # Compare mean squared error.  
  fit_measure = c(summary(fit_sparse)$adj.r.squared, summary(fit_ns)$adj.r.squared, summary(fit_gam)$r.sq)  # Compare R2-style fit measures.  
)  # End fit table.  
# Extract GAM smooth-term information because EDF helps describe whether the nonlinearity is substantively large.  
edf_table <- as_tibble(summary(fit_gam)$s.table, rownames = "smooth_term") %>%  # Convert the GAM smooth summary into a tidy table.  
  mutate(  # Add an exam-friendly interpretation label.  
    effect_type = case_when(  # Translate EDF magnitude into plain-language model behavior.  
      edf < 1.2 ~ "approximately linear",  # EDF near 1 means the smooth behaves almost like a straight line.  
      edf < 2 ~ "mild curvature",  # Slightly above 1 suggests modest nonlinearity.  
      TRUE ~ "clear nonlinearity"  # Larger EDF suggests meaningfully flexible structure.  
    )  # End interpretation rules.  
  )  # End mutate.  
fit_compare  # Print the sparse-versus-nonlinear fit comparison.  
edf_table  # Print the GAM smooth-term interpretation table.  
```

This comparison matters because sparsity is not automatically the best answer if the underlying relationship is nonlinear. If the spline or GAM clearly improves fit, the BAN404 conclusion should acknowledge that a simpler sparse linear story may be missing functional-form structure. At the same time, if the improvement is only marginal, the sparse answer may still be preferable because it is easier to explain, easier to defend, and less likely to overcomplicate an exam script.

## (e) Write the final Task 1 recommendation paragraph

```{r}
# Identify the best penalized model because the final memo should make one explicit recommendation.  
best_penalized <- cv_table %>% slice_min(cv_mse, n = 1, with_ties = FALSE)  # Best penalized model by shared-fold CV-MSE.  
# Decide whether the nonlinear upgrade is strong enough to deserve emphasis in the final answer.  
nonlinearity_note <- if_else(min(fit_compare$mse[fit_compare$model != "Sparse linear"]) < fit_compare$mse[fit_compare$model == "Sparse linear"], "A nonlinear benchmark improves fit enough to be mentioned as an alternative if interpretation time allows", "The sparse linear answer remains the easiest main recommendation to defend")  # Compact nonlinear-model judgment.  
# Build the final Task 1 memo because a polished exam answer should end with one integrated conclusion.  
final_task1_memo <- tibble(  # Start Task 1 memo table.  
  recommendation = paste0(  # Build one BAN404-style recommendation sentence.  
    "Main reported model: ", best_penalized$model, ". ",  # State the preferred regularized model.  
    "For interpretation, I would emphasize lambda.1se because it supports a simpler and more defensible active set than chasing the absolute minimum CV error. ",  # Explain the conservative penalty logic.  
    "Repeated-CV and bootstrap evidence suggest that selection stability should be discussed explicitly, especially for weaker predictors. ",  # Connect the stability evidence.  
    nonlinearity_note, "."  # Add the final nonlinearity judgment.  
  )  # End recommendation sentence.  
)  # End Task 1 memo table.  
final_task1_memo  # Print the integrated Task 1 recommendation.  
```

This memo shows the kind of compact synthesis that often earns BAN404 points. Instead of repeating every number, it turns the outputs into a defendable judgment about sparsity, stability, and whether added flexibility is genuinely worth the complexity. In an exam, this is where you show that you can move from **technical output** to **statistical reasoning**.

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Build a deployment-aware churn dataset

```{r}
# Read the churn dataset because Task 2 is an end-to-end binary prediction problem.  
df_c_raw <- read_csv("../customer_data.csv", show_col_types = FALSE)  # Import the raw churn data.  
# Clean names and recode types because deployment-ready pipelines should be explicit and tidyverse-readable.  
df_c <- df_c_raw %>%  # Start churn-preparation pipeline.  
  rename_with(~ gsub("[\\. ]", "_", tolower(.x))) %>%  # Convert names to lower-case snake_case.  
  mutate(  # Recode variables into modeling-friendly types.  
    churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")),  # Make churn an explicit binary factor.  
    international_plan = factor(international_plan),  # Treat the plan flag as categorical rather than numeric.  
    voice_mail_plan = factor(voice_mail_plan),  # Treat the voice-mail indicator as categorical.  
    area_code = factor(area_code)  # Treat area code as a category, not a numeric scale.  
  ) %>%  # End type recoding.  
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)  # Remove geography and deterministic charge duplicates.  
# Set a seed because the train/test split should be reproducible and reviewable.  
set.seed(907)  # Reproducible hold-out split for Task 2.  
# Sample the training indices because all EDA, tuning, and model selection should happen before final testing.  
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))  # Draw 80 percent of rows for training.  
# Create the training set because this is where the modeling workflow will be learned.  
train_c <- df_c[idx, ]  # Training data.  
# Create the test set because it must stay untouched until the final comparison.  
test_c <- df_c[-idx, ]  # Held-out test data.  
# Summarize class balance because churn prediction typically involves class imbalance.  
class_split <- bind_rows(  # Start class-balance summary.  
  train = prop.table(table(train_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),  # Training-set class proportions.  
  test = prop.table(table(test_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"),  # Test-set class proportions.  
  .id = "split"  # Keep the split origin visible.  
)  # End class-balance summary.  
class_split  # Print the train-versus-test class balance.  
```

This pipeline is designed to be defendable in deployment terms rather than only convenient for modeling. Removing deterministic charge columns avoids redundant information, and leaving the test set untouched preserves the meaning of “unseen data.” In BAN404 language, this is the difference between an **honest predictive workflow** and one that quietly leaks future information into model choice.

## (b) Use EDA to build retention personas and hypotheses

```{r}
# Create interpretable personas because descriptive churn analysis is stronger when it maps to likely retention logic.  
train_persona <- train_c %>%  # Start persona-construction pipeline.  
  mutate(  # Create grouped customer personas.  
    service_band = case_when(  # Group customers by how often they contact customer service.  
      customer_service_calls >= 4 ~ "Repeated service friction",  # High-friction group likely linked to dissatisfaction.  
      customer_service_calls >= 2 ~ "Moderate service friction",  # Middle-friction group.  
      TRUE ~ "Low service friction"  # Customers with little visible service strain.  
    ),  # End service-band definition.  
    day_band = case_when(  # Group customers by daytime usage intensity.  
      total_day_minutes >= quantile(total_day_minutes, 0.75) ~ "Heavy day use",  # High daytime usage segment.  
      total_day_minutes <= quantile(total_day_minutes, 0.25) ~ "Light day use",  # Low daytime usage segment.  
      TRUE ~ "Mid day use"  # Middle-usage segment.  
    )  # End day-band definition.  
  )  # End mutate.  
# Summarize persona-level churn because the prompt asks for profile-driven interpretation.  
persona_summary <- train_persona %>%  # Start persona summary table.  
  group_by(service_band, day_band, international_plan, voice_mail_plan) %>%  # Group by an interpretable customer profile structure.  
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>%  # Compute churn rate and segment size.  
  filter(n >= 30) %>%  # Keep only segments large enough to discuss credibly.  
  arrange(desc(churn_rate))  # Put the highest-risk profiles first.  
# Translate patterns into actions because BAN404 rewards linking EDA to business response ideas.  
intervention_table <- tibble(  # Start intervention table.  
  observed_pattern = c(  # Summarize the main descriptive risk signals.  
    "Repeated service friction is associated with higher churn",  # Service-quality pattern.  
    "International-plan customers often appear riskier",  # Plan-linked pattern.  
    "Heavy daytime users may be commercially valuable but still at risk"  # Usage-linked pattern.  
  ),  # End pattern vector.  
  suggested_action = c(  # Map those patterns into plausible retention actions.  
    "Trigger proactive support after repeated service contacts",  # Service-recovery action.  
    "Offer plan review or price clarification to international-plan customers",  # Plan-optimization action.  
    "Protect heavy-use customers with targeted retention offers"  # High-value retention action.  
  )  # End action vector.  
)  # End intervention table.  
# Plot churn share by service friction because this is often the clearest operational signal.  
ggplot(train_persona, aes(x = service_band, fill = churn)) +  # Start first persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions rather than raw counts.  
  scale_y_continuous(labels = scales::percent_format()) +  # Show the vertical axis as percentages.  
  labs(title = "Churn proportion by service-friction band", x = "Service-friction band", y = "Proportion")  # Add clear plot labels.  
# Plot churn share by usage band and international plan because multiple segment dimensions can reveal richer structure.  
ggplot(train_persona, aes(x = day_band, fill = churn)) +  # Start second persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions across usage bands.  
  facet_wrap(~ international_plan) +  # Separate the pattern by international-plan status.  
  scale_y_continuous(labels = scales::percent_format()) +  # Show the vertical axis as percentages.  
  labs(title = "Churn proportion by day-use band and international plan", x = "Day-use band", y = "Proportion")  # Add clear plot labels.  
persona_summary  # Print the persona-level churn summary.  
intervention_table  # Print the persona-to-action mapping.  
```

This is the kind of EDA BAN404 rewards because it does more than list descriptive summaries. It converts visible patterns into **customer personas**, **retention hypotheses**, and **candidate interventions** that later models can either confirm or complicate. The main limitation is that persona-level summaries remain descriptive, so they can still reflect confounding structure that only multivariate models can separate properly.

## (c) Build an interpretable logistic decision rule

```{r}
# Fit logistic regression because it is the standard interpretable churn baseline in BAN404.  
logit_fit <- glm(  # Start logistic model specification.  
  churn ~ account_length + area_code + international_plan + voice_mail_plan + number_vmail_messages + total_day_minutes + total_day_calls + total_eve_minutes + total_night_minutes + total_intl_minutes + customer_service_calls,  # Use the cleaned deployment-ready predictors.  
  data = train_c,  # Fit only on the training data to preserve honest final evaluation.  
  family = binomial()  # Choose logistic regression for the binary outcome.  
)  # End logistic fit.  
# Predict test-set probabilities because ranking and threshold selection should be assessed out of sample.  
p_logit <- predict(logit_fit, newdata = test_c, type = "response")  # Test-set churn probabilities from the logistic model.  
# Compute AUC because ranking quality should be stated separately from the final intervention threshold.  
logit_auc <- auc_rank(test_c$churn, p_logit)  # Test-set AUC for logistic regression.  
# Build a threshold grid because the business-optimal cutoff is not automatically 0.5 under asymmetric costs.  
threshold_grid <- seq(0.10, 0.70, by = 0.05)  # Candidate decision thresholds under the 5:1 cost ratio.  
# Evaluate the threshold grid because sensitivity, specificity, and expected cost depend on the cutoff.  
logit_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_logit, threshold = .x))  # Threshold-by-threshold logistic performance table.  
# Choose the cost-minimizing threshold because the prompt emphasizes business loss rather than default classification.  
best_logit <- logit_thresholds %>% slice_min(expected_cost, n = 1, with_ties = FALSE)  # Best logistic operating threshold under the 5:1 cost ratio.  
# Convert coefficients to odds ratios because BAN404 interpretations are usually written in multiplicative-odds language.  
logit_or <- tidy(logit_fit) %>%  # Tidy coefficient table.  
  mutate(odds_ratio = exp(estimate)) %>%  # Translate log-odds estimates into odds multipliers.  
  arrange(desc(abs(estimate)))  # Sort by effect magnitude for easier discussion.  
# Keep a few exam-friendly terms because the prompt asks for interpretation rather than a full coefficient dump.  
key_terms <- logit_or %>% filter(term %in% c("international_planYes", "voice_mail_planYes", "customer_service_calls"))  # Select interpretable odds-ratio examples.  
logit_auc  # Print the logistic AUC.  
best_logit  # Print the best threshold under the asymmetric loss structure.  
key_terms  # Print the selected odds-ratio interpretations.  
```

This chunk highlights a central BAN404 distinction: a model can rank customers well and still need a separate decision rule before it becomes operational. Under a 5:1 false-negative cost ratio, the chosen threshold is usually lower than `0.5`, because the company prefers extra outreach over missing genuinely risky churners. Logistic regression remains valuable even if a later ensemble wins on AUC, because it gives a transparent story about **which predictors increase or decrease churn odds**.

## (d) Compare flexible ensembles against the logistic baseline

```{r}
# Set a seed because Random Forest uses random sampling and should be reproducible.  
set.seed(1007)  # Reproducible Random Forest tuning seed.  
# Create inner training folds because tuning must happen only inside the training data.  
rf_fold_id <- sample(rep(1:5, length.out = nrow(train_c)))  # Five-fold CV inside the training set.  
# Define the mtry grid because this parameter controls how much feature competition happens at each split.  
mtry_grid <- c(2, 4, 6, 8)  # Compact but informative tuning grid for Random Forest.  
# Evaluate each mtry value with cross-validated AUC because ranking quality matters before thresholding.  
rf_cv <- map_dfr(mtry_grid, function(m) {  # Loop over candidate mtry values.  
  aucs <- map_dbl(1:5, function(k) {  # Loop over the inner folds.  
    tr <- train_c[rf_fold_id != k, ]  # Training portion for fold k.  
    va <- train_c[rf_fold_id == k, ]  # Validation portion for fold k.  
    rf_k <- randomForest(churn ~ ., data = tr, ntree = 450, mtry = m)  # Fit the forest for this fold and mtry.  
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]  # Predict validation probabilities.  
    auc_rank(va$churn, p_k)  # Return the fold-level AUC.  
  })  # End inner fold loop.  
  tibble(mtry = m, cv_auc = mean(aucs))  # Store the mean validation AUC for this mtry value.  
})  # End Random Forest tuning loop.  
# Choose the best mtry because the final forest should use the strongest validated setting.  
best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1, with_ties = FALSE) %>% pull(mtry)  # Extract the best Random Forest tuning value.  
# Fit the final Random Forest on all training data.  
rf_fit <- randomForest(churn ~ ., data = train_c, ntree = 600, mtry = best_mtry, importance = TRUE)  # Final tuned forest with importance tracking.  
# Score the test set because final comparison should use unseen data.  
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]  # Random Forest probabilities on the held-out test set.  
# Evaluate the Random Forest threshold grid because operational policy should reflect the same loss structure.  
rf_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_rf, threshold = .x))  # Threshold performance table for the forest.  
# Choose the best Random Forest threshold under the 5:1 cost ratio.  
best_rf <- rf_thresholds %>% slice_min(expected_cost, n = 1, with_ties = FALSE)  # Best Random Forest operating point.  
# Extract variable importance because tree-based interpretation relies on relevance rankings instead of coefficients.  
rf_imp <- importance(rf_fit) %>%  # Pull the forest importance matrix.  
  as.data.frame() %>%  # Convert the matrix to a data frame.  
  rownames_to_column("variable") %>%  # Preserve variable names explicitly.  
  as_tibble() %>%  # Convert to a tibble for tidy sorting.  
  arrange(desc(MeanDecreaseGini))  # Put the most important variables first.  
# Create numeric outcomes because gbm with Bernoulli loss expects a 0/1 target.  
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Training data with numeric churn outcome.  
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Test data with numeric churn outcome.  
# Create an inner train-validation split because boosting should choose tree count without peeking at the final test set.  
set.seed(1107)  # Reproducible inner split for boosting.  
idx_gbm <- sample(seq_len(nrow(train_g)), size = floor(0.8 * nrow(train_g)))  # Draw rows for the boosting fit subset.  
fit_g <- train_g[idx_gbm, ]  # Inner training subset for gbm estimation.  
val_g <- train_g[-idx_gbm, ]  # Inner validation subset for choosing the number of trees.  
# Fit the boosting model because staged trees can capture complex churn structure gradually.  
gbm_fit <- gbm(  # Start gbm fit.  
  churn01 ~ . - churn,  # Use all predictors except the original factor outcome.  
  distribution = "bernoulli",  # Bernoulli loss is appropriate for binary churn prediction.  
  data = fit_g,  # Fit only on the inner training subset.  
  n.trees = 800,  # Fit a long boosting path so staged selection is meaningful.  
  interaction.depth = 3,  # Allow moderate interaction structure.  
  shrinkage = 0.03,  # Use a small learning rate for gradual fitting.  
  n.minobsinnode = 15,  # Avoid tiny terminal nodes that encourage overfitting.  
  verbose = FALSE  # Keep rendered output clean.  
)  # End gbm fit.  
# Evaluate validation AUC across the boosting path because early stopping should be evidence-based.  
tree_grid <- seq(25, 800, by = 25)  # Candidate numbers of trees along the boosting path.  
val_auc_path <- map_dbl(tree_grid, function(nt) {  # Loop over candidate tree counts.  
  p_nt <- predict(gbm_fit, newdata = val_g, n.trees = nt, type = "response")  # Predict validation probabilities at this stage of boosting.  
  auc_rank(val_g$churn, p_nt)  # Return validation AUC for the current number of trees.  
})  # End staged boosting loop.  
# Choose the best tree count because the final boosting model should stop where validation performance peaks.  
best_trees <- tree_grid[which.max(val_auc_path)]  # Extract the best early-stopping point.  
# Score the untouched test set because final boosting evaluation should be truly out of sample.  
p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")  # Final boosting probabilities on the test set.  
# Evaluate the same threshold grid so all models are compared under the same business objective.  
boost_thresholds <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_boost, threshold = .x))  # Threshold performance table for boosting.  
# Choose the best boosting policy under the 5:1 cost ratio.  
best_boost <- boost_thresholds %>% slice_min(expected_cost, n = 1, with_ties = FALSE)  # Best boosting operating point.  
# Build a final comparison table because the prompt asks for a joint model judgment rather than isolated outputs.  
model_compare <- bind_rows(  # Start all-model comparison table.  
  best_logit %>% mutate(model = "Logistic"),  # Logistic regression at its best threshold.  
  best_rf %>% mutate(model = paste0("Random Forest (mtry=", best_mtry, ")")),  # Random Forest at its best threshold.  
  best_boost %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))  # Boosting at its best threshold.  
) %>%  # End row binding.  
  select(model, threshold, accuracy, sensitivity, specificity, auc, expected_cost) %>%  # Keep the decision-relevant metrics.  
  arrange(expected_cost, desc(auc))  # Rank mainly by business cost and secondarily by discrimination.  
# Plot the boosting validation path because early stopping is easier to defend visually than from one number.  
ggplot(tibble(trees = tree_grid, auc = val_auc_path), aes(x = trees, y = auc)) +  # Start boosting-path plot.  
  geom_line(color = "purple") +  # Show validation AUC across the boosting path.  
  geom_vline(xintercept = best_trees, color = "darkgreen", linetype = "dashed") +  # Mark the selected number of trees.  
  labs(title = "Boosting validation AUC path", x = "Number of trees", y = "Validation AUC")  # Add clear labels.  
rf_cv  # Print Random Forest tuning results.  
rf_imp %>% slice_head(n = 10)  # Print the top Random Forest importance values.  
model_compare  # Print the combined logistic-versus-ensemble scorecard.  
```

This chunk shows how flexible models can be judged under the same business objective rather than on AUC alone. Random Forest and boosting may outperform logistic regression when interactions, nonlinear thresholds, or segment effects matter, but they are harder to explain in the simple coefficient language that managers often understand fastest. In BAN404 terms, the strongest recommendation balances **predictive gain**, **decision cost**, and **explanation quality** rather than treating any one of them as sufficient on its own.

## (e) Write the final retention-policy memo

```{r}
# Identify the recommended model because the final memo should make one explicit operational recommendation.  
recommended_row <- model_compare %>% slice_min(expected_cost, n = 1, with_ties = FALSE)  # Best model under the business-loss objective.  
# Build the final deployment note because the prompt asks for a management-facing recommendation.  
final_task2_memo <- tibble(  # Start final Task 2 memo table.  
  memo = paste0(  # Build one BAN404-style deployment recommendation.  
    "Recommended deployment model: ", recommended_row$model, ". ",  # State the chosen model.  
    "Recommended threshold: ", round(recommended_row$threshold, 2), ", because the main objective is to minimize expected cost under a 5:1 false-negative penalty. ",  # State the chosen operating policy.  
    "This policy accepts more false positives in exchange for catching more true churners, which is appropriate when missed churners are especially expensive. ",  # Describe the core trade-off.  
    "After deployment, I would monitor churn prevalence, threshold performance, and whether the key risk signals remain stable over time."  # Add a practical monitoring plan.  
  )  # End memo text.  
)  # End final memo table.  
final_task2_memo  # Print the final retention-policy memo.  
```

This closing memo is the kind of final step that strengthens a BAN404 answer because it connects the statistics directly to decision-making. The recommendation is not merely “pick the model with the best generic score,” but “pick the model whose threshold policy best fits the company’s loss structure.” In exam language, that is the difference between reporting metrics mechanically and showing that you understand **what decision the model is supposed to support**.
