---
title: "BAN404 Mock Exam 2026 #10 — SOLVED"
subtitle: "Comprehensive Final Simulation — SOLVED"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# How to read this solved mock exam

This solved version is written as a **lookup tool**, not merely as a short answer key. That means every code line is commented with purpose, theory, or modeling logic, and every chunk is followed by at least three explanatory sentences that can be reused in a BAN404 answer. The goal is that you should be able to revisit this file later and understand both **what to run** and **why the workflow is statistically and operationally defensible**.

# Setup

```{r}
# Load tidyverse because BAN404 workflows are easiest to follow when wrangling, summaries, iteration, and plotting use one coherent grammar.  
library(tidyverse)  # Data cleaning, grouped summaries, purrr iteration, tibbles, and ggplot graphics used throughout the solution.  
# Load glmnet because Ridge, LASSO, and Elastic Net are fit through penalized regression paths in this package.  
library(glmnet)  # Penalized regression with built-in cross-validation support.  
# Load mgcv because GAMs are the standard BAN404 tool for structured nonlinear regression.  
library(mgcv)  # Generalized Additive Models with smooth terms and EDF summaries.  
# Load randomForest because tree ensembles can capture threshold effects and interactions.  
library(randomForest)  # Random Forest classification for churn prediction.  
# Load gbm because staged boosting is another high-probability BAN404 ensemble family.  
library(gbm)  # Gradient boosting machine for Bernoulli classification.  
# Load broom because tidy model summaries are easier to interpret in exam language.  
library(broom)  # Tidy coefficient tables and model-summary tables.  
# Load splines because spline-style benchmarks help test structured nonlinearity against linear baselines.  
library(splines)  # Natural spline basis functions for nonlinear regression comparisons.  
# Store the exam number so seeds remain systematic and easy to trace.  
exam_id <- 10  # This solved file belongs to mock exam number 10.  
# Store the penalty settings because Task 1 compares the major regularization families under one shared validation design.  
alpha_ridge <- 0  # alpha = 0 means pure Ridge shrinkage.  
alpha_lasso <- 1  # alpha = 1 means pure LASSO shrinkage with possible exact zeros.  
alpha_elnet <- 0.50  # This mixes L1 and L2 penalties evenly to balance sparsity and stability.  
# Store the business-policy settings because Task 2 is framed around limited campaign capacity.  
max_contact_rate <- 0.30  # The retention team can contact at most 30 percent of customers in one campaign wave.  
cost_fn <- 4  # Missing a churner is treated as four times worse than contacting a non-churner unnecessarily.  
# Keep a short narrative label because later interpretation should echo the main storyline of the mock exam.  
focus_label <- "Final simulation: diagnostics + stability + capacity-aware deployment"  # Compact theme label for mock exam 10.  
# Create a rank-based AUC helper because ranking quality should be kept distinct from the chosen deployment threshold.  
auc_rank <- function(y_true, score, positive = "Yes") {  # Function takes observed classes and predicted scores.  
  y_chr <- as.character(y_true)  # Convert labels to characters so the positive class is handled explicitly.  
  y_bin <- if_else(y_chr == positive, 1, 0)  # Encode the positive class as 1 for the Mann-Whitney AUC identity.  
  n1 <- sum(y_bin == 1)  # Count positives because AUC compares positive and negative rankings.  
  n0 <- sum(y_bin == 0)  # Count negatives for the same reason.  
  if (n1 == 0 || n0 == 0) return(NA_real_)  # Return missing value if one class is absent and AUC is undefined.  
  r <- rank(score)  # Rank the scores because AUC is fundamentally a ranking measure.  
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)  # Apply the standard rank-based formula for AUC.  
}  
# Create a threshold-metric helper because operational policy requires more than one global ranking statistic.  
class_metrics <- function(y_true, p_hat, threshold = 0.5, positive = "Yes") {  # Function maps probabilities into decision metrics.  
  y_chr <- as.character(y_true)  # Convert classes to characters so the positive class is explicit.  
  negative <- setdiff(sort(unique(y_chr)), positive)[1]  # Identify the non-positive class as the reference level.  
  actual <- factor(y_chr, levels = c(negative, positive))  # Lock class order so confusion-matrix indexing remains stable.  
  pred <- factor(if_else(p_hat >= threshold, positive, negative), levels = c(negative, positive))  # Turn probabilities into class decisions.  
  cm <- table(actual = actual, pred = pred)  # Build the confusion matrix because threshold metrics come from these counts.  
  tn <- cm[negative, negative]  # Extract true negatives because they determine specificity.  
  fp <- cm[negative, positive]  # Extract false positives because they reflect unnecessary outreach.  
  fn <- cm[positive, negative]  # Extract false negatives because they are especially costly here.  
  tp <- cm[positive, positive]  # Extract true positives because they are correctly identified churners.  
  tibble(  # Return a one-row tibble so many thresholds can be compared cleanly.  
    threshold = threshold,  # Store the evaluated cutoff.  
    accuracy = (tp + tn) / sum(cm),  # Compute accuracy for completeness even though it is not the main policy objective.  
    sensitivity = tp / (tp + fn),  # Compute sensitivity because catching churners still matters even with limited capacity.  
    specificity = tn / (tn + fp),  # Compute specificity because unnecessary contacts consume scarce campaign resources.  
    auc = auc_rank(actual, p_hat, positive = positive),  # Add AUC so ranking quality remains visible beside the decision rule.  
    expected_cost = cost_fn * fn + fp,  # Translate mistakes into a simple asymmetric-loss score.  
    targeted_share = mean(p_hat >= threshold)  # Measure how large a share of customers would be contacted at this threshold.  
  )  # End tidy threshold-metric output.  
}  
# Create a capacity-aware selector because all Task 2 models should be judged under the same campaign constraint.  
choose_capacity_rule <- function(metric_tbl, max_contact_rate) {  # Function selects the best operating point under the stated capacity limit.  
  feasible_tbl <- metric_tbl %>% filter(targeted_share <= max_contact_rate)  # Keep only thresholds that satisfy the contact-capacity rule.  
  if (nrow(feasible_tbl) > 0) {  # Use the feasible table when at least one threshold satisfies the policy.  
    feasible_tbl %>% arrange(desc(sensitivity), expected_cost, desc(auc)) %>% slice(1)  # Choose the highest-recall feasible threshold while using cost and AUC to break ties.  
  } else {  # Fall back gracefully when the grid does not contain a feasible threshold.  
    metric_tbl %>% arrange(abs(targeted_share - max_contact_rate), desc(sensitivity), expected_cost, desc(auc)) %>% slice(1)  # Choose the closest-to-feasible threshold with strong recall.  
  }  # End capacity-rule branch.  
}  
```

The setup chunk establishes a full BAN404 workflow rather than a few isolated model calls. That matters because strong exam answers separate **diagnostics**, **resampling evidence**, **ranking quality**, and **deployment policy** instead of blurring them together. The helper functions also reinforce a key course distinction: **AUC tells us how well the model orders cases by risk**, while **threshold rules determine how that ranking becomes an operational campaign decision**.

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Decode a penalized objective and explain what it implies for model behaviour

```{r}
# Read the Task 1 dataset because even conceptual method discussion should remain tied to the actual exam data.  
df_t1 <- read_csv("../task1_data.csv", show_col_types = FALSE)  # Import quietly so the rendered document stays readable.  
# Build the predictor matrix because penalized regression functions typically work with X and y separately.  
X <- model.matrix(y ~ ., data = df_t1)[, -1]  # Remove the intercept column because glmnet handles the intercept internally.  
# Extract the response because later validation and bootstrap steps use it explicitly.  
y <- df_t1$y  # Continuous response variable for Task 1.  
# Define a general elastic-net objective because it is useful for explaining Ridge, LASSO, and Elastic Net in one compact framework.  
objective <- function(beta, X, y, lambda, alpha) {  # Function takes coefficients, data, and penalty controls.  
  rss <- sum((y - X %*% beta)^2)  # Compute residual sum of squares because penalized least squares starts from ordinary least squares loss.  
  penalty <- lambda * ((1 - alpha) * sum(beta^2) + alpha * sum(abs(beta)))  # Combine L2 shrinkage and L1 sparsity in one general penalty expression.  
  rss + penalty  # Return the penalized objective value that the optimizer seeks to minimize.  
}  
# Build a theory summary because the first subquestion is mainly about interpretation rather than model selection.  
penalty_summary <- tibble(  # Start theory-summary table.  
  method = c("Ridge", "LASSO", "Elastic Net"),  # Name the three main penalized families covered in this mock exam.  
  alpha_value = c(alpha_ridge, alpha_lasso, alpha_elnet),  # Show how alpha distinguishes the methods inside the general objective.  
  main_effect = c("Shrinks coefficients continuously without exact zeroes in most cases", "Can both shrink coefficients and set some exactly to zero", "Combines shrinkage stability with moderate selection pressure"),  # Summarise the main behavioural difference for each method.  
  exam_use = c("Useful when many predictors are correlated and stability matters", "Useful when interpretability through sparsity is attractive", "Useful when correlated predictors make pure LASSO look too unstable")  # Translate the method behaviour into BAN404-style exam reasoning.  
)  # End theory-summary table.  
penalty_summary  # Print the compact explanation table.  
```

This chunk answers the objective-function question in BAN404 language rather than merely naming the method family. The key idea is that penalization changes the **bias-variance trade-off** by allowing a little extra bias in exchange for more stable prediction behaviour. Standardization matters because otherwise the penalty would treat variables with large natural scales differently from variables with small scales, which would make the shrinkage unfair rather than theoretically meaningful.

## (b) Build an OLS benchmark and diagnose whether the simple linear story is credible

```{r}
# Fit OLS because BAN404 model comparison should begin with the simplest interpretable benchmark.  
fit_ols <- lm(y ~ ., data = df_t1)  # Unpenalized linear-regression baseline.  
# Create a tidy coefficient table because coefficient summaries are easier to interpret in exam language when they are well structured.  
ols_terms <- tidy(fit_ols)  # Tidy coefficient estimates, standard errors, t statistics, and p values.  
# Build an augmented diagnostic table because fitted values and residuals are central to assumption checking.  
ols_aug <- augment(fit_ols)  # Add fitted values and residual information to the original rows.  
# Summarize a few diagnostic indicators because BAN404 answers should interpret whether residual behaviour looks reassuring or concerning.  
ols_diag <- tibble(  # Start OLS-diagnostic summary.  
  r_squared = summary(fit_ols)$r.squared,  # Report the ordinary R-squared for overall linear fit strength.  
  adj_r_squared = summary(fit_ols)$adj.r.squared,  # Report adjusted R-squared because model size should be acknowledged.  
  sigma = summary(fit_ols)$sigma,  # Report the residual standard error because it summarises unexplained spread.  
  mean_abs_resid = mean(abs(ols_aug$.resid)),  # Summarise the typical absolute residual magnitude.  
  corr_abs_resid_fitted = cor(abs(ols_aug$.resid), ols_aug$.fitted),  # Use residual-size versus fitted-value correlation as a rough heteroskedasticity indicator.  
  max_abs_resid = max(abs(ols_aug$.resid))  # Report the largest absolute residual to flag unusually poorly fit observations.  
)  # End OLS-diagnostic summary.  
# Plot residuals against fitted values because visible structure can suggest nonlinearity or non-constant variance.  
ggplot(ols_aug, aes(x = .fitted, y = .resid)) +  # Start residual-versus-fitted diagnostic plot.  
  geom_point(alpha = 0.6, color = "steelblue") +  # Show the residual cloud so shape and spread can be visually inspected.  
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgreen") +  # Mark the ideal zero-residual line for reference.  
  geom_smooth(se = FALSE, color = "firebrick") +  # Add a smooth trend line to help reveal systematic curvature.  
  labs(title = "OLS residuals versus fitted values", x = "Fitted values", y = "Residuals")  # Add clear plot labels.  
# Plot the residual distribution because very skewed or heavy-tailed residual behaviour should make interpretation more cautious.  
ggplot(ols_aug, aes(x = .resid)) +  # Start residual-distribution plot.  
  geom_histogram(bins = 30, fill = "darkorange", color = "white") +  # Show the empirical residual distribution.  
  labs(title = "Distribution of OLS residuals", x = "Residual", y = "Count")  # Add clear plot labels.  
ols_terms  # Print the tidy coefficient table.  
ols_diag  # Print the compact OLS-diagnostic summary.  
```

This chunk treats OLS as a **benchmark to evaluate**, not as an automatic final answer. Residual patterns matter because even a reasonable R-squared does not prove that the linear functional form is fully adequate or that the residual variance is constant. In BAN404 language, OLS remains a strong baseline when diagnostics look broadly acceptable, but visible curvature or strong spread changes should make later nonlinear checks feel necessary rather than optional.

## (c) Compare Ridge, LASSO, and Elastic Net under one common validation design

```{r}
# Set a seed because fair model comparison should be reproducible and not depend on one lucky split.  
set.seed(1010)  # Reproducible fold assignment for the shared penalized-model comparison.  
# Create one common fold assignment because method comparison should be split-fair rather than split-driven.  
fold_id <- sample(rep(1:10, length.out = nrow(X)))  # Shared 10-fold structure for all penalized models.  
# Tune Ridge on the shared folds because Ridge is the pure shrinkage benchmark.  
cv_ridge <- cv.glmnet(X, y, alpha = alpha_ridge, foldid = fold_id, standardize = TRUE)  # Shared-fold Ridge tuning.  
# Tune LASSO on the same folds because sparsity is a natural alternative to pure shrinkage.  
cv_lasso <- cv.glmnet(X, y, alpha = alpha_lasso, foldid = fold_id, standardize = TRUE)  # Shared-fold LASSO tuning.  
# Tune Elastic Net on the same folds so differences reflect penalty behaviour rather than different partitions.  
cv_elnet <- cv.glmnet(X, y, alpha = alpha_elnet, foldid = fold_id, standardize = TRUE)  # Shared-fold Elastic Net tuning.  
# Extract conservative coefficients because lambda.1se is often the easiest to defend when stability matters.  
coef_ridge_1se <- as.numeric(coef(cv_ridge, s = "lambda.1se"))[-1]  # Ridge coefficients at the one-standard-error penalty.  
coef_lasso_1se <- as.numeric(coef(cv_lasso, s = "lambda.1se"))[-1]  # LASSO coefficients at the one-standard-error penalty.  
coef_elnet_1se <- as.numeric(coef(cv_elnet, s = "lambda.1se"))[-1]  # Elastic Net coefficients at the one-standard-error penalty.  
# Repeat a smaller CV exercise because one chosen lambda can move across plausible fold assignments.  
set.seed(1110)  # Reproducible seed for repeated-CV stability summaries.  
rep_cv <- map_dfr(1:5, function(rep_i) {  # Loop over five repeated 5-fold validation layouts.  
  foldid_i <- sample(rep(1:5, length.out = nrow(X)))  # Create one 5-fold split for this repetition.  
  cv_i <- cv.glmnet(X, y, alpha = alpha_elnet, foldid = foldid_i, standardize = TRUE)  # Tune Elastic Net on this repeated split.  
  tibble(rep = rep_i, lambda_min = cv_i$lambda.min, lambda_1se = cv_i$lambda.1se, cvm_min = min(cv_i$cvm))  # Store the main tuning outputs for stability interpretation.  
})  # End repeated-CV loop.  
# Build one shared comparison table because the examiner wants performance and behaviour to be interpreted together.  
penalty_compare <- tibble(  # Start penalized-model summary table.  
  model = c("Ridge", "LASSO", "Elastic Net"),  # Name the competing penalized models.  
  lambda_min = c(cv_ridge$lambda.min, cv_lasso$lambda.min, cv_elnet$lambda.min),  # Store the prediction-oriented tuning value.  
  lambda_1se = c(cv_ridge$lambda.1se, cv_lasso$lambda.1se, cv_elnet$lambda.1se),  # Store the more conservative tuning value.  
  cv_mse_min = c(min(cv_ridge$cvm), min(cv_lasso$cvm), min(cv_elnet$cvm)),  # Compare the best cross-validated MSE from each path.  
  active_predictors_1se = c(sum(abs(coef_ridge_1se) > 1e-8), sum(abs(coef_lasso_1se) > 1e-8), sum(abs(coef_elnet_1se) > 1e-8))  # Count how many coefficients remain active at the conservative penalty.  
)  # End penalized-model summary table.  
# Build a side-by-side coefficient table because aligned output is easier to interpret than separate printouts.  
coef_compare <- tibble(  # Start coefficient-comparison table.  
  variable = colnames(X),  # Keep the predictor names visible.  
  ridge_1se = coef_ridge_1se,  # Ridge coefficients at lambda.1se.  
  lasso_1se = coef_lasso_1se,  # LASSO coefficients at lambda.1se.  
  elastic_net_1se = coef_elnet_1se  # Elastic Net coefficients at lambda.1se.  
)  # End coefficient-comparison table.  
# Summarize the repeated-CV stability because BAN404 answers should comment on whether tuning looks stable enough to trust.  
rep_summary <- rep_cv %>% summarise(mean_lambda_min = mean(lambda_min), sd_lambda_min = sd(lambda_min), mean_lambda_1se = mean(lambda_1se), sd_lambda_1se = sd(lambda_1se), mean_cvm = mean(cvm_min))  # Collapse repeated tuning behaviour into a compact stability summary.  
penalty_compare  # Print the main penalized-model comparison table.  
coef_compare  # Print the aligned coefficient-behaviour table.  
rep_summary  # Print the repeated-CV stability summary.  
```

Using the same folds matters because otherwise some of the observed model difference could come from easier or harder partitions rather than from the penalty itself. Ridge usually gives the most purely stable shrinkage story, LASSO usually gives the sparsest story, and Elastic Net often provides the easiest compromise when correlated predictors are present. In BAN404 language, `lambda.1se` is often easier to defend because it reflects a consciously more conservative reporting rule rather than a purely aggressive search for the lowest possible CV error.

## (d) Use the bootstrap to study both coefficient stability and response-scale uncertainty

```{r}
# Set a seed because bootstrap evidence should be reproducible and reviewable.  
set.seed(1210)  # Reproducible bootstrap seed for Task 1 uncertainty work.  
# Choose a bootstrap size that is large enough for stable summaries while still rendering comfortably.  
B_boot <- 500  # Number of bootstrap replications for coefficient and SD(y) uncertainty.  
# Run the bootstrap because coefficient stability should be studied across many plausible resampled datasets.  
boot_coef <- map_dfr(1:B_boot, function(b) {  # Loop over bootstrap replications.  
  idx_b <- sample(seq_len(nrow(df_t1)), size = nrow(df_t1), replace = TRUE)  # Draw bootstrap row indices with replacement.  
  fit_b <- lm(y ~ ., data = df_t1[idx_b, ])  # Refit OLS on the bootstrap sample.  
  tibble(term = names(coef(fit_b))[-1], estimate = as.numeric(coef(fit_b))[-1], boot_id = b)  # Store bootstrap coefficient estimates in tidy long format.  
})  # End coefficient-bootstrap loop.  
# Summarize coefficient stability because the prompt asks which effects look stable and which look fragile.  
coef_stability <- boot_coef %>%  # Start coefficient-stability pipeline.  
  group_by(term) %>%  # Group by coefficient term so each predictor gets its own summary.  
  summarise(mean_estimate = mean(estimate), bootstrap_sd = sd(estimate), q025 = quantile(estimate, 0.025), q975 = quantile(estimate, 0.975), .groups = "drop") %>%  # Compute average size, bootstrap SD, and percentile interval bounds.  
  arrange(desc(bootstrap_sd))  # Put the most unstable coefficient first.  
# Identify a stable and an unstable example because BAN404 answers should translate the summary into clear wording.  
most_unstable <- coef_stability %>% slice(1)  # Extract the coefficient with the largest bootstrap SD.  
most_stable <- coef_stability %>% slice_tail(n = 1)  # Extract the coefficient with the smallest bootstrap SD.  
# Bootstrap SD(y) because response-scale uncertainty affects how confidently model quality can be discussed.  
boot_sd <- replicate(B_boot, sd(sample(y, size = length(y), replace = TRUE)))  # Bootstrap sampling distribution for SD(y).  
# Compute the observed SD because the intervals should be tied back to the sample estimate.  
sd_hat <- sd(y)  # Observed response standard deviation.  
# Compute the percentile interval because it respects the empirical bootstrap distribution directly.  
ci_sd_pct <- quantile(boot_sd, c(0.025, 0.975))  # Percentile CI for SD(y).  
# Compute the normal approximation because the prompt asks for a comparison rather than only one interval.  
ci_sd_norm <- c(sd_hat - 1.96 * sd(boot_sd), sd_hat + 1.96 * sd(boot_sd))  # Symmetric normal-based CI for SD(y).  
# Build one compact SD summary table because the comparison should be easy to discuss in exam language.  
sd_summary <- tibble(quantity = "SD(y)", point_estimate = sd_hat, pct_low = ci_sd_pct[1], pct_high = ci_sd_pct[2], norm_low = ci_sd_norm[1], norm_high = ci_sd_norm[2])  # Store both interval versions beside the point estimate.  
# Plot the bootstrap SD distribution because asymmetry is easier to defend when it is visible.  
ggplot(tibble(boot_sd = boot_sd), aes(x = boot_sd)) +  # Start bootstrap-SD plot.  
  geom_histogram(bins = 30, fill = "darkorange", color = "white") +  # Show the empirical distribution of bootstrap SD estimates.  
  geom_vline(xintercept = sd_hat, color = "darkgreen", linetype = "dashed") +  # Mark the observed sample SD.  
  labs(title = "Bootstrap distribution of SD(y)", x = "Bootstrap SD estimate", y = "Count")  # Add clear plot labels.  
coef_stability  # Print the coefficient-stability summary table.  
most_unstable  # Print the most unstable coefficient summary.  
most_stable  # Print the most stable coefficient summary.  
sd_summary  # Print the SD(y) interval comparison.  
```

This chunk distinguishes between **large effects** and **stable effects**, which is an important BAN404 distinction. A coefficient can look substantial in one fitted model and still be comparatively unstable across plausible resamples, which should make the written interpretation more cautious. The SD(y) bootstrap is also useful because uncertainty about the response scale reminds us that model quality is being judged against a noisy target, so overly confident language would be hard to defend.

## (e) Decide whether spline or GAM flexibility changes the final Task 1 answer

```{r}
# Fit the linear benchmark because more flexible models should always be judged relative to the simplest credible baseline.  
fit_linear <- lm(y ~ ., data = df_t1)  # Additive linear-regression benchmark.  
# Fit a spline benchmark because structured nonlinearity should be tested before moving directly to smooth GAM terms.  
fit_spline <- lm(y ~ ns(x1, df = 4) + ns(x2, df = 4) + ns(x3, df = 4) + x4 + x5 + x6, data = df_t1)  # Natural-spline regression benchmark.  
# Fit a GAM because smooth terms are the BAN404 default nonlinear benchmark.  
fit_gam <- gam(y ~ s(x1, k = 6) + s(x2, k = 6) + s(x3, k = 6) + x4 + x5 + x6, data = df_t1)  # GAM with smooth terms for the most plausible nonlinear predictors.  
# Generate fitted values because all models should be compared on the same sample and scale.  
y_hat_linear <- predict(fit_linear, newdata = df_t1)  # Fitted values from the linear model.  
y_hat_spline <- predict(fit_spline, newdata = df_t1)  # Fitted values from the spline benchmark.  
y_hat_gam <- predict(fit_gam, newdata = df_t1)  # Fitted values from the GAM.  
# Build a fit table because the question asks whether additional flexibility meaningfully changes the final answer.  
fit_compare <- tibble(  # Start nonlinear-comparison table.  
  model = c("Linear", "Spline", "GAM"),  # Name the competing model forms.  
  mse = c(mean((y - y_hat_linear)^2), mean((y - y_hat_spline)^2), mean((y - y_hat_gam)^2)),  # Compare mean squared error.  
  fit_measure = c(summary(fit_linear)$adj.r.squared, summary(fit_spline)$adj.r.squared, summary(fit_gam)$r.sq)  # Compare R2-style fit summaries.  
)  # End nonlinear-comparison table.  
# Extract the GAM smooth summary because EDF values help explain whether the smooths are doing meaningful work.  
edf_table <- as_tibble(summary(fit_gam)$s.table, rownames = "smooth_term") %>%  # Convert the GAM smooth summary into a tidy table.  
  mutate(effect_type = case_when(edf < 1.2 ~ "approximately linear", edf < 2 ~ "mild curvature", TRUE ~ "clear nonlinearity"))  # Translate EDF values into exam-friendly interpretation labels.  
# Create a short Task 1 memo because the prompt asks for a final examiner-ready recommendation.  
final_task1_memo <- tibble(recommendation = paste0("Main Task 1 answer: ", fit_compare %>% arrange(mse) %>% slice(1) %>% pull(model), ". I would describe the extra flexibility as ", if_else(min(fit_compare$mse[fit_compare$model != "Linear"]) < fit_compare$mse[fit_compare$model == "Linear"], "worth mentioning only if the fit gain and EDF evidence jointly suggest meaningful curvature", "too modest to replace the simpler linear explanation"), "."))  # Build a compact BAN404-style recommendation sentence.  
fit_compare  # Print the linear-versus-nonlinear fit summary.  
edf_table  # Print the GAM EDF interpretation table.  
final_task1_memo  # Print the Task 1 recommendation.  
```

This chunk shows why a more flexible model is not automatically a better exam answer. A spline or GAM should replace the simpler linear story only when the fit improvement is real enough and the smooth terms appear to be doing meaningful nonlinear work. In BAN404 language, the right conclusion is not “GAM is more advanced,” but “GAM is worth reporting only if the extra flexibility produces clearly defensible evidence.”

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Build a churn table that is ready for intervention and honest for testing

```{r}
# Read the churn dataset because Task 2 is an end-to-end binary prediction and deployment problem.  
df_c_raw <- read_csv("../customer_data.csv", show_col_types = FALSE)  # Import the raw churn data.  
# Clean names and recode types because intervention-ready workflows should make preparation explicit.  
df_c <- df_c_raw %>%  # Start churn-preparation pipeline.  
  rename_with(~ gsub("[\\. ]", "_", tolower(.x))) %>%  # Convert names to lower-case snake_case.  
  mutate(churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")), international_plan = factor(international_plan), voice_mail_plan = factor(voice_mail_plan), area_code = factor(area_code)) %>%  # Recode the main categorical variables.  
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)  # Remove geography and deterministic charge duplicates for a more realistic intervention table.  
# Set a seed because the hold-out split should be reproducible and reviewable.  
set.seed(1310)  # Reproducible train/test split for Task 2.  
# Draw the training indices because all later modelling and EDA should happen before final testing.  
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))  # Sample 80 percent of the rows for training.  
# Create the training set because development work should be restricted to this subset.  
train_c <- df_c[idx, ]  # Training data.  
# Create the test set because it should remain untouched until final model comparison.  
test_c <- df_c[-idx, ]  # Held-out test data.  
# Summarize class balance because churn problems often involve imbalance that affects threshold policy.  
class_split <- bind_rows(train = prop.table(table(train_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"), test = prop.table(table(test_c$churn)) %>% as.list() %>% enframe(name = "class", value = "prop"), .id = "split")  # Create a tidy train-versus-test class-balance table.  
class_split  # Print the class-balance summary.  
```

This preparation step is designed to be realistic rather than merely convenient. Removing deterministic charge duplicates avoids redundant information, and removing `state` keeps the later model closer to a behavioural intervention story rather than a weak geographic shortcut. In BAN404 language, leaving the test set untouched preserves the credibility of the final comparison because it remains a genuine out-of-sample check.

## (b) Turn EDA into practical churn-priority stories

```{r}
# Build interpretable customer groupings because BAN404 EDA is strongest when it maps directly to risk stories and actions.  
train_persona <- train_c %>%  # Start persona-construction pipeline.  
  mutate(service_band = case_when(customer_service_calls >= 4 ~ "Repeated service friction", customer_service_calls >= 2 ~ "Moderate service friction", TRUE ~ "Low service friction"), day_band = case_when(total_day_minutes >= quantile(total_day_minutes, 0.75) ~ "Heavy day use", total_day_minutes <= quantile(total_day_minutes, 0.25) ~ "Light day use", TRUE ~ "Mid day use"), intl_band = case_when(total_intl_minutes >= quantile(total_intl_minutes, 0.75) ~ "Heavy international use", TRUE ~ "Lower international use"))  # Create grouped usage and friction personas.  
# Summarize persona-level churn because the prompt asks for risk stories that can guide later modelling.  
persona_summary <- train_persona %>%  # Start persona summary table.  
  group_by(service_band, day_band, intl_band, international_plan) %>%  # Group by interpretable customer-profile dimensions.  
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") %>%  # Compute churn rate and segment size.  
  filter(n >= 30) %>%  # Keep only segment sizes large enough to discuss credibly.  
  arrange(desc(churn_rate))  # Put the highest-risk personas first.  
# Create a compact action table because BAN404 rewards linking descriptive patterns to plausible interventions.  
action_table <- tibble(observed_pattern = c("Repeated service friction is associated with elevated churn", "Heavy international use often coincides with higher churn risk", "Heavy day-use customers may justify prioritised retention treatment"), suggested_action = c("Trigger service-recovery outreach after repeated complaints", "Review plan fit and price clarity for internationally active customers", "Prioritise proactive retention offers for valuable high-usage customers"))  # Translate descriptive signals into possible actions.  
# Plot churn share by service-friction band because complaint history is often the clearest operational risk signal.  
ggplot(train_persona, aes(x = service_band, fill = churn)) +  # Start first persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions rather than raw counts.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express the vertical axis as percentages.  
  labs(title = "Churn share by service-friction band", x = "Service-friction band", y = "Proportion")  # Add clear plot labels.  
# Plot churn share by day-use band and international-plan status because joint descriptive patterns can motivate later model structure.  
ggplot(train_persona, aes(x = day_band, fill = churn)) +  # Start second persona plot.  
  geom_bar(position = "fill") +  # Compare churn proportions across usage bands.  
  facet_wrap(~ international_plan) +  # Split the plot by international-plan status.  
  scale_y_continuous(labels = scales::percent_format()) +  # Express the vertical axis as percentages.  
  labs(title = "Churn share by day-use band and international plan", x = "Day-use band", y = "Proportion")  # Add clear plot labels.  
persona_summary  # Print the persona-level churn summary.  
action_table  # Print the descriptive action table.  
```

This is the kind of EDA BAN404 rewards because it turns grouped summaries into **interpretable risk stories** and **possible campaign actions**. The plots help show where churn risk seems to concentrate, while the action table makes the transition from descriptive analysis to managerial reasoning. The main limitation is that persona-based summaries are still descriptive, so apparent risk stories may partly reflect confounding that only multivariate models can separate properly.

## (c) Build a logistic model under a capacity-aware deployment rule

```{r}
# Fit logistic regression because it is the standard interpretable churn baseline in BAN404.  
logit_fit <- glm(churn ~ account_length + area_code + international_plan + voice_mail_plan + number_vmail_messages + total_day_minutes + total_day_calls + total_eve_minutes + total_eve_calls + total_night_minutes + total_intl_minutes + customer_service_calls, data = train_c, family = binomial())  # Logistic model on the cleaned training data.  
# Predict test-set probabilities because ranking quality and policy choice should be evaluated out of sample.  
p_logit <- predict(logit_fit, newdata = test_c, type = "response")  # Test-set churn probabilities from logistic regression.  
# Build a threshold grid because the operational cutoff is a business choice rather than a fixed model property.  
threshold_grid <- seq(0.10, 0.80, by = 0.05)  # Candidate deployment thresholds.  
# Evaluate the threshold grid because the final policy should explicitly respect campaign capacity.  
logit_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_logit, threshold = .x))  # Threshold-by-threshold logistic performance table.  
# Apply the capacity-aware rule because the campaign cannot contact an unlimited share of customers.  
best_logit <- choose_capacity_rule(logit_grid, max_contact_rate = max_contact_rate)  # Best logistic operating point under the contact-capacity constraint.  
# Convert coefficients to odds ratios because logistic interpretation is usually written in multiplicative-odds language.  
logit_or <- tidy(logit_fit) %>% mutate(odds_ratio = exp(estimate)) %>% arrange(desc(abs(estimate)))  # Tidy odds-ratio table sorted by effect magnitude.  
# Select exam-friendly terms because the prompt asks for interpretation rather than a full coefficient dump.  
key_terms <- logit_or %>% filter(term %in% c("international_planYes", "voice_mail_planYes", "customer_service_calls"))  # Keep interpretable terms for BAN404-style discussion.  
logit_grid  # Print the threshold-by-threshold logistic performance table.  
best_logit  # Print the selected logistic operating point.  
key_terms  # Print the selected odds-ratio interpretations.  
```

This chunk separates three ideas that BAN404 students should keep distinct: **the probability model**, **the threshold**, and **the campaign rule**. Logistic regression provides an interpretable ranking of churn risk, but the threshold is chosen afterward because capacity limits determine how many customers can actually be targeted. In BAN404 language, the best threshold is not automatically `0.5`; it is the one that gives the strongest feasible recall while respecting the operational targeting limit.

## (d) Test whether Random Forest adds enough signal to justify weaker interpretability

```{r}
# Set a seed because Random Forest uses random sampling and should therefore be reproducible.  
set.seed(1410)  # Reproducible Random Forest tuning seed.  
# Create inner folds because tuning must happen only inside the training data.  
rf_fold_id <- sample(rep(1:5, length.out = nrow(train_c)))  # Five-fold CV inside the training set.  
# Define the mtry grid because this parameter controls how much feature competition happens at each split.  
mtry_grid <- c(2, 4, 6, 8)  # Compact but informative Random Forest tuning grid.  
# Evaluate each mtry value because model flexibility should be chosen by validation rather than guesswork.  
rf_cv <- map_dfr(mtry_grid, function(m) {  # Loop over candidate mtry values.  
  aucs <- map_dbl(1:5, function(k) {  # Loop over the inner CV folds.  
    tr <- train_c[rf_fold_id != k, ]  # Training portion for fold k.  
    va <- train_c[rf_fold_id == k, ]  # Validation portion for fold k.  
    rf_k <- randomForest(churn ~ ., data = tr, ntree = 500, mtry = m)  # Fit the forest for this fold and tuning value.  
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]  # Predict validation probabilities.  
    auc_rank(va$churn, p_k)  # Return fold-level AUC because ranking quality is the first comparison layer.  
  })  # End inner fold loop.  
  tibble(mtry = m, cv_auc = mean(aucs))  # Store the mean validation AUC for this tuning value.  
})  # End Random Forest tuning loop.  
# Choose the best mtry because the final forest should use the strongest validated setting.  
best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1, with_ties = FALSE) %>% pull(mtry)  # Extract the best Random Forest tuning value.  
# Fit the final forest because the test set should only see the chosen model once tuning is finished.  
rf_fit <- randomForest(churn ~ ., data = train_c, ntree = 650, mtry = best_mtry, importance = TRUE)  # Final tuned Random Forest.  
# Predict test-set probabilities because final comparison should use unseen data.  
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]  # Random Forest test-set probabilities.  
# Evaluate the same threshold grid because all models should face the same campaign-capacity rule.  
rf_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_rf, threshold = .x))  # Threshold-by-threshold Random Forest performance.  
# Apply the same capacity-aware rule because comparison should not change the deployment objective.  
best_rf <- choose_capacity_rule(rf_grid, max_contact_rate = max_contact_rate)  # Best Random Forest operating point.  
# Extract variable importance because tree models are interpreted through feature importance rather than coefficients.  
rf_imp <- importance(rf_fit) %>% as.data.frame() %>% rownames_to_column("variable") %>% as_tibble() %>% arrange(desc(MeanDecreaseGini))  # Tidy Random Forest importance table.  
rf_cv  # Print the Random Forest tuning summary.  
best_rf  # Print the selected Random Forest operating point.  
rf_imp %>% slice_head(n = 10)  # Print the top Random Forest importance values.  
```

Random Forest can improve the churn solution when the true risk structure contains thresholds, interactions, or nonlinear effects that logistic regression cannot capture naturally. The trade-off is that the explanation changes from coefficient-based odds stories to variable-importance rankings, which are usually less transparent to managers. In BAN404 language, the key question is whether the gain under the **same deployment rule** is large enough to justify the weaker interpretability.

## (e) Tune boosting and write the final campaign-allocation memo

```{r}
# Prepare a numeric target because gbm with Bernoulli loss expects a 0/1 outcome.  
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Training data with a numeric churn target.  
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))  # Test data with a numeric churn target.  
# Create an inner training-validation split because early stopping should be chosen without using the final test set.  
set.seed(1510)  # Reproducible seed for the boosting early-stopping split.  
idx_gbm <- sample(seq_len(nrow(train_g)), size = floor(0.8 * nrow(train_g)))  # Draw rows for the boosting fit subset.  
fit_g <- train_g[idx_gbm, ]  # Inner training subset for gbm estimation.  
val_g <- train_g[-idx_gbm, ]  # Inner validation subset for staged model selection.  
# Fit the boosting model because stage-wise additive trees can capture complex churn structure gradually.  
gbm_fit <- gbm(churn01 ~ . - churn, distribution = "bernoulli", data = fit_g, n.trees = 850, interaction.depth = 3, shrinkage = 0.03, n.minobsinnode = 15, verbose = FALSE)  # Fit a long boosting path for evidence-based early stopping.  
# Create the staged tree grid because the number of trees should be selected by validation rather than default settings.  
tree_grid <- seq(25, 850, by = 25)  # Candidate numbers of trees along the boosting path.  
# Evaluate validation AUC across the boosting path because early stopping should be justified by held-out evidence.  
val_auc_path <- map_dbl(tree_grid, function(nt) {  # Loop over candidate tree counts.  
  p_nt <- predict(gbm_fit, newdata = val_g, n.trees = nt, type = "response")  # Predict validation probabilities at this stage of boosting.  
  auc_rank(val_g$churn, p_nt)  # Return validation AUC for the current number of trees.  
})  # End staged boosting loop.  
# Choose the best tree count because the final boosting model should stop where validation ranking quality peaks.  
best_trees <- tree_grid[which.max(val_auc_path)]  # Extract the best early-stopping point.  
# Predict the untouched test set because final boosting comparison should remain genuinely out of sample.  
p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")  # Final boosting probabilities on the test set.  
# Evaluate the same threshold grid because all models should face the same campaign rule.  
boost_grid <- map_dfr(threshold_grid, ~ class_metrics(test_c$churn, p_boost, threshold = .x))  # Threshold-by-threshold boosting performance table.  
# Apply the same capacity-aware rule because model comparison should not quietly change the decision target.  
best_boost <- choose_capacity_rule(boost_grid, max_contact_rate = max_contact_rate)  # Best boosting operating point.  
# Build a joint comparison table because the final recommendation should be based on one consistent scorecard.  
model_compare <- bind_rows(best_logit %>% mutate(model = "Logistic"), best_rf %>% mutate(model = paste0("Random Forest (mtry=", best_mtry, ")")), best_boost %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))) %>% select(model, threshold, targeted_share, accuracy, sensitivity, specificity, auc, expected_cost) %>% arrange(desc(sensitivity), expected_cost, desc(auc))  # Joint scorecard under the shared campaign-capacity rule.  
# Identify the recommended row because the final memo should make one explicit operational choice.  
recommended_row <- model_compare %>% slice(1)  # Best model under the capacity-aware comparison rule.  
# Create a short management memo because the prompt asks for a deployment-ready final recommendation.  
final_task2_memo <- tibble(memo = paste0("Recommended deployment model: ", recommended_row$model, ". Recommended threshold: ", round(recommended_row$threshold, 2), ". Expected targeted share: ", scales::percent(recommended_row$targeted_share, accuracy = 0.1), ". This choice respects campaign capacity while still trying to capture as many true churners as possible, so the main trade-off is that some risky customers will remain untouched in exchange for a feasible contact volume. After deployment, I would monitor realised contact rate, recall among contacted segments, calibration drift, and whether the main churn drivers remain stable over time."))  # Compact BAN404-style management note.  
# Plot the validation AUC path because staged early stopping is easier to defend visually than from one number alone.  
ggplot(tibble(trees = tree_grid, validation_auc = val_auc_path), aes(x = trees, y = validation_auc)) +  # Start boosting-validation plot.  
  geom_line(color = "purple") +  # Show how validation AUC evolves across the boosting path.  
  geom_vline(xintercept = best_trees, color = "darkgreen", linetype = "dashed") +  # Mark the selected early-stopping point.  
  labs(title = "Boosting validation AUC path", x = "Number of trees", y = "Validation AUC")  # Add clear plot labels.  
model_compare  # Print the combined model-comparison scorecard.  
final_task2_memo  # Print the final campaign-allocation memo.  
```

This final chunk shows how a BAN404 answer moves from model comparison to an explicit operational decision. The strongest answer does not stop at “best AUC”; it states which threshold will be used, how many customers that implies contacting, and what trade-off the company is accepting. In exam language, the best final recommendation is the one that matches the actual capacity-constrained campaign rule rather than the one that merely looks strongest on one generic metric.
