---
title: "BAN404 Mock Exam 2026 #10"
subtitle: "Comprehensive Final Simulation"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Read algorithm from code and describe assumptions (10 points)

Given custom penalized objective + optimizer, identify method and assumptions.

## (b) OLS baseline and diagnostics (10 points)

Fit OLS, evaluate assumptions, and identify signs of nonlinearity.

```{r}
# Your code here
```

## (c) Penalized alternatives + tuning (10 points)

Tune Ridge/LASSO/Elastic Net and select model by cross-validated error and simplicity.

```{r}
# Your code here
```

## (d) Bootstrap uncertainty (10 points)

Bootstrap variance and provide confidence interval interpretation suitable for exam write-up.

```{r}
# Your code here
```

## (e) GAM enhancement (10 points)

Fit GAM and assess whether it improves enough to justify lower interpretability.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data prep strategy (8 points)

Design prediction-ready dataset and explain each exclusion choice.

```{r}
# Your code here
```

## (b) EDA and business framing (8 points)

Link patterns to retention actions and define what model errors are most costly.

```{r}
# Your code here
```

## (c) Logistic model (12 points)

Train and evaluate logistic model with threshold sweep.

```{r}
# Your code here
```

## (d) Random forest (12 points)

Tune and evaluate RF, interpret importance, and compare to logistic coefficients.

```{r}
# Your code here
```

## (e) Boosting and final recommendation (10 points)

Tune boosting, compare all candidates, and provide final justified deployment choice.

```{r}
# Your code here
```
