---
title: "BAN404 Mock Exam 2026 #9"
subtitle: "Validation Discipline and Robust Prediction"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Theory: LOOCV vs K-fold (10 points)

Discuss bias-variance-computation tradeoff between LOOCV and K-fold CV for n = 5000.

## (b) Leakage-safe preprocessing (10 points)

Implement scaling inside CV folds for regularized regression and compare with naive scaling.

```{r}
# Your code here
```

## (c) Ridge and LASSO tuning (10 points)

Tune both models and compare best CV-MSE values.

```{r}
# Your code here
```

## (d) Bootstrap CI for SD(y) (10 points)

Compute normal and percentile CIs for SD(y) with B = 1000.

```{r}
# Your code here
```

## (e) OLS vs GAM fit quality (10 points)

Compare models and explain whether added flexibility is justified.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data prep and split (8 points)

Prepare a strict train/test process with no peeking.

```{r}
# Your code here
```

## (b) EDA and candidate predictors (8 points)

Summarize churn patterns and identify predictors likely to survive multivariate modeling.

```{r}
# Your code here
```

## (c) Logistic model diagnostics (12 points)

Evaluate metrics at different thresholds and discuss trade-offs.

```{r}
# Your code here
```

## (d) Random forest (12 points)

Tune `mtry`, evaluate AUC, and compare with logistic.

```{r}
# Your code here
```

## (e) Boosting and final choice (10 points)

Use staged evaluation to choose tree count and justify final model for production.

```{r}
# Your code here
```
