---
title: "BAN404 Mock Exam 2026 #1"
subtitle: "Ridge-first Generalization and Churn Operations"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Decode the objective: Ridge variant (10 points)

Given

```{r}
f <- function(beta, X, y, lambda) {
  sum((y - X %*% beta)^2) + lambda * sum(beta^2)
}
```

Identify the method and explain the shrinkage effect as `lambda` increases.

## (b) Baseline vs shrinkage fit comparison (10 points)

Fit OLS and Ridge on the dataset and compare coefficient stability under multicollinearity.

```{r}
# Your code here
```

## (c) Ten-fold CV tuning with robust choice rule (10 points)

Tune lambda with 10-fold CV and report both `lambda.min` and `lambda.1se`. Argue which choice is safer for exam-style deployment.

```{r}
# Your code here
```

## (d) Bootstrap uncertainty for variance (10 points)

Use bootstrap (`B = 1500`) for `Var(y)`, show histogram, and compare percentile vs normal CI.

```{r}
# Your code here
```

## (e) Linear vs additive model improvement (10 points)

Fit OLS and GAM. Compare MSE and adjusted R² and explain if smooth terms are justified.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Prediction-ready data design (8 points)

Prepare variables for realistic deployment, remove deterministic duplicates, and split train/test using a fixed seed.

```{r}
# Your code here
```

## (b) EDA for churn risk framing (8 points)

Build churn-rate summaries and two plots. State three testable churn hypotheses.

```{r}
# Your code here
```

## (c) Logistic model with FN-cost thresholding (12 points)

Fit logistic regression and choose threshold when FN cost is 3x FP cost. Report confusion metrics and AUC.

```{r}
# Your code here
```

## (d) Random forest hyperparameter tuning (12 points)

Tune `mtry` with CV, evaluate on test set, and relate top importance variables to your EDA hypotheses.

```{r}
# Your code here
```

## (e) Boosting selection and deployment recommendation (10 points)

Tune number of trees with staged AUC and provide final model recommendation for retention operations.

```{r}
# Your code here
```
