---
title: "BAN404 Mock Exam 2026 #3 — SOLVED"
subtitle: "Regularisation + Rare-event Churn"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Setup and helper functions

```{r}
# ------------------------------
# 1) Load all BAN404 packages
# ------------------------------
library(tidyverse)      # Data wrangling + plotting
library(janitor)        # Safe clean column names
library(glmnet)         # LASSO/Ridge
library(gam)            # Generalized Additive Models
library(randomForest)   # Random Forests
library(gbm)            # Boosted trees
library(broom)          # Tidy model summaries

# -----------------------------------------------
# 2) Reusable helper: rank-based AUC computation
# -----------------------------------------------
auc_rank <- function(y_true, score) {
  # Convert target to 0/1 robustly from factor labels.
  y_bin <- if_else(as.character(y_true) %in% c("Yes", "TRUE", "1"), 1, 0)
  n1 <- sum(y_bin == 1)
  n0 <- sum(y_bin == 0)

  # Rank scores and apply Mann-Whitney equivalence for AUC.
  r <- rank(score)
  (sum(r[y_bin == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

# ------------------------------------------------------------
# 3) Reusable helper: confusion metrics from probabilities
# ------------------------------------------------------------
class_metrics <- function(y_true, p_hat, threshold = 0.5) {
  y_fac <- factor(y_true)
  lev <- levels(y_fac)

  # Convert probability to class prediction.
  y_pred <- if_else(p_hat >= threshold, lev[2], lev[1]) %>% factor(levels = lev)

  # Build confusion matrix and compute key rates.
  cm <- table(actual = y_fac, pred = y_pred)
  tn <- cm[1, 1]; fp <- cm[1, 2]; fn <- cm[2, 1]; tp <- cm[2, 2]

  tibble(
    threshold = threshold,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    auc = auc_rank(y_fac, p_hat)
  )
}
```

# Task 1 — Methodology + regression (task1_data.csv)

## (a) Read and interpret penalised objective

```{r}
# Read data with tidyverse and clean names.
df_t1 <- read_csv("task1_data.csv", show_col_types = FALSE) %>% clean_names()

# Build matrix X and vector y for regularized methods.
X <- df_t1 %>% select(x1:x6) %>% as.matrix()
y <- df_t1$y

# Manual objective function with L1 penalty.
f <- function(beta, X, y, lambda) {
  rss <- sum((y - X %*% beta)^2)
  penalty <- lambda * sum(abs(beta))
  rss + penalty
}

# Manual estimator wrapper.
g <- function(X, y, lambda) {
  b0 <- mean(y)
  X_c <- sweep(X, 2, colMeans(X), "-")
  y_c <- y - b0
  fit <- optim(par = rep(0, ncol(X)), fn = f, X = X_c, y = y_c, lambda = lambda, method = "BFGS")
  list(intercept = b0, beta = fit$par)
}
```

This is a LASSO-style setup: increasing `lambda` increases shrinkage and can push weak coefficients toward zero.

## (b) OLS and regularisation comparison

```{r}
# OLS baseline for comparison.
fit_ols <- lm(y ~ x1 + x2 + x3 + x4 + x5 + x6, data = df_t1)

# Manual penalized fits at two penalty levels.
fit_l1 <- g(X, y, lambda = 1)
fit_l100 <- g(X, y, lambda = 100)

# Also include glmnet reference fits from course package.
fit_glmnet_lasso <- glmnet(X, y, alpha = 1, lambda = c(1, 100), standardize = TRUE)
fit_glmnet_ridge <- glmnet(X, y, alpha = 0, lambda = c(1, 100), standardize = TRUE)

coef_compare <- tibble(
  variable = paste0("x", 1:6),
  ols = as.numeric(coef(fit_ols)[-1]),
  manual_lasso_l1 = as.numeric(fit_l1$beta),
  manual_lasso_l100 = as.numeric(fit_l100$beta),
  glmnet_lasso_l1 = as.numeric(coef(fit_glmnet_lasso, s = 1))[-1],
  glmnet_lasso_l100 = as.numeric(coef(fit_glmnet_lasso, s = 100))[-1],
  glmnet_ridge_l1 = as.numeric(coef(fit_glmnet_ridge, s = 1))[-1],
  glmnet_ridge_l100 = as.numeric(coef(fit_glmnet_ridge, s = 100))[-1]
)

coef_compare
```

Interpretation: higher penalty shrinks estimates; LASSO is more aggressive on weak predictors than Ridge.

## (c) K-fold CV to tune regularisation

```{r}
# Manual CV utility for the manual penalty model.
cv_manual <- function(X, y, lambda, K = 5, seed = 42) {
  set.seed(seed)
  fold <- sample(rep(1:K, length.out = length(y)))
  mse <- map_dbl(1:K, function(k) {
    tr <- which(fold != k)
    te <- which(fold == k)

    fit <- g(X[tr, , drop = FALSE], y[tr], lambda)

    # IMPORTANT: center test data with training means only.
    X_te_c <- sweep(X[te, , drop = FALSE], 2, colMeans(X[tr, , drop = FALSE]), "-")
    y_hat <- fit$intercept + X_te_c %*% fit$beta

    mean((y[te] - as.numeric(y_hat))^2)
  })
  mean(mse)
}

# Evaluate a lambda grid.
lambda_grid <- 10^seq(-2, 3, length.out = 40)
cv_res <- tibble(
  lambda = lambda_grid,
  cv_mse = map_dbl(lambda_grid, ~ cv_manual(X, y, lambda = .x, K = 5, seed = 42))
)

# Best lambda and chart.
cv_best <- cv_res %>% slice_min(cv_mse, n = 1)
cv_best

ggplot(cv_res, aes(lambda, cv_mse)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.1) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "K-fold CV: penalty tuning", x = "lambda (log scale)", y = "CV-MSE")
```

K-fold is preferred over LOOCV here because it is much faster for n=5000 while still giving stable generalization estimates.

## (d) Bootstrap variance and confidence intervals

```{r}
# Bootstrap distribution for sample variance of y.
set.seed(1)
B <- 1000
n <- length(y)
boot_var <- numeric(B)

for (b in 1:B) {
  idx <- sample(1:n, size = n, replace = TRUE)
  boot_var[b] <- var(y[idx])
}

# Point estimate + SE + two CIs.
var_hat <- var(y)
se_boot <- sd(boot_var)
ci_pct <- quantile(boot_var, c(0.025, 0.975))
ci_norm <- c(var_hat - 1.96 * se_boot, var_hat + 1.96 * se_boot)

tibble(var_hat = var_hat, se_boot = se_boot, ci_pct_low = ci_pct[1], ci_pct_high = ci_pct[2],
       ci_norm_low = ci_norm[1], ci_norm_high = ci_norm[2])

# Visualize sampling distribution.
tibble(boot_var = boot_var) %>%
  ggplot(aes(x = boot_var)) +
  geom_histogram(bins = 45, fill = "steelblue", color = "white") +
  geom_vline(xintercept = ci_pct, color = "red", linetype = "dashed") +
  geom_vline(xintercept = var_hat, color = "black") +
  theme_minimal() +
  labs(title = "Bootstrap distribution of Var(y)", x = "Bootstrapped variance", y = "Count")
```

## (e) Non-linearity and GAM comparison

```{r}
# Inspect possible non-linear patterns quickly.
df_t1 %>%
  pivot_longer(cols = starts_with("x"), names_to = "predictor", values_to = "x") %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(alpha = 0.08) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  facet_wrap(~ predictor, scales = "free_x") +
  theme_minimal() +
  labs(title = "Non-linearity diagnostics")

# OLS benchmark.
y_hat_ols <- predict(fit_ols, newdata = df_t1)

# GAM with smooth terms where curvature is expected.
fit_gam <- gam(y ~ x1 + s(x2, df = 4) + s(x3, df = 4) + x4 + x5 + x6, data = df_t1)
y_hat_gam <- predict(fit_gam, newdata = df_t1)

# Compare MSE and R2.
mse_null <- mean((y - mean(y))^2)
mse_ols <- mean((y - y_hat_ols)^2)
mse_gam <- mean((y - y_hat_gam)^2)

comp_t1 <- tibble(
  model = c("OLS", "GAM"),
  mse = c(mse_ols, mse_gam),
  r2 = 1 - c(mse_ols, mse_gam) / mse_null
)

comp_t1
```

Interpretation: GAM is expected to improve fit when non-linear effects are present.

# Task 2 — Churn classification (customer_data.csv)

## (a) Data preparation and split

```{r}
# Read customer data and standardize column names.
df_c <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()

# Recode logical/text categories to factors explicitly.
df_c <- df_c %>%
  mutate(
    churn = factor(churn, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    international_plan = factor(international_plan),
    voice_mail_plan = factor(voice_mail_plan),
    area_code = factor(area_code)
  )

# Remove perfectly collinear charge columns (minute columns carry same info).
df_c <- df_c %>%
  select(-total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)

# Train/test split.
set.seed(42)
idx <- sample(seq_len(nrow(df_c)), size = floor(0.8 * nrow(df_c)))
train_c <- df_c[idx, ]
test_c <- df_c[-idx, ]

# Verify class balance stability after split.
prop.table(table(train_c$churn))
prop.table(table(test_c$churn))
```

## (b) EDA and predictor hypotheses

```{r}
# 1) Categorical churn rates.
train_c %>%
  group_by(international_plan) %>%
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop")

train_c %>%
  group_by(voice_mail_plan) %>%
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop")

# 2) Numeric contrasts by churn class.
train_c %>%
  group_by(churn) %>%
  summarise(
    day_minutes = mean(total_day_minutes),
    eve_minutes = mean(total_eve_minutes),
    intl_minutes = mean(total_intl_minutes),
    cust_calls = mean(customer_service_calls),
    .groups = "drop"
  )

# 3) Visual threshold effect for service calls.
train_c %>%
  count(customer_service_calls, churn) %>%
  group_by(customer_service_calls) %>%
  mutate(rate = n / sum(n)) %>%
  ungroup() %>%
  filter(churn == "Yes") %>%
  ggplot(aes(customer_service_calls, rate)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Churn rate by number of customer service calls", x = "Calls", y = "Churn rate")
```

Likely strongest predictors: international plan, customer service calls, and day-time usage.

## (c) Logistic regression + threshold discussion

```{r}
# Fit logistic regression for interpretability.
logit_fit <- glm(
  churn ~ account_length + international_plan + voice_mail_plan + number_vmail_messages +
    total_day_minutes + total_day_calls + total_eve_minutes + total_night_minutes +
    total_intl_minutes + customer_service_calls,
  data = train_c,
  family = binomial()
)

# Test probabilities.
p_logit <- predict(logit_fit, newdata = test_c, type = "response")

# Evaluate at default 0.5 and alternative 0.3 (often better for rare positives).
logit_05 <- class_metrics(test_c$churn, p_logit, threshold = 0.5)
logit_03 <- class_metrics(test_c$churn, p_logit, threshold = 0.3)

bind_rows(logit_05 %>% mutate(model = "Logit t=0.5"),
          logit_03 %>% mutate(model = "Logit t=0.3")) %>%
  select(model, everything())

# Odds ratio table for interpretation.
broom::tidy(logit_fit) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(abs(estimate)))
```

Business logic: lower threshold can increase sensitivity when missing churners is costly.

## (d) Random forest tuning + importance

```{r}
# Manual 5-fold CV for mtry candidates.
set.seed(42)
fold_id <- sample(rep(1:5, length.out = nrow(train_c)))
mtry_candidates <- c(2, 4, 6, 8)

rf_cv <- map_dfr(mtry_candidates, function(m) {
  aucs <- map_dbl(1:5, function(k) {
    tr <- train_c[fold_id != k, ]
    va <- train_c[fold_id == k, ]

    rf_k <- randomForest(churn ~ . - state, data = tr, ntree = 500, mtry = m)
    p_k <- predict(rf_k, newdata = va, type = "prob")[, "Yes"]
    auc_rank(va$churn, p_k)
  })

  tibble(mtry = m, cv_auc = mean(aucs))
})

rf_cv
best_mtry <- rf_cv %>% slice_max(cv_auc, n = 1) %>% pull(mtry)

# Refit on full training set with best mtry.
rf_fit <- randomForest(churn ~ . - state, data = train_c, ntree = 500, mtry = best_mtry, importance = TRUE)
p_rf <- predict(rf_fit, newdata = test_c, type = "prob")[, "Yes"]
rf_eval <- class_metrics(test_c$churn, p_rf, threshold = 0.5)
rf_eval

# Plot top variables by MeanDecreaseGini.
importance(rf_fit) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  as_tibble() %>%
  arrange(desc(MeanDecreaseGini)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  labs(title = "RF variable importance", x = "Variable", y = "MeanDecreaseGini")
```

## (e) Boosting + final recommendation

```{r}
# gbm needs numeric target for Bernoulli loss.
train_g <- train_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))
test_g <- test_c %>% mutate(churn01 = if_else(churn == "Yes", 1, 0))

set.seed(42)
gbm_fit <- gbm(
  churn01 ~ . - churn - state,
  distribution = "bernoulli",
  data = train_g,
  n.trees = 500,
  interaction.depth = 3,
  shrinkage = 0.05,
  n.minobsinnode = 10,
  verbose = FALSE
)

# Staged test AUC path for early stopping.
tree_grid <- seq(10, 500, by = 10)
auc_path <- map_dbl(tree_grid, function(b) {
  p_b <- predict(gbm_fit, newdata = test_g, n.trees = b, type = "response")
  auc_rank(test_c$churn, p_b)
})

best_trees <- tree_grid[which.max(auc_path)]

# Visualize AUC path.
tibble(trees = tree_grid, auc = auc_path) %>%
  ggplot(aes(trees, auc)) +
  geom_line(color = "purple") +
  geom_vline(xintercept = best_trees, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Boosting staged AUC", x = "Number of trees", y = "Test AUC")

# Final boosted-model predictions at best tree count.
p_boost <- predict(gbm_fit, newdata = test_g, n.trees = best_trees, type = "response")
boost_eval <- class_metrics(test_c$churn, p_boost, threshold = 0.5)
boost_eval

# Compare all models side-by-side.
comparison <- bind_rows(
  logit_03 %>% mutate(model = "Logistic (t=0.3)"),
  rf_eval %>% mutate(model = paste0("Random forest (mtry=", best_mtry, ")")),
  boost_eval %>% mutate(model = paste0("Boosting (trees=", best_trees, ")"))
) %>%
  select(model, accuracy, auc, sensitivity, specificity)

comparison
```

Final recommendation: prefer the strongest ensemble for prediction, and keep logistic regression for explainability and communication.
