---
title: "BAN404 Mock Exam 2026 #3"
subtitle: "Elastic Net Trade-offs and Segment-aware Churn"
author: "Candidate"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Identify Elastic Net objective mechanics (10 points)

Interpret this objective and explain how `alpha` controls Ridge/LASSO balance:

```{r}
f <- function(beta, X, y, lambda, alpha) {
  sum((y - X %*% beta)^2) + lambda * ((1 - alpha) * sum(beta^2) + alpha * sum(abs(beta)))
}
```

## (b) Ridge vs LASSO vs Elastic Net comparison (10 points)

Fit all three and compare CV-MSE and sparsity.

```{r}
# Your code here
```

## (c) Validation discipline under preprocessing risk (10 points)

Implement fold-specific preprocessing and explain how leakage inflates performance.

```{r}
# Your code here
```

## (d) Bootstrap coefficient stability (10 points)

Bootstrap OLS coefficients and identify the least stable predictor.

```{r}
# Your code here
```

## (e) GAM smoothness interpretation (10 points)

Fit GAM and interpret EDF values to decide which predictors are effectively non-linear.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data preparation and deployment assumptions (8 points)

Prepare a prediction-ready dataset and justify each exclusion in terms of real-time availability.

```{r}
# Your code here
```

## (b) Segment-centric exploratory analysis (8 points)

Construct risk segments and identify one high-risk persona for intervention.

```{r}
# Your code here
```

## (c) Logistic baseline with odds interpretation (12 points)

Fit logistic regression and explain at least three coefficients in odds-ratio language.

```{r}
# Your code here
```

## (d) Random forest interaction gains (12 points)

Tune and evaluate RF; explain where nonlinear interactions likely improve over logistic baseline.

```{r}
# Your code here
```

## (e) Boosting for final scorecard (10 points)

Tune staged boosting and deliver model recommendation with clear threshold policy.

```{r}
# Your code here
```
