---
title: "BAN404 Mock Exam 2026 #7"
subtitle: "LASSO Sparsity and Cost-Sensitive Churn"
author: "Candidate"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  eval: true
  echo: true
---

# Exam Information

- **Duration:** 6 hours
- **Datasets:** `../task1_data.csv`, `../customer_data.csv`
- **Point split:** Task 1 (50), Task 2 (50)

# Task 1 (50 points) — Methodology + Regression (`task1_data.csv`)

## (a) Method identification from custom code (10 points)

Interpret the following objective and discuss variable selection behavior.

```{r}
f <- function(beta, X, y, lambda) {
  sum((y - X %*% beta)^2) + lambda * sum(abs(beta))
}
```

## (b) OLS vs LASSO sparsity path (10 points)

Fit OLS and LASSO over a lambda grid. Show coefficient paths and identify first predictors to shrink to zero.

```{r}
# Your code here
```

## (c) Repeated K-fold CV (10 points)

Use repeated 5-fold CV to reduce variance in lambda choice. Explain why single split can be unstable.

```{r}
# Your code here
```

## (d) Bootstrap SD(y) and CI comparison (10 points)

Bootstrap SD(y) with B = 1200 and compare percentile and normal CIs.

```{r}
# Your code here
```

## (e) Basis expansion vs GAM (10 points)

Compare polynomial regression and GAM for non-linear effects. Comment on overfitting risk.

```{r}
# Your code here
```

# Task 2 (50 points) — Churn (`customer_data.csv`)

## (a) Data cleaning and leakage prevention (8 points)

Prepare factors, remove non-actionable columns, and define an 80/20 split.

```{r}
# Your code here
```

## (b) EDA and imbalance diagnosis (8 points)

Quantify class imbalance and produce churn-rate comparisons for key categorical and numeric predictors.

```{r}
# Your code here
```

## (c) Logistic model with cost matrix (12 points)

Choose threshold by minimizing expected cost where FN cost is 4x FP cost.

```{r}
# Your code here
```

## (d) Random forest (12 points)

Tune model and discuss top importance variables versus logistic coefficient signs.

```{r}
# Your code here
```

## (e) Gradient boosting (10 points)

Tune number of trees and provide final model recommendation for churn retention campaign.

```{r}
# Your code here
```
