---
title: "R from Zero — BAN404 Exam Foundation"
author: "BAN404 Self-Study"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
execute:
  echo: true
  eval: false
  warning: false
---

# R from Zero — Everything You Need for BAN404 (and Nothing Else)

> **This file assumes zero prior coding knowledge.** It teaches only the R concepts that appear on BAN404 exams. Read it once before studying the method-specific files.

---

# PART 1: R Language Basics

## 1.1 Running Code

Code runs line by line. In RStudio, place cursor on a line and press **Ctrl+Enter** (Windows) or **Cmd+Enter** (Mac).

```{r}
# This is a comment — R ignores it
2 + 2          # R calculates and shows: [1] 4
```

---

## 1.2 Storing Values (Variables)

```{r}
x <- 5          # Store the number 5 in a variable called x
x               # Show x: [1] 5
y <- x * 2      # y = 10
name <- "hello" # Text (strings) need quote marks
```

The `<-` arrow means "store the right side in the left side."

---

## 1.3 Vectors (a list of numbers)

```{r}
v <- c(1, 2, 3, 4, 5)    # c() combines values into a vector
length(v)                  # How many elements? [1] 5
v[2]                       # Get the 2nd element: [1] 2
v[c(1,3)]                  # Get elements 1 and 3: [1] 1 3
v[-2]                      # All elements EXCEPT 2nd: [1] 1 3 4 5
mean(v)                    # Average: [1] 3
sd(v)                      # Standard deviation
var(v)                     # Variance
sum(v)                     # Sum: [1] 15
```

---

## 1.4 Data Frames (a table of data — what exams give you)

```{r}
# A data frame is like a spreadsheet
df <- data.frame(
  age    = c(25, 30, 35),
  income = c(40000, 60000, 80000),
  churn  = c(0, 1, 0)
)

nrow(df)         # Number of rows
ncol(df)         # Number of columns
dim(df)          # Both: rows x columns
head(df)         # First 6 rows
str(df)          # Structure: column types
summary(df)      # Summary statistics per column

# Access a column with $ 
df$age           # [1] 25 30 35

# Access a specific cell: df[row, column]
df[2, 1]         # Row 2, column 1: [1] 30
df[, 1]          # All rows, column 1 (same as df$age)
df[2, ]          # All columns, row 2
```

---

## 1.5 Reading Data Files (how exams start)

```{r}
# Read a CSV file (comma-separated)
data <- read.csv("myfile.csv")

# Read a semicolon-separated file (European format)
data <- read.csv2("myfile.csv")

# Check what you got
head(data)
dim(data)
names(data)    # Column names
```

---

## 1.6 Factors (categorical variables — very important for models)

```{r}
# A "factor" is a categorical variable (like gender, class, zone)
x <- c("male", "female", "female", "male")
x_factor <- as.factor(x)
levels(x_factor)    # [1] "female" "male"

# EXAM TIP: Always convert category columns to factors before fitting models
df$gender <- as.factor(df$gender)
df$zone   <- as.factor(df$zone)
```

---

## 1.7 Logical Conditions (TRUE/FALSE)

```{r}
x <- 5
x > 3      # [1] TRUE
x == 5     # [1] TRUE  (== means "is equal to")
x != 4     # [1] TRUE  (! means NOT)

# Use on vectors
v <- c(1, 2, 3, 4, 5)
v > 3      # [1] FALSE FALSE FALSE TRUE TRUE
v[v > 3]   # Keep only elements > 3: [1] 4 5
```

---

## 1.8 For Loops (the engine of LOOCV and Bootstrap)

```{r}
# A for loop repeats code for each value in a sequence
for(i in 1:5) {
  print(i)    # Prints 1, then 2, then 3, then 4, then 5
}

# Typical exam pattern: pre-allocate a container, fill it in the loop
n <- 10
results <- numeric(n)    # Create an empty vector of 0s with n slots
for(i in 1:n) {
  results[i] <- i^2      # Store i-squared in slot i
}
results                  # [1]  1  4  9 16 25 36 49 64 81 100
```

---

## 1.9 Writing Functions

```{r}
# Functions let you package code to reuse
my_function <- function(x, y) {
  result <- x + y
  return(result)    # return() sends the answer back
}

my_function(3, 4)   # [1] 7

# Functions with default arguments
my_power <- function(x, power = 2) {
  return(x ^ power)
}
my_power(3)     # Uses default: 3^2 = 9
my_power(3, 3)  # Override: 3^3 = 27
```

---

## 1.10 Matrices (how X is stored in exam code)

```{r}
# A matrix is a 2D table of numbers
M <- matrix(c(1,2,3,4,5,6), nrow=2, ncol=3)
#      [,1] [,2] [,3]
# [1,]    1    3    5
# [2,]    2    4    6

M[1, 2]      # Row 1, column 2: [1] 3
nrow(M)      # [1] 2
ncol(M)      # [1] 3

# Exam code often does this: 
X <- as.matrix(df[, -1])   # All columns except the first → predictor matrix
y <- as.matrix(df[,  1])   # First column only → response vector

# Matrix multiply: %*%
A <- matrix(c(1,2,3,4), 2, 2)
b <- c(1, 0)
A %*% b     # Matrix × vector
```

---

## 1.11 sapply — Apply a Function to a List

```{r}
# sapply runs a function on each element of a vector/list
# This is how λ-grids and K-grids are searched on exams
lambdas <- c(0, 10, 100, 1000)
result <- sapply(lambdas, function(la) la^2)
# result = [1]      0    100  10000 1000000

# Equivalent to:
result2 <- numeric(length(lambdas))
for(i in 1:length(lambdas)) result2[i] <- lambdas[i]^2
```

---

# PART 2: The BAN404 Code Translator

> **This is the most important part for the exam.** When you see a code pattern, look it up here.

## 2.1 The Master Recognition Table

| If you see this in code | It is doing | Section |
|-------------------------|-------------|---------|
| `sample(..., replace=TRUE)` | Bootstrap (resampling with replacement) | 3.1 |
| `for(i in 1:n)` + `X[-i,]` | LOOCV (leave-one-out cross-validation) | 3.2 |
| `for(i in 1:K)` + `folds == i` | K-fold cross-validation | 3.3 |
| `sum((y - X%*%b)^2) + la*sum(b^2)` | Ridge regression objective | 3.4 |
| `sum((y - X%*%b)^2) + la*sum(abs(b))` | Lasso regression objective | 3.5 |
| `solve(t(X)%*%X) %*% t(X)%*%y` | OLS closed-form solution | 3.6 |
| `abs(x-x0)` + `order(d)[1:K]` | KNN (K-nearest neighbors) | 3.7 |
| `y - f2` / `y - f1` inside loop + `smooth.spline` | Backfitting (GAM) | 3.8 |
| `exp(eta)/(1+exp(eta))` | Logistic sigmoid → probability | 3.9 |
| `prob > 0.5` + `table(truth, predicted)` | Threshold + confusion matrix | 3.10 |
| `randomForest(y ~ ., mtry=...)` | Random Forest | 3.11 |
| `gbm(y ~ ., distribution="bernoulli"...)` | Gradient Boosting | 3.12 |
| `r <- y - fhat` inside loop | Boosting residual update | 3.12 |
| `for(s in candidates)` + RSS or Gini | Tree split selection | 3.13 |
| `nlminb(b1start, f, ...)` | Numerical optimizer (minimise a function) | 3.4 |

---

# PART 3: Minimum Runnable Examples

## 3.1 Bootstrap — Standard Error and 95% CI

```{r}
# Goal: estimate standard error of any statistic (e.g., mean, variance, coefficient)
set.seed(1)
n   <- length(y)          # Number of observations
B   <- 1000               # Number of bootstrap repetitions
res <- numeric(B)         # Container for bootstrap statistics

for(b in 1:B) {
  idx     <- sample(1:n, size=n, replace=TRUE)   # Sample WITH replacement
  yb      <- y[idx]                              # Bootstrap sample
  res[b]  <- var(yb)                             # YOUR STATISTIC HERE (e.g., var, mean, coef)
}

# Standard error
se_boot <- sd(res)

# 95% CI — Normal approximation
ci_normal <- mean(res) + c(-1, 1) * 1.96 * se_boot

# 95% CI — Percentile (no normality assumption — more robust)
ci_percentile <- quantile(res, c(0.025, 0.975))

# Histogram of sampling distribution
hist(res, breaks=50, main="Bootstrap distribution", xlab="Statistic")
```

**In plain English:** We fake-create 1000 new datasets by resampling rows from the original data (with replacement). We compute our statistic on each fake dataset. The spread (sd) of these 1000 values is the bootstrap standard error.

---

## 3.2 LOOCV — Leave-One-Out Cross-Validation

```{r}
# Goal: estimate test error when n is small, or tune a hyperparameter
n      <- nrow(X)
preds  <- numeric(n)

for(i in 1:n) {
  train_X <- X[-i, ]          # Train on all EXCEPT row i
  train_y <- y[-i]
  test_X  <- X[i, , drop=FALSE]
  
  b       <- solve(t(train_X)%*%train_X) %*% t(train_X)%*%train_y  # Fit OLS
  preds[i] <- test_X %*% b                                           # Predict row i
}

testMSE <- mean((y - preds)^2)
```

**In plain English:** We hold out one row at a time, train on the rest, predict the held-out row. We repeat for every row. Average error = LOOCV test MSE.

---

## 3.3 K-Fold Cross-Validation

```{r}
K     <- 5
n     <- nrow(X)
folds <- sample(rep(1:K, length.out=n))   # Randomly assign rows to K groups
errors <- numeric(K)

for(i in 1:K) {
  train_X <- X[folds != i, ]
  train_y <- y[folds != i]
  test_X  <- X[folds == i, ]
  test_y  <- y[folds == i]
  
  b        <- solve(t(train_X)%*%train_X) %*% t(train_X)%*%train_y
  preds    <- test_X %*% b
  errors[i] <- mean((test_y - preds)^2)
}

cv_error <- mean(errors)
```

---

## 3.4 Ridge Regression — Manual Implementation

```{r}
# Ridge objective function: RSS + lambda * sum(beta^2)
f <- function(b1, X, y, la) {
  sum((y - X %*% b1)^2) + la * sum(b1^2)
}

# Ridge estimator: demean X and y, then minimise f numerically
g <- function(X, y, la) {
  q      <- ncol(X)
  b0     <- mean(y)          # Intercept = mean(y) after demeaning
  yd     <- y - b0           # Demean y
  Xd     <- X
  for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])   # Demean each predictor
  
  b1start <- rep(0, q)       # Start optimiser at zero
  opt     <- nlminb(b1start, f, X=Xd, y=yd, la=la)
  return(opt$par)            # Return ridge coefficients
}

# LOOCV to find optimal lambda
lavec <- seq(0, 5000, length.out=20)
mses  <- sapply(lavec, function(la) {
  n  <- nrow(X)
  q  <- ncol(X)
  yd <- y - mean(y)
  Xd <- X; for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])
  preds <- numeric(n)
  for(i in 1:n) {
    b1       <- g(X=X[-i,], y=y[-i], la=la)
    preds[i] <- Xd[i,] %*% b1
  }
  mean((yd - preds)^2)
})
la_opt <- lavec[which.min(mses)]
```

---

## 3.5 Lasso with glmnet (Practical exam approach)

```{r}
library(glmnet)

# glmnet requires matrices, alpha=1 for Lasso, alpha=0 for Ridge
lasso_cv <- cv.glmnet(X, y, alpha=1, standardize=TRUE)
la_opt   <- lasso_cv$lambda.min     # λ with smallest CV error

# Fit final model at optimal λ
lasso_fit <- glmnet(X, y, alpha=1, lambda=la_opt)
coef(lasso_fit)   # Coefficients (zeros = selected out)

# Predict on new data
preds <- predict(lasso_fit, newx=X_test, s=la_opt)
```

---

## 3.6 OLS — Linear Regression

```{r}
# Method 1: lm() function (standard exam approach)
model <- lm(y ~ x1 + x2 + x3, data=df)
summary(model)               # Coefficients, p-values, R²
coef(model)                  # Extract coefficients
predict(model, newdata=df_test)   # Predictions on test data

# Method 2: Matrix formula (may appear in Task 1)
X_mat <- cbind(1, as.matrix(df[, c("x1","x2","x3")]))  # Add intercept column
b     <- solve(t(X_mat) %*% X_mat) %*% t(X_mat) %*% y
```

---

## 3.7 KNN Local Regression

```{r}
# KNN local regression (from 2024 exam)
f_knn <- function(x0, x, y, K=3) {
  d  <- abs(x - x0)          # Distance from x0 to all points
  o  <- order(d)[1:K]        # Indices of K nearest points
  xl <- x[o]; yl <- y[o]
  reg   <- lm(yl ~ xl, data=data.frame(xl, yl))
  ypred <- predict(reg, newdata=data.frame(xl=x0))
  return(ypred)
}

# LOOCV to find optimal K
Kgrid <- 1:10
mseK  <- sapply(Kgrid, function(K) {
  mean(sapply(1:n, function(i) {
    pred <- f_knn(x[i], x[-i], y[-i], K=K)
    (y[i] - pred)^2
  }))
})
K_opt <- Kgrid[which.min(mseK)]
```

---

## 3.8 GAM with Smoothing Splines

```{r}
library(gam)

# Fit GAM — s(x) means "smooth function of x"
gam_fit <- gam(y ~ s(x1) + x2 + x3, data=train)

# Compare MSE: OLS vs GAM
ols_fit  <- lm(y ~ x1 + x2 + x3, data=train)
MSE_ols  <- mean(residuals(ols_fit)^2)    # In-sample (training) MSE
MSE_gam  <- mean(residuals(gam_fit)^2)

# R² = fraction of variance explained
# Formula: 1 - MSE_model / Var(y)
MSE_null <- var(train$y)
R2_ols   <- 1 - MSE_ols / MSE_null
R2_gam   <- 1 - MSE_gam / MSE_null

# Plot smooth terms
plot(gam_fit, se=TRUE)
```

---

## 3.9 Logistic Regression

```{r}
# family=binomial → logistic regression
logreg <- glm(churn ~ age + income + tenure, data=train, family=binomial())
summary(logreg)          # Coefficients on log-odds scale

# Predict probabilities
prob <- predict(logreg, newdata=test, type="response")  # type="response" → probabilities

# Convert to classes with a threshold
threshold <- 0.5
pred_class <- ifelse(prob > threshold, 1, 0)

# Confusion matrix and accuracy
cm  <- table(Actual=test$churn, Predicted=pred_class)
cm
acc <- sum(diag(cm)) / sum(cm)
prop.table(cm, margin=1)    # Row proportions (sensitivity/specificity)
```

**Interpreting coefficients:** A coefficient of 0.3 on variable x means: holding other predictors constant, a 1-unit increase in x increases the **log-odds** of the outcome by 0.3. To get the odds ratio: `exp(0.3)` = 1.35 → 35% higher odds.

---

## 3.10 Confusion Matrix + Threshold Selection

```{r}
# Why use a different threshold?
# Default 0.5 predicts "no" when events are rare (e.g., 2% churn rate)
# A lower threshold catches more true positives at cost of more false positives

# Check class balance first
table(train$churn)   # If very imbalanced, use low threshold

# Try several thresholds, evaluate sensitivity (true positive rate)
thresholds <- seq(0.01, 0.5, by=0.01)
sensitivity <- sapply(thresholds, function(th) {
  pred <- prob > th
  cm   <- table(Actual=test$churn, Predicted=pred)
  # Sensitivity = TP / (TP + FN) = row 2, col 2 / row 2 sum
  if(nrow(cm)==2 && ncol(cm)==2) cm[2,2] / sum(cm[2,]) else NA
})
best_th <- thresholds[which.max(sensitivity)]

# Final confusion matrix at chosen threshold
pred_final <- prob > best_th
table(Actual=test$churn, Predicted=pred_final)
```

---

## 3.11 Random Forest

```{r}
library(randomForest)

# For classification: outcome must be a factor
train$churn <- as.factor(train$churn)
test$churn  <- as.factor(test$churn)

rf <- randomForest(churn ~ ., data=train, 
                   mtry=sqrt(ncol(train)-1),  # Default: sqrt(p) for classification
                   ntree=500,
                   importance=TRUE)

# Predict classes
pred_rf <- predict(rf, newdata=test)
table(Actual=test$churn, Predicted=pred_rf)

# Predict probabilities (for threshold tuning)
prob_rf <- predict(rf, newdata=test, type="prob")   # Matrix, 1 col per class

# Variable importance plot
varImpPlot(rf)   # Higher = more important
importance(rf)   # Numeric values: MeanDecreaseAccuracy, MeanDecreaseGini
```

---

## 3.12 Boosting (gbm)

```{r}
library(gbm)

# Binary classification: distribution="bernoulli", y must be 0/1 numeric
train$churn <- as.numeric(as.character(train$churn))   # Factor → 0/1 numeric

bt <- gbm(churn ~ ., data=train,
          distribution="bernoulli",   # Binary outcome
          n.trees=1000,
          interaction.depth=3,        # Tree depth
          shrinkage=0.01,             # Learning rate (smaller = slower but better)
          cv.folds=5)                 # Cross-validation to find optimal n.trees

best_trees <- which.min(bt$cv.error)   # Optimal n.trees from CV

# Predict probabilities
prob_bt <- predict(bt, newdata=test, n.trees=best_trees, type="response")
pred_bt <- prob_bt > 0.5

# Confusion matrix
table(Actual=test$churn, Predicted=pred_bt)
```

---

## 3.13 Decision Tree

```{r}
library(tree)

# Regression tree
tree_reg <- tree(y ~ ., data=train)
plot(tree_reg); text(tree_reg, pretty=0)   # Visualize
pred_tree <- predict(tree_reg, newdata=test)
MSE_tree  <- mean((test$y - pred_tree)^2)

# Classification tree (outcome must be factor)
tree_cls <- tree(as.factor(churn) ~ ., data=train)
plot(tree_cls); text(tree_cls, pretty=0)
pred_cls  <- predict(tree_cls, newdata=test, type="class")
table(Actual=test$churn, Predicted=pred_cls)
```

---

# PART 4: Common R Errors and Fixes

## Error lookup table (exam panic guide)

| Error message | Likely cause | Fix |
|---------------|-------------|-----|
| `object 'X' not found` | Variable doesn't exist yet | Check spelling, run the line that creates X |
| `non-conformable arguments` | Matrix dimensions don't match | Check `dim(X)` and `length(y)` |
| `factor has different levels` | Train/test have different factor levels | Use `droplevels()` or refit with full data |
| `$ operator is invalid for atomic vectors` | Using `$` on a non-data-frame | Use `[` indexing instead |
| `system is computationally singular` | Matrix can't be inverted (collinear predictors) | Remove duplicated/collinear columns |
| `Error in predict: object is not a model` | Fitted object stored incorrectly | Check that model was saved with `<-` |
| `subscript out of bounds` | Accessing an index that doesn't exist | Check `length()` or `nrow()` of the object |
| `package not found` | Package not installed | `install.packages("packagename")` |
| NaN or NA in predictions | Data contains missing values | `na.omit(df)` or `complete.cases()` |

## Quick fixes checklist

```{r}
# Check for missing values
sum(is.na(df))
df <- na.omit(df)          # Remove rows with any NA

# Make sure factor columns are factors
str(df)                    # Check types
df$col <- as.factor(df$col)  # Convert if needed

# Check matrix dimensions before multiplying
dim(X); length(y)

# Make predictions safe: type="response" for glm, type="class" for tree classification
predict(model, newdata=test, type="response")
```

---

# PART 5: R² Formula Reference

```{r}
# R² measures how much variation the model explains
# Formula: R² = 1 - MSE_model / MSE_null
# where MSE_null = Var(y) = mean squared deviation from the mean

MSE_null  <- mean((y - mean(y))^2)   # = var(y) * (n-1)/n
MSE_model <- mean((y - fitted(model))^2)

R2 <- 1 - MSE_model / MSE_null
cat("R² =", round(R2, 4))

# Note: lm() reports adjusted R², which is similar but penalises extra predictors
# For exam purposes, the formula above is sufficient
```

---

# PART 6: Packages You Must Have Installed

```{r}
# Run this BEFORE the exam to install all needed packages:
pkgs <- c("MASS", "ISLR2", "car", "class", "e1071", "leaps", "glmnet",
          "pls", "splines", "gam", "tree", "randomForest", "gbm",
          "ROCR", "Matrix", "insuranceData")

for(p in pkgs) {
  if(!requireNamespace(p, quietly=TRUE)) install.packages(p)
}
```
