---
title: "BAN404 Exam Foundation — Practical Cheatsheet in Tidyverse"
author: "BAN404 Self-Study"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

# Why this file exists

This file is for the last days before the exam. It is written to be:

- **tidyverse-first** so the code looks like the style you are used to
- **runnable** so you can copy, run, inspect output, and learn from it
- **over-commented** so you do not have to guess why a line is there
- **exam-focused** so it starts with the most likely practical tasks

> Important: the examples below use small built-in datasets so that the code is self-contained. On the exam, you will keep the same structure and only swap dataset names and variable names.

---

# 1. Setup you can trust

```{r}
# Load the packages used most often in the exam workflow                 # tidyverse = read, clean, plot, summarise
library(tidyverse)                                                       # gives dplyr, ggplot2, readr, tibble, stringr, purrr, etc.

# Set a seed before splits or resampling so your results are reproducible # important when you must explain your exact output
set.seed(123)
```

## A small runnable demo dataset

```{r}
# Start from mtcars because it exists in base R installations            # safe built-in dataset for practice
cars_tbl <- mtcars |>
  tibble::rownames_to_column(var = "model") |>
  as_tibble() |>
  mutate(
    transmission = factor(am, levels = c(0, 1), labels = c("automatic", "manual")), # binary target example
    cylinders    = factor(cyl),                                                        # discrete count -> category for many plots
    engine_shape = factor(vs, levels = c(0, 1), labels = c("v-shaped", "straight")), # another factor for EDA
    gears        = factor(gear)                                                        # categorical version of number of gears
  ) |>
  select(model, mpg, disp, hp, wt, qsec, transmission, cylinders, engine_shape, gears) # keep a compact practice table

cars_tbl |> glimpse()                                                     # always inspect structure before doing anything else
cars_tbl |> head()                                                        # first rows help you understand meaning and formatting
```

**What to notice:**
- `transmission` is now a clean binary target for classification practice
- some columns stay numeric (`mpg`, `hp`, `wt`)
- some columns are made into factors because they are categories in practice

---

# 2. Read real exam data without getting stuck

## 2.1 Read the file

```{r}
# Comma-separated file                                                    # use this if the file looks like a normal CSV
exam_data <- readr::read_csv("data.csv")

# Semicolon-separated file                                                # use this if numbers use comma decimals or the file is European-style
exam_data <- readr::read_delim("data.csv", delim = ";")
```

## 2.2 First inspection checklist

```{r}
exam_data |> glimpse()                                                    # fastest way to see column names + types + sample values
exam_data |> head()                                                       # first rows to catch weird formatting early
exam_data |> summary()                                                    # quick univariate summary for all variables
names(exam_data)                                                          # useful if you want to copy exact variable names safely
nrow(exam_data)                                                           # number of rows = sample size
ncol(exam_data)                                                           # number of columns = feature count
```

## 2.3 Questions to ask immediately

1. **What is the target?**
2. **Is the target already binary, or must you create it?**
3. **Which variables are categories and should become factors?**
4. **Which variables are leakage, IDs, or post-event variables and should be removed?**
5. **Are there missing values you must handle before modeling?**

---

# 3. Tidyverse cleaning and recoding

## 3.1 Basic mutate / select / filter patterns

```{r}
# Create a cleaned practice copy                                          # work on a new object so you can always go back if needed
cars_clean <- cars_tbl |>
  mutate(
    weight_ton = wt / 2.205,                                              # example of creating a new variable from an old one
    heavy_car  = if_else(wt > median(wt), "heavy", "light"),             # binary label from a numeric split
    heavy_car  = factor(heavy_car)                                         # convert character label to factor for plotting/modeling
  ) |>
  filter(mpg > 10) |>                                                     # keep only rows satisfying a condition
  select(-model)                                                          # remove a column when it is only an ID / label helper

cars_clean |> glimpse()
```

## 3.2 Convert categories to factors

```{r}
# This is one of the most common exam clean-up steps                      # models should know these are categories, not numbers to average
cars_model <- cars_tbl |>
  mutate(
    across(c(transmission, cylinders, engine_shape, gears), as.factor)    # tidyverse way to convert several columns at once
  )

cars_model |> glimpse()
```

## 3.3 Remove leakage or useless columns

```{r}
# Example pattern for exam data                                           # adapt the variable names to the actual dataset
leakage_vars <- c("claim_amount_after_event", "raw_target_count")         # values known only after the outcome happened
id_vars      <- c("customer_id", "policy_number")                        # unique identifiers usually add no predictive signal

# exam_model_data <- exam_data |>
#   select(-any_of(c(leakage_vars, id_vars)))                              # any_of() is safe even if one name is missing
```

**What to write in the exam:**
“I removed `[variable]` because it is a leakage variable / post-event variable / pure identifier. Keeping it would let the model use information that would not be available at prediction time.”

## 3.4 Create a binary target if needed

```{r}
# Example pattern if the raw variable is a count                          # common BAN404 move: turn event count into yes/no target
# exam_model_data <- exam_data |>
#   mutate(target = if_else(number_of_claims >= 1, "yes", "no"),         # define the event of interest clearly
#          target = factor(target, levels = c("no", "yes")))              # set reference level explicitly
```

---

# 4. Train / test split in tidyverse-friendly style

```{r}
set.seed(123)                                                             # same split every time if you re-run the code

n_obs      <- nrow(cars_model)                                            # count rows once so the split code is readable
train_index <- sample.int(n_obs, size = floor(0.5 * n_obs))               # 50/50 split is very common in BAN404 examples

train_data <- cars_model |> slice(train_index)                            # training set = sampled rows
test_data  <- cars_model |> slice(-train_index)                           # test set = all remaining rows

train_data |> count(transmission) |> mutate(prop = n / sum(n))            # always check class balance in training data
test_data  |> count(transmission) |> mutate(prop = n / sum(n))            # and also in test data
```

**Why this step matters:**
You fit models on `train_data` and evaluate them on `test_data`. That is how you estimate whether the model generalizes to new data.

---

# 5. EDA — which plot should you use?

## 5.1 Decision guide

| Situation | Good first choice | Why |
|---|---|---|
| Numeric predictor vs binary target | Boxplot or violin plot | Shows whether the distributions differ across groups |
| Numeric predictor alone | Histogram + density | Shows skewness, spread, and possible outliers |
| Categorical predictor vs binary target | Proportion table + bar plot | Shows whether event rates differ by category |
| Numeric vs numeric | Scatter plot | Shows direction, shape, and possible non-linearity |
| Many numeric predictors together | Correlation heatmap | Quick scan for strong relationships and collinearity |

> Exam rule: if you are unsure, start with a simple plot that helps you compare groups clearly.

---

# 6. EDA — full runnable examples

## 6.1 Histogram for one numeric variable

```{r}
ggplot(cars_model, aes(x = mpg)) +                                        # choose the numeric variable you want to inspect
  geom_histogram(bins = 12, fill = "steelblue", color = "white") +      # histogram = first look at shape, spread, and outliers
  labs(
    title = "Distribution of mpg",                                       # title should say exactly what the plot shows
    x = "Miles per gallon",
    y = "Count"
  ) +
  theme_minimal()
```

**What you are looking for:**
- Is the variable symmetric or skewed?
- Are there extreme values?
- Does it look multi-modal?

## 6.2 Histogram split by target group

```{r}
ggplot(cars_model, aes(x = mpg, fill = transmission)) +                   # fill by target to compare the two groups directly
  geom_histogram(alpha = 0.60, bins = 10, position = "identity") +       # overlap shows whether groups live in different regions
  labs(
    title = "mpg by transmission group",
    x = "Miles per gallon",
    y = "Count",
    fill = "Transmission"
  ) +
  theme_minimal()
```

**When this helps:**
Use it when you want to see whether one class tends to have higher or lower values than the other.

## 6.3 Boxplot for numeric predictor vs binary target

```{r}
ggplot(cars_model, aes(x = transmission, y = hp, fill = transmission)) +  # boxplot is usually the clearest first choice in Task 2
  geom_boxplot(alpha = 0.80, outlier.color = "red") +                    # red outliers help you spot extreme observations fast
  labs(
    title = "Horsepower by transmission",
    x = "Transmission",
    y = "Horsepower"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**What to write after this plot:**
“`hp` looks useful because the distributions differ across the target groups. The medians are not the same, and the manual group tends to have higher horsepower.”

## 6.4 Group summary table that supports the boxplot

```{r}
cars_model |>
  group_by(transmission) |>
  summarise(
    mean_hp   = mean(hp),                                                  # average helps quantify the visual difference
    median_hp = median(hp),                                                # median is robust to outliers
    sd_hp     = sd(hp),                                                    # spread matters too
    n         = n(),                                                       # always useful to know how many observations support the pattern
    .groups   = "drop"
  )
```

## 6.5 Bar plot for categorical predictor vs binary target

```{r}
cars_model |>
  count(cylinders, transmission) |>
  group_by(cylinders) |>
  mutate(prop = n / sum(n)) |>                                             # within each cylinder group, compute class proportions
  ungroup() |>
  ggplot(aes(x = cylinders, y = prop, fill = transmission)) +
  geom_col(position = "dodge") +                                          # side-by-side bars make comparison easier than stacked bars here
  scale_y_continuous(labels = scales::percent_format()) +                  # show proportions as percentages
  labs(
    title = "Transmission mix inside each cylinder group",
    x = "Cylinders",
    y = "Share of cars",
    fill = "Transmission"
  ) +
  theme_minimal()
```

**What to write after this plot:**
“`cylinders` seems informative because the transmission proportions change noticeably across cylinder groups. That suggests the category may help predict the target.”

## 6.6 Scatter plot for numeric vs numeric

```{r}
ggplot(cars_model, aes(x = wt, y = mpg, color = transmission)) +          # start by checking direction and rough shape of the relationship
  geom_point(size = 2.5, alpha = 0.85) +                                   # each point is one observation
  geom_smooth(method = "lm", se = FALSE) +                                # straight fitted lines help you judge linear tendency quickly
  labs(
    title = "Weight vs mpg",
    x = "Weight",
    y = "Miles per gallon",
    color = "Transmission"
  ) +
  theme_minimal()
```

**What you are looking for:**
- negative or positive association
- possible curvature
- outliers that may influence the model

## 6.7 Correlation heatmap for collinearity check

```{r}
corr_tbl <- cars_model |>
  select(where(is.numeric)) |>                                             # keep only numeric columns because correlation needs numbers
  cor() |>                                                                  # compute the correlation matrix
  as.data.frame() |>                                                        # turn matrix into data frame for tidy reshaping
  tibble::rownames_to_column(var = "var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")     # long format is easier for ggplot heatmaps

ggplot(corr_tbl, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +                                                             # one tile per variable pair
  geom_text(aes(label = round(correlation, 2)), size = 3) +                # print the correlation on top so you do not have to guess shades
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  labs(
    title = "Correlation heatmap for numeric predictors",
    x = NULL,
    y = NULL,
    fill = "Corr"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**How to use this in the exam:**
- very high absolute correlations suggest possible collinearity
- if two predictors tell almost the same story, OLS/logistic coefficients can become unstable
- this is where you may mention **ridge** or **lasso** as regularized alternatives if needed

**Example interpretation sentence:**
“Some predictors appear strongly correlated, so coefficient-based models may become unstable. If that causes problems, a regularized model such as ridge or lasso could be more robust.”

## 6.8 Quick numeric screening table

```{r}
numeric_screen <- cars_model |>
  group_by(transmission) |>
  summarise(
    across(where(is.numeric), list(mean = mean, median = median), .names = "{.col}_{.fn}"),
    .groups = "drop"
  )

numeric_screen
```

This is useful when you want a compact table to support statements like:

If the table becomes too wide, keep only the few variables you actually want to discuss, or reshape it after summarising so it is easier to read under exam pressure.

- “manual cars have higher average `hp`”
- “automatic cars have higher average `wt`”

---

# 7. Missing values and quick checks

```{r}
# Count missing values per column                                           # do this before modeling so you know if imputation is needed
cars_model |>
  summarise(across(everything(), ~ sum(is.na(.x))))

# Share of missing values per column                                        # more interpretable than raw counts when n is large
cars_model |>
  summarise(across(everything(), ~ mean(is.na(.x))))
```

If you find missing values, your next question is:
- can I drop a few rows safely?
- do I need to impute?
- should I mention that the missingness may affect conclusions?

---

# 8. Fast interpretation helpers

## 8.1 After a histogram

Write about:
- centre
- spread
- skewness
- outliers

## 8.2 After a boxplot

Write about:
- median differences
- overlap between groups
- spread / IQR
- outliers
- whether the predictor seems useful

## 8.3 After a bar plot with proportions

Write about:
- which category has the highest event rate
- whether the differences are small or large
- whether the variable looks promising for prediction

## 8.4 After a correlation heatmap

Write about:
- strong positive or negative relationships
- possible redundancy among predictors
- whether regularization may help

---

# 9. A copy-adjust-use starter workflow for Task 2

```{r}
# 1. Read data                                                             # start simple and make sure the file loads correctly
# data_raw <- readr::read_csv("your_file.csv")

# 2. Inspect                                                               # never skip this or you may model the wrong variable types
# data_raw |> glimpse()
# data_raw |> summary()

# 3. Clean
# data_clean <- data_raw |>
#   mutate(across(c(var1, var2, var3), as.factor)) |>                      # replace with actual categorical variables
#   select(-any_of(c("id", "leakage_var")))                               # remove IDs and leakage if present

# 4. Split
# set.seed(123)
# idx <- sample.int(nrow(data_clean), size = floor(0.5 * nrow(data_clean)))
# train <- data_clean |> slice(idx)
# test  <- data_clean |> slice(-idx)

# 5. EDA
# ggplot(train, aes(x = target, y = numeric_var, fill = target)) +
#   geom_boxplot()

# train |>
#   count(category_var, target) |>
#   group_by(category_var) |>
#   mutate(prop = n / sum(n))

# 6. Only after this move to the modeling files                             # EDA should tell you what looks promising
```

---

# 10. Final exam reminders

1. Start with a **simple clean workflow** before trying advanced models.
2. Use plots that directly answer the question you need to answer.
3. Support plots with one small summary table if it helps interpretation.
4. If predictors look highly correlated, mention ridge/lasso as a sensible next step.
5. Keep the loop short: **run -> inspect -> interpret -> adjust -> re-run**.
