---
title: "Task 1 Mastery — Method-by-Method Exam Preparation"
author: "BAN404 Self-Study"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  echo: true
  eval: false
  warning: false
---

# How to Use This File

Every Task 1 on BAN404 exams follows this formula:
1. You are **given R functions** (you did not write them)
2. You must **explain what they do**
3. You must **use them** on data
4. You must **modify or extend** them

This file teaches you to do all four steps for every high-probability method.

**For each method:**
- **Recognition cue** — the single line or pattern that identifies the method instantly
- **Plain English** — what the function is doing, step by step
- **Why it is used** — the statistical motivation
- **Key assumptions** — what must be true for the method to be valid
- **Exam follow-up patterns** — the typical modifications/extensions examiners ask
- **Full R code** — working implementation you can use
- **Written answer template** — fill-in-the-blank exam answer

---

# METHOD 1: Ridge Regression ⭐ HIGHEST PRIORITY

## Recognition Cue
```{r}
sum((y - X%*%b1)^2) + la * sum(b1^2)    # THE TELL: b1^2 penalty
```
The key is `sum(b1^2)` — squaring the coefficients. This is the L2 penalty.

Also look for `nlminb(b1start, f, ...)` — numerical optimizer minimising this function.

## Plain English Explanation
The function `f` computes a score (objective value) for a set of regression coefficients `b1`.
- The first part `sum((y - X%*%b1)^2)` is the **residual sum of squares (RSS)** — it measures how well the coefficients fit the data.
- The second part `la * sum(b1^2)` is the **penalty** — it punishes large coefficients.
- The optimizer finds the `b1` that makes this score as small as possible.
- The function `g` prepares the data by **demeaning** (centering) y and each column of X, so the penalty applies fairly to all predictors.

## Why It Is Used
Ridge regression reduces **variance** in the coefficient estimates. When:
- Predictors are correlated (multicollinearity)
- The number of predictors is large relative to observations
- The regular OLS estimates are unstable

Ridge shrinks coefficients toward zero without setting them exactly to zero.

## Key Assumptions
1. Predictors and response should be **demeaned** (centered) before applying ridge so the intercept is handled separately
2. Ideally predictors should be on the same scale (standardized), though the 2025 exam version only demeans, not scales
3. A **tuning parameter λ (lambda)** must be chosen, typically via LOOCV

## Exam Follow-Up Patterns

### Pattern A: Compare OLS vs Ridge
```{r}
# OLS via lm
b_ols <- lm(y ~ X)$coef[-1]

# Ridge with la=0 (should equal OLS)
b_r0 <- g_ridge(X, y, la=0)

# Ridge with la=1000 (shrinkage)
b_r1000 <- g_ridge(X, y, la=1000)

# Compare all three
round(cbind(OLS=b_ols, Ridge_0=b_r0, Ridge_1000=b_r1000), 4)

# Quantify shrinkage
sum(b_ols^2)      # Large
sum(b_r1000^2)    # Smaller — ridge shrinks
```

### Pattern B: LOOCV to find optimal λ
```{r}
# Broad grid first
lavec <- seq(0, 10000, length.out=20)
mses  <- sapply(lavec, function(la) loo_ridge(la, X, y))
plot(lavec, mses, type="l", main="LOOCV MSE vs lambda")

# Refine around the minimum
la_min_broad <- lavec[which.min(mses)]
lavec2 <- seq(la_min_broad - 1000, la_min_broad + 1000, length.out=20)
mses2  <- sapply(lavec2, function(la) loo_ridge(la, X, y))
la_opt <- lavec2[which.min(mses2)]
cat("Optimal lambda:", la_opt)
```

### Pattern C: Is standardization necessary?
Answer: The demeaning in `g` ensures the intercept is handled correctly (= mean(y)). Demeaning is **necessary** for the intercept trick. Full standardization (dividing by sd) is needed if predictors are on very different scales, so the penalty treats them equally. The 2025 exam's `g` function only demeans (not divides by sd), so it applies a slightly unfair penalty to variables with larger scales.

## Written Answer Template

> "The function `f` computes the **ridge regression objective function**: the sum of squared residuals (RSS) plus a penalty term `λ · Σβ²`. This is **ridge regression** (L2-penalized regression).
>
> The function `g` estimates the ridge regression coefficients for a given penalty parameter `λ`. It first demeans both y and each column of X so that the intercept equals the mean of y. It then uses numerical optimization (`nlminb`) to minimize `f` and returns the optimal coefficients.
>
> When `λ = 0`, ridge reduces to OLS and the coefficients are identical. When `λ` increases, coefficients are **shrunk toward zero** — this reduces variance at the cost of a small increase in bias, which often improves test performance when predictors are numerous or correlated."

---

# METHOD 2: LOOCV — Leave-One-Out Cross-Validation ⭐ HIGHEST PRIORITY

## Recognition Cue
```{r}
for(i in 1:n) {
  b1 <- g(X=X[-i,], y=y[-i], la)   # THE TELL: X[-i,] removes row i
  preds[i] <- X[i,] %*% b1
}
```
The key is `X[-i,]` — excluding row i from training.

## Plain English Explanation
The loop runs once for every observation (n times total).
- In each iteration i: remove observation i from the dataset
- Fit the model on the remaining n-1 observations
- Predict observation i using the fitted model
- Store the prediction error `(y[i] - pred[i])^2`
At the end, return the average squared prediction error across all n iterations.

## Why It Is Used
LOOCV gives an **almost unbiased** estimate of how well the model generalizes to new data. It is especially used when:
- The dataset is small (n is small)
- You need to choose a tuning parameter (λ for ridge, K for KNN)
- You want to compare models honestly

## Key Assumptions
- Observations are **independent** (i.i.d.)
- The data is shuffled or randomly ordered (so leaving out one row gives a random test case)

## Exam Follow-Up Patterns

### Pattern A: Why demean inside the LOOCV loop?
The demeaning must happen on the **training fold only** in each iteration. The test prediction uses the demeaned X values from the full dataset because the means of the training fold were used to center X during model fitting. This ensures consistency between training and prediction.

### Pattern B: λ grid search with LOOCV
```{r}
# This is the most common exam task
lavec <- seq(0, 5000, length.out=20)
mses  <- sapply(lavec, function(la) loo_ridge(la, X, y))
la_opt <- lavec[which.min(mses)]
cat("Optimal lambda:", la_opt)

# Best model using optimal lambda
b_opt <- g_ridge(X, y, la=la_opt)
```

## Written Answer Template

> "The function `loo` implements **leave-one-out cross-validation (LOOCV)**. In each iteration `i`, it removes observation `i` from the dataset, fits the ridge model on the remaining `n-1` observations using function `g`, and predicts observation `i` using the held-out predictor values. The function returns the average squared prediction error (testMSE) across all `n` iterations.
>
> LOOCV is used here to **choose the optimal λ**: we evaluate the LOOCV test MSE for a grid of λ values and select the λ that minimizes it. This is the standard method for hyperparameter selection when n is small, as it uses the data efficiently."

---

# METHOD 3: Bootstrap ⭐ HIGHEST PRIORITY

## Recognition Cue
```{r}
idx <- sample(1:n, size=n, replace=TRUE)   # THE TELL: replace=TRUE
```
The key is `replace=TRUE` in `sample()`.

## Plain English Explanation
The bootstrap estimates the **sampling distribution** of any statistic (mean, variance, coefficient, etc.).
- Create a "fake" dataset of size n by drawing rows **with replacement** (so some rows appear twice, others not at all)
- Compute the statistic on this fake dataset
- Repeat B times (typically B=1000 or 10000)
- The spread of the B estimates is the **standard error**
- The 2.5% and 97.5% percentiles give a **95% confidence interval**

## Why It Is Used
The bootstrap is used when:
- The analytic formula for the standard error is complex or unknown
- You want to estimate uncertainty without distributional assumptions
- You want a confidence interval for any statistic

## Exam Follow-Up Patterns

### Pattern A: Bootstrap for variance of y (from 2025 exam)
```{r}
set.seed(1)
n <- length(y); B <- 1000
vboot <- numeric(B)
for(b in 1:B) {
  yb      <- sample(y, size=n, replace=TRUE)
  vboot[b] <- var(yb)
}
hist(vboot, breaks=50, main="Bootstrap distribution of Var(y)")
# 95% CI — percentile method (no normality assumption)
quantile(vboot, c(0.025, 0.975))
# 95% CI — normal approximation
mean(vboot) + c(-1,1)*1.96*sd(vboot)
```

### Pattern B: Bootstrap for regression coefficient (from 2021 exam)
```{r}
set.seed(1); B <- 1000
coef_boot <- numeric(B)
for(b in 1:B) {
  idx <- sample(1:n, replace=TRUE)
  mod <- lm(y ~ x, data=data[idx, ])
  coef_boot[b] <- coef(mod)[2]   # Slope coefficient
}
sd(coef_boot)   # Bootstrap SE
# Compare to lm SE:
summary(lm(y ~ x, data=data))$coef[2, 2]
```

### Pattern C: Two types of CI
- **Normal approximation:** `mean(res) ± 1.96 × sd(res)` — assumes bootstrap distribution is bell-shaped
- **Percentile method:** `quantile(res, c(0.025, 0.975))` — uses empirical quantiles, more robust for skewed distributions

## Written Answer Template

> "This function implements the **bootstrap** method. It estimates the sampling distribution of [INSERT STATISTIC] by resampling.
>
> In each of the `B` iterations, a random sample of size `n` is drawn **with replacement** from the data (`replace=TRUE`). The statistic is computed on each bootstrap sample. The standard deviation of the `B` bootstrap values is the **bootstrapped standard error**.
>
> A 95% confidence interval can be computed in two ways: (1) the **normal approximation** assumes the bootstrap distribution is approximately normal; (2) the **percentile method** uses the 2.5th and 97.5th percentiles of the bootstrap distribution directly, requiring no distributional assumption."

---

# METHOD 4: GAM — Generalized Additive Model ⭐ HIGH PRIORITY

## Recognition Cue
```{r}
f1 <- smooth.spline(x1, y - f2)$y   # THE TELL: smooth.spline on residuals
f2 <- smooth.spline(x2, y - f1)$y   # ...and y-f1, y-f2 in alternating loop
```
Also: `gam(y ~ s(x1) + x2)` — `s()` means smooth term.

## Plain English Explanation

### Backfitting algorithm (when shown as code):
The loop alternates between fitting two smooth functions:
1. Start: both f1 and f2 set to 0 (flat line)
2. Fit f1 to the partial residuals (y minus the current f2)
3. Update f2 to the partial residuals (y minus the updated f1)
4. Repeat until changes are tiny
5. Return f1 + f2 as the final GAM prediction

### Using gam library:
`gam(y ~ s(x1) + x2)` fits a smooth function for x1 and a linear function for x2. The `s()` function tells GAM to be flexible (non-linear) for x1.

## Why It Is Used
- When some predictors have **non-linear** relationships with y
- A GAM extends linear regression by allowing each predictor to have its own smooth function
- Retains interpretability: you can plot each predictor's effect separately
- If the relationship is truly linear, the smooth s(x) will be approximately linear

## How to Detect Non-Linearity (Exam Task 1e/2e style)
```{r}
# Scan all predictors with scatter plots
par(mfrow=c(4, 5))
for(j in 1:ncol(X)) {
  plot(X[,j], y, main=paste("X", j), xlab="", ylab="y")
}
# Look for curved patterns, U-shapes, step patterns
```

## Full Example
```{r}
library(gam)

# Fit GAM with variable X3 as smooth (identified as non-linear)
gam1 <- gam(y ~ s(X3) + X1 + X2 + X4, data=train)
plot(gam1, se=TRUE)   # Shows partial effects

# Compare MSE
ols  <- lm(y ~ ., data=train)
MSE_ols  <- mean(residuals(ols)^2)
MSE_gam  <- mean(residuals(gam1)^2)
MSE_null <- var(train$y)

R2_ols <- 1 - MSE_ols/MSE_null
R2_gam <- 1 - MSE_gam/MSE_null
cat("R² OLS:", round(R2_ols,4), "  R² GAM:", round(R2_gam,4))
```

## Written Answer Template

> "This code implements a **Generalized Additive Model (GAM)** using the backfitting algorithm. GAMs extend linear regression by replacing each linear term β_j × x_j with a smooth function f_j(x_j). The model is: y = f_1(x_1) + f_2(x_2) + ε.
>
> Backfitting estimates each smooth function iteratively. In each iteration, f_1 is fitted on the partial residuals (y - f_2), then f_2 is fitted on the updated partial residuals (y - f_1). This continues until the functions stabilize.
>
> GAMs can improve predictions when predictors have non-linear relationships with y. Here, the GAM reduces training MSE from [OLS MSE] to [GAM MSE], increasing R² from [OLS R²] to [GAM R²]."

---

# METHOD 5: KNN Local Regression ⭐ MED-HIGH PRIORITY

## Recognition Cue
```{r}
d <- abs(x - x0)            # THE TELL: distance from x0
o <- order(d)[1:K]          # K nearest indices
```

## Plain English Explanation
For a query point x0:
1. Compute the distance from x0 to every other data point
2. Find the K closest data points (`order(d)[1:K]`)
3. Fit a small linear regression using only those K neighbors
4. Return the predicted value at x0 from that local regression

## Why It Is Used
KNN local regression is **non-parametric**: it does not assume a global linear relationship. It adapts to local patterns in the data. Small K = more flexible (low bias, high variance). Large K = smoother (high bias, low variance).

## Exam Follow-Up Patterns

### Pattern A: Extend to multiple predictors
Instead of `abs(x - x0)` (one predictor), use Euclidean distance:
```{r}
d <- sqrt(rowSums((X - matrix(x0, nrow(X), ncol(X), byrow=TRUE))^2))
```
Or for standardized predictors:
```{r}
d <- sqrt(rowSums(((X - matrix(x0, nrow(X), ncol(X), byrow=TRUE)) / 
                   matrix(apply(X,2,sd), nrow(X), ncol(X), byrow=TRUE))^2))
```

### Pattern B: Backfitting for multi-predictor KNN
Backfitting with KNN is analogous to GAM backfitting: alternately fit KNN on residuals for each predictor until convergence.

## Written Answer Template

> "The function `f` implements **K-nearest-neighbor local regression**. Given a query point `x0`, it:
> 1. Computes the absolute distance from `x0` to all observations
> 2. Selects the `K` closest observations
> 3. Fits a local linear regression on those `K` neighbors
> 4. Returns the predicted value at `x0`
>
> The parameter K controls flexibility: small K = high flexibility (low bias, high variance); large K = smoother fit (higher bias, lower variance). LOOCV is used to choose the optimal K."

---

# METHOD 6: Logistic Regression ⭐ HIGHEST PRIORITY

## Recognition Cue
```{r}
glm(y ~ ., data=train, family=binomial())    # THE TELL: family=binomial
# or: exp(eta) / (1 + exp(eta))             # THE TELL: sigmoid function
```

## Coefficient Interpretation (Key Exam Skill)
```{r}
mod <- glm(churn ~ age + income, data=train, family=binomial())
coef(mod)["age"]    # e.g., -0.05

# Interpretation:
# On the LOG-ODDS scale: age increases by 1 → log-odds decreases by 0.05
# On the ODDS scale: multiply by exp(-0.05) = 0.951 → 4.9% lower odds
exp(coef(mod))      # Odds ratios — more intuitive to interpret
```

**Template:** "A one-unit increase in [variable] is associated with a [increase/decrease] in the log-odds of [outcome] of [coefficient], holding other variables constant. The odds ratio is exp([coefficient]) = [value], meaning the odds of [outcome] are [value] times [higher/lower]."

## Full Exam Workflow
```{r}
# 1. Fit
logreg <- glm(as.factor(y) ~ ., data=train, family=binomial())
summary(logreg)

# 2. Check significance and consider reducing model
logreg_reduced <- glm(y ~ x1 + x3 + x5, data=train, family=binomial())

# 3. Predict probabilities on TEST
prob <- predict(logreg, newdata=test, type="response")

# 4. Choose threshold (default 0.5, but tune if imbalanced)
pred <- ifelse(prob > 0.5, 1, 0)

# 5. Evaluate
cm  <- table(Actual=test$y, Predicted=pred)
acc <- sum(diag(cm)) / sum(cm)
prop.table(cm, margin=1)
```

---

# METHOD 7: Random Forest ⭐ HIGHEST PRIORITY

## Recognition Cue
```{r}
randomForest(y ~ ., mtry=3, ...)    # THE TELL: randomForest + mtry
```

## Plain English Explanation
Random forests grow many decision trees. Each tree:
1. Is trained on a **bootstrap sample** of the training data
2. At each split, only considers a **random subset of mtry predictors**

The final prediction averages all trees (regression) or takes a majority vote (classification). The `mtry < p` restriction decorrelates the trees, improving the variance reduction from averaging.

## Variable Importance — How to Explain It
```{r}
varImpPlot(rf)    # Plot
importance(rf)    # Numbers
```
**MeanDecreaseAccuracy:** When this predictor is randomly permuted (its values shuffled), how much does out-of-bag accuracy drop? Higher = more important.
**MeanDecreaseGini:** How much does this predictor reduce Gini impurity across all splits in all trees? Higher = more important.

**Written template:** "The variable importance plot shows [variable] is the most important predictor, as randomizing its values leads to the largest decrease in prediction accuracy. [Variable 2] and [variable 3] are also important. [Variable 4] has near-zero importance, suggesting it contributes little to predictions."

---

# METHOD 8: Boosting ⭐ HIGH PRIORITY

## Recognition Cue
```{r}
r <- y - fhat    # THE TELL: compute residuals each iteration
fhat <- fhat + nu * predict(tree_b, X)   # Update with shrunken tree
```
Also: `gbm(..., distribution="bernoulli")` or `distribution="gaussian"`

## Plain English Explanation
Boosting builds trees **sequentially**:
1. Start with a flat prediction (zeros or mean(y))
2. Fit a small tree to the **residuals** (errors)
3. Update predictions by adding a small fraction (shrinkage `nu`) of the tree's predictions
4. Repeat B times
Each tree corrects the errors of all previous trees. Boosting reduces **bias** (unlike bagging which reduces variance).

## Key Parameters
| Parameter | Effect | Typical value |
|-----------|--------|---------------|
| `n.trees` | More trees = more flexible but risk overfitting | 500–2000 |
| `interaction.depth` | Tree depth = complexity | 1–5 |
| `shrinkage` | Learning rate; smaller = need more trees but better | 0.001–0.1 |

## Written Answer Template

> "The boosted tree model uses the `gbm` function with `distribution='bernoulli'` for binary classification. It sequentially fits small trees, each correcting the errors of the previous ones.
>
> On test data, the model achieves [confusion matrix result]. The threshold was set to [threshold]. Compared to logistic regression, boosting [performs better/similarly/worse] on the [class] class. The key difference is that boosting builds up prediction accuracy gradually through many weak learners, while logistic regression fits a single global model."

---

# METHOD 9: Decision Tree — Fit, Interpret, Prune

## Recognition Cue
```{r}
tree(y ~ ., data=train)    # THE TELL: tree() function
# or: for(s in candidates) RSS_left + RSS_right   # split search loop
```

## Reading a Tree Plot
- Each **internal node** shows the split condition (e.g., "age < 30")
- Follow **left branch if TRUE**, right branch if FALSE
- **Terminal nodes (leaves)** show the predicted value (regression) or class (classification)

## Full Code
```{r}
library(tree)
t <- tree(y ~ ., data=train)
plot(t); text(t, pretty=0)

# Prune using CV
cv_t    <- cv.tree(t)
best_sz <- cv_t$size[which.min(cv_t$dev)]
t_pruned <- prune.tree(t, best=best_sz)
plot(t_pruned); text(t_pruned, pretty=0)

# Evaluate
pred_t <- predict(t_pruned, newdata=test)
MSE_t  <- mean((test$y - pred_t)^2)
```

---

# METHOD 10: OLS — Baselines and Comparisons

## Key role in exams
OLS always appears as the **baseline** to compare ridge/lasso/GAM/tree against.

## What to show
```{r}
ols <- lm(y ~ ., data=train)
summary(ols)             # See R², p-values
MSE_ols_train <- mean(residuals(ols)^2)
pred_ols_test <- predict(ols, newdata=test)
MSE_ols_test  <- mean((test$y - pred_ols_test)^2)

# R² computation
R2_train <- summary(ols)$r.squared         # From lm output
R2_manual <- 1 - MSE_ols_train / var(train$y)   # Manual formula
```

---

# Quick Answer Templates Glossary

| Situation | Template start |
|-----------|---------------|
| Identify a method | "This code implements **[METHOD]**, which..." |
| Explain a loop | "The for loop runs n/K/B times. In each iteration..." |
| Compare models | "Ridge reduces the sum of squared coefficients from [X] to [Y], indicating strong shrinkage. This reduces variance at the cost of a small increase in bias." |
| Threshold choice | "With a class imbalance of [X]% positive cases, the default threshold of 0.5 classifies almost all observations as negative. A lower threshold of [T] was chosen to balance sensitivity and specificity." |
| GAM vs OLS | "The GAM improves training R² from [X] to [Y], suggesting that variable [Z] has a non-linear relationship with y. The GAM is preferred when the linear model leaves systematic patterns in the residuals." |
| Variable importance | "The variable [X] shows the largest decrease in accuracy when permuted, making it the most important predictor. [Y] and [Z] are also important. [W] contributes little." |
