---
title: "Task 2 Playbooks — Dataset Analysis End-to-End"
author: "BAN404 Self-Study"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  echo: true
  eval: false
  warning: false
---

# How to Use This File

Task 2 gives you a real dataset and asks you to analyze it. Every Task 2 follows the same 7-step pipeline. This file gives you:
- **Playbook A**: Classification (most common — 2021, 2022, 2023, 2024, 2025 Task 2)
- **Playbook B**: Regression (appears in 2021, 2022, 2023 Task 1 — but can appear in Task 2 too)
- **Decision guides**: How to make the choices the exam asks about
- **Written templates**: Fill-in-the-blank answers for each step

> **Key insight:** The content changes (different dataset, different predictors). The **structure never changes.** Master the structure, adapt to the content.

---

# THE 7-STEP PIPELINE (applies to every Task 2)

```
Step 1: DATA PREP        → clean, factor-encode, train/test split
Step 2: EDA              → find useful predictors with plots/tables
Step 3: BASELINE MODEL   → logistic (class) or OLS (regression)
Step 4: EVALUATE         → confusion matrix + accuracy, or MSE
Step 5: ENSEMBLE 1       → random forest + variable importance
Step 6: ENSEMBLE 2       → boosting (or bagging)
Step 7: COMPARE + INTERPRET → business answer
```

---

# PLAYBOOK A — Classification Pipeline

## Step 1: Data Preparation

### 1a. Load and inspect data
```{r}
# Read data (use read.csv2 for semicolon-separated; read.csv for comma)
data <- read.csv("data.csv")
str(data)          # Check column types
head(data)         # First rows
summary(data)      # Summary statistics
names(data)        # Column names
dim(data)          # Rows x columns
```

### 1b. Create binary target (if needed)
```{r}
# Example: recode a count variable to binary
data$target <- as.numeric(data$count_variable >= 1)  # 0/1
# or as factor:
data$target <- factor(ifelse(data$count_variable >= 1, "yes", "no"))
```

### 1c. Remove leakage variables
**Rule:** Remove any variable that:
- IS the target (e.g., the raw count if target = "count >= 1")
- Is only observable AFTER the event (e.g., claim cost is only known if there was a claim)
- Is a perfect proxy for the target (a different ID/code for the same thing)

```{r}
# Remove by name
data <- data[, !(names(data) %in% c("id", "raw_count", "cost_after_event"))]
```

**Written template for exam:** "I removed [variable] because it is [directly derived from the target / only available after the event / a unique identifier with no predictive power]. Keeping it would lead to data leakage — the model would 'cheat' by using information not available at prediction time."

### 1d. Factor encoding
```{r}
# Always convert categorical variables to factors
data$gender  <- as.factor(data$gender)
data$zone    <- as.factor(data$zone)
data$class   <- as.factor(data$class)

# Verify
str(data)   # Should show "Factor" for categorical columns
```

**When to make something a factor:** If it takes a small number of discrete values and represents a category (gender, zone, class, type), it should be a factor. If it is an ordered number (age, income, score), leave as numeric.

### 1e. Train/test split
```{r}
set.seed(123)   # Always set seed for reproducibility
n     <- nrow(data)
idx   <- sample(1:n, size=floor(n/2))  # 50/50 split
train <- data[idx, ]
test  <- data[-idx, ]

# Check proportions of target
prop.table(table(train$target))
prop.table(table(test$target))
```

**Written template:** "I split the data 50/50 into training (n=[X]) and test sets (n=[Y]) using `set.seed([seed])` for reproducibility. Models are fitted on training data and evaluated on test data to estimate generalization performance."

---

## Step 2: EDA — Finding Useful Predictors

### The EDA Decision Guide

| Predictor type | Target type | Best plot/table |
|----------------|-------------|-----------------|
| Continuous | Binary | Boxplot by target group |
| Continuous | Binary | Mean/SD comparison by group |
| Categorical | Binary | prop.table — proportion of target=1 per category |
| Continuous | Continuous | Scatter plot, correlation |
| Categorical | Continuous | Boxplot of response by category |

```{r}
# ------- Continuous predictor vs binary target -------
# Boxplot: do values differ between target=0 and target=1?
boxplot(age ~ target, data=train, main="Age by target")
# Means by group:
tapply(train$age, train$target, mean)

# ------- Categorical predictor vs binary target -------
# Proportions: what fraction of each category has target=1?
prop.table(table(train$gender, train$target), margin=1)
# Column margin=1 means: within each gender, what fraction has target=1?

# ------- Quick scan with sapply -------
num_vars <- names(train)[sapply(train, is.numeric)]
for(v in num_vars) {
  boxplot(train[[v]] ~ train$target, 
          main=paste(v, "by target"), xlab="target")
}

# ------- Correlation with target (for numeric target or numeric predictors) -------
round(cor(train[, sapply(train, is.numeric)]), 2)
```

**Written template:** "From the boxplots, [variable A] shows a clear difference between target=0 and target=1 groups (median [X] vs [Y]), suggesting it is a useful predictor. [Variable B] shows similar distributions and is unlikely to add predictive power. The cross-table shows that [category X] of [variable C] has a much higher proportion of target=1 ([P]%) compared to [category Y] ([Q]%)."

---

## Step 3: Baseline Model — Logistic Regression

```{r}
# Full model (all predictors)
logreg_full <- glm(target ~ ., data=train, family=binomial())
summary(logreg_full)

# Reduced model (significant predictors only, based on EDA and summary)
logreg_red <- glm(target ~ age + gender + zone, data=train, family=binomial())
summary(logreg_red)

# Predict on TEST data
prob_full <- predict(logreg_full, newdata=test, type="response")
prob_red  <- predict(logreg_red,  newdata=test, type="response")
```

**Interpretation guide for coefficients:**
```{r}
# On the log-odds scale (raw summary output)
coef(logreg_full)["age"]    # e.g., -0.05

# On the odds ratio scale (more interpretable)
round(exp(coef(logreg_full)), 3)
```
**Written template:** "The coefficient on [variable] is [coef], corresponding to an odds ratio of exp([coef]) = [OR]. This means a one-unit increase in [variable] multiplies the odds of [target] by [OR]. [Variable] is [statistically significant / not significant] at the 5% level."

---

## Step 4: Evaluate — Confusion Matrix and Threshold

### The threshold problem
```{r}
# Check class balance FIRST
prop.table(table(train$target))
# If target=1 is rare (< 10%), use a low threshold
```

**Decision rule for threshold:**
- Class balance close to 50/50 → use threshold = 0.5
- Class is rare (e.g., 1–5% positive) → experiment with lower thresholds (0.01–0.1)
- Goal is prediction precision → optimize accuracy on training set
- Goal is catching true positives → optimize sensitivity

```{r}
# Evaluate at chosen threshold
threshold <- 0.5   # adjust as needed

pred_class <- ifelse(prob_full > threshold, 1, 0)
cm         <- table(Actual=test$target, Predicted=pred_class)
cm
prop.table(cm, margin=1)    # Row-wise: sensitivity per class
acc <- sum(diag(cm)) / sum(cm)
cat("Accuracy:", round(acc, 4))

# If imbalanced: scan thresholds
thresholds <- seq(0.01, 0.5, by=0.01)
acc_vec    <- sapply(thresholds, function(th) {
  pred <- ifelse(prob_full > th, 1, 0)
  mean(pred == as.numeric(as.character(test$target)))
})
best_th <- thresholds[which.max(acc_vec)]
```

**Written template:** "Using a threshold of [T], the logistic model achieves overall accuracy of [A]%. The confusion matrix shows that [X]% of actual positives (target=1) are correctly classified (sensitivity) and [Y]% of actual negatives (target=0) are correctly classified (specificity). The threshold was chosen [because the positive class is rare / to maximize accuracy / to balance sensitivity and specificity]."

---

## Step 5: Random Forest

```{r}
library(randomForest)

# Target must be a factor for classification
train$target <- as.factor(train$target)
test$target  <- as.factor(test$target)

rf <- randomForest(target ~ ., data=train,
                   mtry=floor(sqrt(ncol(train)-1)),   # sqrt(p) for classification
                   ntree=500,
                   importance=TRUE)
rf    # Print OOB error rate

# Predict classes
pred_rf <- predict(rf, newdata=test)
cm_rf   <- table(Actual=test$target, Predicted=pred_rf)
prop.table(cm_rf, margin=1)

# Or predict probabilities (for threshold tuning)
prob_rf   <- predict(rf, newdata=test, type="prob")[, 2]  # P(target=1)
pred_rf2  <- ifelse(prob_rf > threshold, 1, 0)
cm_rf2    <- table(Actual=test$target, Predicted=pred_rf2)

# Variable importance
varImpPlot(rf, main="Variable Importance")
importance(rf)
```

**Written template:** "The random forest achieves [accuracy]% accuracy on test data. The variable importance plot shows [variable A] is the most important predictor (largest decrease in accuracy when permuted), followed by [B] and [C]. [Variable D] has near-zero importance. Compared to logistic regression ([X]% accuracy), the random forest [performs better/similarly/worse]."

---

## Step 6: Boosting

```{r}
library(gbm)

# Target must be 0/1 numeric for gbm with distribution="bernoulli"
train_bt        <- train
train_bt$target <- as.numeric(as.character(train$target))

bt <- gbm(target ~ ., data=train_bt,
          distribution="bernoulli",   # Binary classification
          n.trees=1000,
          interaction.depth=3,
          shrinkage=0.01,
          cv.folds=5)

best_trees <- which.min(bt$cv.error)
cat("Optimal n.trees:", best_trees)

# Predict
prob_bt  <- predict(bt, newdata=test, n.trees=best_trees, type="response")
pred_bt  <- ifelse(prob_bt > threshold, 1, 0)
cm_bt    <- table(Actual=test$target, Predicted=pred_bt)
prop.table(cm_bt, margin=1)
```

**Written template:** "The boosted tree model achieves [accuracy]% accuracy with CV-optimal n.trees=[N]. The confusion matrix shows sensitivity of [S]% and specificity of [Sp]%. [Comparison to logistic/RF]."

---

## Step 7: Compare All Models and Give Business Answer

```{r}
# Summary table
results <- data.frame(
  Model    = c("Logistic full", "Logistic reduced", "Random Forest", "Boosting"),
  Accuracy = c(acc_full, acc_red, acc_rf, acc_bt),
  row.names = NULL
)
print(results)
```

**Business answer template:** "Based on the analysis, [variable A] is the strongest predictor of [target]. Customers/policies/individuals with [characteristic] are significantly more likely to [outcome] ([odds ratio/accuracy evidence]). [Variable B] is the second most important factor. The [best model] achieves the highest accuracy of [X]% on test data. To [business goal], the company should focus on [actionable insight]."

---

# PLAYBOOK B — Regression Pipeline

## Differences from Classification

| Step | Classification | Regression |
|------|---------------|------------|
| Target | Factor (0/1, yes/no) | Numeric (continuous) |
| Baseline model | `glm(..., family=binomial)` | `lm(y ~ .)` |
| Evaluation | Confusion matrix, accuracy | MSE, RMSE, R² |
| RF | `randomForest(as.factor(y)~.)` | `randomForest(y~.)` |
| Boosting | `distribution="bernoulli"` | `distribution="gaussian"` |
| Threshold | Yes (convert prob → class) | No threshold needed |

## Key Code Changes
```{r}
# Step 3: OLS baseline
ols <- lm(y ~ ., data=train)
MSE_ols  <- mean((test$y - predict(ols, test))^2)
R2_ols   <- 1 - MSE_ols / var(test$y)

# Step 5: Regression RF
rf_reg <- randomForest(y ~ ., data=train,
                       mtry=floor((ncol(train)-1)/3),   # p/3 for regression
                       ntree=500)
MSE_rf <- mean((test$y - predict(rf_reg, test))^2)

# Step 6: Regression boosting
bt_reg <- gbm(y ~ ., data=train, distribution="gaussian",
              n.trees=1000, interaction.depth=3, shrinkage=0.01)
MSE_bt <- mean((test$y - predict(bt_reg, test, n.trees=1000))^2)

# Compare
cat("MSE: OLS=", round(MSE_ols,2), "RF=", round(MSE_rf,2), "BT=", round(MSE_bt,2))
```

---

# EDA Choice Logic — Quick Reference

> "What descriptive statistics should I use?" — Use this table:

| What you want to know | How to show it |
|-----------------------|----------------|
| Distribution of a numeric variable | Histogram, boxplot, `summary()` |
| Association between numeric predictor and binary target | Boxplot by target group |
| Association between categorical predictor and binary target | `prop.table(table(cat, target), margin=1)` |
| Association between two numeric variables | Scatter plot, `cor()` |
| Class balance (how rare is the positive class?) | `table(y)` and `prop.table(table(y))` |
| Missing values | `sum(is.na(data))`, `colSums(is.na(data))` |
| Outliers | Boxplot, or check `summary()` for max vs Q3 |

---

# Model Comparison Logic

After fitting all models, choose the best by:
1. **Test accuracy** (classification) or **test MSE/RMSE** (regression) — lower MSE, higher accuracy = better
2. If nearly equal: prefer the **simpler model** (logistic regression over RF if similar accuracy)
3. If the goal is **interpretability** (which variables matter?): logistic regression or tree
4. If the goal is **prediction accuracy**: RF or boosting usually wins

```{r}
# Model comparison table (fill in the values you computed)
model_comparison <- data.frame(
  Model    = c("OLS / Logistic", "Tree", "Bagging/RF", "Boosting"),
  Metric   = c(acc1, acc2, acc3, acc4),   # accuracy or MSE
  Comments = c("Interpretable", "Visualizable", "Best for variance reduction", "Often best accuracy")
)
```

---

# Common Exam Writing Phrases

**Starting EDA:** "To identify potentially useful predictors, I used [boxplots / cross-tables / scatter plots]. The results show..."

**Explaining train/test choice:** "I fit the model on training data to avoid overfitting and evaluate on test data to estimate how the model performs on unseen observations."

**Explaining threshold choice:** "The default threshold of 0.5 is inappropriate here because [the positive class represents only X% of observations / the cost of missing a true positive is high]. I chose threshold [T] which achieves [sensitivity]% sensitivity and [specificity]% specificity."

**Comparing to a benchmark:** "The random forest improves accuracy from [X]% (logistic regression) to [Y]%, a gain of [Z] percentage points. This suggests the model benefits from capturing non-linear interactions between predictors."

**Variable importance interpretation:** "According to the random forest variable importance measure, [variable] contributes most to prediction accuracy. When its values are permuted, model accuracy drops by [X] percentage points. This aligns with the EDA finding that [variable] showed the largest difference between target groups."
