---
title: "BAN404 Mock Exams — Three Full Practice Exams"
author: "BAN404 Self-Study"
format:
  html:
    toc: true
    toc-depth: 2
    theme: flatly
execute:
  echo: true
  eval: false
  warning: false
---

# How to Use These Mock Exams

**Instructions:**
- Use a timer. Treat as real exam conditions.
- Mock Exam 1 and 2: School exam format (6 hours, ~10 sub-questions)
- Mock Exam 3: Home exam format (8 hours, more sub-questions, more explanation required)
- All written materials permitted. Your `snippets/ban404_core_snippets.R` and compendium files are allowed.
- After finishing, self-grade using `08_self_grading_rubric.md`

---

# MOCK EXAM 1 (School format — 6 hours)

*Style: 2025 exam (school, 6h, one CSV file + one package dataset)*

**Dataset for Task 1:** `creditdata` (simulate below)  
**Dataset for Task 2:** `OJ` from ISLR2 package

---

## Task 1 (50 points) — Methodological questions

### Setup
```{r}
# Simulate Task 1 dataset
set.seed(42)
n  <- 80; p <- 10
X  <- matrix(rnorm(n*p), n, p)
y  <- X[,1]*2 + X[,3]^2 + rnorm(n, sd=2)
Xy <- as.data.frame(cbind(y, X))
names(Xy)[1] <- "y"
X  <- as.matrix(Xy[,-1])
y  <- as.matrix(Xy[,1])
```

### Task 1a (10p) — Explain the function

> Consider the following R function:
> ```r
> f <- function(b1, X, y, la) {
>   fval <- sum((y - X %*% b1)^2) + la * sum(b1^2)
>   return(fval)
> }
> g <- function(X, y, la) {
>   q  <- ncol(X); b0 <- mean(y); yd <- y - b0
>   Xd <- X; for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])
>   opt <- nlminb(rep(0,q), f, X=Xd, y=yd, la=la)
>   return(opt$par)
> }
> ```
> (a) What method do `f` and `g` implement? Identify the key line that tells you.
> (b) What is `la`? What happens when la=0?
> (c) Why does `g` subtract the mean from each column of X?

*[Write your answer here. 15 minutes suggested.]*

### Task 1b (10p) — Apply and compare

> Using the dataset above:
> (a) Fit OLS using `lm()`. Extract and show the coefficients.
> (b) Fit ridge using `g(X, y, la=100)`. Compare the sum of squared coefficients.
> (c) Explain why ridge coefficients are smaller in sum.

```{r}
# Your code here:
```

*[Write your answer here.]*

### Task 1c (10p) — LOOCV for optimal λ

> Write a LOOCV function and use it to find the optimal λ on the grid `seq(0, 500, length.out=15)`.
> Plot LOOCV MSE vs λ. What is the optimal λ?

```{r}
# Your code here:
```

### Task 1d (10p) — Bootstrap

> Use the bootstrap (B=2000) to estimate the sampling distribution of the **mean** of y.
> (a) Plot the bootstrap distribution as a histogram.
> (b) Compute a 95% CI using the percentile method.
> (c) Compare to the normal-theory CI: `mean(y) ± 1.96 * sd(y)/sqrt(n)`

```{r}
# Your code here:
```

### Task 1e (10p) — GAM

> (a) Create scatter plots of y against each of the 10 predictors. Identify any non-linear patterns.
> (b) Fit an OLS and at least one GAM. Compare training MSE and R².
> (c) Which model would you prefer? Justify.

```{r}
# Your code here:
```

---

## Task 2 (50 points) — OJ dataset classification

```{r}
library(ISLR2)
data(OJ)
# Target: Purchase (CH or MM brand)
```

### Task 2a (10p) — Data prep and EDA

> (a) Check the data using `str()` and `summary()`. Are there variables to remove?
> (b) Split 50/50, set.seed(8655).
> (c) Use descriptive statistics to find useful predictors for Purchase.

```{r}
# Your code here:
```

*[Written answer: What did you find? Which predictors look useful?]*

### Task 2b (10p) — Logistic regression

> Fit a logistic regression with promising predictors from 2a. Interpret 2 coefficients.

```{r}
# Your code here:
```

### Task 2c (10p) — Evaluate predictions

> Predict on test data. Choose a threshold. Build a confusion matrix. Report accuracy and row proportions.

```{r}
# Your code here:
```

### Task 2d (10p) — Random forest

> Fit a random forest. Evaluate predictions. Plot variable importance. Compare to logistic regression.

```{r}
# Your code here:
```

### Task 2e (10p) — Boosting

> Fit a boosted tree. Evaluate predictions. Compare with logistic and RF.

```{r}
# Your code here:
```

---

# MOCK EXAM 1 — SOLUTIONS

## Task 1 Solutions

### 1a Solution
**(a)** `f` and `g` implement **ridge regression**. The key line is `la * sum(b1^2)` — squaring the coefficients gives the L2 (ridge) penalty.
**(b)** `la` is the regularization parameter λ. When la=0, the penalty disappears and `g` returns OLS coefficients.
**(c)** Subtracting the mean from each column (demeaning) centers the predictors. This is necessary so the intercept equals mean(y) after demeaning y as well. The penalty then only applies to slope coefficients, not the intercept — which is standard practice in ridge regression.

```{r}
# 1b solution
f <- function(b1,X,y,la) sum((y-X%*%b1)^2) + la*sum(b1^2)
g <- function(X,y,la) {
  q=ncol(X); b0=mean(y); yd=y-b0
  Xd=X; for(k in 1:q) Xd[,k]=X[,k]-mean(X[,k])
  nlminb(rep(0,q),f,X=Xd,y=yd,la=la)$par
}
b_ols   <- lm(y ~ X)$coef[-1]
b_r100  <- g(X, y, la=100)
cat("Sum sq OLS:", round(sum(b_ols^2),4))
cat("Sum sq Ridge(100):", round(sum(b_r100^2),4))
```

```{r}
# 1c solution
loo_fn <- function(la, X, y) {
  n=nrow(X); q=ncol(X)
  yd=y-mean(y); Xd=X; for(k in 1:q) Xd[,k]=X[,k]-mean(X[,k])
  preds=numeric(n)
  for(i in 1:n) { b1=g(X[-i,,drop=F],y[-i],la); preds[i]=Xd[i,]%*%b1 }
  mean((yd-preds)^2)
}
lavec <- seq(0, 500, length.out=15)
mses  <- sapply(lavec, function(la) loo_fn(la, X, y))
plot(lavec, mses, type="b", main="LOOCV MSE vs lambda")
cat("Optimal lambda:", lavec[which.min(mses)])
```

```{r}
# 1d solution
set.seed(1); n=length(y); B=2000; res=numeric(B)
for(b in 1:B) res[b] <- mean(y[sample(1:n,replace=T)])
hist(res, breaks=50, main="Bootstrap distribution of mean(y)")
ci_pct <- quantile(res, c(0.025, 0.975))
ci_norm_theory <- mean(y) + c(-1,1)*1.96*sd(y)/sqrt(n)
cat("Bootstrap CI:", round(ci_pct,3), "\n")
cat("Normal-theory CI:", round(ci_norm_theory,3))
```

```{r}
# 1e solution
library(gam)
# Scan plots
par(mfrow=c(2,5))
for(j in 1:10) plot(X[,j], y, main=paste("X",j))
# Fit OLS and GAM
ols   <- lm(y ~ ., data=Xy)
gam1  <- gam(y ~ s(V3) + V1 + V2 + V4 + V5 + V6 + V7 + V8 + V9 + V10, data=Xy)
MSE_null <- var(y)
R2_ols   <- 1 - mean(residuals(ols)^2)/MSE_null
R2_gam   <- 1 - mean(residuals(gam1)^2)/MSE_null
cat("R2 OLS:", round(R2_ols,4), " | R2 GAM:", round(R2_gam,4))
```

---

# MOCK EXAM 2 (School format — 6 hours)

*Style: 2024 exam (KNN + backfitting in Task 1, two-perspective Task 2)*

**Dataset for Task 1:** Simulate below  
**Dataset for Task 2:** `Carseats` from ISLR2

---

## Task 1 (50 points) — Methodological questions

### Task 1a (10p) — Explain the function

> ```r
> h <- function(x0, x, y, K=5) {
>   d <- sqrt((x - x0)^2)
>   o <- order(d)[1:K]
>   predict(lm(y[o] ~ x[o]), newdata=data.frame(x=x0))
> }
> ```
> What method is this? What does each step do?

### Task 1b (10p) — LOOCV for K

```{r}
# Use this data
set.seed(7); n=30
x <- runif(n, 0, 10)
y <- sin(x) + rnorm(n, sd=0.3)
```

> Use LOOCV on x and y to find the optimal K (search K = 1:15). Plot LOOCV MSE vs K.

### Task 1c (10p) — Plot with optimal K

> Plot y vs x. Add a smooth line showing KNN predictions for the optimal K.

### Task 1d (10p) — Multi-predictor extension

> How would you extend h to 2 predictors? Write the modified distance formula.

### Task 1e (10p) — Backfitting explanation

> Without code: Explain the backfitting algorithm. When is it equivalent to OLS regression? When is it not?

---

## Task 2 (50 points) — Carseats dataset

```{r}
library(ISLR2)
data(Carseats)
# Target: create binary variable High = (Sales > 8)
Carseats$High <- as.factor(ifelse(Carseats$Sales > 8, "Yes", "No"))
Carseats <- Carseats[, names(Carseats) != "Sales"]
```

### Task 2a (8p) — EDA and prep

> Examine the data. Is any variable inappropriate to include? Split 70/30, set.seed(100).
> Use EDA to find predictors of `High`.

### Task 2b (7p) — Logistic regression

> Fit logistic regression. Interpret two coefficients. Evaluate on test data.

### Task 2c (7p) — Classification tree

> Fit a classification tree. Plot it. Interpret one branch. Compare accuracy to logistic regression.

### Task 2d (10p) — Random forest

> Fit a random forest. Evaluate and plot variable importance. Compare to previous models.

### Task 2e (8p) — Threshold CV

> Use cross-validation to find the threshold that maximises accuracy for the logistic regression model.

### Task 2f (10p) — Business answer

> Based on all analysis: What predicts high sales? What would you recommend to the store manager?

---

## Mock Exam 2 — Key Solutions

```{r}
# Task 1b: LOOCV for KNN
h <- function(x0, x, y, K=5) {
  o <- order((x-x0)^2)[1:K]
  predict(lm(y[o]~x[o]), newdata=data.frame(x=x0))
}
Kgr  <- 1:15
mseK <- sapply(Kgr, function(K)
  mean(sapply(1:n, function(i) (y[i]-h(x[i],x[-i],y[-i],K))^2))
)
K_opt <- Kgr[which.min(mseK)]
plot(Kgr, mseK, type="b", main="LOOCV MSE vs K"); abline(v=K_opt, col="red")
cat("Optimal K:", K_opt)

# Task 2 pipeline
library(ISLR2); data(Carseats)
Carseats$High <- as.factor(ifelse(Carseats$Sales > 8, "Yes", "No"))
Carseats <- Carseats[, names(Carseats)!="Sales"]
set.seed(100); idx <- sample(1:nrow(Carseats), floor(0.7*nrow(Carseats)))
train <- Carseats[idx,]; test <- Carseats[-idx,]
logreg <- glm(High ~ ., data=train, family=binomial()); summary(logreg)
prob   <- predict(logreg, test, type="response")
pred   <- ifelse(prob>0.5, "Yes", "No")
table(Actual=test$High, Pred=pred)

library(randomForest)
rf <- randomForest(High ~ ., data=train, ntree=300, importance=TRUE)
varImpPlot(rf)
```

---

# MOCK EXAM 3 (Home exam format — 8 hours)

*Style: 2023 exam (two tasks, both with regression + classification, more explanation required)*

**Dataset:** `Churn.csv` from `/home/runner/work/BAN404/BAN404/Previous Exam/2023/Churn.csv`

---

## Task 1 (50 points) — Predict average bill (regression)

```{r}
Churn <- read.csv("/home/runner/work/BAN404/BAN404/Previous Exam/2023/Churn.csv")
```

### 1a — Data cleaning and split (10p)
> Remove variables not suitable as predictors. Convert factors. Split 50/50, `set.seed(86554354)`. Explain your choices.

### 1b — EDA for bill_avg (10p)
> Find useful predictors for `bill_avg` using plots and tables. Write which R functions you used.

### 1c — OLS regression (8p)
> Fit OLS. Motivate variable choices. Evaluate test MSE.

### 1d — Lasso (10p)
> Fit lasso using `cv.glmnet`. Find optimal λ. Compare selected variables to OLS. Evaluate test MSE.

### 1e — Regression tree (5p)
> Fit, plot, and interpret a regression tree. Evaluate test MSE.

### 1f-1i — Random forest (7p)
> Fit RF. Variable importance plot + interpretation. Evaluate test MSE. Final: which customers have high/low bills?

---

## Task 2 (50 points) — Predict churn (classification)

### 2a — EDA for churn (8p)
> Use tables, plots, common sense to find promising predictors. Motivate variable removal.

### 2b — Bootstrap CI (10p)
> Let p = P(churn=1). Use first 50 observations. Bootstrap 95% CI for p.
> Compare to normal approximation: `p_hat ± 1.96 * sqrt(p_hat*(1-p_hat)/n)`.

### 2c — Logistic regression (8p)
> Fit logistic on all variables. Interpret `is_tv_subscriber` and `is_movie_package_subscriber`. Evaluate.

### 2d — Predict and evaluate logistic (7p)
> Predict churn on test data. Build confusion matrix. Improve by removing variables?

### 2e — Random forest (7p)
> Fit RF. Evaluate. Compare to logistic.

### 2f — Final answer (10p)
> What are the typical features of customers who churn?

---

## Mock Exam 3 — Key Solutions for Task 2b Bootstrap

```{r}
# Task 2b: Bootstrap CI for p = P(churn=1)
Churn <- read.csv("/home/runner/work/BAN404/BAN404/Previous Exam/2023/Churn.csv")
churn50 <- Churn$churn[1:50]   # First 50 observations

B <- 10000; pboot <- numeric(B); set.seed(1)
for(b in 1:B) pboot[b] <- mean(sample(churn50, replace=TRUE))

# Bootstrap CI (percentile)
ci_boot <- quantile(pboot, c(0.025, 0.975))

# Normal approximation
p_hat  <- mean(churn50)
n_s    <- length(churn50)
ci_norm <- p_hat + c(-1,1)*1.96*sqrt(p_hat*(1-p_hat)/n_s)

cat("Bootstrap CI:", round(ci_boot, 4), "\n")
cat("Normal approx CI:", round(ci_norm, 4))

# The bootstrap CI is preferred when p is close to 0 or 1 (skewed distribution)
# or when n is small (normality assumption weaker)
```
