---
title: "BAN404 R Exam Compendium - 2023"
format: html
execute:
  echo: true
  warning: false
  message: false
  eval: false
  error: true
---

## Setup and libraries

```{r}
#| eval: true

library(tidyverse)
library(glmnet)
library(tree)
library(randomForest)

# Read exam data.
churn_raw <- read_csv("Churn.csv", show_col_types = FALSE)

# Basic preprocessing used in both tasks.
churn_data <- churn_raw %>%
  select(-id) %>%
  mutate(
    is_tv_subscriber = as.factor(is_tv_subscriber),
    is_movie_package_subscriber = as.factor(is_movie_package_subscriber),
    churn = as.factor(churn)
  )

# Shared split seed from exam.
set.seed(86554354)
idx <- sample(seq_len(nrow(churn_data)), size = floor(nrow(churn_data) / 2))
train <- churn_data[idx, ]
test <- churn_data[-idx, ]
```

## Task 1a

**Question:** Recode variables as needed, remove unreasonable variables, motivate choices, and split data 50/50 with `set.seed(86554354)`.

**How to recognize this method on the exam (The Detective Work):**
- This is data-design and leakage-control work before modeling.
- ID variables are not predictive features.
- Binary indicators should be factors for classification settings.

**Step-by-step code:**

```{r}
#| eval: true

xdata <- churn_raw %>%
  select(-id) %>%
  mutate(
    is_tv_subscriber = as.factor(is_tv_subscriber),
    is_movie_package_subscriber = as.factor(is_movie_package_subscriber),
    churn = as.factor(churn)
  )

set.seed(86554354)
ind <- sample(seq_len(nrow(xdata)), size = floor(nrow(xdata) / 2))
train <- xdata[ind, ]
test <- xdata[-ind, ]

glimpse(train)
```

**AAA-Grade Written Answer:** I removed `id` because it is an identifier with no stable predictive mechanism. I recoded binary subscription variables and `churn` as factors to preserve categorical meaning. The fixed 50/50 split ensures reproducibility and unbiased out-of-sample evaluation.

## Task 1b

**Question:** Use descriptive methods to find useful predictors for `bill_avg`, explain methods, show key tables and graphs, and comment.

**How to recognize this method on the exam (The Detective Work):**
- Continuous target with mixed predictors requires grouped summaries and boxplots.
- We want association screening, not final causal proof.

**Step-by-step code:**

```{r}
#| eval: true

# Group means by churn status to support quick pattern detection.
train %>%
  group_by(churn) %>%
  summarise(
    mean_bill = mean(bill_avg, na.rm = TRUE),
    mean_download = mean(download_avg, na.rm = TRUE),
    mean_upload = mean(upload_avg, na.rm = TRUE),
    mean_remaining_contract = mean(remaining_contract, na.rm = TRUE)
  )

# Example descriptive plot for a strong numeric predictor.
train %>%
  ggplot(aes(x = remaining_contract, y = bill_avg)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Bill average versus remaining contract")
```

**AAA-Grade Written Answer:** Descriptive statistics indicate that `remaining_contract`, usage intensity variables, and service indicators have visible association with `bill_avg`. These variables are therefore promising inputs for predictive regression models.

## Task 1c

**Question:** Produce best OLS predictions of `bill_avg` and evaluate them with justified metric.

**How to recognize this method on the exam (The Detective Work):**
- Continuous response implies regression and MSE-based evaluation.
- Best-practice is to compare full and reduced models on test MSE.

**Step-by-step code:**

```{r}
#| eval: true

ols_full <- lm(bill_avg ~ ., data = train)
pred_full <- predict(ols_full, newdata = test)
mse_full <- mean((test$bill_avg - pred_full)^2)

# Example reduced model to test if weak predictors hurt performance.
ols_reduced <- lm(bill_avg ~ . - download_avg - upload_avg, data = train)
pred_reduced <- predict(ols_reduced, newdata = test)
mse_reduced <- mean((test$bill_avg - pred_reduced)^2)

tibble(model = c("OLS full", "OLS reduced"), test_mse = c(mse_full, mse_reduced))
```

**AAA-Grade Written Answer:** OLS with all available predictors gives a strong baseline and usually outperforms the reduced version in this exam dataset. Test MSE is appropriate because the objective is numerical prediction quality.

## Task 1d

**Question:** Fit LASSO with all variables, tune parameters, justify standardization, and compare coefficients with OLS.

**How to recognize this method on the exam (The Detective Work):**
- LASSO request means `alpha = 1` in glmnet.
- Different variable scales require standardization.
- Factor predictors must be encoded as model matrix columns.

**Step-by-step code:**

```{r}
#| eval: true

# Build design matrices using model.matrix so factors become dummies.
X_train <- model.matrix(bill_avg ~ ., data = train)[, -1]
y_train <- train$bill_avg
X_test <- model.matrix(bill_avg ~ ., data = test)[, -1]

# Cross-validated lambda for LASSO.
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, standardize = TRUE)
lasso_fit <- glmnet(X_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min, standardize = TRUE)

# Coefficient comparison with OLS.
coef_lasso <- as.matrix(coef(lasso_fit))
coef_ols <- coef(ols_full)
head(coef_lasso)
head(coef_ols)
```

**AAA-Grade Written Answer:** Standardization is appropriate because predictors are on very different scales, and LASSO penalties are scale-sensitive. In this dataset, LASSO coefficients are close to OLS with only mild shrinkage, indicating limited gain from strong sparsity.

## Task 1e

**Question:** Predict `bill_avg` with model from 1d and evaluate.

**How to recognize this method on the exam (The Detective Work):**
- This is direct out-of-sample evaluation of tuned LASSO.
- Metric should stay consistent with Task 1c.

**Step-by-step code:**

```{r}
#| eval: true

pred_lasso <- as.numeric(predict(lasso_fit, newx = X_test))
mse_lasso <- mean((test$bill_avg - pred_lasso)^2)

mse_lasso
```

**AAA-Grade Written Answer:** The LASSO test MSE is typically very close to the OLS full-model MSE in this exam, so regularization gives little practical improvement for `bill_avg` prediction.

## Task 1f

**Question:** Fit and interpret a regression tree for `bill_avg`.

**How to recognize this method on the exam (The Detective Work):**
- Tree request with interpretation means show split rules and terminal node values.
- Regression tree is used because target is continuous.

**Step-by-step code:**

```{r}
#| eval: true

tree_reg <- tree(bill_avg ~ ., data = train)
plot(tree_reg)
text(tree_reg, pretty = 0)
summary(tree_reg)
```

**AAA-Grade Written Answer:** The regression tree identifies high- and low-bill customer segments through split rules, often emphasizing usage and contract variables. Terminal node means provide direct, business-interpretable bill predictions.

## Task 1g

**Question:** Predict `bill_avg` with tree from 1f and evaluate.

**How to recognize this method on the exam (The Detective Work):**
- Same target and metric as prior regression tasks.

**Step-by-step code:**

```{r}
#| eval: true

pred_tree <- predict(tree_reg, newdata = test)
mse_tree <- mean((test$bill_avg - pred_tree)^2)

mse_tree
```

**AAA-Grade Written Answer:** The regression tree provides competitive performance but is often slightly weaker than OLS for this dataset unless strong nonlinear effects dominate.

## Task 1h

**Question:** Fit random forest for `bill_avg`, plot variable importance, interpret, and explain measure.

**How to recognize this method on the exam (The Detective Work):**
- Ensemble request with importance implies random forest regression.
- For regression RF, importance often uses decrease in RSS node impurity.

**Step-by-step code:**

```{r}
#| eval: true

rf_reg <- randomForest(bill_avg ~ ., data = train, ntree = 50)
pred_rf_reg <- predict(rf_reg, newdata = test)
mse_rf_reg <- mean((test$bill_avg - pred_rf_reg)^2)

mse_rf_reg
varImpPlot(rf_reg)
importance(rf_reg)
```

**AAA-Grade Written Answer:** Random forest usually improves prediction by averaging many decorrelated trees, reducing variance relative to one tree. Importance values summarize average impurity reduction attributable to each predictor.

## Task 1i

**Question:** Use random forest model from 1h to predict `bill_avg` and evaluate.

**How to recognize this method on the exam (The Detective Work):**
- This confirms performance from ensemble model in explicit evaluation step.

**Step-by-step code:**

```{r}
#| eval: true

# Reuse predictions and report MSE clearly.
tibble(model = "Random forest", test_mse = mse_rf_reg)
```

**AAA-Grade Written Answer:** The random forest generally gives the lowest test MSE among models in Task 1, indicating it captures nonlinearities and interactions effectively.

## Task 1j

**Question:** Based on 1a to 1i, what features characterize customers with high versus low average bills?

**How to recognize this method on the exam (The Detective Work):**
- This is synthesis from descriptive stats and model outputs.
- Use signs from linear models and split logic from trees.

**Step-by-step code:**

```{r}
#| eval: true

# Coefficient-based directional evidence.
tidy_coefs <- broom::tidy(ols_full)
head(tidy_coefs)

# Segment means for practical summary.
train %>%
  mutate(bill_group = if_else(bill_avg >= median(bill_avg, na.rm = TRUE), "high", "low")) %>%
  group_by(bill_group) %>%
  summarise(
    mean_download = mean(download_avg, na.rm = TRUE),
    mean_upload = mean(upload_avg, na.rm = TRUE),
    mean_remaining_contract = mean(remaining_contract, na.rm = TRUE)
  )
```

**AAA-Grade Written Answer:** High-bill customers generally show heavier usage patterns and distinct contract/subscription profiles, while low-bill customers show lower usage intensity and lower service load. The exact pattern is consistent across OLS and tree-based summaries.

## Task 2a

**Question:** Ensure right formats, remove unhelpful variables with motivation, and use descriptive statistics to find promising churn predictors.

**How to recognize this method on the exam (The Detective Work):**
- Binary outcome requires careful factor encoding.
- You must justify feature relevance and possible leakage.

**Step-by-step code:**

```{r}
#| eval: true

# Reconfirm formats for churn classification.
churn_cls <- churn_raw %>%
  select(-id) %>%
  mutate(
    is_tv_subscriber = as.factor(is_tv_subscriber),
    is_movie_package_subscriber = as.factor(is_movie_package_subscriber),
    churn = as.factor(churn)
  )

# Group summaries by churn class.
churn_cls %>%
  group_by(churn) %>%
  summarise(
    tv_rate = mean(as.numeric(as.character(is_tv_subscriber))),
    movie_rate = mean(as.numeric(as.character(is_movie_package_subscriber))),
    mean_remaining_contract = mean(remaining_contract, na.rm = TRUE),
    mean_download_over_limit = mean(download_over_limit, na.rm = TRUE)
  )
```

**AAA-Grade Written Answer:** Promising churn predictors include contract horizon, service failures, over-limit behavior, and subscription package indicators. Format corrections ensure logistic and tree models treat categorical predictors correctly.

## Task 2b

**Question:** Using the first 50 original churn observations, bootstrap a 95% confidence interval for churn probability and compare with normal approximation.

**How to recognize this method on the exam (The Detective Work):**
- Parameter is Bernoulli probability p.
- Bootstrap percentile interval and Wald approximation should be compared.

**Step-by-step code:**

```{r}
#| eval: true

# First 50 original values.
y <- as.numeric(as.character(churn_cls$churn[1:50]))
B <- 1000

# Bootstrap sample proportions.
set.seed(1)
p_boot <- replicate(B, mean(sample(y, size = length(y), replace = TRUE) == 1))

# Percentile interval.
ci_boot <- quantile(p_boot, probs = c(0.025, 0.975))

# Standard normal approximation interval.
phat <- mean(y == 1)
n <- length(y)
ci_norm <- c(
  phat - 1.96 * sqrt(phat * (1 - phat) / n),
  phat + 1.96 * sqrt(phat * (1 - phat) / n)
)

ci_boot
ci_norm
```

**AAA-Grade Written Answer:** The bootstrap and normal-approximation intervals are usually very close in this task, indicating the normal approximation works reasonably well for this sample size and estimated churn fraction.

## Task 2c

**Question:** Fit logistic regression with all variables for churn, interpret coefficients for TV and movie package subscription, and discuss signs.

**How to recognize this method on the exam (The Detective Work):**
- Binary target with coefficient interpretation in odds scale.
- Requested variables should be translated from log-odds to odds ratios.

**Step-by-step code:**

```{r}
#| eval: true

logit_all <- glm(churn ~ ., data = train, family = binomial())
summary(logit_all)

# Convert key effects to odds ratios.
exp(coef(logit_all)[c("is_tv_subscriber1", "is_movie_package_subscriber1")])
```

**AAA-Grade Written Answer:** A negative TV-subscriber coefficient implies lower churn odds for TV subscribers, holding other variables fixed. The movie-package effect is weaker but often also negative. Most signs are directionally plausible given customer commitment and contract context.

## Task 2d

**Question:** Predict churn with logistic model from 2c, evaluate appropriately, and test whether removing variables improves predictions.

**How to recognize this method on the exam (The Detective Work):**
- Classification evaluation should include confusion matrix and class-wise rates.
- Reduced model check tests robustness against non-significant predictors.

**Step-by-step code:**

```{r}
#| eval: true

# Full model predictions.
prob_full <- predict(logit_all, newdata = test, type = "response")
pred_full <- if_else(prob_full > 0.5, "1", "0") %>% factor(levels = levels(test$churn))
cm_full <- table(actual = test$churn, predicted = pred_full)
cm_full

# Reduced model without weak terms from marking guidance.
logit_red <- glm(churn ~ . - bill_avg - upload_avg, data = train, family = binomial())
prob_red <- predict(logit_red, newdata = test, type = "response")
pred_red <- if_else(prob_red > 0.5, "1", "0") %>% factor(levels = levels(test$churn))
cm_red <- table(actual = test$churn, predicted = pred_red)
cm_red
```

**AAA-Grade Written Answer:** Logistic predictions are strong with threshold 0.5 in this dataset. Removing weak variables usually changes confusion-matrix performance only marginally, so the gain from simplification is limited.

## Task 2e

**Question:** Use random forest to predict churn and evaluate.

**How to recognize this method on the exam (The Detective Work):**
- Ensemble trees are a standard performance upgrade check after logistic baseline.

**Step-by-step code:**

```{r}
#| eval: true

rf_cls <- randomForest(churn ~ ., data = train, ntree = 50)
pred_rf_cls <- predict(rf_cls, newdata = test)
cm_rf_cls <- table(actual = test$churn, predicted = pred_rf_cls)
cm_rf_cls
sum(diag(cm_rf_cls)) / sum(cm_rf_cls)
```

**AAA-Grade Written Answer:** Random forest typically improves churn classification noticeably, especially for the majority class and often for the minority class as well, by capturing interaction and nonlinear structure.

## Task 2f

**Question:** Based on 2a to 2e, describe typical features of customers who churn.

**How to recognize this method on the exam (The Detective Work):**
- Integrate descriptive differences with model signs and feature importance.

**Step-by-step code:**

```{r}
#| eval: true

# Support narrative with grouped means.
train %>%
  mutate(churn_num = as.numeric(as.character(churn))) %>%
  group_by(churn) %>%
  summarise(
    mean_subscription_age = mean(subscription_age, na.rm = TRUE),
    mean_remaining_contract = mean(remaining_contract, na.rm = TRUE),
    mean_service_failures = mean(service_failure_count, na.rm = TRUE),
    mean_over_limit = mean(download_over_limit, na.rm = TRUE)
  )
```

**AAA-Grade Written Answer:** Typical churners tend to have shorter relationship horizon, lower remaining contract commitment, higher service friction, and behavior indicating usage-limit tension. These patterns are consistent across logistic and random-forest analyses.

# 💡 Generalizable Tips and Tricks (2023 Exam)

| Tip | Why it matters | Reusable snippet |
|---|---|---|
| Remove IDs early | Prevents noise and false signal from identifier columns | `df <- df %>% select(-id)` |
| Keep metric consistent | Fair model comparison needs same test metric | `mean((y_true - y_pred)^2)` |
| Encode factors before GLM/RF | Correct handling of categorical predictors | `mutate(across(c(var1,var2), as.factor))` |
| Use model.matrix for glmnet | LASSO needs numeric matrix with dummy coding | `X <- model.matrix(y ~ ., data=df)[,-1]` |
| Compare full vs reduced logistic | Tests whether simplification improves generalization | `glm(churn ~ . - weak_var1 - weak_var2, family=binomial())` |
| Always show confusion matrix | Accuracy alone can hide class imbalance behavior | `table(actual, predicted)` |
