---
title: "BAN404 Practical Cheatsheet"
subtitle: "Exam-first, tidyverse-first, and built for runnable adaptation"
author: "William"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  eval: false
editor: visual
---

# How to use this file

Use this file when you need a **full practical workflow** instead of isolated code lines.

The idea is:

1. read data,
2. inspect data,
3. clean only what is needed,
4. do EDA with sensible plots,
5. choose models,
6. evaluate and explain.

Everything below is written so it is easy to adapt during the exam.

# 1. Setup

```{r}
library(tidyverse)      # main exam workflow: readr + dplyr + ggplot2 + tibble + forcats
library(randomForest)   # strong nonlinear benchmark model
library(gbm)            # boosting model if asked for or useful

set.seed(123)           # replace only if the exam gives a different seed
```

# 2. Read and understand the data

```{r}
# Use read_csv() for comma separated files.
# Use read_csv2() for semicolon separated files.
data_raw <- read_csv("[REPLACE_WITH_FILE_NAME].csv")
# data_raw <- read_csv2("[REPLACE_WITH_FILE_NAME].csv")

glimpse(data_raw)       # best first look at names, types, and example values
summary(data_raw)       # helps spot strange ranges, missing values, and skew
head(data_raw, 5)       # good for understanding the row structure
```

## What to write after this step

- what the target variable seems to be,
- which variables look numeric,
- which variables look categorical,
- whether any columns look like IDs or leakage.

# 3. Clean the data

```{r}
data <- data_raw %>%
  rename_with(~ gsub("\\.", "_", make.names(.x)), everything()) %>%   # easier names make the exam workflow less error-prone
  mutate(across(where(is.character), as.factor))               # character columns are often true categories

missing_tbl <- data %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  arrange(desc(n_missing))

missing_tbl
```

## Important reminders

- Convert a numeric variable to factor if it is really a category code.
- Remove a variable only if you can explain why.
- Good removal reasons: unique ID, leakage, or information unavailable at prediction time.

# 4. Split train and test

```{r}
index <- sample(seq_len(nrow(data)), size = floor(0.5 * nrow(data)))
train <- data %>% slice(index)
test  <- data %>% slice(-index)

tibble(sample = c("train", "test"), n = c(nrow(train), nrow(test)))
```

# 5. EDA — what plot to use when

## 5.1 One numeric variable: histogram

```{r}
train %>%
  ggplot(aes(x = [REPLACE_NUMERIC_VAR])) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of [REPLACE_NUMERIC_VAR]",
    x = "[REPLACE_NUMERIC_VAR]",
    y = "Count"
  )
```

Use this when you want to check:

- skewness,
- outliers,
- unusual spikes,
- whether transformations may help.

## 5.2 Numeric variable vs target group: boxplot

```{r}
train %>%
  ggplot(aes(x = target, y = [REPLACE_NUMERIC_VAR], fill = target)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "[REPLACE_NUMERIC_VAR] by target",
    x = "Target",
    y = "[REPLACE_NUMERIC_VAR]"
  )
```

Use this when the target is binary or categorical and you want to see whether groups differ.

## 5.3 Two numeric variables: scatter plot

```{r}
train %>%
  ggplot(aes(x = [REPLACE_X], y = [REPLACE_Y])) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(
    title = "[REPLACE_Y] against [REPLACE_X]",
    x = "[REPLACE_X]",
    y = "[REPLACE_Y]"
  )
```

Use this when you want to see trend, direction, and possible nonlinearity.

## 5.4 Categorical variable vs target: within-group proportions

```{r}
train %>%
  count([REPLACE_FACTOR_VAR], target) %>%
  group_by([REPLACE_FACTOR_VAR]) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = [REPLACE_FACTOR_VAR], y = prop, fill = target)) +
  geom_col() +
  labs(
    title = "Target composition within [REPLACE_FACTOR_VAR]",
    x = "[REPLACE_FACTOR_VAR]",
    y = "Within-group proportion"
  )
```

Use this when you want to see whether outcome rates differ by category.

## 5.5 Collinearity check: correlation heatmap

```{r}
cor_tbl <- train %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

cor_tbl %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "navy", midpoint = 0) +
  labs(title = "Correlation heatmap") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## What to remember about collinearity

- high collinearity can make OLS coefficients unstable,
- **ridge** is often better when the goal is prediction with correlated predictors,
- **lasso** is useful when you want a smaller selected set of predictors.

# 6. Baseline classification model: logistic regression

```{r}
log_fit <- glm(target ~ ., data = train, family = binomial())
summary(log_fit)
exp(coef(log_fit))

log_prob <- predict(log_fit, newdata = test, type = "response")
log_pred <- if_else(log_prob > 0.5, "yes", "no") %>%
  factor(levels = levels(test$target))

log_cm <- table(Actual = test$target, Predicted = log_pred)
log_cm
prop.table(log_cm, margin = 1)
log_accuracy <- sum(diag(log_cm)) / sum(log_cm)
log_accuracy
```

# 7. Flexible benchmark: random forest

```{r}
rf_fit <- randomForest(
  target ~ .,
  data = train,
  ntree = 500,
  mtry = floor(sqrt(ncol(train) - 1)),
  importance = TRUE
)

rf_pred <- predict(rf_fit, newdata = test)
rf_cm <- table(Actual = test$target, Predicted = rf_pred)
rf_cm
prop.table(rf_cm, margin = 1)
rf_accuracy <- sum(diag(rf_cm)) / sum(rf_cm)
rf_accuracy

importance(rf_fit)
varImpPlot(rf_fit)
```

# 8. Flexible benchmark: boosting

```{r}
train_bt <- train %>%
  mutate(target_num = if_else(target == "yes", 1, 0)) %>%
  select(-target)

test_bt <- test %>%
  mutate(target_num = if_else(target == "yes", 1, 0))

boost_fit <- gbm(
  target_num ~ .,
  data = train_bt,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  bag.fraction = 0.8
)

boost_prob <- predict(boost_fit, newdata = test_bt, n.trees = 1000, type = "response")
boost_pred <- if_else(boost_prob > 0.5, 1, 0)
boost_cm <- table(Actual = test_bt$target_num, Predicted = boost_pred)
boost_cm
prop.table(boost_cm, margin = 1)
boost_accuracy <- sum(diag(boost_cm)) / sum(boost_cm)
boost_accuracy
```

# 9. Final comparison

```{r}
tibble(
  model = c("logistic", "random_forest", "boosting"),
  accuracy = c(log_accuracy, rf_accuracy, boost_accuracy)
) %>%
  arrange(desc(accuracy))
```

# 10. Final writing checklist

When you write the answer, include:

- what data preparation you did,
- what EDA suggested,
- which variables looked important,
- which model performed best,
- and one short reason why that result makes sense.
