---
title: "Task 2 Cheatsheet"
subtitle: "Fill this in during the exam, then keep only the parts you actually use"
author: "Candidate [REPLACE: number]"
date: "[REPLACE: exam date]"
format:
  html:
    toc: true
    toc-depth: 2
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# How to use this file under exam pressure

1. Replace all `[REPLACE: ...]` markers first.
2. Keep the setup chunk, data-cleaning chunk, and EDA chunk.
3. Delete model sections you do not need.
4. After each code chunk, write **2-5 lines** explaining what you found and why it matters.
5. Only claim something in text if your code output supports it.

## Mental workflow for Task 2

- First understand the dataset
- Then clean only what is necessary
- Then do EDA to find promising predictors and possible problems
- Then fit a simple baseline model first
- Then fit one or two stronger models
- Then compare and conclude in plain language

# 1. Setup and first look

```{r setup}
library(tidyverse)      # dplyr, ggplot2, readr, tibble, stringr, forcats, purrr
library(randomForest)   # random forest for a strong nonlinear baseline
library(gbm)            # boosting if the task asks for it or if you want comparison

set.seed(123)           # replace if the exam gives a specific seed

# Read the file exactly as delivered in the exam.
# Use read_csv() for comma separated files and read_csv2() for semicolon separated files.
data_raw <- read_csv("[REPLACE: filename.csv]")     # simplest tidyverse import for comma-separated data
# data_raw <- read_csv2("[REPLACE: filename.csv]")   # use this instead if the file is semicolon-separated

# First inspection: do this before changing anything.
glimpse(data_raw)       # quick overview of variable names, types, and a few values
summary(data_raw)       # useful first summary to spot strange ranges and missing values
head(data_raw, 5)       # first rows so you can visually understand the data structure
```

**Write after running:**

- What is the response / target variable?
- Which columns look categorical?
- Which columns may be IDs, leakage variables, or post-outcome variables?
- Are there obvious missing-value or coding issues?

# 2. Cleaning and exam-safe preprocessing

```{r cleaning}
# Start from the raw data so your workflow stays easy to explain.
data <- data_raw %>%
  rename_with(~ gsub("\\.", "_", make.names(.x)), everything()) %>%   # safer column names without needing an extra package
  mutate(
    across(where(is.character), as.factor)                     # character columns are often categorical in exam datasets
  )

# If the exam dataset has known categorical numeric codes, convert them manually.
# data <- data %>%
#   mutate(region = as.factor(region))                         # use this when a numeric variable is really a category

# Remove columns only when you can justify it in words.
# Typical reasons: unique ID, leakage, post-event information, or text field not used in the model.
# data <- data %>%
#   select(-customer_id, -claim_amount_after_event)

# Create a binary target only if the exam task asks for it.
# data <- data %>%
#   mutate(target = factor(if_else(num_claims > 0, "yes", "no")))

# Quick missing-value check.
missing_overview <- data %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "n_missing") %>%
  arrange(desc(n_missing))

missing_overview
```

**Write after running:**

- What did you convert to factor, and why?
- What did you remove, and why?
- Are missing values small enough to ignore, or do they need handling?

# 3. Train / test split

```{r split}
# Keep the split simple and easy to explain.
set.seed(123)
index <- sample(seq_len(nrow(data)), size = floor(0.5 * nrow(data)))

train <- data %>% slice(index)
test  <- data %>% slice(-index)

tibble(
  sample = c("train", "test"),
  n = c(nrow(train), nrow(test))
)

# If the target is binary, always inspect class balance.
# train %>% count(target) %>% mutate(prop = n / sum(n))
```

**Write after running:**

- How many observations are in train and test?
- Is the class balance reasonable or very skewed?
- If skewed: mention that threshold choice will matter later.

# 4. EDA — understand the data before modeling

## 4.1 Numeric summaries

```{r numeric-summary}
train %>%
  select(where(is.numeric)) %>%
  summary()
```

## 4.2 Histograms for single numeric variables

```{r histograms}
# Replace the variable names with the most important numeric columns.
train %>%
  ggplot(aes(x = [REPLACE_NUMERIC_VAR])) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of [REPLACE_NUMERIC_VAR]",
    x = "[REPLACE_NUMERIC_VAR]",
    y = "Count"
  )
```

## 4.3 Boxplots for numeric variable vs target

```{r boxplot}
# Use this when the target is binary or categorical and you want to compare groups.
train %>%
  ggplot(aes(x = target, y = [REPLACE_NUMERIC_VAR], fill = target)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(
    title = "[REPLACE_NUMERIC_VAR] by target group",
    x = "Target group",
    y = "[REPLACE_NUMERIC_VAR]"
  )
```

## 4.4 Bar plot for categorical variable vs target

```{r barplot-cat}
# Use this when you want to compare how the target is distributed across categories.
train %>%
  count([REPLACE_FACTOR_VAR], target) %>%
  group_by([REPLACE_FACTOR_VAR]) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = [REPLACE_FACTOR_VAR], y = prop, fill = target)) +
  geom_col(position = "stack") +
  labs(
    title = "Target mix within [REPLACE_FACTOR_VAR]",
    x = "[REPLACE_FACTOR_VAR]",
    y = "Within-category proportion"
  )
```

## 4.5 Correlation heatmap for numeric predictors

```{r corr-heatmap}
# Use this to check multicollinearity among numeric predictors.
cor_tbl <- train %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column(var = "var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

cor_tbl %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "navy", midpoint = 0) +
  labs(
    title = "Correlation heatmap for numeric predictors",
    x = NULL,
    y = NULL
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Interpretation reminder:**

- Histograms help you see skewness, outliers, and unrealistic values.
- Boxplots help you see whether a numeric predictor separates outcome groups.
- Stacked proportions help you see whether a categorical predictor changes target rates.
- Very high absolute correlations suggest multicollinearity; if OLS becomes unstable, **ridge is usually safer for prediction**, while **lasso is useful when you want variable selection**.

# 5. Baseline model — logistic regression

```{r logistic}
# Start simple because it is easy to explain and often gives a strong baseline.
log_fit <- glm(target ~ ., data = train, family = binomial())
summary(log_fit)

# Odds ratios make coefficient interpretation easier in text.
exp(coef(log_fit))
```

**Write after running:**

- Which variables are clearly important?
- What does the sign of the coefficient imply?
- Which odds ratios are meaningfully above or below 1?

# 6. Evaluate logistic regression

```{r logistic-eval}
log_prob <- predict(log_fit, newdata = test, type = "response")
threshold <- 0.5                                               # lower this if the positive class is rare and you want more sensitivity
log_pred <- if_else(log_prob > threshold, "yes", "no") %>%
  factor(levels = levels(test$target))

log_cm <- table(Actual = test$target, Predicted = log_pred)
log_cm
prop.table(log_cm, margin = 1)

log_accuracy <- sum(diag(log_cm)) / sum(log_cm)
log_accuracy
```

**Write after running:**

- State the threshold used and why it is reasonable.
- Mention sensitivity / specificity if the task is about classification quality.
- Do not only report accuracy if class imbalance is strong.

# 7. Stronger nonlinear model — random forest

```{r rf}
# Random forest is often a good comparison because it can capture nonlinearities and interactions automatically.
rf_fit <- randomForest(
  target ~ .,                                                    # formula interface is easy to explain in the written answer
  data = train,
  ntree = 500,
  mtry = floor(sqrt(ncol(train) - 1)),
  importance = TRUE
)

rf_fit
rf_pred <- predict(rf_fit, newdata = test)
rf_cm <- table(Actual = test$target, Predicted = rf_pred)
rf_cm
prop.table(rf_cm, margin = 1)

rf_accuracy <- sum(diag(rf_cm)) / sum(rf_cm)
rf_accuracy

importance(rf_fit)
varImpPlot(rf_fit)
```

**Write after running:**

- Compare random forest with logistic regression.
- Mention whether performance improved and which variables seem most important.
- If RF is better, say it may capture nonlinearities / interactions missed by logistic regression.

# 8. Optional model — boosting

```{r boosting}
# Boosting is useful when the task explicitly asks for it or when you want a second flexible model.
train_bt <- train %>%
  mutate(target_num = if_else(target == "yes", 1, 0)) %>%
  select(-target)

test_bt <- test %>%
  mutate(target_num = if_else(target == "yes", 1, 0))

boost_fit <- gbm(
  target_num ~ .,                                                # numeric 0/1 target for bernoulli boosting
  data = train_bt,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  bag.fraction = 0.8
)

boost_prob <- predict(boost_fit, newdata = test_bt, n.trees = 1000, type = "response")
boost_pred <- if_else(boost_prob > threshold, 1, 0)
boost_cm <- table(Actual = test_bt$target_num, Predicted = boost_pred)
boost_cm
prop.table(boost_cm, margin = 1)

boost_accuracy <- sum(diag(boost_cm)) / sum(boost_cm)
boost_accuracy
```

# 9. Final comparison table

```{r compare}
tibble(
  model = c("logistic_regression", "random_forest", "boosting"),
  accuracy = c(log_accuracy, rf_accuracy, boost_accuracy)
) %>%
  arrange(desc(accuracy))
```

## Final writing reminder

In your final answer, say:

1. what you did,
2. what the main findings were,
3. which model performed best,
4. why that result makes sense,
5. and any practical caveat (class imbalance, leakage risk, multicollinearity, small sample, etc.).
