---
title: "Task 1 compendium"
author: "William"
format: html
editor: visual
---

# TASK 1 CHEAT SHEET: "Explain the Code" Dictionary

**How to use this section in the exam:** 1. Look at the code the professor gives you. 2. Scan the code for "The Tell" (the specific keyword or math symbol that gives away the secret). 3. Find the matching block below. 4. Copy the "Plug-and-Play Answer" and adjust the variable names in the brackets `[ ]`.

------------------------------------------------------------------------

## GROUP 1: The Resampling Loops (Shuffling the Data)

### 1. The Bootstrap

**The Logic:** The computer is trying to find the uncertainty (variance, standard error, etc.) of a specific metric. It does this by creating thousands of fake datasets from the original data, making sure to put data points back in the pile after picking them so they can be picked again.

**The Raw Code you might see:**

```{r, eval=FALSE}

f_boot <- function(data, B=1000) {
  n <- nrow(data)
  res <- numeric(B)
  for(i in 1:B) {
    # THE TELL is on the next line!
    idx <- sample(1:n, size=n, replace=TRUE) 
    fake_data <- data[idx, ]
    model <- lm(y ~ x, data=fake_data)
    res[i] <- coef(model)[2]
  }
  return(sd(res))
}
```

**The Tell:** Look for `sample(...)` combined with `replace = TRUE` (or `replace = T`). If you see `replace=TRUE`, it is **100% the Bootstrap**.

**Plug-and-Play Answer:**

> "This R-function implements the **Bootstrap** method. It estimates the standard error (or uncertainty) of the \[INSERT METRIC, e.g., regression coefficient\]. It does this by running a `for` loop `B` times. In each iteration, it draws a random sample of size `n` from the original dataset **with replacement** (indicated by `replace=TRUE`). It then fits the model on this simulated dataset, stores the result, and finally returns the standard deviation of all the simulated results."

### **2. K-Fold Cross Validation**

**The Logic:** The computer is trying to find out how accurate a model is on *unseen* data. It chops the data into K blocks (usually 5 or 10). It holds one block back, trains on the rest, and tests on the holdout block. It repeats this until every block has been the test set.

```{r}
#| eval: false
f_kfold <- function(X, y, K=5) {
  n <- length(y)
  # THE TELL is on the next line!
  folds <- sample(rep(1:K, length=n)) 
  cv_errors <- numeric(K)
  
  for(i in 1:K) {
    train_X <- X[folds != i, ]
    train_y <- y[folds != i]
    test_X <- X[folds == i, ]
    test_y <- y[folds == i]
    
    b <- solve(t(train_X) %*% train_X) %*% t(train_X) %*% train_y
    preds <- test_X %*% b
    cv_errors[i] <- mean((test_y - preds)^2)
  }
  return(mean(cv_errors))
}
```

**The Tell:** Look for the data being split into `K` pieces, and a loop `for(i in 1:K)`. Look closely at `folds != i` (training data) and `folds == i` (test data).

**Plug-and-Play Answer:**

> "This R-function performs **K-Fold Cross-Validation**. The goal is to estimate the Test MSE of the model. The function randomly divides the dataset into `K` equally sized groups (folds). It uses a `for` loop to iterate `K` times. In each step `i`, it treats the `i`-th fold as the test set (`folds == i`) and the remaining folds as the training set (`folds != i`). It fits the model on the training data, calculates the squared errors on the test data, and finally returns the average error across all `K` folds."

## **GROUP 2: Loss Functions & Optimization (Phase 3 Methods)**

### **3. Ridge Regression (Used in 2025 Exam)**

**The Logic:** The computer is trying to fit a linear regression, but it wants to shrink the variables to prevent overfitting. It does this by adding a penalty for having large coefficients. The Ridge penalty squares the coefficients.

**The Raw Code you might see:**

```{r, eval=FALSE}

f_ridge <- function(b, X, y, la) {
  # THE TELL is on the next line!
  loss <- sum((y - X %*% b)^2) + la * sum(b^2) 
  return(loss)
}

```

**Part A: Loss function**

```{r, eval=FALSE}

# Part A: Loss function

f <-function(b1,X,y,la){
  fval <- sum ((y-X%*%b1)ˆ2)+la*sum(b1ˆ2)
  return(fval)
}

```

```{r, eval=FALSE}

g <- function(X, y, la) {
  q <- ncol(X)
  
  # 1. THE INTERCEPT TRICK
  b0 <- mean(y)   
  yd <- y - b0    
  
  # 2. THE DEMEANING LOOP (The biggest exam trap!)
  Xd <- X
  for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])
  
    # 3. THE OPTIMIZER
  b1start <- rep(0, q)
  opt <- nlminb(b1start, f, X=Xd, y=yd, la=la)
  b1 <- opt$par
  return(b1)
}
```

*Explanation:* `nlminb` is an R function that means "Non-Linear Minimization with Box Constraints". The professor is telling the computer: *"Start with all coefficients at 0 (`b1start`). Keep guessing new coefficients and plugging them into function `f`. When you find the coefficients that make `fval` as small as possible, stop and give me those numbers (`opt$par`)."*

**The Tell:** Look for the basic OLS math `sum((y - ...)^2)` PLUS a penalty that looks like `la * sum(b^2)`. The key is `b^2` (squaring the coefficients).

**Plug-and-Play Answer:**

> If you see this exact two-part chunk (`f` and `g`) on the exam, here is the flawless answer that proves you understand everything:
>
> "These functions work together to manually implement **Ridge Regression**.
>
> The `f` function calculates the objective loss: the standard sum of squared errors plus the **L2 shrinkage penalty** (`la * sum(b1^2)`), which penalizes large coefficients to reduce model variance.
>
> The `g` function prepares the data and finds the optimal coefficients. Crucially, it calculates the mean of y (`b0`) and subtracts it to remove the intercept. It then uses a `for` loop to **demean (center) every X variable**. This is mathematically required for Ridge Regression because the shrinkage penalty must be applied fairly; if variables are not on the same scale, the penalty will disproportionately affect variables with smaller numerical scales. Finally, the function uses `nlminb`, an optimization solver, to find the specific β1 coefficients that minimize the loss function `f`, and returns them."

### **4. Lasso Regression (The logical next step for 2026)**

**The Logic:** Exactly the same as Ridge, but instead of shrinking everything near zero, it wants to delete useless variables completely by forcing them to exactly zero. It does this by using the Absolute Value instead of the square.

**The Raw Code you might see:**

```{r, eval=FALSE}
f_lasso <- function(b, X, y, la) {
  # THE TELL is on the next line!
  loss <- sum((y - X %*% b)^2) + la * sum(abs(b)) 
  return(loss)
}
```

**The Tell:** Look for `abs(b)` (Absolute Value) instead of `b^2`.

**Plug-and-Play Answer:**

> "This R-function calculates the objective loss function for **Lasso Regression** (L1 Regularization). The first part is the standard Residual Sum of Squares. The second part `la * sum(abs(b))` is the **L1 shrinkage penalty**. By using the absolute value of the coefficients instead of the square, the Lasso algorithm is able to force some coefficients to become exactly zero. This means this function mathematically performs both regularization and **variable selection**."

### **5. Ordinary Least Squares (OLS) via Matrix Math**

**The Logic:** This is just standard Linear Regression, but instead of using the friendly `lm()` function, the professor wrote out the raw linear algebra formula.

**The Raw Code you might see:**

```{r, eval=FALSE}
f_ols <- function(X, y) {
  # THE TELL is on the next line!
  b <- solve(t(X) %*% X) %*% t(X) %*% y 
  return(b)
}
```

**The Tell:** Look for `solve()`, which in R means "invert a matrix", combined with `%*%` (matrix multiplication).

**Plug-and-Play Answer:**

> "This R-function computes the exact mathematical coefficients for **Ordinary Least Squares (OLS) Linear Regression**. Instead of using optimization to estimate the coefficients, it uses the closed-form matrix algebra solution: β\^=(XTX)−1XTy. In R, `t(X)` transposes the matrix, `%*%` performs matrix multiplication, and `solve()` computes the inverse of the matrix. It returns the regression coefficients."

This tells RStudio: *"Display this code with pretty colors, but DO NOT run it."* This way, your compendium knits perfectly into an HTML or PDF file without needing the datasets to actually exist!

------------------------------------------------------------------------

## GROUP 3: The 2024 & 2025 Actual Exam Functions

### 6. Leave-One-Out Cross-Validation (LOOCV) (Used in 2025 Exam)

**The Logic:** This is the extreme version of K-Fold CV. Instead of chopping the data into 5 blocks, it leaves exactly *one single row* out. It trains the model on all other rows, tests on the 1 row left out, and repeats this for every single row in the dataset.

**The Raw Code you might see (From 2025):**

```{r, eval=FALSE}


loo <- function(la,X,y)
{
  n <- nrow(X)
  q <- ncol(X)
  b0 <- mean(y)
  yd <- y- b0
  Xd <- X
# THE TELL is on the next line!
  for(k in 1:q) Xd[,k] <- X[,k]- mean(X[,k])
  ydpred <- matrix(0,n,1)
  for(i in 1:n)
  {
  b1 <- g(X=X[-i,],y=y[-i],la)
  ydpred[i] <- Xd[i,]%*%b1
  }
  testMSE <- mean((yd-ydpred)ˆ2)
  return(testMSE)
}
```

**The Tell:** Look for `for(i in 1:n)` (looping exactly as many times as there are rows). Then look for `X[-i, ]` (training data: everything EXCEPT row i) and `X[i, ]` (test data: ONLY row i).

**Plug-and-Play Answer:**

> "This R-function implements **Leave-One-Out Cross-Validation (LOOCV)** to estimate the Test MSE for a specific tuning parameter `la`. It uses a `for` loop that iterates exactly `n` times (the total number of observations). In each iteration `i`, it drops the `i`-th observation from the training data (`X[-i, ]`, `y[-i]`), fits the model using the helper function `g`, and then predicts the value for the single held-out observation `X[i, ]`. Finally, it computes and returns the average squared prediction error across all `n` iterations."

### **7. K-Nearest Neighbors (KNN) Local Regression (Used in 2024 Exam)**

**The Logic:** To predict a new point, the computer looks at all the data, finds the K points that are physically closest to the new point, and builds a tiny local regression just on those neighbors.

**The Raw Code you might see (From 2024):**

```{r, eval = FALSE}


f_knn <- function(x0, x, y, K=3) {
  # THE TELL is the next 2 lines!
  d <- abs(x - x0)          
  o <- order(d)[1:K]        
  
  xl <- x[o]
  yl <- y[o]
  xydata <- data.frame(xl=xl, yl=yl)
  
  reg <- lm(yl ~ xl, data=xydata)
  ypred <- predict(reg, newdata=data.frame(xl=x0, yl=1))
  return(ypred)
}
```

**The Tell:** Look for `abs(x - x0)` (calculating the distance between points) and `order(d)[1:K]` (sorting the distances and picking the K closest ones).

**Plug-and-Play Answer:**

> "This R-function implements a form of **K-Nearest Neighbors (KNN) Local Regression**. To predict the value for a target point `x0`, the function first calculates the absolute distance between `x0` and all other `x` observations (`abs(x - x0)`). It then orders these distances and selects the `K` closest observations (`order(d)[1:K]`). Finally, instead of just averaging the neighbors, it fits a simple linear regression (`lm`) strictly on those `K` nearest neighbors, and uses that local line to predict the value for `x0`."

## **GROUP 4: Advanced Non-Linear Mechanics (Highly Probable Targets)**

### **8. The Backfitting Algorithm (GAMs)**

**The Logic:** How does a GAM fit multiple curvy lines at the same time? It guesses the curve for X1, subtracts it from Y, and fits the curve for X2 on the leftovers (residuals). It repeats this ping-pong process until the curves stop changing. *(Note: You were asked to explain this in theory in 2022 and 2024!)*

**The Raw Code you might see:**

```{r, eval = FALSE }

f_backfit <- function(x1, x2, y) {
  f1 <- rep(0, length(y))
  f2 <- rep(0, length(y))
  
  # THE TELL is on the next lines!
  for(i in 1:10) {
    f1 <- smooth.spline(x1, y - f2)$y
    f2 <- smooth.spline(x2, y - f1)$y
  }
  return(f1 + f2)
}
```

**The Tell:** Look for `y - f2` and `y - f1` inside a smoothing function. Fitting a model on "Y minus the other variable" is the definition of backfitting.

**Plug-and-Play Answer:**

> "This R-function implements the **Backfitting Algorithm**, which is used to fit Generalized Additive Models (GAMs). Because it cannot estimate the non-linear effects of multiple variables simultaneously, it uses an iterative approach. It holds the estimate for `f2` constant, subtracts it from `y` to isolate the residual effect, and fits a smoothing spline for `f1` on that residual. It then holds `f1` constant and does the same for `f2`. This ping-pong process repeats (here, 10 times) until the non-linear curves converge."

### **9. Bootstrap Aggregation (Bagging) Loop**

**The Logic:** To fix the instability of a single Decision Tree, the computer bootstraps 500 fake datasets, grows a tree on each one, and averages the predictions.

**The Raw Code you might see:**

```{r, eval=FALSE}

f_bagging <- function(X, y, B=100) {
  n <- nrow(X)
  preds <- matrix(0, nrow=n, ncol=B)
  
  for(i in 1:B) {
    # THE TELL: Bootstrap sample + fitting a tree inside a loop!
    idx <- sample(1:n, replace=TRUE) 
    tree_mod <- tree(y ~ ., data=X[idx, ])
    preds[, i] <- predict(tree_mod, newdata=X)
  }
  
  return(rowMeans(preds))
}
```

**The Tell:** A combination of two things we've seen: `sample(..., replace=TRUE)` PLUS a model like `tree()` being fit *inside* the loop, followed by `rowMeans()` at the end.

**Plug-and-Play Answer:**

> "This R-function manually implements **Bagging (Bootstrap Aggregating)** for decision trees. It runs a `for` loop `B` times. In each iteration, it draws a bootstrapped sample of the data (with replacement). It then fits a highly flexible decision tree on that specific bootstrapped sample and records the predictions. Because single trees have high variance, the function concludes by using `rowMeans()` to average the predictions of all `B` trees together, which drastically reduces the variance and improves predictive accuracy without increasing bias."
