---
title: "Task 1 compendium"
author: "William"
format: html
editor: visual
---

# TASK 1 CHEAT SHEET: "Explain the Code" Dictionary

**How to use this section in the exam:**
1. Look at the code the professor gives you.
2. Scan the code for "The Tell" (the specific keyword or math symbol that gives away the secret).
3. Find the matching block below.
4. Copy the "Plug-and-Play Answer" and adjust the variable names in the brackets `[ ]`.

------------------------------------------------------------------------

## GROUP 1: The Resampling Loops (Shuffling the Data)

### 1. The Bootstrap
**Math / Theory:** We want the sampling distribution and standard error of a statistic. We simulate $B$ datasets of size $n$ **with replacement** and compute the statistic each time.

**The Raw Code you might see:**

```{r}
#| eval: false

f_boot <- function(data, B=1000) {              # Function takes data and number of simulations
  n <- nrow(data)                               # Number of observations
  res <- numeric(B)                             # Container for the B simulated statistics
  for(i in 1:B) {                               # Loop over simulations
    idx <- sample(1:n, size=n, replace=TRUE)    # THE TELL: sample WITH replacement (bootstrap)
    fake_data <- data[idx, ]                    # Create a bootstrap sample
    model <- lm(y ~ x, data=fake_data)          # Fit model on bootstrap sample
    res[i] <- coef(model)[2]                    # Save the statistic (e.g., beta1)
  }
  return(sd(res))                               # Standard error of the statistic
}
```

**The Tell:** `sample(..., replace = TRUE)`.

**Plug-and-Play Answer:**
> "This R-function implements the **Bootstrap** method. It estimates the standard error of the [INSERT METRIC]. It runs a `for` loop `B` times, and in each iteration it draws a random sample of size `n` **with replacement** (`replace=TRUE`). It fits the model on this simulated dataset, stores the statistic, and finally returns the standard deviation of the simulated statistics, which is the bootstrapped standard error."

------------------------------------------------------------------------

### 2. K-Fold Cross Validation
**Math / Theory:** Estimate Test MSE by splitting data into $K$ folds, training on $K-1$, testing on 1, and averaging errors.

```{r}
#| eval: false

f_kfold <- function(X, y, K=5) {                # X predictors, y response, K folds
  n <- length(y)                                # Number of observations
  folds <- sample(rep(1:K, length=n))           # THE TELL: assign each row to a fold
  cv_errors <- numeric(K)                       # Container for fold errors

  for(i in 1:K) {                               # Loop through each fold
    train_X <- X[folds != i, ]                  # Training data (all but fold i)
    train_y <- y[folds != i]
    test_X  <- X[folds == i, ]                  # Test data (fold i)
    test_y  <- y[folds == i]

    b <- solve(t(train_X)%*%train_X) %*% t(train_X)%*%train_y  # Fit OLS
    preds <- test_X %*% b                       # Predict on test fold
    cv_errors[i] <- mean((test_y - preds)^2)    # MSE for this fold
  }
  return(mean(cv_errors))                       # Average CV error
}
```

**The Tell:** `for(i in 1:K)` with `folds != i` vs `folds == i`.

**Plug-and-Play Answer:**
> "This R-function performs **K-Fold Cross-Validation**. The data are split into `K` folds. For each fold `i`, the model is trained on all other folds and tested on fold `i`, producing a test MSE. The function returns the average test MSE across the K folds."

------------------------------------------------------------------------

### 3. Leave-One-Out Cross-Validation (LOOCV)
**Math / Theory:** Special case of K-fold with $K=n$. Train on $n-1$, test on 1, repeat for all rows.

```{r}
#| eval: false

loo <- function(X, y) {                          # X predictors, y response
  n <- nrow(X)                                   # Number of observations
  ydpred <- matrix(0, n, 1)                      # Store 1 prediction per loop

  for(i in 1:n) {                                # THE TELL: loop over every row
    train_X <- X[-i, ]                           # Train on all but row i
    train_y <- y[-i]
    b1 <- g(X = train_X, y = train_y)            # Fit model on n-1 observations
    ydpred[i] <- X[i, ] %*% b1                   # Predict for left-out row i
  }

  testMSE <- mean((y - ydpred)^2)                # LOOCV MSE
  return(testMSE)
}
```

**The Tell:** `for(i in 1:n)` and `X[-i, ]`.

**Plug-and-Play Answer:**
> "This R-function implements **LOOCV**. It trains the model on all observations except one, predicts the left-out observation, and repeats for every row. The returned value is the mean squared prediction error across all left-out observations."

------------------------------------------------------------------------

## GROUP 2: Loss Functions & Optimization (Phase 3 Methods)

### 4. Ridge Regression Engine (Exam 2025 style)
**Math / Theory:** Ridge minimizes $RSS + \lambda \sum \beta_j^2$. All variables must be centered so the penalty is fair.

```{r}
#| eval: false

# PART A: Loss function
f <- function(b1, X, y, la) {                    # b1 = coefficients, la = lambda
  fval <- sum((y - X%*%b1)^2) + la * sum(b1^2)   # THE TELL: RSS + lambda*sum(beta^2)
  return(fval)
}

# PART B: Data prep + optimizer
g <- function(X, y, la) {
  q <- ncol(X)                                   # Number of predictors
  b0 <- mean(y)                                  # Intercept trick: mean of y
  yd <- y - b0                                   # Center y to remove intercept

  Xd <- X
  for(k in 1:q) {                                # Demeaning loop (critical!)
    Xd[,k] <- X[,k] - mean(X[,k])                # Center each predictor
  }

  b1start <- rep(0, q)                           # Start betas at 0
  opt <- nlminb(b1start, f, X=Xd, y=yd, la=la)   # Optimize loss function
  return(opt$par)                                # Return ridge coefficients
}
```

**The Tell:** `sum((y - X%*%b)^2) + la * sum(b^2)` and the demeaning loop.

**Plug-and-Play Answer:**
> "These functions manually implement **Ridge Regression**. `f` calculates the loss $RSS + \lambda \sum \beta^2$. `g` centers `y` to remove the intercept, **demeans each column of X** so the penalty applies fairly, and then uses `nlminb` to find coefficients that minimize the ridge loss."

------------------------------------------------------------------------

### 5. Lasso Regression Engine (Probable 2026)
**Math / Theory:** Lasso minimizes $RSS + \lambda \sum |\beta_j|$ and can shrink coefficients to exactly 0.

```{r}
#| eval: false

f_lasso <- function(b1, X, y, la) {              # b1 = coefficients
  fval <- sum((y - X%*%b1)^2) + la * sum(abs(b1))# THE TELL: absolute value penalty
  return(fval)
}
```

**The Tell:** `abs(b1)` instead of `b1^2`.

**Plug-and-Play Answer:**
> "This function computes the **Lasso** objective: RSS plus the L1 penalty $\lambda \sum |\beta|$. The absolute value penalty allows coefficients to become exactly zero, so Lasso performs shrinkage and variable selection."

------------------------------------------------------------------------

### 6. Ordinary Least Squares (OLS) Matrix Formula
**Math / Theory:** $\hat{\beta} = (X^T X)^{-1} X^T y$.

```{r}
#| eval: false

f_ols <- function(X, y) {
  b <- solve(t(X) %*% X) %*% t(X) %*% y          # THE TELL: (X'X)^-1 X'y
  return(b)
}
```

**The Tell:** `solve()` with `t(X) %*% X`.

**Plug-and-Play Answer:**
> "This code computes **OLS regression coefficients** using the closed-form matrix solution $(X^T X)^{-1} X^T y$. The `solve()` function inverts the matrix."

------------------------------------------------------------------------

## GROUP 3: The 2024 & 2025 Actual Exam Functions

### 7. K-Nearest Neighbors (KNN) Local Regression (Exam 2024 style)
**Math / Theory:** Find the K closest points to $x_0$ and fit a local model on only those neighbors.

```{r}
#| eval: false

f_knn <- function(x0, x, y, K=3) {
  d <- abs(x - x0)                               # THE TELL: distance from x0 to all x
  o <- order(d)[1:K]                             # Indices of K closest points

  xl <- x[o]                                     # K closest x values
  yl <- y[o]                                     # K closest y values
  xydata <- data.frame(xl=xl, yl=yl)

  reg <- lm(yl ~ xl, data=xydata)                # Fit local regression on neighbors
  ypred <- predict(reg, newdata=data.frame(xl=x0, yl=1))
  return(ypred)
}
```

**The Tell:** `abs(x - x0)` and `order(d)[1:K]`.

**Plug-and-Play Answer:**
> "This function implements **KNN local regression**. It calculates distances from `x0`, selects the K nearest observations, fits a local regression on those K points, and predicts the value at `x0`."

------------------------------------------------------------------------

## GROUP 4: Advanced Non-Linear Mechanics (Highly Probable Targets)

### 8. Backfitting Algorithm (GAMs)
**Math / Theory:** Iteratively fit each smooth term while holding others fixed: $y = f_1(x_1) + f_2(x_2)$.

```{r}
#| eval: false

f_backfit <- function(x1, x2, y) {
  f1 <- rep(0, length(y))                        # Initialize f1
  f2 <- rep(0, length(y))                        # Initialize f2

  for(i in 1:10) {                               # Iterate to convergence
    f1 <- smooth.spline(x1, y - f2)$y            # Fit f1 on residuals (y - f2)
    f2 <- smooth.spline(x2, y - f1)$y            # Fit f2 on residuals (y - f1)
  }
  return(f1 + f2)                                # Final additive fit
}
```

**The Tell:** `y - f2` and `y - f1` inside a smooth.

**Plug-and-Play Answer:**
> "This code implements **backfitting** for a GAM. It alternates between fitting a smooth function for $x_1$ on the residuals after removing $f_2$, then fitting $f_2$ on residuals after removing $f_1$. Repeating this yields the additive model $f_1(x_1)+f_2(x_2)$."

------------------------------------------------------------------------

### 9. Bagging Loop (Bootstrap Aggregation)
**Math / Theory:** Build many trees on bootstrap samples, then average predictions to reduce variance.

```{r}
#| eval: false

f_bagging <- function(X, y, B=100) {
  n <- nrow(X)
  preds <- matrix(0, nrow=n, ncol=B)             # Store B predictions

  for(i in 1:B) {
    idx <- sample(1:n, replace=TRUE)             # Bootstrap sample
    tree_mod <- tree(y ~ ., data=X[idx, ])       # Fit tree on bootstrap sample
    preds[, i] <- predict(tree_mod, newdata=X)   # Predict on full data
  }

  return(rowMeans(preds))                        # Average predictions across trees
}
```

**The Tell:** `sample(..., replace=TRUE)` inside a loop + model fit + `rowMeans()`.

**Plug-and-Play Answer:**
> "This function implements **Bagging**. It creates many bootstrap samples, fits a decision tree to each, and averages the predictions. Averaging reduces variance and improves prediction accuracy compared to a single tree."
