---
title: "BAN404 Exam — Task 2 Working Template"
subtitle: "Replace all placeholders before final submission"
author: "Candidate [REPLACE]"
date: "[REPLACE]"
format:
  pdf:
    toc: false
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# How to use this template during the exam

1. Open this file **immediately** when Task 2 starts.
2. Replace the placeholders in the setup chunk first.
3. Run the setup and inspection chunks before writing anything important.
4. Do **not** jump straight to models. First decide: target, factor variables, leakage variables, and split.
5. Keep code that you use. Delete code you do not use only near the end if you want a cleaner PDF.

> This template is intentionally over-explained. The goal is that you should know **why each step exists** while you are working.

---

# 1. Setup and first inspection

```{r setup}
# Load packages used in the standard BAN404 Task 2 workflow                 # tidyverse handles reading, wrangling, plotting, and summaries
library(tidyverse)
library(randomForest)                                                       # common comparison model for classification tasks
library(gbm)                                                                # common boosting package in course material

set.seed(123)                                                               # replace if the exam gives you another seed

file_path      <- "[REPLACE_WITH_FILE_NAME.csv]"                            # exact file name provided in the exam
file_delimiter <- ","                                                      # change to ";" if the file is semicolon-separated

target_var     <- "[REPLACE_WITH_TARGET]"                                   # the variable you want to predict
positive_class <- "[REPLACE_WITH_POSITIVE_CLASS]"                           # e.g. "yes", "1", "claim", "satisfied"
negative_class <- "[REPLACE_WITH_NEGATIVE_CLASS]"                           # make the opposite class explicit to avoid guessing later

factor_vars    <- c("[REPLACE_FACTOR_1]", "[REPLACE_FACTOR_2]")            # list every categorical variable you want as a factor
id_vars        <- c("[REPLACE_ID_IF_ANY]")                                  # IDs usually add no predictive value
leakage_vars   <- c("[REPLACE_LEAKAGE_IF_ANY]")                             # post-event / target-derived variables must be removed

# Read the data in a way that works for both comma and semicolon files      # this keeps the file-reading logic explicit and easy to change
if (file_delimiter == ";") {
  data_raw <- readr::read_delim(file_path, delim = ";")                     # European-style csv
} else {
  data_raw <- readr::read_csv(file_path)                                     # normal comma-separated csv
}

data_raw |> glimpse()                                                        # first mandatory check: names + types + sample values
data_raw |> head()                                                           # second mandatory check: make sure rows look sensible
data_raw |> summary()                                                        # quick scan for strange ranges and missing values
```

## What you should decide before moving on

- Which variable is the target?
- Is the target already in the format you need?
- Which variables are categories?
- Which variables are leakage or IDs?
- Are there missing values that need handling?

---

# 2. Cleaning and recoding

## 2.1 Count missing values first

```{r missing-check}
data_raw |>
  summarise(across(everything(), ~ sum(is.na(.x))))                          # count missing values for every column before you change anything
```

## 2.2 Create the modeling dataset

```{r clean-data}
# If the target must be created from another variable, do it here first     # uncomment and adapt only if needed
# data_raw <- data_raw |>
#   mutate(
#     target = if_else(number_of_claims >= 1, "yes", "no"),                # example binary target creation from a count
#     target = factor(target, levels = c("no", "yes"))                     # set the non-event as the first level
#   )

data_model <- data_raw |>
  mutate(across(any_of(factor_vars), as.factor)) |>                          # convert known categorical variables to factors
  select(-any_of(c(id_vars, leakage_vars)))                                  # remove IDs and leakage variables safely

# If the target is character, convert it to factor explicitly               # factor target is safer for tables and tree-based classification models
# data_model <- data_model |>
#   mutate(!!target_var := factor(.data[[target_var]]))

data_model |> glimpse()                                                      # inspect again after cleaning so you know the modeling types are correct
```

**Written answer starter:**
“I converted `[factor variables]` to factors because they represent categories, not continuous measurements. I removed `[removed variables]` because they are IDs / leakage variables / post-event variables and would otherwise distort the prediction task.”

---

# 3. Train / test split

```{r split}
set.seed(123)                                                                # repeat here so the split is easy to find when reading your file
n_obs <- nrow(data_model)                                                    # count rows once for readability
idx   <- sample.int(n_obs, size = floor(0.5 * n_obs))                        # 50/50 split is often expected in BAN404 examples

train_data <- data_model |> slice(idx)                                       # training data used to fit models
test_data  <- data_model |> slice(-idx)                                      # test data used only for evaluation

train_data |> count(.data[[target_var]]) |> mutate(prop = n / sum(n))        # always inspect class balance before choosing threshold
test_data  |> count(.data[[target_var]]) |> mutate(prop = n / sum(n))
```

**Why this step exists:**
You need an honest test set. If you evaluate on the same data you trained on, the performance can look much better than it really is.

**Written answer starter:**
“I split the cleaned data into training and test sets using a 50/50 split and a fixed seed for reproducibility. The models are fitted on the training data and evaluated on the test data to estimate out-of-sample performance.”

---

# 4. EDA — do this before fitting models

## 4.1 Create variable lists

```{r variable-lists}
numeric_vars <- train_data |>
  select(where(is.numeric), -any_of(target_var)) |>                          # numeric predictors only; remove target if it is numeric 0/1
  names()

categorical_vars <- train_data |>
  select(where(~ is.factor(.x) || is.character(.x)), -any_of(target_var)) |> # category-like predictors only
  names()

numeric_vars
categorical_vars
```

## 4.2 Numeric predictor vs binary target — boxplots

```{r eda-boxplots}
# Replace "hp" with one of your real numeric variables first               # start with the variables you think are most likely to matter
eda_numeric_var <- "[REPLACE_NUMERIC_VAR]"

ggplot(train_data, aes(x = .data[[target_var]], y = .data[[eda_numeric_var]], fill = .data[[target_var]])) +
  geom_boxplot(alpha = 0.80, outlier.color = "red") +                      # boxplot is usually the clearest first EDA plot for numeric vs binary target
  labs(
    title = paste(eda_numeric_var, "by", target_var),
    x = target_var,
    y = eda_numeric_var,
    fill = target_var
  ) +
  theme_minimal()

train_data |>
  group_by(.data[[target_var]]) |>
  summarise(
    mean_value   = mean(.data[[eda_numeric_var]], na.rm = TRUE),             # average difference supports what you see in the plot
    median_value = median(.data[[eda_numeric_var]], na.rm = TRUE),           # median is robust and often more stable to discuss
    sd_value     = sd(.data[[eda_numeric_var]], na.rm = TRUE),               # spread matters for overlap and separation
    n            = n(),
    .groups      = "drop"
  )
```

**What to ask yourself:**
- Do the medians differ clearly?
- Is one group much more spread out than the other?
- Are there strong outliers?
- Does the predictor look useful for distinguishing the classes?

## 4.3 Categorical predictor vs binary target — proportions

```{r eda-categories}
# Replace "cylinders" with one real categorical predictor                  # choose the variable that seems most interesting after inspection
eda_categorical_var <- "[REPLACE_CATEGORICAL_VAR]"

train_data |>
  count(.data[[eda_categorical_var]], .data[[target_var]]) |>
  group_by(.data[[eda_categorical_var]]) |>
  mutate(prop = n / sum(n)) |>                                               # within each category, what share belongs to each target class?
  ungroup()

train_data |>
  count(.data[[eda_categorical_var]], .data[[target_var]]) |>
  group_by(.data[[eda_categorical_var]]) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = .data[[eda_categorical_var]], y = prop, fill = .data[[target_var]])) +
  geom_col(position = "dodge") +                                            # dodge is easier to compare than stacked bars for this purpose
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = paste("Target proportions by", eda_categorical_var),
    x = eda_categorical_var,
    y = "Share within category",
    fill = target_var
  ) +
  theme_minimal()
```

## 4.4 Correlation heatmap for numeric predictors

```{r eda-correlation}
correlation_tbl <- train_data |>
  select(where(is.numeric), -any_of(target_var)) |>                          # keep numeric predictors only
  cor(use = "pairwise.complete.obs") |>                                      # safer if missing values exist
  as.data.frame() |>
  tibble::rownames_to_column(var = "var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(correlation_tbl, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +                                                               # heatmap is the fastest visual collinearity check
  geom_text(aes(label = round(correlation, 2)), size = 3) +                  # print values so you can quote them directly if needed
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  labs(title = "Correlation heatmap", x = NULL, y = NULL, fill = "Corr") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**What to write if correlations are high:**
“Some predictors are strongly correlated, which may make coefficient estimates less stable in linear or logistic models. If necessary, a regularized model such as ridge or lasso could help.”

---

# 5. Baseline model — logistic regression

```{r logistic}
# Replace the formula with the predictors you decided to keep               # do not be afraid to start with a sensible smaller model
log_mod <- glm(
  formula = as.formula("[REPLACE_TARGET] ~ [REPLACE_PREDICTORS]"),          # example: target ~ age + income + zone + class
  data = train_data,
  family = binomial()
)

summary(log_mod)                                                             # inspect sign, size, and significance of coefficients
exp(coef(log_mod))                                                           # odds ratios are easier to explain than raw log-odds
```

**How to interpret one coefficient:**
If a coefficient is positive, the predictor increases the log-odds of the positive class. If its odds ratio is above 1, the predictor increases the odds. If the odds ratio is below 1, it decreases the odds.

**Written answer starter:**
“The coefficient for `[variable]` is `[estimate]`, which corresponds to an odds ratio of `[OR]`. This means that a one-unit increase in `[variable]` multiplies the odds of `[positive class]` by `[OR]`, holding the other variables fixed.”

---

# 6. Evaluate the logistic model

```{r logistic-eval}
prob_log <- predict(log_mod, newdata = test_data, type = "response")        # predicted probability of the positive class

threshold <- 0.50                                                            # start here; lower it if the positive class is rare
pred_log <- if_else(prob_log >= threshold, positive_class, negative_class)   # convert probabilities into class labels using explicit class names

pred_log <- factor(pred_log, levels = c(negative_class, positive_class))     # fix the class order so the confusion matrix stays readable
actual_y <- factor(test_data[[target_var]], levels = levels(pred_log))       # align levels before building the confusion matrix

cm_log <- table(Actual = actual_y, Predicted = pred_log)
cm_log
prop.table(cm_log, margin = 1)                                               # row-wise performance = class-wise accuracy
sum(diag(cm_log)) / sum(cm_log)                                              # overall accuracy
```

## Optional threshold scan

```{r threshold-scan}
threshold_tbl <- tibble(threshold = seq(0.05, 0.95, by = 0.05)) |>
  mutate(
    pred = map(threshold, ~ if_else(prob_log >= .x, positive_class, negative_class)),
    accuracy = map_dbl(pred, ~ mean(factor(.x, levels = levels(actual_y)) == actual_y))
  )

threshold_tbl
```

**When to lower the threshold:**
If the positive class is rare and you care more about catching positives than about avoiding false positives, a lower threshold can be justified.

---

# 7. Random forest comparison

```{r random-forest}
rf_formula <- as.formula("[REPLACE_TARGET] ~ [REPLACE_PREDICTORS]")         # define the formula once so you can inspect or reuse it easily
rf_predictor_count <- length(attr(terms(rf_formula), "term.labels"))        # count the predictors named in the formula, not raw data columns

rf_mod <- randomForest(
  formula = rf_formula,                                                      # use the same response and a reasonable predictor set
  data = train_data,
  ntree = 500,                                                               # many trees gives a more stable forest
  mtry = max(1, floor(sqrt(rf_predictor_count))),                            # use sqrt(p), where p is the number of predictors in the formula; if p = 1, RF offers little extra value
  importance = TRUE
)

rf_mod                                                                       # prints OOB error and class summary
pred_rf <- predict(rf_mod, newdata = test_data)
cm_rf <- table(Actual = actual_y, Predicted = pred_rf)
cm_rf
prop.table(cm_rf, margin = 1)
sum(diag(cm_rf)) / sum(cm_rf)

varImpPlot(rf_mod)                                                           # use this to discuss which variables matter most
```

**Written answer starter:**
“The random forest achieves `[accuracy]` accuracy on the test set. The variable importance plot suggests that `[variable]` is the most important predictor, followed by `[variable]`. Compared with logistic regression, the random forest performs `[better / similarly / worse]`.”

---

# 8. Boosting comparison

```{r boosting}
# gbm usually wants a numeric 0/1 response for bernoulli classification      # therefore we create a temporary training copy
train_boost <- train_data |>
  mutate(boost_target = if_else(.data[[target_var]] == positive_class, 1, 0))

test_boost <- test_data |>
  mutate(boost_target = if_else(.data[[target_var]] == positive_class, 1, 0))

boost_mod <- gbm(
  formula = as.formula("boost_target ~ [REPLACE_PREDICTORS]"),              # same predictor logic as the earlier models
  data = train_boost,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

prob_boost <- predict(boost_mod, newdata = test_boost, n.trees = 1000, type = "response")
pred_boost <- if_else(prob_boost >= threshold, 1, 0)                        # reuse the chosen threshold for fairer comparison

cm_boost <- table(Actual = test_boost$boost_target, Predicted = pred_boost)
cm_boost
prop.table(cm_boost, margin = 1)
sum(diag(cm_boost)) / sum(cm_boost)
```

**Written answer starter:**
“Boosting builds trees sequentially, where later trees focus more on observations that earlier trees handled poorly. In this analysis, the boosted model achieves `[accuracy]` accuracy and performs `[better / similarly / worse]` than the other models.”

---

# 9. Final comparison and business-style conclusion

```{r compare-models}
comparison_tbl <- tibble(
  model = c("Logistic regression", "Random forest", "Boosting"),
  accuracy = c(
    sum(diag(cm_log)) / sum(cm_log),
    sum(diag(cm_rf)) / sum(cm_rf),
    sum(diag(cm_boost)) / sum(cm_boost)
  )
)

comparison_tbl
```

**What your conclusion should answer:**
1. Which model performed best?
2. Which variables looked most important?
3. What does that mean in the practical context of the dataset?
4. What action or recommendation follows from the result?

**Written answer starter:**
“Based on the test-set comparison, `[best model]` performed best with an accuracy of `[value]`. The most important predictors were `[variables]`, which suggests that `[practical interpretation]`. Therefore, if the goal is to improve prediction / reduce churn / identify risk / target interventions, attention should first be directed toward `[actionable conclusion]`.”
