---
title: "BAN404 Exam — Task 2 Template"
subtitle: "[REPLACE: Dataset description]"
author: "Candidate [REPLACE: number]"
date: "[REPLACE: exam date]"
format:
  pdf:
    toc: false
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# INSTRUCTIONS FOR USE
# 1. Open this file at the START of the exam
# 2. Replace all [REPLACE: ...] markers with actual values
# 3. Delete this instruction block
# 4. Delete code sections you do not need
# 5. Add your written answers after each code chunk

---

# Setup

```{r setup}
# Load packages
library(MASS); library(ISLR2)
library(e1071); library(glmnet)
library(gam); library(tree)
library(randomForest); library(gbm)
library(splines)

# Set seed (use seed given in exam, or 123)
set.seed(123)

# Read data
data <- read.csv("[REPLACE: filename.csv]")
# OR: data <- read.csv2("[REPLACE: filename.csv]")  # for semicolon-separated

str(data)
head(data, 3)
dim(data)
```

---

# Task 2a — Data Cleaning and EDA (10p)

## Data Cleaning

```{r clean}
# STEP 1: Inspect unique values / types for all columns
summary(data)

# STEP 2: Convert categorical variables to factors
# [REPLACE: list the columns that need to be factors]
# data$col1 <- as.factor(data$col1)
# data$col2 <- as.factor(data$col2)

# STEP 3: Remove leakage / non-predictive variables
# [REPLACE: list variables to remove and explain why]
data <- data[, !(names(data) %in% c("[REPLACE: var1]", "[REPLACE: var2]"))]

# STEP 4: Create binary target (if needed)
# data$target <- as.numeric(data$raw_count >= 1)
# data$target <- factor(ifelse(data$raw >= 1, "yes", "no"))

# STEP 5: Train/test split (50/50)
n     <- nrow(data)
idx   <- sample(1:n, size=floor(n/2))
train <- data[idx, ]
test  <- data[-idx, ]

cat("Train size:", nrow(train), "| Test size:", nrow(test))
prop.table(table(train$target))  # Class balance check
```

**Written answer — data cleaning:**
I removed [variable X] because [it is the raw target / only available after the event / has no predictive value as a unique ID]. The variable [Y] was converted to a factor because it represents [a categorical group / a zone / a class]. The data was split 50/50 into training (n=[]) and test (n=[]) sets.

## EDA — Finding Useful Predictors

```{r eda-numeric}
# Continuous predictors vs binary target: boxplots
# Uncomment and adjust as needed:

par(mfrow=c(2,3))  # Adjust grid size for your number of plots

# boxplot(var1 ~ target, data=train, main="var1 by target")
# boxplot(var2 ~ target, data=train, main="var2 by target")
# boxplot(var3 ~ target, data=train, main="var3 by target")

# Means by group
# tapply(train$var1, train$target, mean)
```

```{r eda-categorical}
# Categorical predictors vs binary target: cross-tables
# prop.table(table(train$cat_var, train$target), margin=1)
```

**Written answer — EDA:**
From the boxplots, [variable A] shows a clear difference between groups (median [X] for target=0 vs [Y] for target=1). [Variable B] shows similar distributions in both groups. The cross-table for [variable C] shows that [category X] has [P]% positive cases vs [Q]% for [category Y], suggesting it is a useful predictor. I conclude that the most promising predictors are: [A, B, C].

---

# Task 2b — Logistic Regression (10p)

```{r logistic}
# Full model
logreg_full <- glm(target ~ ., data=train, family=binomial())
summary(logreg_full)

# Reduced model (based on EDA + significance)
# [REPLACE: list selected predictors]
# logreg_red <- glm(target ~ var1 + var2 + var3, data=train, family=binomial())

# Odds ratios
round(exp(coef(logreg_full)), 3)
```

**Written answer — logistic regression:**
The full logistic model shows that [variable A] is strongly significant (p < 0.001) with a coefficient of [X], corresponding to an odds ratio of [OR]. This means a one-unit increase in [A] multiplies the odds of [target] by [OR]. [Variable B] with coefficient [Y] (OR = [OR2]) is [significant/not significant]. Based on the summary, I selected the reduced model with [list variables] as these showed the strongest and most significant associations.

---

# Task 2c — Evaluate Predictions (10p)

```{r evaluate}
# Predict probabilities on test data
prob_full <- predict(logreg_full, newdata=test, type="response")

# Check class imbalance to decide threshold
prop.table(table(train$target))

# Choose threshold (0.5 default; lower if positive class is rare)
threshold <- 0.5   # [REPLACE: adjust if needed]
pred_class <- ifelse(prob_full > threshold, 1, 0)

# Confusion matrix
cm <- table(Actual=test$target, Predicted=pred_class)
cm
prop.table(cm, margin=1)  # Row proportions

# Accuracy
acc_logistic <- sum(diag(cm)) / sum(cm)
cat("Accuracy:", round(acc_logistic, 4))
```

**Written answer — evaluation:**
Using a threshold of [T], the logistic regression achieves an overall accuracy of [A]% on test data. The confusion matrix shows that [X]% of actual positives (sensitivity) and [Y]% of actual negatives (specificity) are correctly classified. The threshold of [T] was chosen [because the positive class represents only [Z]% of observations / to maximize accuracy]. [Add comparison if applicable].

---

# Task 2d — Random Forest (10p)

```{r random-forest}
library(randomForest)

# Ensure target is a factor
train$target <- as.factor(train$target)
test$target  <- as.factor(test$target)

rf <- randomForest(target ~ ., data=train,
                   mtry=floor(sqrt(ncol(train)-1)),
                   ntree=500,
                   importance=TRUE)
rf   # Print summary including OOB error

# Predict
pred_rf <- predict(rf, newdata=test)
cm_rf   <- table(Actual=test$target, Predicted=pred_rf)
cm_rf
prop.table(cm_rf, margin=1)
acc_rf <- sum(diag(cm_rf)) / sum(cm_rf)
cat("RF Accuracy:", round(acc_rf, 4))

# Variable importance
varImpPlot(rf)
```

**Written answer — random forest:**
The random forest model achieves [accuracy]% accuracy on test data, [better than / similar to / slightly worse than] logistic regression ([logistic accuracy]%). The variable importance plot shows [variable A] is the most important predictor, causing the largest drop in accuracy when permuted. This is consistent with the EDA finding that [A] had the largest group differences. The model [correctly classifies / struggles with] the positive class, achieving [sensitivity]% sensitivity.

---

# Task 2e — Boosted Tree (10p)

```{r boosting}
library(gbm)

# Target must be 0/1 numeric for bernoulli
train_bt        <- train
train_bt$target <- as.numeric(as.character(train$target))
test_bt         <- test

bt <- gbm(target ~ ., data=train_bt,
          distribution="bernoulli",
          n.trees=1000,
          interaction.depth=3,
          shrinkage=0.01)

# Predict
prob_bt  <- predict(bt, newdata=test_bt, n.trees=1000, type="response")
pred_bt  <- ifelse(prob_bt > threshold, 1, 0)
cm_bt    <- table(Actual=as.numeric(as.character(test$target)), Predicted=pred_bt)
cm_bt
prop.table(cm_bt, margin=1)
acc_bt <- sum(diag(cm_bt)) / sum(cm_bt)
cat("Boosting Accuracy:", round(acc_bt, 4))
```

**Written answer — boosting:**
The boosted tree model achieves [accuracy]% accuracy on test data. [Compare to logistic and RF.] Boosting builds trees sequentially, each correcting the errors of previous trees. With shrinkage=0.01, learning is slow but stable. [Conclude which model performs best overall and why.]

---

# Summary Comparison

```{r summary-table}
cat("Model Comparison (Test Accuracy):\n")
cat("Logistic Regression:", round(acc_logistic, 4), "\n")
cat("Random Forest:      ", round(acc_rf, 4), "\n")
cat("Boosted Tree:       ", round(acc_bt, 4), "\n")
```

**Final written answer:**
Across all models, [best model] achieves the highest accuracy of [X]%. The analysis suggests that [key predictors] are the most important drivers of [target]. [Business interpretation: what this means for the company/policy/decision]. [Optional: recommendation for which model to use in practice and why.]
