---
title: "BAN404 R Exam Compendium"
subtitle: "Comprehensive solution guide — 2023, 2024, and 2025 school exams"
author: "Self-study preparation file"
date: "2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
  error: true
---

# How to use this compendium

This file documents **every sub-question** from the 2023, 2024, and 2025 BAN404 school exams. For each sub-question you will find:

1.  **The question** — verbatim or closely paraphrased from the exam.
2.  **Theory connection** — which method or concept the question tests, and why.
3.  **Step-by-step code** — fully commented R code that answers the question.
4.  **Written answer** — model text that could appear in an exam submission.

The 2025 and 2024 exams are **highest priority** (same format expected). 2023 is included because it tests most of the same methods and has a different dataset angle.

------------------------------------------------------------------------

## Preparing and loading libraries and packages

Before starting any tasks, it is best practice to load all the libraries you might need for the exam. This prevents interruptions later. If a library is missing on your exam computer, you can install it using `install.packages("PackageName")`.

```{r setup_libraries}
# | eval: true
# | message: false
# | warning: false

# 1. Dataset Packages (Used in previous BAN404 exams)
library(ISLR)          # Contains OJ, Smarket datasets
library(ISLR2)         # Updated datasets for Statistical Learning
library(insuranceData) # Contains dataOhlsson, dataCar

# 2. Linear Model Selection and Regularization
library(glmnet)        # For Lasso and Ridge regression (matrix-based)

# 3. Non-linear Modeling
library(gam)           # For Generalized Additive Models (GAMs) and smoothing splines
library(splines)       # For basis functions and splines

# 4. Tree-Based Methods
library(tree)          # For basic Classification and Regression Trees
library(randomForest)  # For Random Forests and Bagging
library(gbm)           # For Gradient Boosting Machines

# 5. Advanced Classifiers (Just in case)
library(MASS)          # For LDA and QDA
library(e1071)         # For Support Vector Machines (SVM) and Naive Bayes

#6. tidyverse style
library(tidyverse)     # For read_csv2, dplyr, ggplot2, etc.
```

# 2025 School Exam — `data_task1.csv` + `insuranceData::dataOhlsson`

## Exam overview

-   **Date**: 2025-06-02, 6 hours
-   **Task 1** (50 p): methodological questions — ridge regression, LOOCV, bootstrap, GAM
-   **Task 2** (50 p): real-data classification with insurance claims data
-   **Allowed**: all written self-made material, calculator
-   **Submission**: PDF + R script or R Markdown

------------------------------------------------------------------------

## Task 1 — Methodological questions

### Task 1a (10p) — Explain the ridge regression functions

**Question:**\
You are given two R functions `f` and `g`. Explain what they do. Which method treated in the course is this a version of?

```{r}
f <- function(b1, X, y, la) {
  fval <- sum((y - X %*% b1)^2) + la * sum(b1^2)
  return(fval)
}

g <- function(X, y, la) {
  q <- ncol(X)
  b0 <- mean(y)
  yd <- y - b0
  Xd <- X
  for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])
  b1start <- rep(0, q)
  opt <- nlminb(b1start, f, X=Xd, y=yd, la=la)
  b1 <- opt$par
  return(b1)
}
```

**Theory connection:** This is **ridge regression** (L2-penalized regression). Ridge minimises RSS + λ·Σβ². When λ = 0 you get OLS. As λ increases, coefficients are shrunk toward zero. The functions also demean both y and X (centre around their means) before fitting — this handles the intercept analytically and is standard practice for ridge. This tests your understanding of **Ridge Regression** (L2-penalized regression). The core concept of Ridge is minimizing the Residual Sum of Squares (RSS) plus a shrinkage penalty $\lambda \sum \beta_j^2$. Because we do not want to penalize the intercept, it is standard practice to center (demean) the data first.

**How to recognize this method on the exam (The Detective Work):** When the exam gives you a custom optimization function, you must deduce what it does by looking at its mathematical components. Here is the logical checklist to arrive at the answer:

1.  **Is it a Tree-based method (Trees, RF, Boosting)?** No. There is no splitting logic, no Gini/Entropy, and no recursive partitioning. Instead, we see matrix multiplication (`X %*% b1`), which definitively means this is a **Linear Model**.
2.  **Is it Logistic Regression?** No. Logistic regression uses a logit link function (e.g., `exp(x)/(1+exp(x))`) and maximizes likelihood. Here, the first part of `f` is `sum((y - X %*% b1)^2)`, which is the **Residual Sum of Squares (RSS)**. RSS is used for continuous regression, not classification.
3.  **Is it standard Ordinary Least Squares (OLS)?** Close, but no. Standard OLS *only* minimizes the RSS. If the function was just `sum((y - X %*% b1)^2)`, it would be OLS. But this function adds a penalty term (`+ la *`), meaning it is a **Regularized (Shrinkage) method**.
4.  **Is it Lasso or Ridge?** Look exactly at the penalty term: `la * sum(b1^2)`.
    -   If it was **Lasso**, it would penalize the absolute value: `la * sum(abs(b1))`.
    -   Because it penalizes the **squared coefficients** (`b1^2`), it is the $L_2$ norm.
    -   **Conclusion:** A linear model minimizing RSS + squared coefficients is mathematically the exact definition of **Ridge Regression**.

**Step-by-step code :**

```{r task1a_functions}
#| eval: true

# f(b1, X, y, la) computes the Ridge objective function to be minimized
f <- function(b1, X, y, la) {
  # sum((y - X %*% b1)^2) is the Residual Sum of Squares (RSS)
  # la * sum(b1^2) is the L2 penalty (lambda * sum of squared coefficients)
  fval <- sum((y - X %*% b1)^2) + la * sum(b1^2)
  return(fval)
}

# g(X, y, la) prepares the data and runs the optimization to find the best coefficients
g <- function(X, y, la) {
  q <- ncol(X)                   # Number of predictors (p)
  
  # Intercept trick: By demeaning y, the intercept naturally becomes mean(y)
  b0 <- mean(y)                  
  yd <- y - b0                   # Demean (center) the response variable
  
  Xd <- X
  for(k in 1:q) {
    Xd[,k] <- X[,k] - mean(X[,k]) # Demean (center) each predictor column
  }
  
  b1start <- rep(0, q)           # Initialize optimization at zero for all coefficients
  
  # nlminb numerically minimizes the objective function 'f' 
  opt <- nlminb(b1start, f, X=Xd, y=yd, la=la)
  
  b1 <- opt$par                  # Extract the optimal coefficients
  return(b1)
}
```

**Written Answer:**\
These functions implement a version of **Ridge Regression** (also known as L2-regularized regression).

The first function, `f`, represents the **Ridge objective function**. It calculates the value that needs to be minimized: the Residual Sum of Squares `sum((y - X %*% b1)^2)` plus the L2 shrinkage penalty `la * sum(b1^2)`. The parameter `la` (lambda) controls the strength of the penalty; larger values force the regression coefficients (`b1`) closer to zero to reduce model variance.

The second function, `g`, is the **estimation engine**. Before estimating the coefficients, it explicitly **demeans (centers)** both the response variable `y` and every predictor column in `X`. This is a crucial step in Ridge regression because the intercept ($\beta_0$) should *not* be penalized. By centering the data, the intercept simply becomes the mean of `y` (`b0 <- mean(y)`), allowing the function to drop the intercept from the optimization process. Finally, it uses the `nlminb` numerical optimizer, starting from zero, to find the specific coefficient values (`b1`) that minimize the objective function `f`.

------------------------------------------------------------------------

### Task 1b (10p) — Compare OLS vs ridge, la=0 vs la=1000

**Question:**\
Consider the dataset in `data_task1.csv`. Extract `X` and `y`. Fit the model by ordinary least squares (OLS) and by using `g` with `la=1000`. Comment on the difference between the results. What is the explanation for the difference? Also, fit the model using `g` and let `la=0`.

**How to recognize this method on the exam (The Detective Work):** The exam is asking us to compare Ridge Regression (`g`) against standard OLS (`lm`). - **What happens when** $\lambda = 0$ ? The penalty term `la * sum(b1^2)` becomes zero. The objective function becomes exactly the OLS objective function (minimizing only RSS). Therefore, Ridge with `la=0` should yield identical coefficients to OLS. - **What happens when** $\lambda = 1000$? A massive penalty is applied to the size of the coefficients. The optimizer will shrink the coefficients toward zero to keep the penalty term small. - **Why do this?** The dataset has $n=100$ observations and $p=20$ predictors. A ratio of 5 observations per predictor is quite low, meaning OLS will likely have **high variance** (overfitting). Ridge deliberately introduces a little bit of bias (shrinking coefficients) to heavily reduce variance, leading to better predictions on unseen data.

**Step-by-step code:**

```{r task1b_comparison}
#| eval: true

# 1. Read data using the correct relative path 
# (read.csv2 handles European format: semicolon separators, comma decimals)
Xy <- read.csv2("2025/data_task1.csv")

# Extract the X matrix and y vector (as required by the g function)
X <- as.matrix(Xy[, -1])  # All columns except the 1st
y <- as.matrix(Xy[,  1])  # Only the 1st column

# 2. Fit OLS model
# We drop the intercept (position 1) because the 'g' function only returns slope coefficients
b1_ols <- lm(y ~ X)$coefficients[-1]

# 3. Fit Ridge with lambda = 0
b1_ridge_0 <- g(X, y, la = 0)

# 4. Fit Ridge with lambda = 1000
b1_ridge_1000 <- g(X, y, la = 1000)

# 5. Combine results into a clean table for comparison
comparison_table <- data.frame(
  OLS         = b1_ols,
  Ridge_la0   = b1_ridge_0,
  Ridge_la1000= b1_ridge_1000
)
# Print the first few rows to show the effect
head(round(comparison_table, 4))

# 6. Quantify the overall shrinkage by summing the squared coefficients
sum_sq_ols <- sum(b1_ols^2)
sum_sq_ridge1000 <- sum(b1_ridge_1000^2)

cat("\nSum of squared coefficients (OLS):", round(sum_sq_ols, 2))
cat("\nSum of squared coefficients (Ridge la=1000):", round(sum_sq_ridge1000, 2))
```

That table is a goldmine for an exam answer! You can see the shrinkage happening right before your eyes. Look at **X4** dropping from 61.55 to 39.41, or **X2** shrinking from -17.16 to -5.46. If you want, you can drop a line into your exam text saying: *"For example, the coefficient for X4 is shrunk from 61.55 in the OLS model down to 39.41 in the Ridge model."*

*(Notice how in the written text I used inline R code r round()so the text automatically updates with the real numbers! This is a massive flex on an exam).*

**AAA-Grade Written Answer:**\
When we estimate the model using the `g` function with $\lambda = 0$, the resulting coefficients are identical to the Ordinary Least Squares (OLS) coefficients obtained via the `lm()` function. This is mathematically expected: when the penalty parameter $\lambda$ is zero, the L2 penalty term disappears, and the Ridge objective function reduces to simply minimizing the Residual Sum of Squares (RSS), which is exactly what OLS does.

However, when we fit the model using `g` with $\lambda = 1000$, the estimated coefficients are substantially different. Specifically, their magnitudes are severely reduced. We can quantify this by calculating the sum of the squared coefficients ($\sum \beta^2$), which drops from approximately `r round(sum(b1_ols^2), 0)` in the OLS model to approximately `r round(sum(b1_ridge_1000^2), 0)` in the Ridge model.

**The explanation for this difference** lies in the bias-variance tradeoff. This dataset contains 20 predictors for only 100 observations. With this many predictors relative to the sample size, OLS is highly susceptible to high variance and overfitting (chasing the noise in the training data). By applying a strict penalty ($\lambda = 1000$), Ridge regression deliberately shrinks the coefficients toward zero. This introduces a small amount of bias into the estimates, but significantly reduces the variance, leading to more stable and robust coefficient estimates.

------------------------------------------------------------------------

### Task 1c (10p) — LOOCV to find optimal lambda

**Question:**\
Explain what the `loo` function does and use it to determine an optimal `la` for this dataset. Is it necessary to demean the x-variables? If so, why?

**How to recognize this method on the exam (The Detective Work):** 1. **Look at the loop:** `for(i in 1:n)`. It runs exactly $n$ times (once for every observation). 2. **Look at the training data:** `g(X=X[-i,], y=y[-i], la)`. The `[-i,]` syntax means it trains the model on all observations *except* the $i$-th one. 3. **Look at the prediction:** `ydpred[i] <- Xd[i,] %*% b1`. It predicts the value of the single left-out observation. 4. **Conclusion:** This is **Leave-One-Out Cross-Validation (LOOCV)**. It is used to estimate the Test MSE of the model. By testing many values of `la`, we can find the one that minimizes this Test MSE.

**Step-by-step code (Runnable Definition & Execution):**

```{r task1c_loocv}
#| eval: true

# --- PART 1: The LOOCV function provided in the exam ---
loo <- function(la, X, y) {
  n <- nrow(X)
  q <- ncol(X)
  b0 <- mean(y)
  yd <- y - b0
  
  Xd <- X
  for(k in 1:q) Xd[,k] <- X[,k] - mean(X[,k])
  
  ydpred <- matrix(0, n, 1)
  
  for(i in 1:n) {
    # Train the Ridge model on n-1 observations
    b1 <- g(X=X[-i,], y=y[-i], la)
    # Predict the left-out observation using the demeaned predictors
    ydpred[i] <- Xd[i,] %*% b1
  }
  
  # Calculate and return the Mean Squared Error (MSE)
  testMSE <- mean((yd - ydpred)^2)
  return(testMSE)
}

# --- PART 2: Our code to find the best Lambda ---

# "Grid search" is not a package. It just means: "create a sequence of numbers 
# and try them all to see which works best." We use base R functions for this.

# Step A: Create a broad sequence of candidate lambdas.
# seq() creates a list of 20 evenly spaced numbers between 0 and 10,000.
lavec_broad <- seq(0, 10000, length.out = 20)

# Step B: Test all of them.
# sapply() is a built-in R loop. It says: "For every 'la' in 'lavec_broad', 
# run the loo() function, and save the resulting MSE."
mse_broad <- sapply(lavec_broad, function(la) loo(la, X, y))

# Step C: Refine the search.
# If we plotted 'mse_broad', we'd see the lowest MSE is between 4000 and 6000.
# So, we create a new, narrower sequence of lambdas to get a more precise answer.
lavec_fine <- seq(4000, 6000, length.out = 20)
mse_fine <- sapply(lavec_fine, function(la) loo(la, X, y))

# Step D: Extract the exact lambda that gave the lowest MSE
optimal_la <- lavec_fine[which.min(mse_fine)]
min_mse <- min(mse_fine)

cat("Optimal lambda:", round(optimal_la, 2), "\n")
cat("Minimum LOOCV MSE:", round(min_mse, 2), "\n")

# Step E: Plot the U-shaped curve to prove we found the minimum
plot(lavec_fine, mse_fine, type = "b", col = "blue", lwd = 2,
     main = "LOOCV MSE vs Lambda (Refined)",
     xlab = "Lambda (Penalty Parameter)", 
     ylab = "Leave-One-Out Cross-Validation MSE")
abline(v = optimal_la, col = "red", lty = 2) # Mark the exact minimum with a red line
```

**AAA-Grade Written Answer:**\
The `loo` function implements **Leave-One-Out Cross-Validation (LOOCV)** to estimate the out-of-sample prediction error (Test MSE) for the Ridge regression model. It iterates $n$ times. In each iteration $i$, it holds out the $i$-th observation, trains the Ridge model (using the `g` function) on the remaining $n-1$ observations, and then uses the resulting coefficients to predict the held-out observation. Finally, it computes and returns the average squared prediction error across all $n$ iterations.

To find the optimal tuning parameter without using external packages, I performed a manual grid search. First, I generated a sequence of candidate $\lambda$ values using the `seq()` function. Then, I used `sapply()` to pass each candidate $\lambda$ into the `loo()` function to calculate its Test MSE. After a broad search, I refined the grid to search between 4000 and 6000. The LOOCV MSE forms a distinct U-shape, reaching its absolute minimum at an **optimal** $\lambda$ of 4736.84 (yielding an MSE of 1,637,998).

**Is it necessary to demean the x-variables?**\
Yes, it is absolutely necessary. In Ridge regression, the penalty $\lambda \sum \beta_j^2$ is applied to the magnitude of the coefficients. We only want to penalize the slope coefficients (the relationships between $X$ and $Y$); we **do not** want to penalize the intercept ($\beta_0$), as the base level of $Y$ should not be shrunk towards zero.

By demeaning both $y$ and the $X$-variables (centering them around zero), the intercept mathematically becomes the mean of the original $y$. This "intercept trick" allows the optimization algorithm `g` to completely ignore the intercept and solely focus on penalizing the slope coefficients `b1` fairly. If the variables were not demeaned, variables with larger absolute scales would arbitrarily skew the intercept, and the Ridge penalty would distort the predictions.

### Task 1d (10p) — Bootstrap variance and 95% CI

**Question:**\
Use the bootstrap method to estimate the sampling distribution of the sample variance of the variable y in `data_task1.csv`, $S^2 = \frac{1}{n-1} \sum_{i=1}^n (y_i - \bar{y})^2$. Show the result as a histogram. Also, use the bootstrap to compute a 95% confidence interval of the variance of y, $\sigma_y^2 = Var(y)$.

**How to recognize this method on the exam (The Detective Work):** - **The trigger word:** The question explicitly asks for the "bootstrap method" to estimate a "sampling distribution" and a "confidence interval." - **Why use the Bootstrap?** Traditional statistics require formulas (like standard errors) that rely on strict assumptions, usually that the data is perfectly normally distributed. The variance of a sample is particularly sensitive to non-normality. The Bootstrap bypasses this entirely: it treats our actual sample as the "population," and repeatedly draws "fake samples" from it *with replacement*. - **The Code "Tell":** Any time you see or need to write sample(dotdotdot, replace = TRUE) inside a `for` loop, you are doing a Bootstrap.

**Step-by-step code (Runnable Execution):**

```{r task1d_bootstrap}
#| eval: true

# 1. Setup the parameters
n <- length(y)               # Number of observations in our original sample
B <- 10000                   # Number of bootstrap replications (10,000 is standard for high accuracy)
vboot <- numeric(B)          # Create an empty vector of 10,000 zeros to store our results

# Always set a seed before generating random numbers for the exam so results are reproducible!
set.seed(123)

# 2. The Bootstrap Loop
for(b in 1:B) {
  # Step A: Resample WITH REPLACEMENT. 
  # This creates a "fake" dataset of the same size, where some original rows 
  # might appear 2 or 3 times, and others might be left out entirely.
  y_boot <- sample(y, size = n, replace = TRUE)  
  
  # Step B: Calculate the statistic of interest (the sample variance)
  vboot[b] <- var(y_boot)                          
}

# 3. Visualise the bootstrap distribution
# We use breaks=50 to make the histogram more detailed
hist(vboot, breaks = 50, col = "lightblue", border = "white",
     main = "Bootstrap Distribution of the Sample Variance of y",
     xlab = "Estimated Sample Variance",
     ylab = "Frequency")

# 4. Calculate the 95% Confidence Interval
# The most robust way is the "Percentile Method": we simply chop off the bottom 2.5% 
# and the top 2.5% of our 10,000 bootstrap estimates.
ci_95 <- quantile(vboot, probs = c(0.025, 0.975))

# Print the results nicely
cat("Original sample variance of y: ", round(var(y), 2), "\n")
cat("95% Bootstrap CI for Var(y):   [", round(ci_95[1], 2), ", ", round(ci_95[2], 2), "]\n")

# Add the Confidence Interval to the plot as red dashed lines
abline(v = ci_95[1], col = "red", lty = 2, lwd = 2)
abline(v = ci_95[2], col = "red", lty = 2, lwd = 2)
# Add the original sample variance as a black line
abline(v = var(y), col = "black", lwd = 2)
```

**Written Answer:**\
To estimate the sampling distribution of the sample variance of the variable $y$, I implemented the **Bootstrap resampling method**. Because theoretical confidence intervals for variance rely heavily on the assumption that the underlying data is normally distributed, the bootstrap offers a robust, non-parametric alternative.

The procedure works by treating our original 100 observations of $y$ as the population. I ran a loop $B = 10,000$ times. In each iteration, I generated a new bootstrap sample by drawing $n=100$ observations from $y$ **with replacement** (sample(dotdotdot, replace = TRUE)). I then calculated the sample variance for this bootstrap sample and stored it.

The resulting histogram visualizes the empirical sampling distribution of the variance. It is approximately bell-shaped and centered around the original sample variance (the solid black line).

To compute the **95% confidence interval**, I used the *percentile method*. By extracting the 2.5th and 97.5th percentiles of my 10,000 bootstrap variance estimates (`quantile(dotdotdot, probs = c(0.025, 0.975))`), I obtained a 95% confidence interval of **\[`r format(round(ci_95[1], 2), scientific=FALSE)`, `r format(round(ci_95[2], 2), scientific=FALSE)`\]** (represented by the red dashed lines on the histogram). This means we are 95% confident that the true population variance of $y$ lies within this range, without having to make strict assumptions about the underlying distribution of the data.

------------------------------------------------------------------------

### Task 1e (10p) — GAM for nonlinearity + R²

**Question:**\
There might be one or more variables that have non-linear relationships to y. Try to identify such variable(s). Can you improve the in-sample (training) MSE by using a generalized additive model (GAM)? Compare the training MSE with at least one GAM. Also, use these MSE’s to compute the share of explained variation, R², for the linear model estimated by OLS and for the GAM-model.

**How to recognize this method on the exam (The Detective Work):** 1. **"Identify non-linear relationships":** We plot the response $y$ against the predictors. If the dots form a U-shape or wave, it's non-linear. 2. **"Generalized Additive Model (GAM)":** A GAM extends linear regression. Instead of a straight line, it allows a flexible curve using the `s()` function (smoothing spline) from the `gam` package. 3. **"Residuals":** A residual is simply the error ($Actual - Predicted$). Mean Squared Error (MSE) is just the average of the squared residuals. 4. **"Compute R² manually":** The formula for $R^2$ is $1 - \frac{MSE_{model}}{MSE_{null}}$. The $MSE_{null}$ is simply the variance of $y$, calculated as `mean((y - mean(y))^2)`.

**Step-by-step code (Runnable Execution):**

```{r task1e_gam}
#| eval: true

library(gam)

# 1. Identify non-linear relationships visually
# We plot y against ALL 20 predictors to hunt for non-linearity.
# mfrow = c(4, 5) creates a 4-row, 5-column grid.
# mar = c(2, 2, 1, 1) shrinks the margins so the 20 plots fit cleanly on one page.
par(mfrow = c(4, 5), mar = c(2, 2, 1, 1)) 

for(j in 2:21) {   # Columns 2 through 21 are X1 to X20
  # Plot each X variable against y (Column 1)
  plot(Xy[, j], Xy[, 1], 
       main = names(Xy)[j],   # Title is the variable name
       pch = 20, col = "darkgray", cex = 0.5) # cex=0.5 makes dots smaller
}
# OBSERVATION: Scanning the grid, X3 clearly stands out with a wavy/U-shaped 
# non-linear relationship. The others look like random clouds or straight lines.

# 2. Fit the baseline OLS model (assumes all variables are straight lines)
ols_model <- lm(y ~ ., data = Xy)
mse_ols <- mean(residuals(ols_model)^2)

# 3. Fit the GAM model
# Shortcut formula: 'y ~ . - X3 + s(X3)'
# Meaning: "Predict y using all variables (.), remove linear X3, add smoothed s(X3)"
gam_model <- gam(y ~ . - X3 + s(X3), data = Xy)
mse_gam <- mean(residuals(gam_model)^2)

# 4. Calculate the share of explained variation (R-squared)
# Calculate the "Total MSE" (variance of y dividing by n, not n-1)
mse_all <- mean((Xy$y - mean(Xy$y))^2)

r2_ols <- 1 - (mse_ols / mse_all)
r2_gam <- 1 - (mse_gam / mse_all)

# Reset plot window and plot the smooth effect of X3 from the GAM
par(mfrow = c(1, 1))
plot(gam_model, se = TRUE, col = "blue", main = "GAM: Smooth Partial Effect of X3")

# Print the final matched results
cat("In-sample MSE (OLS):", round(mse_ols, 2), "\n")
cat("In-sample MSE (GAM):", round(mse_gam, 2), "\n\n")

cat("R-squared (OLS):", round(r2_ols, 4), "\n")
cat("R-squared (GAM):", round(r2_gam, 4), "\n")
```

**AAA-Grade Written Answer:**\
To identify non-linear relationships, I plotted the response variable $y$ against the predictors. Visually inspecting the scatterplots reveals that predictor `X3` clearly exhibits a non-linear, wavy relationship with $y$, whereas other variables (like `X1`) do not.

To capture this non-linearity, I fitted a **Generalized Additive Model (GAM)** using the `gam` package. In this model, all predictors entered linearly except for `X3`, which was modeled using a smoothing spline (`s(X3)`). A smoothing spline allows the model to flexibly follow the natural curve of the data rather than forcing a straight line. I compared this GAM to a standard Ordinary Least Squares (OLS) model where all variables are forced to be strictly linear.

The inclusion of the smoothing spline substantially improved the model's fit on the training data. The **in-sample MSE dropped** from `r format(round(mse_ols, 0), scientific=FALSE)` (OLS) down to `r format(round(mse_gam, 0), scientific=FALSE)` (GAM).

We can translate this improvement into the share of explained variation ($R^2$). The standard linear OLS model explains approximately **`r round(r2_ols * 100, 1)`%** of the variation in $y$. By simply allowing `X3` to be flexible via the GAM, the share of explained variation jumps to **`r round(r2_gam * 100, 1)`%**. This proves that the strict linearity assumption of OLS was inadequate for `X3`, and modelling the non-linearity via a smoothing spline successfully captures systematic patterns that OLS missed.

## Task 2 — Insurance claims (`dataOhlsson`)

### Task 2a (10p) — Create claim variable, remove leakage, EDA

**Question:**\
Create a binary variable `claim` (antskad ≥ 1 = TRUE). Remove variables not available at prediction time. Split into training/test. Use descriptive statistics to find useful predictors.

**Theory connection:** Before fitting any model, we must remove *target leakage* — variables that contain direct information about the outcome that would not be available in a real prediction scenario. `antskad` (number of claims) determines `claim` directly; `skadkost` (claim cost) is only known after a claim occurs. Both must be removed.\
Encode categorical variables as factors so that R handles them correctly in `glm`, `tree`, etc.

**Step-by-step code:**

```{r}
library(insuranceData)
data("dataOhlsson")
ydata <- dataOhlsson

# Create binary target: claim = TRUE if antskad >= 1
ydata$claim <- ydata$antskad >= 1

# Remove variables that are NOT available at prediction time:
# antskad = number of claims  -> IS the (raw) target, not a predictor
# skadkost = claim cost       -> only known AFTER a claim has occurred
ydata <- ydata[, !(names(ydata) %in% c("antskad", "skadkost"))]

# Encode ordinal/categorical variables as factors
ydata$zon     <- as.factor(ydata$zon)     # geographic zone
ydata$bonuskl <- as.factor(ydata$bonuskl) # bonus class
ydata$mcklass <- as.factor(ydata$mcklass) # motorcycle class

# 50/50 train/test split
set.seed(123)
n     <- nrow(ydata)
ind   <- sample(1:n, size = floor(n/2))
train <- ydata[ind,]
test  <- ydata[-ind,]

# --- Descriptive statistics: numeric predictors ---
# Boxplot of numeric vars by claim status
plot(duration ~ as.factor(claim), data = train, main = "Duration by claim")
plot(agarald  ~ as.factor(claim), data = train, main = "Owner age by claim")
plot(fordald  ~ as.factor(claim), data = train, main = "Vehicle age by claim")

# --- Descriptive statistics: categorical predictors ---
# Proportion of claim=TRUE within each category
prop.table(table(train$kon,     train$claim), margin = 1)
prop.table(table(train$bonuskl, train$claim), margin = 1)
prop.table(table(train$zon,     train$claim), margin = 1)
prop.table(table(train$mcklass, train$claim), margin = 1)
```

**Written answer:**\
`antskad` and `skadkost` are removed because they are either the outcome itself or only observable after the claim. Categorical variables `zon`, `bonuskl`, and `mcklass` are encoded as factors. Descriptive analysis shows that: (1) younger owner age (`agarald`) is associated with more claims; (2) newer vehicles (`fordald`) make more claims; (3) male policy-holders (`kon = M`) claim more often; (4) zone 1 has the highest claim rate; (5) bonus class 7 has the highest claim rate. `duration` shows little visible difference between claim groups.

------------------------------------------------------------------------

### Task 2b (10p) — Logistic regression: full and reduced models

**Question:**\
Fit a logistic regression with all variables and one with a selection of variables. Base the selection on task 2a and on the significance of coefficients in the full model.

**Theory connection:** Logistic regression models P(claim=1 \| X) = sigmoid(β₀ + β'X). We use `family = binomial()` in `glm`. A *reduced model* that drops non-significant predictors may improve parsimony and generalisation.

**Step-by-step code:**

```{r}
# Full logistic model with all available predictors
logreg1 <- glm(claim ~ ., data = train, family = binomial())
summary(logreg1)
# Key significant predictors: agarald (***), konM (***), zon (***),
#                             fordald (***), duration (***), bonuskl7 (*)
# Non-significant: most mcklass levels, bonuskl2-6

# Reduced model: keep the most important predictors based on significance
logreg2 <- glm(claim ~ agarald + kon + zon + fordald + duration,
               data = train, family = binomial())
summary(logreg2)
```

**Written answer:**\
The full logistic model shows that `agarald` (owner age), `kon` (gender), most `zon` levels, `fordald` (vehicle age), and `duration` are strongly significant (p \< 0.001). Most `mcklass` and `bonuskl` levels are not significant. A reduced model using only the significant continuous and categorical predictors retains near-equivalent fit while being more parsimonious.

------------------------------------------------------------------------

### Task 2c (10p) — Evaluate predictions: threshold + confusion matrix

**Question:**\
Evaluate the model predictions from 2b with an appropriate measure. Determine a threshold to convert probabilities to predicted classes.

**Theory connection:** With rare events (claim rate ≈ 1%), the default 0.5 threshold classifies almost everything as "no claim." A lower threshold catches more true claims (higher sensitivity) at the cost of more false positives. We can experiment on the training set to find a pragmatic threshold.

**Step-by-step code:**

```{r}
# --- Determine threshold by experimenting on TRAINING data ---
prob_train <- predict(logreg1, newdata = train, type = "response")

# Try threshold = 0.01 (very low, because claim rate is ~1%)
th <- 0.01
pred_train <- prob_train > th
prop.table(table(train$claim, pred_train), margin = 1)
# Look at row 2 (claim=TRUE): we want this to be high (~sensitivity)

# --- Evaluate on TEST data ---
prob_test <- predict(logreg1, newdata = test, type = "response")
pred_test <- prob_test > th

# Confusion matrix (rows = actual, columns = predicted)
cm1 <- table(test$claim, pred_test)
cm1

# Row-wise proportions (how well do we classify each class?)
prop.table(cm1, margin = 1)
# Example output:
#        FALSE  TRUE
# FALSE  0.686  0.314   <- 68.6% of no-claim correctly predicted
# TRUE   0.283  0.717   <- 71.7% of claims correctly predicted

# Overall accuracy
sum(diag(cm1)) / sum(cm1)
```

**Written answer:**\
Because claims are rare (\~1% of observations), a 0.5 threshold classifies almost everything as "no claim." Experimenting on training data with a threshold of 0.01 gives a more balanced classifier: approximately 71.7% of actual claims are correctly identified (sensitivity) and 68.6% of non-claims are correctly identified (specificity). A good model answer acknowledges this class imbalance and justifies the threshold choice.

------------------------------------------------------------------------

### Task 2d (10p) — Random forest

**Question:**\
Fit a random forest to predict `claim` and evaluate predictions. Compare with the logistic model.

**Theory connection:** Random forests grow many decision trees on bootstrap samples of the training data, each using a random subset of `mtry` predictors at each split. The prediction is the majority vote (classification) or average (regression). They typically outperform single logistic models on complex patterns and class-imbalanced data.

**Step-by-step code:**

```{r, eval=FALSE}
library(randomForest)

# Fit random forest; convert claim to factor for classification
rf <- randomForest(as.factor(claim) ~ ., data = train, mtry = 7, ntree = 500)

# Predict probabilities and convert to class using a threshold
prob_rf <- predict(rf, newdata = test, type = "prob")  # columns: [FALSE, TRUE]
pred_rf  <- prob_rf[, 2] > 0.02  # low threshold: same reasoning as logistic

# Confusion matrix
cm_rf <- table(test$claim, pred_rf)
prop.table(cm_rf, margin = 1)
# Typically: FALSE ~83.5% correct, TRUE ~43.2% correct
# Better at no-claim class, worse at claim class vs logistic
```

**Written answer:**\
The random forest improves accuracy for the no-claim class compared to logistic regression but at the cost of somewhat lower sensitivity for the claim class. The optimal threshold may differ from the logistic model. The random forest has no single coefficient to interpret, but variable importance can be examined with `varImpPlot(rf)`.

------------------------------------------------------------------------

### Task 2e (10p) — Boosted tree

**Question:**\
Fit a boosted tree and evaluate predictions. Compare with logistic regression and random forest.

**Theory connection:** Gradient boosting (`gbm`) builds trees sequentially, each one fitting the residuals (errors) of all previous trees. Key hyperparameters: `n.trees` (number of iterations), `interaction.depth` (tree depth), `shrinkage` (learning rate — smaller = more trees needed but better generalisation).

**Step-by-step code:**

```{r}
library(gbm)

# For binary outcome use distribution = "bernoulli"
# Convert logical claim to numeric 0/1 for gbm
train_bt <- train
train_bt$claim <- as.numeric(train$claim)  # TRUE -> 1, FALSE -> 0

bt <- gbm(claim ~ ., data = train_bt,
          distribution = "bernoulli",  # binary outcome
          n.trees = 1000,
          interaction.depth = 3,       # max depth of each tree
          shrinkage = 0.01)            # learning rate

# Predict: type="response" gives predicted probabilities
prob_bt <- predict(bt, newdata = test, n.trees = 1000, type = "response")
pred_bt  <- prob_bt > 0.01  # same threshold approach

cm_bt <- table(test$claim, pred_bt)
prop.table(cm_bt, margin = 1)
# Typically similar to logistic regression performance
```

**Written answer:**\
The boosted tree yields performance broadly similar to logistic regression, with approximately 70% of claims and 67% of non-claims correctly classified at a threshold of 0.01. It does not clearly outperform logistic regression here, which may reflect the difficulty of predicting rare claims from the available predictors. The random forest is the strongest performer among the three methods on the no-claim class, while logistic regression and boosting show higher sensitivity.
