---
title: "BAN404 Master Snippet Library 2026"
subtitle: "Task 1 + Task 2 exam-ready blocks from ExamPrep sources"
author: "Exam preparation"
date: "2026"
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
  error: true
---

# Task 1 snippets (from ExamPrep.pdf)

## task1_1_setup_import

Use when the exam says load the Task 1 data, inspect the structure, and prepare a safe starting object before EDA or modeling.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR` if the response is not `y`.
- Add/remove package calls only if the exam task needs them.

```{r}
#| label: task1_1_setup_import
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

glimpse(df_task1)
summary(df_task1)
```

Interpretation template: The Task 1 dataset contains `[N_ROWS]` rows and `[N_COLS]` columns. The response variable is `[TARGET_VAR]`, and the structure check shows `[ANY_IMMEDIATE_RECODING_OR_NOT]`.

## task1_2_missing_values

Use when the exam asks for a data integrity check or whether imputation is needed before regression.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- If the response name differs, update `TARGET_VAR` only if you reference it in your text.

```{r}
#| label: task1_2_missing_values
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

na_counts <- colSums(is.na(df_task1))
na_counts

tibble(variable = names(na_counts), n_missing = as.integer(na_counts)) %>%
  arrange(desc(n_missing))
```

Interpretation template: Missing-value counts show `[VARIABLE]` has `[N_MISSING]` missing values. Therefore, we `[DO / DO NOT]` need imputation before fitting the models.

## task1_3_target_distribution

Use when the exam asks whether the continuous response is skewed, heavy-tailed, or contains outliers.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Change `bins = 40` if the exam requests another granularity.

```{r}
#| label: task1_3_target_distribution
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

ggplot(df_task1, aes(x = .data[[TARGET_VAR]])) +
  geom_histogram(bins = 40, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Target distribution", x = TARGET_VAR, y = "Count")

df_task1 %>%
  summarise(
    min = min(.data[[TARGET_VAR]], na.rm = TRUE),
    q1 = quantile(.data[[TARGET_VAR]], 0.25, na.rm = TRUE),
    median = median(.data[[TARGET_VAR]], na.rm = TRUE),
    mean = mean(.data[[TARGET_VAR]], na.rm = TRUE),
    q3 = quantile(.data[[TARGET_VAR]], 0.75, na.rm = TRUE),
    max = max(.data[[TARGET_VAR]], na.rm = TRUE)
  )
```

Interpretation template: The histogram suggests `[APPROX_SYMMETRY / RIGHT_SKEW / LEFT_SKEW]` with `[OUTLIER_COMMENT]`. This means OLS residual assumptions may be `[REASONABLE / QUESTIONABLE]`, so we `[KEEP_SCALE / CONSIDER_TRANSFORMATION]`.

## task1_4_scale_and_collinearity

Use when the exam asks about feature scaling, multicollinearity, or whether Ridge is likely to help.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Change `CORR_THRESHOLD` if the exam gives a cutoff.

```{r}
#| label: task1_4_scale_and_collinearity
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
CORR_THRESHOLD <- 0.8

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

summary(df_task1)

num_df <- df_task1 %>% select(where(is.numeric))
cor_matrix <- num_df %>% select(-all_of(TARGET_VAR)) %>% cor()
round(cor_matrix, 2)

as_tibble(as.table(cor_matrix), .name_repair = "minimal") %>%
  rename(var1 = Var1, var2 = Var2, corr = n) %>%
  filter(var1 < var2, abs(corr) >= CORR_THRESHOLD)
```

Interpretation template: Predictor scales are `[SIMILAR / VERY_DIFFERENT]`, and the number of highly correlated pairs is `[N_PAIRS]`. Therefore, `[RIDGE_IS_LIKELY_HELPFUL / RIDGE_IS_UNLIKELY_TO_ADD_MUCH]`.

## task1_5_linearity_check

Use when the exam asks whether linear models are appropriate or whether more flexible methods are needed.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Change `PREDICTOR_PREFIX` if predictors do not start with the same prefix.

```{r}
#| label: task1_5_linearity_check
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
PREDICTOR_PREFIX <- "x"

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

df_task1 %>%
  pivot_longer(
    cols = starts_with(PREDICTOR_PREFIX),
    names_to = "predictor",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = .data[[TARGET_VAR]])) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +   # strict linear fit
  geom_smooth(method = "loess", se = FALSE, color = "red") + # flexible fit for comparison
  facet_wrap(~ predictor, scales = "free_x") +
  theme_minimal()
```

Interpretation template: For `[PREDICTOR_NAMES]`, the LOESS fit is `[CLOSE_TO / CLEARLY_DIFFERENT_FROM]` the straight-line fit. This suggests `[LINEAR_MODELS_ARE_FINE / NONLINEAR_METHODS_ARE_NEEDED]`.

## task1_6_eda_findings_summary

Use when the exam asks you to summarise the EDA and explain how it changes model choice.

**How to adapt:**
- Change `NONLINEAR_VARS`, `NOISE_VARS`, and `LINEAR_VARS` after inspecting your plots.
- Remove the rows you do not need.

```{r}
#| label: task1_6_eda_findings_summary
suppressPackageStartupMessages(library(tidyverse))

NONLINEAR_VARS <- c("X2", "X3")
NOISE_VARS <- c("X4", "X6")
LINEAR_VARS <- c("X1", "X5")

tribble(
  ~finding, ~variables, ~model_implication,
  "linear signal", paste(LINEAR_VARS, collapse = ", "), "OLS can model these adequately",
  "non-linear signal", paste(NONLINEAR_VARS, collapse = ", "), "Prefer GAM / trees / manual polynomial fixes",
  "weak or noisy signal", paste(NOISE_VARS, collapse = ", "), "Candidate variables to drop or shrink"
)
```

Interpretation template: EDA suggests `[LINEAR_VARS]` are roughly linear, `[NONLINEAR_VARS]` are non-linear, and `[NOISE_VARS]` look weak. Therefore, a rigid linear baseline is useful, but a more flexible model is likely to perform better.

## task1_7_train_test_split

Use when the exam asks for a holdout split and preparation for glmnet models.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Change `TRAIN_FRAC` and `SEED_VALUE` if specified.

```{r}
#| label: task1_7_train_test_split
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))

train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

x_train <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)[, -1]
y_train <- train_df[[TARGET_VAR]]
x_test  <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = test_df)[, -1]
y_test  <- test_df[[TARGET_VAR]]

dim(train_df)
dim(test_df)
```

Interpretation template: The data was split into `[N_TRAIN]` training and `[N_TEST]` test observations. This gives an out-of-sample benchmark for comparing model flexibility and generalisation.

## task1_8_ols_baseline

Use when the exam asks for a baseline OLS model and a short discussion of its limitations.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Replace `~ .` with an explicit formula if the exam requires selected predictors only.

```{r}
#| label: task1_8_ols_baseline
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

ols_fit <- lm(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)
ols_pred <- predict(ols_fit, newdata = test_df)
ols_mse <- mean((test_df[[TARGET_VAR]] - ols_pred)^2)

summary(ols_fit)
ols_mse
```

Interpretation template: The OLS test MSE is `[OLS_MSE]`. The model is a useful baseline, but it will underfit if the true relationship is non-linear.

## task1_9_polynomial_regression

Use when the exam says fix a specific non-linear predictor without using GAM or trees.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR`.
- Change `NONLINEAR_VAR`, `POLY_DEGREE`, and `CONTROL_VARS`.

```{r}
#| label: task1_9_polynomial_regression
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
NONLINEAR_VAR <- "X_NONLINEAR"
POLY_DEGREE <- 2
CONTROL_VARS <- c("X_LINEAR_1", "X_LINEAR_2")
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

rhs_terms <- c(CONTROL_VARS, paste0("I(", NONLINEAR_VAR, "^", POLY_DEGREE, ")"))
poly_formula <- as.formula(paste(TARGET_VAR, "~", paste(rhs_terms, collapse = " + ")))

poly_fit <- lm(poly_formula, data = train_df)
poly_pred <- predict(poly_fit, newdata = test_df)
poly_mse <- mean((test_df[[TARGET_VAR]] - poly_pred)^2)

summary(poly_fit)
poly_mse
```

Interpretation template: The polynomial model changes the test MSE to `[POLY_MSE]`. If the squared or cubic term is significant and the error falls, that supports a non-linear effect in `[NONLINEAR_VAR]`.

## task1_10_lasso_cv

Use when the exam asks for regularisation, automatic variable selection, or a cross-validated linear benchmark.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Change `CV_FOLDS` if required.
- Use `lambda.1se` instead of `lambda.min` if the exam asks for the simpler CV choice.

```{r}
#| label: task1_10_lasso_cv
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glmnet)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123
CV_FOLDS <- 10

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

x_train <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)[, -1]
y_train <- train_df[[TARGET_VAR]]
x_test  <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = test_df)[, -1]
y_test  <- test_df[[TARGET_VAR]]

set.seed(SEED_VALUE)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = CV_FOLDS)
lasso_fit <- glmnet(x_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min)
lasso_pred <- as.numeric(predict(lasso_fit, newx = x_test))
lasso_mse <- mean((y_test - lasso_pred)^2)

coef(lasso_fit)
lasso_mse
```

Interpretation template: The LASSO test MSE is `[LASSO_MSE]`. Coefficients shown as zero were removed, so LASSO `[DID / DID_NOT]` perform meaningful variable selection.

## task1_11_gam

Use when the exam asks for a flexible but interpretable model for smooth non-linear effects.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Change `LINEAR_TERMS` and `SMOOTH_TERMS`.
- Remove noise variables manually if justified by EDA.

```{r}
#| label: task1_11_gam
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(gam)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
LINEAR_TERMS <- c("X_LINEAR_1", "X_LINEAR_2")
SMOOTH_TERMS <- c("X_SMOOTH_1", "X_SMOOTH_2")
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

rhs_terms <- c(LINEAR_TERMS, paste0("s(", SMOOTH_TERMS, ")"))
gam_formula <- as.formula(paste(TARGET_VAR, "~", paste(rhs_terms, collapse = " + ")))

gam_fit <- gam(gam_formula, data = train_df)
gam_pred <- predict(gam_fit, newdata = test_df)
gam_mse <- mean((test_df[[TARGET_VAR]] - gam_pred)^2)

gam_mse
plot(gam_fit, se = TRUE, col = "blue")
```

Interpretation template: The GAM test MSE is `[GAM_MSE]`. The smooth plots show whether `[SMOOTH_TERMS]` have curved effects, which is why GAM can outperform rigid linear models.

## task1_12_random_forest

Use when the exam asks for a tree-based ensemble and variable importance.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Change `NTREE_VALUE` or `MTRY_VALUE` if specified.

```{r}
#| label: task1_12_random_forest
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(randomForest)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123
NTREE_VALUE <- 500
MTRY_VALUE <- NULL

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

rf_args <- list(
  formula = as.formula(paste(TARGET_VAR, "~ .")),
  data = train_df,
  ntree = NTREE_VALUE,
  importance = TRUE
)
if(!is.null(MTRY_VALUE)) rf_args$mtry <- MTRY_VALUE
rf_fit <- do.call(randomForest, rf_args)

rf_pred <- predict(rf_fit, newdata = test_df)
rf_mse <- mean((test_df[[TARGET_VAR]] - rf_pred)^2)

rf_mse
importance(rf_fit)
varImpPlot(rf_fit)
```

Interpretation template: The Random Forest test MSE is `[RF_MSE]`. The importance plot indicates `[TOP_VARS]` are most useful for prediction, while `[LOW_IMPORTANCE_VARS]` contribute little.

## task1_13_boosting_gbm

Use when the exam asks for gradient boosting and wants you to explain how it differs from Random Forest.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Change `N_TREES`, `INTERACTION_DEPTH`, and `SHRINKAGE` if requested.

```{r}
#| label: task1_13_boosting_gbm
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(gbm)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123
N_TREES <- 1000
INTERACTION_DEPTH <- 3
SHRINKAGE <- 0.01

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>%
  clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

boost_fit <- gbm(
  as.formula(paste(TARGET_VAR, "~ .")),
  data = train_df,
  distribution = "gaussian",
  n.trees = N_TREES,
  interaction.depth = INTERACTION_DEPTH,
  shrinkage = SHRINKAGE,
  verbose = FALSE
)

boost_pred <- predict(boost_fit, newdata = test_df, n.trees = N_TREES)
boost_mse <- mean((test_df[[TARGET_VAR]] - boost_pred)^2)

boost_mse
summary(boost_fit)
```

Interpretation template: The boosting test MSE is `[BOOST_MSE]`. If it beats OLS/LASSO, that suggests sequential bias reduction helped capture non-linearity.

## task1_14_model_leaderboard_mse

Use when the exam asks which model performed best on test MSE.

**How to adapt:**
- Change all placeholder variable names at the top.
- Update the GAM terms and polynomial variable to match your EDA.
- Remove any models you did not fit.

```{r}
#| label: task1_14_model_leaderboard_mse
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glmnet)
  library(gam)
  library(randomForest)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
NONLINEAR_VAR <- "X_NONLINEAR"
CONTROL_VARS <- c("X_LINEAR_1", "X_LINEAR_2")
LINEAR_TERMS <- c("X_LINEAR_1", "X_LINEAR_2")
SMOOTH_TERMS <- c("X_SMOOTH_1", "X_SMOOTH_2")
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]
test_df  <- df_task1[-idx, ]

x_train <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)[, -1]
y_train <- train_df[[TARGET_VAR]]
x_test  <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = test_df)[, -1]
y_test  <- test_df[[TARGET_VAR]]

ols_fit <- lm(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)
mse_ols <- mean((y_test - predict(ols_fit, newdata = test_df))^2)

poly_formula <- as.formula(
  paste(TARGET_VAR, "~", paste(c(CONTROL_VARS, paste0("I(", NONLINEAR_VAR, "^2)")), collapse = " + "))
)
poly_fit <- lm(poly_formula, data = train_df)
mse_poly <- mean((y_test - predict(poly_fit, newdata = test_df))^2)

cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
lasso_fit <- glmnet(x_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min)
mse_lasso <- mean((y_test - as.numeric(predict(lasso_fit, newx = x_test)))^2)

gam_formula <- as.formula(paste(TARGET_VAR, "~", paste(c(LINEAR_TERMS, paste0("s(", SMOOTH_TERMS, ")")), collapse = " + ")))
gam_fit <- gam(gam_formula, data = train_df)
mse_gam <- mean((y_test - predict(gam_fit, newdata = test_df))^2)

rf_fit <- randomForest(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)
mse_rf <- mean((y_test - predict(rf_fit, newdata = test_df))^2)

tibble(
  model = c("OLS", "Polynomial", "LASSO", "GAM", "Random Forest"),
  test_mse = c(mse_ols, mse_poly, mse_lasso, mse_gam, mse_rf)
) %>%
  arrange(test_mse)
```

Interpretation template: The best model by test MSE is `[BEST_MODEL]` with error `[BEST_MSE]`. This supports the claim that `[LINEARITY / FLEXIBILITY / REGULARISATION]` mattered most in this dataset.

## task1_15_rmse_r2_table

Use when the exam asks for RMSE and test R-squared, not only MSE.

**How to adapt:**
- Replace the model metric placeholders in the final tibble with your fitted values if you already calculated them elsewhere.
- If you only fitted some models, keep only those rows.

```{r}
#| label: task1_15_rmse_r2_table
suppressPackageStartupMessages(library(tidyverse))

MODEL_NAMES <- c("OLS", "LASSO", "GAM", "Random Forest")
MODEL_PREDICTIONS <- list(PRED_OLS_HERE, PRED_LASSO_HERE, PRED_GAM_HERE, PRED_RF_HERE)
Y_TEST <- Y_TEST_HERE

tss_test <- sum((Y_TEST - mean(Y_TEST))^2)

metric_tbl <- tibble(
  model = MODEL_NAMES,
  pred = MODEL_PREDICTIONS
) %>%
  mutate(
    mse = map_dbl(pred, ~ mean((Y_TEST - .x)^2)),
    rmse = sqrt(mse),
    sse = map_dbl(pred, ~ sum((Y_TEST - .x)^2)),
    test_r2 = 1 - (sse / tss_test)
  ) %>%
  select(-pred, -sse) %>%
  arrange(desc(test_r2))

metric_tbl
```

Interpretation template: `[BEST_MODEL]` has the highest test R-squared and the lowest / near-lowest RMSE. In exam language, it explains the most variance out-of-sample while keeping average prediction error lowest.

## task1_16_bootstrap_ci

Use when the exam asks for a bootstrap standard error or normal-vs-percentile confidence intervals.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `BOOT_FORMULA` and `COEF_NAME`.
- Change `B_REPS` or confidence level if specified.

```{r}
#| label: task1_16_bootstrap_ci
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
BOOT_FORMULA <- Y_VAR ~ X_VAR
COEF_NAME <- "X_VAR"
B_REPS <- 1000
CONF_LEVEL <- 0.95
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()
alpha <- 1 - CONF_LEVEL

set.seed(SEED_VALUE)
boot_vals <- replicate(B_REPS, {
  idx <- sample(seq_len(nrow(df_task1)), size = nrow(df_task1), replace = TRUE)
  boot_df <- df_task1[idx, ]
  coef(lm(BOOT_FORMULA, data = boot_df))[COEF_NAME]
})

boot_se <- sd(boot_vals)
boot_mean <- mean(boot_vals)
ci_normal <- c(
  boot_mean - qnorm(1 - alpha / 2) * boot_se,
  boot_mean + qnorm(1 - alpha / 2) * boot_se
)
ci_percentile <- quantile(boot_vals, probs = c(alpha / 2, 1 - alpha / 2))

hist(boot_vals, breaks = 30, col = "lightblue", main = "Bootstrap distribution", xlab = COEF_NAME)
abline(v = ci_normal, col = "red", lty = 2, lwd = 2)
abline(v = ci_percentile, col = "blue", lty = 3, lwd = 2)

boot_se
ci_normal
ci_percentile
```

Interpretation template: The bootstrap SE is `[BOOT_SE]`. If the bootstrap distribution is skewed, the percentile interval is preferred because it follows the empirical distribution rather than forcing symmetry.

## task1_17_ridge_vs_lasso

Use when the exam asks you to compare L1 and L2 penalties directly.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Keep the same split for both models so the comparison is fair.

```{r}
#| label: task1_17_ridge_vs_lasso
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glmnet)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]

x_train <- model.matrix(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)[, -1]
y_train <- train_df[[TARGET_VAR]]

cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
lasso_fit <- glmnet(x_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min)

cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0)
ridge_fit <- glmnet(x_train, y_train, alpha = 0, lambda = cv_ridge$lambda.min)

coef(lasso_fit)
coef(ridge_fit)
```

Interpretation template: LASSO can set some coefficients exactly to zero, while Ridge typically only shrinks them toward zero. Therefore, `[LASSO / RIDGE]` is preferable when `[VARIABLE_SELECTION / COEFFICIENT_STABILITY]` is the main goal.

## task1_18_manual_kfold_cv_simple

Use when the exam asks for a hand-written 5-fold CV loop comparing simple polynomial choices.

**How to adapt:**
- Change `DATA_PATH_TASK1`.
- Change `TARGET_VAR` and `POLY_VAR`.
- Change `K_FOLDS` or polynomial degrees if required.

```{r}
#| label: task1_18_manual_kfold_cv_simple
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
POLY_VAR <- "X_VAR"
K_FOLDS <- 5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()

set.seed(SEED_VALUE)
fold_id <- sample(rep(1:K_FOLDS, length.out = nrow(df_task1)))

cv_deg2 <- numeric(K_FOLDS)
cv_deg3 <- numeric(K_FOLDS)

for(k in seq_len(K_FOLDS)) {
  fold_train <- df_task1[fold_id != k, ]
  fold_val   <- df_task1[fold_id == k, ]

  fit_deg2 <- lm(as.formula(paste(TARGET_VAR, "~ poly(", POLY_VAR, ", 2, raw = TRUE)")), data = fold_train)
  fit_deg3 <- lm(as.formula(paste(TARGET_VAR, "~ poly(", POLY_VAR, ", 3, raw = TRUE)")), data = fold_train)

  pred_deg2 <- predict(fit_deg2, newdata = fold_val)
  pred_deg3 <- predict(fit_deg3, newdata = fold_val)

  cv_deg2[k] <- mean((fold_val[[TARGET_VAR]] - pred_deg2)^2)
  cv_deg3[k] <- mean((fold_val[[TARGET_VAR]] - pred_deg3)^2)
}

tibble(model = c("degree_2", "degree_3"), cv_mse = c(mean(cv_deg2), mean(cv_deg3)))
```

Interpretation template: The lower CV MSE belongs to `[BEST_DEGREE]`, so that degree gives the better bias-variance tradeoff out-of-sample.

## task1_19_manual_kfold_cv_best_practice

Use when the exam wants the polynomial comparison while controlling for the other useful predictors.

**How to adapt:**
- Change `DATA_PATH_TASK1`, `TARGET_VAR`, `POLY_VAR`, and `CONTROL_VARS`.
- Change `K_FOLDS` if the exam specifies another value.

```{r}
#| label: task1_19_manual_kfold_cv_best_practice
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
POLY_VAR <- "X_VAR"
CONTROL_VARS <- c("X_LINEAR_1", "X_LINEAR_2", "X_LINEAR_3")
K_FOLDS <- 5
SEED_VALUE <- 123

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()

set.seed(SEED_VALUE)
fold_id <- sample(rep(1:K_FOLDS, length.out = nrow(df_task1)))

cv_deg2 <- numeric(K_FOLDS)
cv_deg3 <- numeric(K_FOLDS)

for(k in seq_len(K_FOLDS)) {
  fold_train <- df_task1[fold_id != k, ]
  fold_val   <- df_task1[fold_id == k, ]

  deg2_formula <- as.formula(
    paste(TARGET_VAR, "~", paste(c(CONTROL_VARS, paste0("poly(", POLY_VAR, ", 2, raw = TRUE)")), collapse = " + "))
  )
  deg3_formula <- as.formula(
    paste(TARGET_VAR, "~", paste(c(CONTROL_VARS, paste0("poly(", POLY_VAR, ", 3, raw = TRUE)")), collapse = " + "))
  )

  fit_deg2 <- lm(deg2_formula, data = fold_train)
  fit_deg3 <- lm(deg3_formula, data = fold_train)

  pred_deg2 <- predict(fit_deg2, newdata = fold_val)
  pred_deg3 <- predict(fit_deg3, newdata = fold_val)

  cv_deg2[k] <- mean((fold_val[[TARGET_VAR]] - pred_deg2)^2)
  cv_deg3[k] <- mean((fold_val[[TARGET_VAR]] - pred_deg3)^2)
}

tibble(model = c("degree_2_controlled", "degree_3_controlled"), cv_mse = c(mean(cv_deg2), mean(cv_deg3)))
```

Interpretation template: After controlling for the other predictors, `[BEST_DEGREE]` still performs best. By parsimony, we should keep the simpler model unless the more complex one meaningfully lowers CV error.

## task1_20_confidence_vs_prediction_interval

Use when the exam asks for both interval types for a new observation and a conceptual distinction.

**How to adapt:**
- Change `DATA_PATH_TASK1` and `TARGET_VAR`.
- Change the predictors in `new_obs` to exactly match your model.
- Change `CONF_LEVEL` if needed.

```{r}
#| label: task1_20_confidence_vs_prediction_interval
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK1 <- "DATA_HERE.csv"
TARGET_VAR <- "Y_VAR"
TRAIN_FRAC <- 0.5
SEED_VALUE <- 123
CONF_LEVEL <- 0.95

df_task1 <- read_csv(DATA_PATH_TASK1, show_col_types = FALSE) %>% clean_names()

set.seed(SEED_VALUE)
idx <- sample(seq_len(nrow(df_task1)), size = floor(TRAIN_FRAC * nrow(df_task1)))
train_df <- df_task1[idx, ]

ols_fit <- lm(as.formula(paste(TARGET_VAR, "~ .")), data = train_df)

new_obs <- tibble(
  X1_VAR = median(train_df$X1_VAR),
  X2_VAR = median(train_df$X2_VAR),
  X3_VAR = median(train_df$X3_VAR)
)

ci_out <- predict(ols_fit, newdata = new_obs, interval = "confidence", level = CONF_LEVEL)
pi_out <- predict(ols_fit, newdata = new_obs, interval = "prediction", level = CONF_LEVEL)

ci_out
pi_out
```

Interpretation template: The confidence interval estimates the mean response for this profile, while the prediction interval estimates one new observation. Therefore, the prediction interval should be wider.

## task1_21_theory_discussion_prompts

Use when the exam asks a short theory/comparison question instead of more code.

**How to adapt:**
- Keep only the rows that match the actual question.
- Replace the bracketed placeholders with your model names, variables, and findings.

```{r}
#| label: task1_21_theory_discussion_prompts
suppressPackageStartupMessages(library(tidyverse))

tribble(
  ~theme, ~answer_stem,
  "GAM vs polynomial", "GAM is more flexible because it uses local smoothers, while a polynomial imposes one global curve.",
  "LASSO vs Ridge", "LASSO is preferred for variable selection; Ridge is preferred for shrinking correlated coefficients.",
  "Bias vs variance", "Rigid linear models risk high bias; flexible models reduce bias but can increase variance.",
  "Why test MSE matters", "Test MSE evaluates generalisation on unseen data and is therefore better than training error for model comparison.",
  "Bootstrap vs asymptotics", "Bootstrap is preferred when theoretical distributional assumptions are weak or clearly violated.",
  "Interpretability vs flexibility", "GAM is more interpretable than tree ensembles because the partial effect of each predictor can be visualised directly."
)
```

Interpretation template: Choose the row matching the exam prompt and then add dataset-specific evidence such as `[BEST_MODEL]`, `[KEY_VARIABLES]`, and `[MODEL_METRIC]`.

# Task 2 snippets (from ExamPrep2)

## task2_1_import_and_raw_summary

Use when the exam asks you to load the churn/classification data and inspect structure, missingness, and descriptive statistics.

**How to adapt:**
- Change `DATA_PATH_TASK2`.
- Change `TARGET_VAR` if the binary response has another name.
- Add/remove summary steps only if the exam requests them.

```{r}
#| label: task2_1_import_and_raw_summary
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"

df_customer_raw <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names()

glimpse(df_customer_raw)
colSums(is.na(df_customer_raw))
summary(df_customer_raw %>% select(where(is.numeric)))

df_customer_raw %>%
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "standard_deviation") %>%
  arrange(desc(standard_deviation))
```

Interpretation template: The raw Task 2 data has `[N_ROWS]` rows and `[N_COLS]` columns with `[NA_STATUS]` missingness. Numeric predictors are on `[SIMILAR / VERY_DIFFERENT]` scales, which matters for methods like KNN or regularised models.

## task2_2_raw_boxplots_and_scale_check

Use when the exam asks for a quick visual inspection of scale, outliers, and suspicious numeric-coded categories.

**How to adapt:**
- Change `DATA_PATH_TASK2`.
- If your dataset has many numeric columns, keep the faceting approach.

```{r}
#| label: task2_2_raw_boxplots_and_scale_check
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"

df_customer_raw <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names()

df_customer_raw %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Interpretation template: The boxplots show `[OUTLIER_COMMENT]` and `[SCALE_COMMENT]`. They also help flag suspicious variables like `[AREA_CODE_OR_SIMILAR]` that may be numeric-coded categories rather than real continuous predictors.

## task2_3_business_logic_checks

Use when the exam asks you to justify why some variables should be recoded or removed before logistic regression.

**How to adapt:**
- Change `DATA_PATH_TASK2`.
- Replace the variable names if your dataset uses different telecom/billing columns.

```{r}
#| label: task2_3_business_logic_checks
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"

df_customer_raw <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names()

unique(df_customer_raw$area_code)
length(unique(df_customer_raw$state))

cor(df_customer_raw$total_day_minutes, df_customer_raw$total_day_charge)
cor(df_customer_raw$total_eve_minutes, df_customer_raw$total_eve_charge)
cor(df_customer_raw$total_night_minutes, df_customer_raw$total_night_charge)
cor(df_customer_raw$total_intl_minutes, df_customer_raw$total_intl_charge)
```

Interpretation template: `area_code` behaves like a category, `state` has high cardinality, and minutes/charge pairs are nearly perfectly correlated. Therefore, we should recode `area_code` as factor and remove `state` plus the redundant charge variables.

## task2_4_collinearity_heatmap

Use when the exam explicitly asks for a correlation matrix or heatmap of the numeric predictors.

**How to adapt:**
- Change `DATA_PATH_TASK2`.
- If the exam wants only predictors, remove the response before `cor()`.

```{r}
#| label: task2_4_collinearity_heatmap
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"

df_customer_raw <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names()

cor_matrix <- cor(df_customer_raw %>% select(where(is.numeric)))
round(cor_matrix, 2)

cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "var1") %>%
  pivot_longer(cols = -var1, names_to = "var2", values_to = "correlation") %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Interpretation template: The heatmap shows `[HIGH_CORR_PAIRS]`, with the strongest correlation around `[MAX_CORR]`. Those variables should not be kept together in the same logistic model.

## task2_5_preprocessing_pipeline

Use when the exam asks you to create the final clean modeling dataset for classification.

**How to adapt:**
- Change `DATA_PATH_TASK2`.
- Edit the removed columns if your dataset differs.
- Change the factor conversions to match your actual categorical variables.

```{r}
#| label: task2_5_preprocessing_pipeline
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"

df_customer_raw <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names()

df_clean <- df_customer_raw %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

dim(df_clean)
names(df_clean)
```

Interpretation template: The cleaned dataset contains `[ROWS]` rows and `[COLS]` columns. The response and all categorical predictors are now safely encoded for classification models.

## task2_6_prior_probabilities

Use when the exam asks for class imbalance, prior probabilities, or why accuracy alone is misleading.

**How to adapt:**
- Change `DATA_PATH_TASK2` and `TARGET_VAR`.
- Keep the same preprocessing choices you justified earlier.

```{r}
#| label: task2_6_prior_probabilities
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

prop.table(table(df_clean[[TARGET_VAR]]))

ggplot(df_clean, aes(x = .data[[TARGET_VAR]], fill = .data[[TARGET_VAR]])) +
  geom_bar(color = "black") +
  theme_minimal() +
  labs(title = "Class distribution", x = TARGET_VAR, y = "Count")
```

Interpretation template: The class proportions are `[MAJORITY]%` versus `[MINORITY]%`. Because the target is `[BALANCED / IMBALANCED]`, we must evaluate later models with the confusion matrix and especially sensitivity/specificity, not accuracy alone.

## task2_7_customer_service_calls_threshold

Use when the exam asks you to investigate a likely threshold effect in a count-type predictor such as customer service calls.

**How to adapt:**
- Change `DATA_PATH_TASK2`, `TARGET_VAR`, and `COUNT_VAR`.
- If the exam wants proportions rather than counts, keep `position = "fill"`.

```{r}
#| label: task2_7_customer_service_calls_threshold
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"
COUNT_VAR <- "customer_service_calls"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

ggplot(df_clean, aes(x = .data[[COUNT_VAR]], fill = .data[[TARGET_VAR]])) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Possible threshold effect", x = COUNT_VAR, y = "Proportion")
```

Interpretation template: The probability of `[TARGET_VAR] = positive` appears to rise sharply after about `[THRESHOLD_VALUE]` calls. That suggests a non-linear threshold effect that logistic regression may miss unless we transform or bin the predictor.

## task2_8_scatter_vs_density_choice

Use when the exam asks you to justify why scatter plots are weak for binary targets and what to use instead.

**How to adapt:**
- Change `DATA_PATH_TASK2`, `TARGET_VAR`, and `CONTINUOUS_VAR`.
- Under time pressure, keep only the density plot.

```{r}
#| label: task2_8_scatter_vs_density_choice
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"
CONTINUOUS_VAR <- "CONTINUOUS_VAR"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

# Jitter plot: often noisy with binary outcomes
p1 <- ggplot(df_clean, aes(x = .data[[TARGET_VAR]], y = .data[[CONTINUOUS_VAR]], color = .data[[TARGET_VAR]])) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_minimal()

# Density plot: usually better for continuous X vs binary Y
p2 <- ggplot(df_clean, aes(x = .data[[CONTINUOUS_VAR]], fill = .data[[TARGET_VAR]])) +
  geom_density(alpha = 0.5) +
  theme_minimal()

p1
p2
```

Interpretation template: The jitter plot suffers from overplotting, while the density plot clearly shows whether the two classes separate. Therefore, density plots are the better exam choice for continuous predictors with binary response.

## task2_9_continuous_density_plots

Use when the exam asks for graphs to identify promising continuous predictors.

**How to adapt:**
- Change `DATA_PATH_TASK2` and `TARGET_VAR`.
- If you only want a few variables, replace `select(where(is.numeric), ...)` with explicit names.

```{r}
#| label: task2_9_continuous_density_plots
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

df_clean %>%
  select(where(is.numeric), all_of(TARGET_VAR)) %>%
  pivot_longer(cols = -all_of(TARGET_VAR), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = .data[[TARGET_VAR]])) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal()
```

Interpretation template: Variables with the clearest class-conditional separation are `[STAR_PREDICTORS]`. Variables with almost complete overlap are `[WEAK_PREDICTORS]` and are unlikely to be useful.

## task2_10_categorical_bar_plot

Use when the exam asks for a graph of a categorical predictor against the binary response.

**How to adapt:**
- Change `DATA_PATH_TASK2`, `TARGET_VAR`, and `CATEGORY_VAR`.
- Keep `position = "fill"` when the exam wants proportions rather than counts.

```{r}
#| label: task2_10_categorical_bar_plot
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"
CATEGORY_VAR <- "international_plan"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

ggplot(df_clean, aes(x = .data[[CATEGORY_VAR]], fill = .data[[TARGET_VAR]])) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Conditional response probability", x = CATEGORY_VAR, y = "Proportion")
```

Interpretation template: The standardized bar chart shows that the churn / positive-class probability is `[HIGHER / LOWER / SIMILAR]` across levels of `[CATEGORY_VAR]`. That makes `[CATEGORY_VAR]` a `[STRONG / WEAK]` candidate predictor.

## task2_11_conditional_means_table

Use when the exam asks for a numeric summary that backs up your Task 2 graphs.

**How to adapt:**
- Change `DATA_PATH_TASK2` and `TARGET_VAR`.
- Replace `mean` with `median` if the question requests a more robust summary.

```{r}
#| label: task2_11_conditional_means_table
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

DATA_PATH_TASK2 <- "DATA_HERE.csv"
TARGET_VAR <- "TARGET_VAR"

df_clean <- read_csv(DATA_PATH_TASK2, show_col_types = FALSE) %>%
  clean_names() %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  mutate(
    !!TARGET_VAR := as.factor(.data[[TARGET_VAR]]),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

df_clean %>%
  group_by(.data[[TARGET_VAR]]) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  t()
```

Interpretation template: The grouped means show the strongest separation for `[TOP_VARS]`. These variables should be expected to matter most in baseline logistic or tree-based classifiers.

## task2_12_master_eda_conclusion

Use when the exam asks for a short final EDA conclusion before modeling.

**How to adapt:**
- Replace the placeholder vectors with the variables you identified from your own graphs/tables.
- Delete any line that is not relevant to the actual exam dataset.

```{r}
#| label: task2_12_master_eda_conclusion
suppressPackageStartupMessages(library(tidyverse))

STAR_PREDICTORS <- c("TOTAL_DAY_MINUTES", "CUSTOMER_SERVICE_CALLS")
WEAK_PREDICTORS <- c("TOTAL_DAY_CALLS", "TOTAL_EVE_CALLS")
NONLINEAR_HINTS <- c("NUMBER_VMAIL_MESSAGES", "CUSTOMER_SERVICE_CALLS")

tribble(
  ~section, ~exam_sentence,
  "star predictors", paste("The strongest candidate predictors are", paste(STAR_PREDICTORS, collapse = ", "), "because they show the clearest class separation."),
  "weak predictors", paste("The weakest predictors are", paste(WEAK_PREDICTORS, collapse = ", "), "because their distributions overlap almost completely."),
  "non-linearity", paste("Potential non-linear effects appear in", paste(NONLINEAR_HINTS, collapse = ", "), "so a flexible tree-based model may outperform a rigid logistic boundary."),
  "evaluation", "Because the response is imbalanced, later model evaluation must emphasise sensitivity/specificity rather than raw accuracy alone."
)
```

Interpretation template: My final EDA conclusion is that `[STAR_PREDICTORS]` look most useful, `[WEAK_PREDICTORS]` look weak, and `[NONLINEAR_HINTS]` may favour flexible classifiers. Therefore, the next modeling step should reflect both imbalance and possible non-linearity.
