---
title: "BAN404 R Exam Compendium"
subtitle: "Comprehensive solution guide — 2023, 2024, and 2025 school exams"
author: "Self-study preparation file"
date: "2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

**General Setup and Package Loading**

Before tackling any specific task, we load the standard suite of BAN404 packages. We use `tidyverse` for data manipulation and visualization, `janitor` for safe column naming, and the core statistical learning packages (`glmnet`, `gam`, `tree`, `randomForest`, `gbm`) so we don't have to interrupt our workflow later.

```{r setup_and_load}
# 0. Load necessary libraries
library(tidyverse)
library(janitor)
library(glmnet)       # Regularization (Ridge/LASSO)
library(gam)          # Generalized Additive Models (Non-linearities)
library(tree)         # Decision Trees
library(randomForest) # Bagging and Random Forests
library(gbm)          # Boosting

# 1. Load the datasets safely
# show_col_types = FALSE keeps the exam output clean. 
# clean_names() ensures no weird spaces or characters in column names.
df_task1 <- read_csv("task1_data.csv", show_col_types = FALSE) %>% clean_names()
df_customer <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()
```

# Task 1: Continuous Outcome (Regression)

Before fitting any models, we must understand the data generating process. The goal of this Exploratory Data Analysis (EDA) is to inform our model selection. We will systematically now do checks for:

1.  Data integrity (Missing values)

2.  Target distribution (Skewness / Outliers)

3.  Feature independence (Multicollinearity)

4.  Functional form (Linearity vs. Non-linearity)

### Data Integrity Check

**Reasoning:** Many machine learning algorithms (like standard Random Forests or `lm()` without `na.action` adjustments) will fail or drop rows entirely if missing values are present. We must verify completeness first.

```{r t1_missing}
# Check for missing values in each column
colSums(is.na(df_task1))
```

*Conclusion:* There are 0 missing values. The dataset is complete, so no imputation is necessary.

### Distribution of the Target Variable (y)

**Reasoning:** The distribution of the continuous outcome `y` tells us if we need to transform the data. Severe skewness can violate OLS residual assumptions and make the model overly sensitive to outliers.

```{r t1_target_dist}
ggplot(df_task1, aes(x = y)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Target Variable (y)", 
       subtitle = "Checking for severe skewness or extreme outliers")
```

*Conclusion:* The histogram shows a right-skewed distribution with a long tail extending towards 50. Normally, this suggests a `log(y)` transformation. However, **the data contains negative values (down to approximately -8)**. Because $\log(x)$ is undefined for $x \le 0$, a standard log transformation is invalid here without shifting the entire dataset. We will proceed with the un-transformed `y`, but we note that our OLS model may yield non-normal residuals.

### Multicollinearity and Feature Scaling

**Reasoning:** If predictor variables are highly correlated (multicollinearity), OLS coefficient estimates become highly unstable (high variance). This is the primary justification for using Ridge Regression. We also check the summary statistics to see if predictors need standardization.

```{r t1_multicollinearity}
# Check summary to see if scales are wildly different
summary(df_task1)

# Check correlation matrix
cor_matrix <- cor(df_task1)
round(cor_matrix, 2)
```

*Conclusion:* The `summary()` shows all predictors are roughly centered around 0 with similar ranges, meaning massive scale disparities won't artificially inflate penalties in regularized models. More importantly, the correlation matrix shows **zero correlation between the independent variables**. Because there is no multicollinearity, **Ridge regression is unlikely to provide any significant variance-reduction benefit over OLS.**

### Checking the Linearity Assumption

**Reasoning:** OLS, Ridge, and LASSO all assume a strictly linear relationship between $X$ and $Y$. If the true relationship is non-linear, these models will severely underfit. We plot each predictor against `y`, overlaying a strict linear fit (blue) and a flexible LOESS fit (red).

```{r t1_linearity}
#| eval: true

df_task1 %>%
  pivot_longer(cols = starts_with("x"), names_to = "predictor", values_to = "value") %>%
  ggplot(aes(x = value, y = y)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +     # Linear OLS fit
  geom_smooth(method = "loess", color = "red", se = FALSE) +   # Flexible non-linear fit
  facet_wrap(~predictor, scales = "free_x") +
  theme_minimal() +
  labs(title = "Predictors vs. y (Blue = Linear, Red = Flexible)",
       subtitle = "Looking for gaps between the straight line and flexible curve")
```

*Conclusion:* This plot dictates our entire modeling strategy: 1. `x1` and `x5` are strictly linear. 2. `x4` and `x6` appear to be pure noise (both lines are flat). 3. **`x2` and `x3` display severe non-linear relationships.** Specifically, `x2` forms a distinct U-shape (quadratic).

Because of this, standard linear models (OLS, LASSO) will falsely interpret `x2` as a weak predictor. To capture the true signal in this dataset, we can "upgrade" from linear models to a **Generalized Additive Model (GAM)** or a **Tree-based model**.

### EDA Findings

Based on the exploratory data analysis, we can draw several critical conclusions that dictate our modeling strategy:

1.  **Pre-Standardization:** The `summary()` output shows that all predictors (`x1` through `x6`) have a mean of approximately 0 and a similar range. Therefore, manual standardization is not strictly necessary before applying penalty-based models like Ridge or LASSO, though `glmnet` handles this automatically anyway.
2.  **No Multicollinearity:** The correlation matrix shows that the independent variables are entirely uncorrelated with one another (correlations are all \~0.00). Because there is no multicollinearity, **Ridge regression is unlikely to provide a massive advantage** over OLS in terms of variance reduction between predictors.
3.  **Severe Non-Linearities:** The scatterplots reveal a massive limitation of standard linear models for this dataset.
    -   `x1` and `x5` have strictly linear relationships with `y` (the blue and red lines match).
    -   **`x2` exhibits a strong U-shaped (quadratic) relationship** with `y`. A standard OLS or LASSO model will assign `x2` a coefficient of near 0 (as seen in the correlation matrix), falsely identifying it as useless.
    -   `x3` also shows a distinct non-linear curve.
    -   `x4` and `x6` appear to be pure noise (flat lines for both fits).
4.  **Modeling Implication:** To achieve optimal predictive performance, a baseline OLS model will underfit the data. We must introduce flexibility. We should proceed with a **Generalized Additive Model (GAM)** to capture the non-linearities in `x2` and `x3` smoothly, or apply a **Tree-based model** (like Random Forest) which naturally handles non-linear relationships without manual specification.

## Data Splitting

**Possible Exam Question:** *"Split the dataset into a training and test set. Explain why this step is necessary before evaluating model performance."*

**Exam-Ready Written Reasoning:** To accurately evaluate how well our models will generalize to unseen data, we must split the dataset into a training set and a test set. If we train and evaluate a model on the exact same data, complex models (like Random Forests) will simply memorize the data (overfitting), leading to an artificially low error rate. By holding out 50% of the data exclusively for testing, we get an honest, unbiased estimate of the model's true predictive power (Test MSE).

```{r t1_split}
# Set a random seed. This locks the random number generator so our split is 
# exactly the same every time we run the code. Crucial for reproducible exam answers!
set.seed(123) 

# Create an index of row numbers. sample() picks random rows. 
# size = floor(nrow()/2) ensures we grab exactly 50% of the data for training.
idx <- sample(seq_len(nrow(df_task1)), size = floor(nrow(df_task1) / 2))

# Use the random row numbers (idx) to pull the training data
train_t1 <- df_task1[idx, ]

# Use the negative index (-idx) to pull all the REMAINING rows for the test data
test_t1  <- df_task1[-idx, ]

# LASSO/Ridge (glmnet package) cannot read normal dataframes. They require a numeric matrix.
# model.matrix() automatically converts the data. The [,-1] drops the intercept column 
# because glmnet automatically creates its own intercept.
X_train <- model.matrix(y ~ ., data = train_t1)[, -1]
y_train <- train_t1$y   # Extract just the target variable for the training set

X_test  <- model.matrix(y ~ ., data = test_t1)[, -1] # Do the exact same for the test set
y_test  <- test_t1$y
```

## Baseline Linear Models (OLS & LASSO)

**Possible Exam Question:** *"Fit an Ordinary Least Squares (OLS) model and a LASSO regression model. Compare their performance. Does LASSO perform variable selection? Explain the limitations of both models given your EDA."*

**Exam-Ready Written Reasoning:** We begin with an OLS regression as our strict baseline. However, our EDA proved that `x2` has a severe U-shaped non-linear relationship with `y`. Because OLS only draws straight lines, we expect it to completely miss the signal in `x2` and perform poorly.

Next, we run a LASSO regression. LASSO applies an "L1 penalty" (absolute value penalty) to the coefficients. Unlike Ridge regression (which only shrinks coefficients towards zero), LASSO's mathematical penalty can force coefficients to exactly zero, effectively dropping them from the model. This performs automatic variable selection, which is perfect here because our EDA suggested `x4` and `x6` are pure noise. However, because LASSO is still a linear model, it will also fail to capture the non-linear curve of `x2`.

```{r t1_baselines}
# --- 1. OLS BASELINE ---
# Fit the linear model. 'y ~ .' means predict y using all other variables in the dataset.
ols_fit <- lm(y ~ ., data = train_t1)

# Use the fitted OLS model to predict the y-values for the unseen test data.
pred_ols <- predict(ols_fit, newdata = test_t1)

# Calculate Mean Squared Error (MSE): The average of the squared differences between 
# the actual test y-values and our predicted y-values. Lower is better.
mse_ols <- mean((test_t1$y - pred_ols)^2)


# --- 2. LASSO REGRESSION ---
set.seed(123) # Lock seed again because cross-validation uses random folds.

# cv.glmnet automatically tests many values of 'lambda' (the penalty size) 
# using Cross-Validation to find the optimal penalty. alpha = 1 tells it to use LASSO.
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)

# Now, fit the final LASSO model using the best lambda (lambda.min) found above.
lasso_fit <- glmnet(X_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min)

# Predict on the test matrix. as.numeric() ensures the output is a simple vector.
pred_lasso <- as.numeric(predict(lasso_fit, newx = X_test))

# Calculate Test MSE for LASSO
mse_lasso <- mean((test_t1$y - pred_lasso)^2)

# Print the LASSO coefficients to prove to the professor that variable selection happened.
# Look for the dots (.), which mean the coefficient was forced to exactly zero.
print("LASSO Coefficients:")
coef(lasso_fit)
```

### Scenario A: Explicit Non-Linearities (GAM)

**Possible Exam Question:** *"Based on your EDA, some variables are non-linear. Fit a model that captures these non-linearities while remaining highly interpretable."*

**Exam-Ready Written Reasoning:** To fix the underfitting of the linear models, we upgrade to a Generalized Additive Model (GAM). A GAM allows us to apply "smoothing splines" to specific variables, letting the model bend to fit curves while still keeping the effects additive and highly interpretable. Based on our EDA and LASSO results, we manually drop the noise variables (`x4`, `x6`). We keep `x1` and `x5` as straight lines (linear), but we wrap `x2` and `x3` in the `s()` function to allow the model to capture their severe non-linear curves.

```{r t1_gam}
# Fit the GAM. 
# Notice we excluded x4 and x6.
# x1 and x5 are entered normally (forcing a straight line).
# s(x2) and s(x3) tell the model to use smoothing splines to fit the curves.
gam_fit <- gam(y ~ x1 + s(x2) + s(x3) + x5, data = train_t1)

# Predict on test data
pred_gam <- predict(gam_fit, newdata = test_t1)

# Calculate Test MSE
mse_gam <- mean((test_t1$y - pred_gam)^2)

# Plot the GAM partial dependence plots. 
# This shows the professor exactly how the model bent to fit x2 and x3.
# mfrow = c(2,2) splits the plot window into a 2x2 grid so all 4 plots fit on one screen.
par(mfrow = c(2,2))
plot(gam_fit, se = TRUE, col = "blue")
par(mfrow = c(1,1)) # Reset plot window back to normal
```

### Scenario B: Tree-Based Ensemble (Random Forest)

**Possible Exam Question:** *"Use a tree-based ensemble method to predict y. Evaluate its performance and explain which variables are most important for making predictions."*

**Exam-Ready Written Reasoning:** Instead of manually specifying which variables are non-linear (like we did in the GAM), we can use a Random Forest. A Random Forest builds hundreds of deep decision trees using bootstrapped samples of the data, and randomly restricts the predictors available at each split (`mtry`). This decorrelates the trees, drastically reducing variance when their predictions are averaged. Decision trees naturally handle non-linearities and ignore useless variables without requiring manual intervention, making it an incredibly powerful "black-box" alternative to OLS.

```{r t1_rf}
set.seed(123) # Lock seed because Random Forest uses random bootstrap sampling

# Fit the Random Forest. importance = TRUE forces it to track which variables 
# reduce the error the most, so we can plot them later.
rf_fit <- randomForest(y ~ ., data = train_t1, importance = TRUE)

# Predict on test data
pred_rf <- predict(rf_fit, newdata = test_t1)

# Calculate Test MSE
mse_rf <- mean((test_t1$y - pred_rf)^2)

# Print the numerical importance scores.
# Increase in Node Purity (IncNodePurity) shows how much worse the model 
# would be if we removed that variable.
importance(rf_fit)

# Plot the variable importance for easy visual interpretation in the exam.
varImpPlot(rf_fit, main = "Random Forest Variable Importance")
```

## Task 1 Final Model Comparison

**Possible Exam Question:** *"Summarize the performance of all models tested. Which model is optimal for this dataset and why?"*

**Exam-Ready Written Reasoning:** By comparing the Test Mean Squared Error (MSE) across all models, we can definitively conclude that capturing non-linearity was the most critical factor for this dataset. OLS and LASSO both performed poorly because their strict linear constraints caused them to completely miss the massive U-shaped signal in `x2`. The flexible models (GAM and Random Forest) achieved a significantly lower Test MSE. If the business requires interpretability to understand exactly *how* each variable affects the outcome, the GAM is optimal. If the sole goal is minimizing predictive error, the Random Forest is the superior choice.

```{r t1_comparison}
# Create a tidy dataframe to cleanly display the MSE of all 4 models
# tibble() creates the table, and we input our calculated MSE variables.
model_results <- tibble(
  Model = c("OLS", "LASSO", "GAM", "Random Forest"),
  Test_MSE = c(mse_ols, mse_lasso, mse_gam, mse_rf)
) %>% 
  arrange(Test_MSE) # Sort from lowest error (best) to highest error (worst)

# Print the final leaderboard
print(model_results)
```

### Task 1 Final Theoretical Discussion

**Possible Exam Question:** *"Based on your results, which model performed best? Explain this outcome using the concepts of bias, flexibility, and the true underlying data structure."*

**Exam-Ready Written Reasoning:** The results cleanly illustrate the limitations of rigid linear assumptions and the power of flexible modeling.

1.  **Failure of Linear Models (OLS & LASSO):** With a Test MSE of \~5.08, the linear models performed the worst. Due to their low flexibility, they suffered from high **bias**. Specifically, they forced a straight line through the quadratic (U-shaped) relationship of `x2`, effectively reducing its coefficient to zero and throwing away highly valuable signal.
2.  **Success of the Tree Ensemble (Random Forest):** By upgrading to a Random Forest, the Test MSE dropped to \~2.31. Trees make no assumptions about linearity; they use recursive binary splitting to approximate curves via step functions. Furthermore, the Random Forest's variable importance plot correctly isolated `x5`, `x3`, `x1`, and `x2` as the true predictors, mathematically confirming that `x4` and `x6` are pure noise.
3.  **Supremacy of the GAM:** The Generalized Additive Model achieved the lowest Test MSE (\~1.54). From a theoretical standpoint, this makes perfect sense. While Random Forests approximate curves using jagged step functions (which requires deep trees and increases variance), a GAM uses continuous smoothing splines to mathematically trace the exact curves of `x2` and `x3`. Because the true data-generating process here lacks complex multi-variable interactions and is purely additive and smooth, the GAM provides the optimal balance: it has enough flexibility to capture the curves (eliminating the bias of OLS) while enforcing an additive structure (keeping variance lower than a Random Forest). Furthermore, the GAM remains highly interpretable, as we can visually inspect the partial effect of each feature.

### Scenario C: RMSE and R-Squared for ALL Models)

**Possible Exam Question:** *"Calculate the Root Mean Squared Error (RMSE) and Test R-squared for all tested models. Create a final summary table and interpret the results."*

**Exam-Ready Written Reasoning:** While Mean Squared Error (MSE) is useful for mathematical optimization, it is in squared units. By calculating RMSE, we return the error to the original units of the target variable ($y$), making it easily interpretable as the "average prediction error." Additionally, calculating the Test $R^2$ tells us the exact percentage of variance in the unseen test data that our models successfully explain. We compute these for all models to provide a comprehensive final comparison.

```{r t1_advanced_metrics}
# =====================================================================
# 1. CALCULATE BASELINE METRIC: TOTAL SUM OF SQUARES (TSS)
# =====================================================================
# TSS is the total variance in the test data. 
# It represents the error if we just predicted the average 'y' for everyone.
# We need this to calculate R-squared for all our models.
tss_test <- sum((test_t1$y - mean(test_t1$y))^2)

# =====================================================================
# 2. CALCULATE SUM OF SQUARED ERRORS (SSE) FOR ALL MODELS
# =====================================================================
# SSE is the total mistakes each model made.
sse_ols   <- sum((test_t1$y - pred_ols)^2)
sse_lasso <- sum((test_t1$y - pred_lasso)^2)
sse_gam   <- sum((test_t1$y - pred_gam)^2)
sse_rf    <- sum((test_t1$y - pred_rf)^2)

# =====================================================================
# 3. BUILD THE FINAL COMPREHENSIVE MASTER TABLE
# =====================================================================
# We use tibble() to create a clean dataframe.
# RMSE = sqrt(MSE)
# R-Squared = 1 - (SSE / TSS)
final_model_results <- tibble(
  Model = c("OLS", "LASSO", "GAM", "Random Forest"),
  
  # Calculate RMSE (square root of the MSEs we already calculated)
  Test_RMSE = c(sqrt(mse_ols), sqrt(mse_lasso), sqrt(mse_gam), sqrt(mse_rf)),
  
  # Calculate R-Squared
  Test_R2 = c(
    1 - (sse_ols / tss_test),
    1 - (sse_lasso / tss_test),
    1 - (sse_gam / tss_test),
    1 - (sse_rf / tss_test)
  )
) %>% 
  arrange(desc(Test_R2)) # Sort by R-Squared, descending (highest/best at the top)

# Print the master table
print(final_model_results)
```

*Written Analysis:* The master table confirms our theoretical expectations. The rigid linear models (OLS, LASSO) explain less than 40% of the variance in the test data (R² \~ 0.36) and are off by an average of \~2.25 units (RMSE). By capturing the non-linearities, the Random Forest explains \~73% of the variance, and the GAM captures an outstanding \~82% of the variance, with an average prediction error (RMSE) of just \~1.24 units.

### Scenario D: The Bootstrap (Normal vs. Percentile CIs)

**Possible Exam Question:** *"Use the Bootstrap method (B=1000) to estimate the standard error for the coefficient of x1. Calculate both the Normal Approximation 95% CI and the Percentile 95% CI. Which is more appropriate if the distribution is skewed?"*

**Written Reasoning:** When model assumptions are violated, theoretical standard errors from `lm()` can be unreliable. We use Bootstrapping (resampling with replacement) to empirically estimate the standard error. We can calculate the Confidence Interval in two ways: 1. **Normal Approximation CI:** Assumes the bootstrap sampling distribution forms a perfect, symmetrical bell curve ($Mean \pm 1.96 \times SE$). 2. **Percentile CI:** Uses the *actual* empirical quantiles of our 1000 bootstrap estimates. If our plot shows the bootstrap distribution is highly skewed or asymmetric, the Normal Approximation is mathematically invalid, and we **must** rely on the Percentile CI because it perfectly traces the actual shape of the data.

```{r t1_bootstrap}
# =====================================================================
# BOOTSTRAP SETUP & LOOP
# =====================================================================
set.seed(123)       # CRITICAL: Locks random sampling for reproducibility.
B <- 1000           # Number of bootstrap samples
n <- nrow(train_t1) # Number of rows in training data

boot_coefs_x1 <- numeric(B) # Empty vector to store our 1000 coefficients

for(i in 1:B) {
  # Step 1: Draw random row numbers WITH REPLACEMENT
  boot_idx <- sample(1:n, size = n, replace = TRUE)
  
  # Step 2: Build the "fake" dataset
  boot_data <- train_t1[boot_idx, ]
  
  # Step 3: Fit model on fake dataset
  boot_model <- lm(y ~ x1, data = boot_data)
  
  # Step 4: Extract slope for x1 (item 2 in the list)
  boot_coefs_x1[i] <- coef(boot_model)[2]
}

# =====================================================================
# CALCULATE STANDARD ERROR & BOTH CONFIDENCE INTERVALS
# =====================================================================

# 1. Bootstrap Standard Error (The standard deviation of our 1000 estimates)
boot_se <- sd(boot_coefs_x1)
boot_mean <- mean(boot_coefs_x1)

# 2. METHOD A: NORMAL APPROXIMATION CI
# Formula: Mean +/- 1.96 * Standard Error
ci_normal_lower <- boot_mean - (1.96 * boot_se)
ci_normal_upper <- boot_mean + (1.96 * boot_se)
ci_normal <- c(ci_normal_lower, ci_normal_upper)

# 3. METHOD B: PERCENTILE CI (ACTUAL DISTRIBUTION)
# Uses the exact 2.5% and 97.5% cutoffs of our 1000 actual estimates.
ci_percentile <- quantile(boot_coefs_x1, probs = c(0.025, 0.975))

# Print outputs cleanly
print(paste("Bootstrap SE for x1:", round(boot_se, 4)))
print("Method A: Normal Approximation 95% CI:")
print(round(ci_normal, 4))
print("Method B: Percentile (Actual Distribution) 95% CI:")
print(round(ci_percentile, 4))

# =====================================================================
# VISUALIZE THE DIFFERENCE
# =====================================================================
hist(boot_coefs_x1, breaks = 30, col = "lightblue", 
     main = "Bootstrap Distribution of x1 Coefficient",
     xlab = "Coefficient Value")

# Add Normal CI lines in RED
abline(v = ci_normal, col = "red", lwd = 2, lty = 2) 

# Add Percentile CI lines in BLUE
abline(v = ci_percentile, col = "blue", lwd = 2, lty = 3) 

legend("topright", legend=c("Normal CI", "Percentile CI"), 
       col=c("red", "blue"), lty=c(2,3), lwd=2)
```

**Written Analysis:** The bootstrap results perfectly illustrate why resampling is necessary when OLS assumptions are violated. Our OLS model assumed perfectly normally distributed residuals, but our EDA showed `y` is highly skewed. Looking at the outputs:

-   **Normal CI:** `[1.0125, 1.2679]`

-   **Percentile CI:** `[1.0139, 1.2632]`

Notice that the Percentile CI is slightly narrower and shifted compared to the Normal CI. The histogram confirms this: the bootstrap sampling distribution is not perfectly symmetrical (it has a slight skew/heavy tail). The **Normal CI** forces a rigid, symmetric bell-curve formula onto the data (Mean±1.96×SE). The **Percentile CI** makes zero assumptions and simply looks at the exact bounds of the empirical data. Because the distribution is slightly skewed, the Normal approximation is mathematically flawed, and the **Percentile CI is the superior, more accurate measure of uncertainty** for this coefficient.

### Scenario E: Polynomial Regression (Fixing OLS Manually)

**Possible Exam Question:** *"Without using a GAM or Tree, modify your OLS model to capture the non-linear relationship in x2. Explain your approach."*

**Exam-Ready Written Reasoning:** Our EDA revealed that `x2` has a distinct U-shaped (quadratic) relationship with `y`. Standard OLS fits a straight line (degree 1), resulting in severe bias (underfitting). Without resorting to complex machine learning models, we can use Polynomial Regression by explicitly adding a squared term for `x2`. This allows the linear model to bend into a parabola for that specific variable, curing the bias while keeping the model entirely parametric and highly interpretable.

```{r t1_polynomial}
# =====================================================================
# POLYNOMIAL REGRESSION
# =====================================================================
# We use I(x2^2) to tell R: "Calculate x2 squared, and treat it as a new variable"
# We could also use poly(x2, 2), but I(x2^2) keeps the coefficients in original units.
poly_fit <- lm(y ~ x1 + x2 + I(x2^2) + x3 + x5, data = train_t1)

# Look at the summary to prove to the professor that the squared term is highly significant
print("Polynomial Model Summary:")
summary(poly_fit)

# Predict and calculate MSE
pred_poly <- predict(poly_fit, newdata = test_t1)
mse_poly <- mean((test_t1$y - pred_poly)^2)
print(paste("Polynomial OLS Test MSE:", round(mse_poly, 4)))
```

*Written Analysis:* By simply adding a quadratic term for `x2`, the previously useless OLS model is "rescued." The Test MSE drops massively compared to the baseline linear OLS, proving that the high error was purely due to functional form misspecification (bias), not a lack of signal in the data.

**Written Analysis:** By manually adding a quadratic term `I(x2^2)` to our OLS model, the p-value for `I(x2^2)` was `< 2e-16`, proving it is highly statistically significant.

-   **OLS Test MSE:** \~5.08

-   **Polynomial Test MSE:** \~3.05

-   **GAM Test MSE:** \~1.54

**The Bias-Variance Tradeoff:** The standard OLS model suffered from massive **Bias** (underfitting) because it forced a straight line through the U-shaped data of `x2`. By using Polynomial Regression, we reduced that Bias significantly, dropping the error from 5.08 to 3.05.

However, the Polynomial model still lost to the GAM (1.54). Why? Mathematically, `I(x2^2)` creates a **global parametric fit**—it forces a single rigid parabola over the entire dataset. A GAM uses `s(x2)` and `s(x3)`, which are **smoothing splines**. Splines are non-parametric and local; they bend flexibly across different regions of the data without forcing a strict global shape. Furthermore, our polynomial model ignored the non-linearity in `x3`. The GAM reduced Bias much more effectively than the manual polynomial, without inflating Variance.

### Scenario F: Boosting (Gradient Boosted Trees)

**Possible Exam Question:** *"Fit a Gradient Boosting model (gbm). Explain how its approach to learning differs fundamentally from a Random Forest."*

**Exam-Ready Written Reasoning:** While Random Forests build deep, high-variance trees independently and average them to reduce Variance, Boosting takes the exact opposite approach. Boosting builds shallow, weak trees *sequentially*. Each new tree is fit to the *residuals* (the mistakes) of the previous trees, slowly learning from its errors. Boosting primarily reduces Bias. Because it learns slowly (controlled by the shrinkage parameter), it is highly resistant to overfitting and often yields the best predictive performance among tree-based methods.

```{r t1_boosting}
# =====================================================================
# BOOSTING (GRADIENT BOOSTED MACHINES)
# =====================================================================
set.seed(123)

# Fit the boosting model using the gbm package
# distribution = "gaussian" is used for continuous regression tasks (like MSE).
# n.trees = 1000 means we build 1000 sequential trees.
# interaction.depth = 3 allows 3 splits per tree (shallow trees).
# shrinkage = 0.01 is the "learning rate" (how fast it learns).
boost_fit <- gbm(y ~ ., data = train_t1, 
                 distribution = "gaussian", 
                 n.trees = 1000, 
                 interaction.depth = 3, 
                 shrinkage = 0.01)

# Predict on test data. For gbm, we MUST specify how many trees to use for prediction.
pred_boost <- predict(boost_fit, newdata = test_t1, n.trees = 1000)

# Calculate Test MSE
mse_boost <- mean((test_t1$y - pred_boost)^2)
print(paste("Boosting Test MSE:", round(mse_boost, 4)))

# Print Variable Importance for Boosting (Relative Influence)
# This will show which variables the boosting algorithm focused on to reduce errors.
print("Boosting Variable Importance:")
summary(boost_fit)
```

**Written Analysis:** Boosting takes a fundamentally different approach to learning than Random Forests. While a Random Forest builds hundreds of deep, high-variance trees independently and averages them to reduce **Variance**, Boosting builds shallow, weak trees *sequentially*. Each new tree is fit to the *residuals* (the mistakes) of the previous trees. Because it attacks its own errors, Boosting is the ultimate **Bias-reduction** algorithm.

**Understanding the Code Parameters:**

-   `n.trees = 1000`: We build 1000 trees sequentially.

-   `interaction.depth = 3`: We force the trees to be very shallow (only 3 splits). This ensures each individual tree is a "weak learner" with low variance.

-   `shrinkage = 0.01`: This is the learning rate. We multiply the predictions of each new tree by a tiny number (0.01) so the model learns very slowly. This slow learning prevents the model from overfitting to the noise.

**Interpreting the Plot:** The `summary(boost_fit)` relative influence plot perfectly mirrors our Random Forest and EDA findings.

-   `x3` (37%) and `x5` (34%) are the massive drivers of the predictions, followed by `x2` and `x1`.

-   The algorithm assigned almost zero relative influence to `x4` (1.08%) and `x6` (0.004%). Because Boosting sequentially attacks residual errors, it quickly "realized" that splitting on `x4` and `x6` did not reduce errors, so it effectively performed automatic variable selection and dropped them.

### Task 1 Final Master Conclusion: The Bias-Variance Tradeoff

```{r t1_final_master_table}
# =====================================================================
# FINAL MASTER COMPARISON (CONCEPTUAL)
# =====================================================================
master_results <- tibble(
  Model = c("GAM", "Random Forest", "Polynomial OLS", "Boosting", "LASSO", "Baseline OLS"),
  Test_MSE = c(mse_gam, mse_rf, mse_poly, mse_boost, mse_lasso, mse_ols),
  Handles_NonLinearity = c("Yes (Splines)", "Yes (Trees)", "Yes (Global)", "Yes (Trees)", "No", "No"),
  Performs_Var_Selection = c("Manual", "Automatic", "Manual", "Automatic", "Automatic", "No")
) %>% arrange(Test_MSE)

print(master_results)
```

**Final Exam-Style Conclusion:** The journey from a Test MSE of \~5.08 down to \~1.54 perfectly demonstrates the core tenets of Statistical Learning:

1.  **Bias Reduction is the priority here:** The baseline OLS had extreme Bias due to functional form misspecification (missing the curves). LASSO dropped the noise variables (`x4`, `x6`), reducing Variance, but failed to fix the Bias.
2.  **Global vs. Local Flexibility:** Adding a polynomial `I(x2^2)` reduced Bias globally, improving MSE to 3.05. However, GAM utilized local smoothing splines to trace the exact curves of `x2` and `x3`, minimizing Bias almost perfectly (MSE 1.54).
3.  **Ensemble Methods:** Random Forest and Boosting performed excellently by automatically detecting the non-linearities and ignoring the noise. However, because the true data generating process here is a purely additive, smooth mathematical function (without crazy interactions between the `x` variables), the GAM's mathematical splines outperformed the jagged, blocky step-functions used by tree-based ensembles.

**Ultimate Recommendation:** For this specific continuous dataset, the Generalized Additive Model (GAM) with manually excluded noise variables is the optimal model, offering both the highest predictive accuracy and full interpretability.
