---
title: "BAN404 R Task 2 Comp"
subtitle: "Comprehensive solution guide based on dataset and possible questions"
author: "Self-study preparation file"
date: "2026"
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    fig-align: center
    fig-width: 6
    fig-height: 4
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: flatly
    highlight-style: kate
execute:
  echo: true
  warning: false
  message: false
  eval: true
  error: true
---

**General Setup and Package Loading**

Before tackling any specific task, we load the standard suite of BAN404 packages. We use `tidyverse` for data manipulation and visualization, `janitor` for safe column naming, and the core statistical learning packages (`glmnet`, `gam`, `tree`, `randomForest`, `gbm`) so we don't have to interrupt our workflow later.

```{r setup_and_load}
# 0. Load necessary libraries
library(tidyverse)
library(janitor)
library(glmnet)       # Regularization (Ridge/LASSO)
library(gam)          # Generalized Additive Models (Non-linearities)
library(tree)         # Decision Trees
library(randomForest) # Bagging and Random Forests
library(gbm)          # Boosting

# 1. Load the datasets safely
# show_col_types = FALSE keeps the exam output clean. 
# clean_names() ensures no weird spaces or characters in column names.
df_task1 <- read_csv("task1_data.csv", show_col_types = FALSE) %>% clean_names()
df_customer <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()
```

# Task 2: Binary Classification (Predicting Churn)

## Historical Exam Context & Strategy

Before touching the code, we must recognize what the professor typically asks for in Classification EDA. Past exams (2023-2025) repeatedly ask to: \* *"Remove variables that cannot plausibly be assumed to be available..."* \* *"If necessary, recode categorical variables as factors. Motivate your choices..."* \* *"Use descriptive statistics and graphs in order to find potentially useful predictors."*

**Our Strategy:** We will systematically verify data types, hunt for "illegal" variables, clean the dataset, evaluate class imbalance, and iteratively visualize the predictors to find the best approach. We follow a strict `Before ➔ Code ➔ After` structure.

## Initial Import and Raw Data Inspection

**Possible Exam Question:** *"Load the dataset. Inspect its dimensions and data types. Check for missing values. Create a visual summary (e.g., boxplots) of all continuous variables in their raw state to identify differences in scale."*

<<<<<<< Updated upstream
**Exam-Ready Written Reasoning:** The very first step in any machine learning pipeline is to import the data safely and inspect its raw structure. Classification algorithms in R require the target variable to be a "factor". We must also check for missing values (`NAs`), as algorithms will either crash or drop rows, distorting our results. Finally, checking the raw scale of continuous variables via boxplots tells us if we need to standardize the data later for distance-based models (like KNN) or regularized models (like LASSO). We keep our original dataset untouched (`df_customer_raw`) as a reliable fallback.
=======
**Exam-Ready Written Reasoning:** The very first step in any machine learning task is to understand the dimensions and data types of the dataset. Many algorithms in R (like `glmnet` or `randomForest`) will fail or behave unpredictably if categorical variables are stored as "character" or "logical" types rather than "factors". We must also check the summary statistics to see if predictors are on vastly different scales, which would require standardization later.

```{r t2_structure}
# glimpse() provides a transposed view of the data: columns become rows.
# This allows us to quickly see the number of rows/cols and the data type of every variable.
glimpse(df_customer)

# Check the summary statistics for all columns
# This helps identify wild differences in scale or obvious outliers in numeric variables
summary(df_customer)

# Convert character/logical categorical variables to factors
# R's classification models strictly require the target (churn) to be a factor.
# We also convert 'international_plan' and 'voice_mail_plan' as they are categorical.
df_customer <- df_customer %>%
  mutate(
    churn = as.factor(churn),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

# Verify the conversion worked by checking the levels of our target variable
# We expect to see "FALSE" and "TRUE"
levels(df_customer$churn)
```

### Data Integrity Check

**Reasoning:** Just like in regression, missing values (NAs) will cause most classification algorithms to drop rows, potentially losing valuable information or crashing the model entirely.

```{r t2_missing}
# colSums(is.na()) counts the number of missing values in every single column
colSums(is.na(df_customer))
```

*Conclusion:* The output shows 0 missing values across all columns. The dataset is complete, so no imputation or row-dropping is necessary.

# Task 2: Binary Classification (Churn Prediction)

## 2.1 Data Familiarization and Logical Inference

**Possible Exam Question:** *"Inspect the customer churn dataset. Identify any variables that are stored as numeric but should logically be treated as categorical. Furthermore, check for potential perfect multicollinearity between the billing variables (e.g., minutes vs. charge). Explain why including perfectly correlated variables or high-cardinality categorical variables (like state) can harm a logistic regression model."*

**Exam-Ready Written Reasoning:** Before running any complex algorithms, we must apply business logic to our variables. First, we must identify "disguised categories." For example, an `area_code` is a geographic label, not a quantity. If left as numeric, a model will incorrectly assume that an increase in area code changes the outcome linearly. Second, we must be wary of "high cardinality" variables like `state` (50 unique values); including it creates 49 dummy variables, which dramatically increases model variance and risks overfitting (the curse of dimensionality). Finally, telecom datasets often contain perfect multicollinearity (e.g., `total_day_charge` is derived directly from `total_day_minutes`). Including both in a logistic regression causes the design matrix to become singular, leading to unstable, highly inflated coefficient estimates. We must test these hypotheses before modeling.

```{r t2_business_logic}
# 1. TEST THE AREA CODE HYPOTHESIS
# We use unique() to see exactly how many different area codes exist in the dataset
# If there are only a few, it proves it is categorical, not a continuous measurement
unique(df_customer$area_code)

# Since we know area code is a category, we overwrite it as a factor
# This prevents the model from calculating a numeric slope for a geographic region
df_customer <- df_customer %>%
  mutate(area_code = as.factor(area_code))

# 2. TEST THE STATE CARDINALITY HYPOTHESIS
# We use length() combined with unique() to count how many distinct states are present
# High numbers (like 50) mean we might want to exclude this variable later to prevent overfitting
length(unique(df_customer$state))

# 3. TEST THE PERFECT MULTICOLLINEARITY HYPOTHESIS
# We use cor() to calculate the Pearson correlation coefficient between minutes and charge
# A value close to 1.0 means they contain the exact same information
cor(df_customer$total_day_minutes, df_customer$total_day_charge)

# We also check the evening minutes vs evening charge just to be sure the pattern holds
# If the correlation is 1, we must remove one of these columns before fitting a Logistic Regression
cor(df_customer$total_eve_minutes, df_customer$total_eve_charge)
```

### Target Distribution (Class Imbalance)

**Reasoning:** In binary classification (like churn, default, or fraud), identifying **class imbalance** is the most critical step. If the vast majority of customers do not churn, a naive model could achieve high overall accuracy simply by predicting "FALSE" for everyone (the accuracy paradox). This dictates that we must evaluate our models later using confusion matrices, sensitivity, and specificity, rather than just raw accuracy.

```{r t2_imbalance}
# Calculate the raw counts for each class in the target variable
table(df_customer$churn)

# Calculate the proportions for each class in the target variable
# This mathematically proves the severity of the class imbalance
prop.table(table(df_customer$churn))

# Create a bar plot to visually display the class imbalance
ggplot(df_customer, aes(x = churn, fill = churn)) +
  geom_bar(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Customer Churn",
       subtitle = "Checking for Class Imbalance",
       x = "Churn (Target Variable)",
       y = "Count of Customers")
```

*Conclusion:* The tables and plot confirm a severe class imbalance. Approximately **85.4% of customers did not churn (FALSE)**, while only **14.6% churned (TRUE)**. Any baseline model we build must beat an 85.4% accuracy to be considered useful, and we must be careful to tune our classification threshold later to catch the rare "TRUE" cases.

### Feature Relationships with the Target

**Reasoning:** Visualizing how specific predictors relate to the target variable helps us form initial hypotheses. If a predictor shows distinctly different distributions for churners vs. non-churners, it will likely be a highly significant variable in our models.

```{r t2_feature_explore}
# Explore a categorical predictor: International Plan vs Churn
# position = "fill" standardizes the bars to 100%, showing relative proportions
ggplot(df_customer, aes(x = international_plan, fill = churn)) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Churn Rate by International Plan Status",
       y = "Proportion of Customers",
       x = "Has International Plan?")

# Explore a continuous predictor: Total Day Minutes vs Churn
# geom_boxplot shows if the median and spread of minutes differ between classes
ggplot(df_customer, aes(x = churn, y = total_day_minutes, fill = churn)) +
  geom_boxplot(color = "black") +
  theme_minimal() +
  labs(title = "Total Day Minutes Distribution by Churn Status",
       x = "Customer Churn",
       y = "Total Day Minutes Used")
```

*Conclusion:* 1. **International Plan:** The stacked bar chart reveals a massive signal. Customers *with* an international plan churn at a drastically higher rate than those without one. This variable will be critical. 2. **Total Day Minutes:** The boxplot shows that customers who churned (TRUE) have a noticeably higher median for total day minutes used compared to those who stayed. High usage might be correlated with higher bills, driving churn.

Here is a summary of our shared understanding of the dataset so far, directly answering your questions:

### **1. Dataset Dimensions & The Outcome Variable**

-   **Dimensions:** We have 2,666 observations (rows) and 20 columns.

-   **Outcome Variable:** The outcome is `churn`, located in the 20th column. It is a binary factor (`FALSE` / `TRUE`).

-   **The Big Catch:** The outcome is severely imbalanced (85.4% FALSE vs. 14.6% TRUE). If we don't account for this later, our models will look highly accurate just by guessing "FALSE" every time.

### **2. The Multicollinearity Problem**

-   We mathematically proved that `total_day_minutes` and `total_day_charge` are perfectly correlated (`cor = 1`). The same is true for the evening variables (`0.9999998`).

-   By logical inference, this pattern will hold for night and international calls too. Telecom companies use a flat rate (e.g., \$0.10/min) to calculate charge from minutes.

-   **Why it matters:** If we put both into a Logistic Regression, the math behind the model (matrix inversion) will collapse or produce wildly inflated standard errors. We *must* drop the `charge` columns (or the `minutes` columns) before modeling.

### **3. Categorical Traps**

-   `area_code` is a category disguised as a number. We fixed this.

-   `state` has 51 unique values (high cardinality). If we keep it, a logistic regression will create 50 dummy variables. With only 388 actual churners (the TRUE cases), adding 50 variables will almost certainly cause **overfitting** (the model memorizes the training data but fails on new data). We should drop it.

### **4. Scatter Plots, Non-linearity, and Normality**

-   **Scatter plots against Y:** Because `churn` is categorical (0 or 1), a standard scatter plot doesn't work well (all dots just line up on the top and bottom). That is why we used a **boxplot** earlier for `total_day_minutes`. Density plots are also great here.

-   **Are we dealing with normal distributions, and does it matter?** This is a classic exam trap! **Logistic regression DOES NOT require the predictor variables (X) to be normally distributed.** It only assumes that the *log-odds* of the outcome relate linearly to X.

-   **Non-linear relationships:** While normality doesn't matter, non-linear relationships *do*. For example, `customer_service_calls`. A customer making 1 or 2 calls might be normal. But if they make 4, 5, or 6 calls, their probability of churning probably skyrockets exponentially. That is a non-linear threshold effect. Linear/Logistic models struggle with this, which is why Tree-based models (Random Forest) often win on churn datasets.

### EDA Master Conclusion for Task 2

1.  **Pre-processing:** We successfully converted `churn`, `international_plan`, and `voice_mail_plan` to factors, fulfilling the strict requirements of R's classification algorithms. There are no missing values.
2.  **Class Imbalance:** The 85/15 split dictates our evaluation strategy. We cannot rely on raw accuracy alone; we must inspect the Confusion Matrix to ensure we are actually catching the 15% who churn.
3.  **Signal Strength:** Predictors like `international_plan` and `total_day_minutes` already show strong visual separation between classes, indicating that a model should be able to find a valid decision boundary.

```{r dropCol}
# 1. DROP PROBLEMATIC COLUMNS
# We remove 'state' to prevent overfitting from 50 dummy variables.
# We remove all 'charge' columns to prevent perfect multicollinearity.
df_clean <- df_customer %>%
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge)

# Verify our new column count (should be 15 now)
ncol(df_clean)

# 2. INVESTIGATE NON-LINEARITY IN CUSTOMER SERVICE CALLS
# We will use a bar chart but split it by churn to see the "threshold" effect
ggplot(df_clean, aes(x = as.factor(customer_service_calls), fill = churn)) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Churn Rate by Number of Customer Service Calls",
       subtitle = "Looking for a non-linear threshold effect",
       x = "Number of Customer Service Calls",
       y = "Proportion of Customers")
```

```{r}
# VISUALIZING ALL NUMERIC VARIABLES VS CHURN
# We take our clean dataset and pipe it into the plotting workflow
df_clean %>%
  # 1. Select only the numeric columns AND the churn column
  # We can't scatter plot factors like 'international_plan' this way
  select(where(is.numeric), churn) %>%
  
  # 2. Pivot the data from "wide" to "long"
  # This collapses all numeric columns into two columns: 'variable' (the name) and 'value'
  # This is the exact trick needed to plot all variables in one go using facet_wrap
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  
  # 3. Build the plot mapping churn to X and the numeric values to Y
  ggplot(aes(x = churn, y = value, color = churn)) +
  
  # 4. Use geom_jitter instead of geom_point!
  # alpha = 0.3 makes dots transparent so we can see density (darker = more dots)
  # width = 0.2 controls how far the dots spread out horizontally
  geom_jitter(alpha = 0.3, width = 0.2) +
  
  # 5. Create a separate mini-plot for every single variable
  # scales = "free_y" allows each plot to have its own y-axis limits (since minutes and calls have different scales)
  facet_wrap(~variable, scales = "free_y") +
  
  # 6. Clean up the aesthetics
  theme_minimal() +
  theme(legend.position = "none") + # Hide legend since the x-axis already says TRUE/FALSE
  labs(title = "Jittered Scatter Plots: All Numeric Variables vs Churn",
       subtitle = "Looking for visual separation between TRUE and FALSE clusters",
       x = "Churn",
       y = "Variable Value")
```

In a binary classification task, the standard, highly effective way to evaluate continuous predictors against the target is using **Overlapping Density Plots** and **Grouped Summary Statistics**.

1.  **Density Plots (`geom_density`):** Instead of plotting individual dots, this draws a smooth curve showing where the data is concentrated. If the curves for TRUE and FALSE perfectly overlap, the variable is useless. If the curves are shifted (different peaks), the variable is a strong predictor.

2.  **Grouped Means:** A quick numerical table showing the average value of each variable for Churners vs. Non-Churners. This gives us hard numbers to back up our visual intuition.

### Evaluating Predictor Distributions (The Right Way)

**Possible Exam Question:** *"Evaluate how the continuous predictors differ between customers who churned and those who did not. Use appropriate visualizations and summary statistics to identify which numerical variables show the strongest potential for predicting churn. Why are density plots or boxplots preferred over scatter plots for this task?"*

**Exam-Ready Written Reasoning:** When exploring how continuous variables relate to a binary target, standard scatter plots suffer from severe overplotting (data points stacking on top of each other), making it impossible to see the true distribution. Furthermore, class imbalance can visually distort scatter plots. Instead, we use **Density Plots**. Density plots abstract away the raw point count and show the *shape* and *proportion* of the distribution. If the density curve for "Churn = TRUE" is significantly shifted horizontally compared to "Churn = FALSE", it indicates that the variable has high discriminatory power (it separates the classes well). Variables where the two curves perfectly overlap provide little to no information for a logistic regression or tree model. We back this visual analysis up by calculating the mean of each variable grouped by the target class.

```{r t2_density_and_means}
# 1. OVERLAPPING DENSITY PLOTS
# We use the same pivot_longer trick, but map 'value' to the x-axis and 'churn' to the fill color
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = churn)) +
  # geom_density creates smooth distribution curves. 
  # alpha = 0.5 makes them semi-transparent so we can see where they overlap
  geom_density(alpha = 0.5) +
  # facet_wrap creates a separate plot for each variable. scales = "free" lets x and y axes adjust per plot.
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Density Plots: Continuous Variables vs Customer Churn",
       subtitle = "Separation between curves indicates a strong predictor",
       x = "Variable Value",
       y = "Density (Concentration of Customers)")

# 2. GROUPED SUMMARY STATISTICS
# Visuals are great, but hard numbers are better for the exam text.
# We group the data by whether they churned or not.
df_clean %>%
  group_by(churn) %>%
  # summarise(across(...)) applies a function to all selected columns.
  # Here, we calculate the mean for all numeric columns, ignoring NAs just in case.
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  # We transpose the output (turn it sideways) using t() just to make it easier to read in the console
  t()
```

### 2.2 Deep Feature Analysis & Conclusions

**Possible Exam Question:** *"Based on your density plots and summary statistics, identify the variables that possess the highest discriminatory power for predicting churn. Conversely, identify the variables that appear to be useless. How does this inform your expectations for a logistic regression model?"*

**Exam-Ready Written Reasoning:** By abstracting away the raw data points into density curves and calculating grouped means, we can evaluate the "discriminatory power" of each predictor. Variables where the density curves perfectly overlap (such as `total_day_calls`, `total_eve_calls`, and `total_night_calls`) provide no mathematical separation between the classes. We expect these variables to have near-zero coefficients and high p-values in a baseline logistic regression.

Conversely, `total_day_minutes` and `customer_service_calls` exhibit strong rightward shifts for the "TRUE" class, confirmed by their higher grouped means (205.2 vs 175.1 minutes; 2.2 vs 1.45 calls). These are our "star" predictors. Finally, `number_vmail_messages` displays severe "zero-inflation" (a massive spike at exactly 0). Standard logistic regression assumes a linear relationship between the predictor and the log-odds of the outcome; this zero-inflated, non-linear distribution suggests that a highly flexible, tree-based model (like a Random Forest) might ultimately extract more signal from this data than a rigid linear boundary.

### THE WHY: Grounding the Extended EDA in Curriculum and Past Exams

**Why is this extensive EDA important in an exam setting?** If you look at the last five years of exams, Task 'a' and 'b' almost always give you 10 to 20 points simply for cleaning the data and proving you understand it *before* running a model. The professors are testing whether you blindly feed data into an algorithm or if you act like a real data scientist. If you feed a character column into `glmnet`, it crashes. If you feed perfectly correlated variables into `glm()`, the math breaks. If you rely on scatter plots for binary data, you miss the actual trends.

**What is the curriculum grounding?** This is deeply rooted in **ISLR Chapter 2 (Statistical Learning)** and **Chapter 4 (Classification)**. The curriculum emphasizes that the "Data Generating Process" must be understood. It explicitly warns about the Curse of Dimensionality (why we drop 50 states), Collinearity (why we drop charges), and the limitations of visualizing qualitative responses.

**How does this help you tomorrow?** By structuring your compendium chronologically—including the "mistakes" like the scatter plot—you create a mental map. Tomorrow, when you read "Use descriptive statistics to find potentially useful predictors" (like the 2025 and 2023 exams), you won't freeze. You will know to do: 1. `glimpse()` for types. 2. `prop.table()` for imbalance. 3. Density plots for continuous variables. 4. Stacked bar charts for categorical variables. 5. Grouped `summarise()` tables for hard numbers.

Here is the **Ultimate EDA Section** for your compendium. It includes the historical exam prompts as a reminder of *why* we do this, the "learning process" with the scatter plots, and the final clean dataset generation.

------------------------------------------------------------------------

## 2.0 Historical Exam Context & Strategy

Before touching the code, we must recognize what the professor typically asks for in Classification EDA. Look at these past exam prompts: \* **2025:** *"Remove variables that cannot plausibly be assumed to be available... Use descriptive statistics and graphs in order to find potentially useful predictors."* \* **2024:** *"If necessary, recode categorical variables as factors. Motivate your choices in words."* \* **2023:** *"Remove variables if you mean that they are unreasonable to include... present results as tables and graphs and comment."*

**Our Strategy:** We will not just plot data blindly. We will systematically verify data types, hunt for "illegal" or "unreasonable" variables (multicollinearity, extreme cardinality, data leakage), visualize the predictors iteratively (showing why some plots fail and others succeed), and output concrete summary tables.

## 2.1 Initial Import and Raw Data Inspection

**Possible Exam Question:** *"Load the dataset. Inspect its dimensions and data types. Check for missing values. Are all variables in the correct format for a classification model?"*

**Exam-Ready Written Reasoning:** The very first step in any machine learning pipeline is to import the data safely and inspect its raw structure. Many R algorithms will fail if categorical variables are stored as "character" or "numeric" rather than "factors". We must also check for missing values (`NAs`), as classification models like logistic regression or random forests will either crash or silently drop rows with missing data, distorting our results. We keep our original dataset untouched (`df_customer_raw`) so we have a reliable fallback.
>>>>>>> Stashed changes

```{r t2_raw_inspect}
# 1. LOAD DATA SAFELY
# read_csv is robust; clean_names() standardizes column names.
df_customer_raw <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()

# 2. DIMENSIONS AND TYPES
# glimpse() allows us to quickly see rows/cols and data types.
glimpse(df_customer_raw)

# 3. MISSING VALUES CHECK
# colSums(is.na()) counts missing values per column.
colSums(is.na(df_customer_raw))

# 4. RAW DESCRIPTIVE BOXPLOTS
# We want to see the scale and outliers of all numeric variables BEFORE cleaning.
df_customer_raw %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightgray", color = "black") +
  coord_flip() + # Turns the plot sideways so variable names are readable
  theme_minimal() +
  labs(title = "Raw Scale and Outliers of All Numeric Variables")
```

**Written analysis of findings/output:** The `glimpse()` output reveals 2,666 observations and 20 columns, and `colSums` confirms there are exactly 0 missing values (meaning no imputation is required). However, the raw boxplots reveal three critical insights that will dictate our preprocessing: 1. **Scale Disparity:** Variables are on vastly different mathematical scales (e.g., `total_day_minutes` spans up to \~350, while `customer_service_calls` is bounded near 10). If we use distance-based models (KNN) or regularized models (LASSO) later, we must ensure these features are standardized so large numbers do not artificially dominate the model. 2. **Presence of Outliers:** There are severe right-tail outliers in almost all usage metrics (the black dots beyond the whiskers). Standard logistic regression can be sensitive to outliers, hinting that robust tree-based models might be safer. 3. **The Categorical Trap:** The boxplot for `area_code` looks highly unnatural, clustered entirely around the 400-500 range. This visually flags that `area_code` is not a continuous measurement, but rather a discrete geographic label (e.g., 408, 415, 510) that must be converted to a categorical factor in our next step.

## Hunting for Traps: Checking Business Logic

**Possible Exam Question:** *"Analyze the variables in the raw dataset. Test for high cardinality, disguised categories, and perfect multicollinearity. Explain your findings."*

**Exam-Ready Written Reasoning:** Before modifying the dataset, we must run diagnostic tests to find "illegal" or problematic variables. 1. **Disguised Categories:** We suspect `area_code` is a geographic label, not a math variable. We will check how many unique values it has. 2. **High Cardinality:** We suspect `state` has too many unique text values, which would create an explosion of dummy variables and risk severe overfitting. We will count them. 3. **Perfect Multicollinearity:** We suspect telecom billing uses a flat rate (Minutes \* Rate = Charge). We will test the correlation; if it is near 1.0, it will make our design matrix singular and crash a logistic regression.

```{r t2_business_tests}
# 1. TEST AREA CODE (Disguised Category Trap)
unique(df_customer_raw$area_code)

# 2. TEST STATE (High Cardinality Trap)
length(unique(df_customer_raw$state))

# 3. TEST MULTICOLLINEARITY (Minutes vs Charge Trap)
cor(df_customer_raw$total_day_minutes, df_customer_raw$total_day_charge)
```

**Written analysis of findings/output:** The tests confirm our three suspicions. First, `area_code` outputs only `415, 408, 510`, proving it is a discrete category, not a continuous number. Second, `state` outputs `51`; leaving it in would force the model to create 50 dummy variables, almost guaranteeing overfitting. Third, the correlation between day minutes and day charge is exactly `1`. Including both would cause perfect multicollinearity.

## Executing the Cleaning Pipeline

**Possible Exam Question:** *"Based on your diagnostic tests, clean the dataset. Remove redundant or harmful variables and ensure all categorical variables are correctly formatted as factors for classification."*

**Exam-Ready Written Reasoning:** Because our diagnostic tests proved the presence of high cardinality and perfect multicollinearity, we must drop the `state` variable and all four `charge` columns. Furthermore, R's classification algorithms (like logistic regression and random forests) strictly require the outcome variable and categorical predictors to be stored as `factors`. We will execute these changes to create our final, safe modeling dataset.

```{r t2_clean_execution}
# EXECUTE THE CLEANING PIPELINE based strictly on prior findings
df_clean <- df_customer_raw %>%
  # Drop state (overfitting) and charge columns (perfect multicollinearity)
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  # Convert target and categorical variables to factors
  mutate(
    churn = as.factor(churn),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

# Verify the final clean dimensions
dim(df_clean)
```

**Written analysis of findings/output:** The cleaning pipeline successfully executed. The output `2666 15` confirms that our row count remains intact (no data was lost), but our column count has been successfully reduced from 20 to 15 by stripping out the 5 harmful variables. `df_clean` is now mathematically safe for modeling.

## Target Distribution (Class Imbalance)

**Possible Exam Question:** *"Analyze the distribution of the target variable. Explain why this distribution dictates how you will evaluate your classification models later."*

**Exam-Ready Written Reasoning:** In binary classification, verifying the balance of the target classes is the most critical pre-modeling step. If a vast majority of the dataset belongs to one class, a naive model can achieve high overall accuracy simply by guessing the majority class every time (the "Accuracy Paradox"). Identifying severe imbalance forces us to abandon raw accuracy as our primary metric and instead rely on the Confusion Matrix, Sensitivity, and Specificity.

```{r t2_imbalance}
# Calculate exact proportions for the target variable
prop.table(table(df_clean$churn))

# Create a clean bar plot to visually display the imbalance
ggplot(df_clean, aes(x = churn, fill = churn)) +
  geom_bar(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Customer Churn", subtitle = "Checking for Class Imbalance")
```

**Written analysis of findings/output:** The output reveals "The Big Catch": severe class imbalance. Approximately 85.4% of the customers belong to the negative class (`FALSE`), while only 14.6% belong to the positive class (`TRUE`). Because our operational goal is to identify churners, any model we build must be evaluated on its ability to correctly catch that 14.6% (Sensitivity). A model that just predicts "FALSE" for everyone achieves 85.4% accuracy but is entirely useless for the business.

## 2.4 Visualizing Predictors: The Learning Process (Scatter vs. Density)

**Possible Exam Question:** *"Use graphs to find potentially useful predictors for churn. Discuss the limitations of scatter plots for binary outcomes and provide a superior alternative. Are we dealing with normal distributions, and does it matter?"*

**Exam-Ready Written Reasoning:** To find useful predictors, we must visualize them against the target. A naive attempt is a jittered scatter plot. However, plotting continuous variables against a binary target results in severe overplotting, and class imbalance heavily distorts the visual. The mathematically superior approach is using **Overlapping Density Plots**, which abstract away point counts and display the normalized *shape* of the distribution. If a variable is a strong predictor, its density curve for the "TRUE" class will be visibly shifted compared to the "FALSE" class. Note: This is a classic exam trap; Logistic regression DOES NOT require predictor variables (X) to be normally distributed, it only assumes the log-odds relate linearly to X.

```{r t2_visual_learning}
# ATTEMPT 1: JITTERED SCATTER PLOTS (The problematic approach)
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = churn, y = value, color = churn)) +
  geom_jitter(alpha = 0.3, width = 0.2) + # Jitter adds horizontal noise, but overplotting is still severe
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Attempt 1: Jittered Scatter Plots (Suffers from Overplotting)")

# ATTEMPT 2: OVERLAPPING DENSITY PLOTS (The optimal approach for continuous variables)
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = churn)) +
  geom_density(alpha = 0.5) + # alpha=0.5 makes curves transparent to see overlap
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Attempt 2: Density Plots (Clear visualization of discriminatory power)")

# ATTEMPT 3: CATEGORICAL PREDICTORS (Stacked Bar Charts)
# For categorical predictors, we use position = "fill" to see standardized proportions
ggplot(df_clean, aes(x = international_plan, fill = churn)) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Churn Rate by International Plan", y = "Proportion of Customers")
```

**Written analysis of findings/output:** The scatter plot proved ineffective due to overplotting. The density plots, however, revealed clear patterns. Variables like `total_day_minutes` show a distinct rightward shift for the blue curve (`TRUE`), indicating that higher daily minutes correlate with higher churn. Conversely, `total_day_calls` shows perfectly overlapping curves, providing no mathematical separation between classes. Furthermore, the stacked bar chart reveals that the proportion of churn is drastically higher for customers who possess an `international_plan`.

## 2.5 Tabular Summary and EDA Master Conclusion

**Possible Exam Question:** *"Present your descriptive findings as a table. Based on all your EDA, summarize which variables have the highest discriminatory power and which appear useless. How might non-linear variables affect model choice?"*

**Exam-Ready Written Reasoning:** Visual intuition must always be backed up by hard mathematical numbers in an exam. By grouping the dataset by churn status and calculating the mean for all numeric columns, we establish a tabular baseline to definitively prove which variables separate the classes.

```{r t2_summary_table}
# CREATE A GROUPED SUMMARY TABLE
df_clean %>%
  group_by(churn) %>%
  # Calculate the mean for every numeric column
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  t() # Transpose for readability in the console
```

<<<<<<< Updated upstream
**Written analysis of findings/output (Master EDA Conclusion):** The grouped means table mathematically confirms our visual findings: 1. **Star Predictors:** The mean for `total_day_minutes` jumps substantially from 175.1 (FALSE) to 205.2 (TRUE), and `customer_service_calls` jumps from 1.45 to 2.2. These will be the primary drivers in our models. 2. **Useless Predictors:** `total_day_calls`, `total_eve_calls`, and `total_night_calls` have nearly identical means across both classes. We expect these to have high p-values and near-zero coefficients in a baseline logistic regression. 3. **Non-Linearity:** `number_vmail_messages` is severely zero-inflated (massive spike at 0 in the density plot), and `customer_service_calls` likely has a threshold effect (churn spikes exponentially after a few calls). Because standard logistic regression assumes a strictly linear relationship between predictors and log-odds, this non-linear structure strongly suggests that a flexible, tree-based model (like a Random Forest) might ultimately extract more signal from this dataset.
=======
# Task 2: Classification (Predicting Churn)

## 2.0 Historical Exam Context & Strategy

Before touching the code, we must recognize the expected steps for classification EDA based on past exams (2023-2025): \* *"Remove variables that cannot plausibly be assumed to be available..."* \* *"If necessary, recode categorical variables as factors. Motivate your choices..."* \* *"Use descriptive statistics and graphs in order to find potentially useful predictors."*

**Our Strategy:** We will follow the standard statistical learning pipeline outlined in ISLR: Data Import, Summary Statistics, Collinearity Analysis, Data Preprocessing (Feature Selection), Response Distribution Check, and Bivariate Analysis. We follow a strict `Before ➔ Code ➔ After` structure.

## 2.1 Data Import and Summary Statistics

**Possible Exam Question:** *"Load the dataset. Output descriptive statistics (mean, median, standard deviation) for all numeric variables. What does this tell you about the scale of the predictors?"*

**Exam-Ready Written Reasoning:** The initial step in statistical learning is inspecting the raw data structure and calculating descriptive statistics. Many R algorithms require the response variable to be a `factor`. Furthermore, calculating the mean, standard deviation, and range for continuous variables tells us if we need to standardize the predictors later. Algorithms that rely on distance metrics (KNN) or penalty terms (Ridge/LASSO) are highly sensitive to the scale of predictors. We keep our original dataset untouched (`df_customer_raw`) as a reliable fallback.

```{r t2_import_and_stats}
# 1. LOAD DATA SAFELY
df_customer_raw <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()

# 2. MISSING VALUES CHECK
# colSums(is.na()) confirms if imputation is necessary
colSums(is.na(df_customer_raw))

# 3. DESCRIPTIVE STATISTICS
# We use summary() to get Min, 1st Qu, Median, Mean, 3rd Qu, and Max
summary(df_customer_raw %>% select(where(is.numeric)))

# We also calculate the Standard Deviation for all numeric columns to observe variance
df_customer_raw %>%
  summarise(across(where(is.numeric), sd)) %>%
  t() %>%
  as.data.frame() %>%
  rename(Standard_Deviation = V1)
```

**Written analysis of findings/output:** There are exactly 0 missing values (`NAs`), meaning no imputation is required. The descriptive statistics mathematically prove severe scale disparities among predictors. For example, `total_day_minutes` has a mean of \~179 and a standard deviation of \~54, whereas `customer_service_calls` has a mean of \~1.5 and a standard deviation of \~1.3. If we utilize regularized methods or distance-based classifiers, we must standardize these variables so the large variance of minutes does not artificially dominate the model.

## 2.2 Correlation Analysis and Collinearity

**Possible Exam Question:** *"Construct a correlation matrix for the numeric predictors. Identify any pairs with high collinearity and explain why including them in a logistic regression model is problematic."*

**Exam-Ready Written Reasoning:** Collinearity refers to the situation where two or more predictor variables are closely related to one another. According to ISLR, collinearity causes the standard errors of coefficient estimates to grow, reducing the statistical power of the model (variance inflation). In extreme cases (perfect collinearity), the design matrix becomes singular, and logistic regression cannot compute unique coefficient estimates. We use a correlation heatmap to detect and subsequently eliminate highly correlated features.

```{r t2_collinearity}
# 1. CALCULATE CORRELATION MATRIX
# We select only numeric columns and compute the Pearson correlation matrix
cor_matrix <- cor(df_customer_raw %>% select(where(is.numeric)))

# 2. VISUALIZE WITH A HEATMAP
# We pivot the correlation matrix to a long format for ggplot2
cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "var1") %>%
  pivot_longer(cols = -var1, names_to = "var2", values_to = "correlation") %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  # Red means positive correlation, Blue means negative, White means zero
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap of Numeric Predictors",
       subtitle = "Dark red squares outside the diagonal indicate severe collinearity")
```

**Written analysis of findings/output:** The heatmap clearly identifies four instances of perfect positive collinearity (correlation = 1.0). `total_day_minutes` is perfectly correlated with `total_day_charge`, and this pattern repeats for evening, night, and international calls. This is a structural artifact of telecom billing (minutes dictate charge). We must remove all `charge` variables to prevent multicollinearity from destabilizing our logistic regression model. No other numeric predictors show significant correlation with one another.

## 2.3 Data Preprocessing and Feature Selection

**Possible Exam Question:** *"Based on your collinearity analysis and variable inspections, preprocess the data. Remove invalid predictors, recode categorical variables, and output the final list of features to be used in modeling."*

**Exam-Ready Written Reasoning:** Data preprocessing transforms raw data into a mathematically safe format for statistical learning. First, we remove the four `charge` columns due to perfect collinearity. Second, we must address categorical variables. `area_code` contains only three discrete numbers, proving it is a categorical variable, not continuous. `state` contains 51 distinct text values; if included, it would generate 50 dummy variables, risking the Curse of Dimensionality and severe overfitting given our sample size. We drop `state` and coerce the remaining categorical features into `factor` format.

```{r t2_preprocessing}
# EXECUTE DATA PREPROCESSING
df_clean <- df_customer_raw %>%
  # Remove high cardinality and collinear variables
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  # Format target and categorical predictors as factors
  mutate(
    churn = as.factor(churn),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

# Output the final list of variables available for modeling
names(df_clean)
```

**Written analysis of findings/output:** The preprocessing pipeline successfully removed 5 harmful variables. The resulting dataset, `df_clean`, contains 15 variables. The single response variable is `churn`. The 14 available predictors are: `account_length`, `area_code`, `international_plan`, `voice_mail_plan`, `number_vmail_messages`, `total_day_minutes`, `total_day_calls`, `total_eve_minutes`, `total_eve_calls`, `total_night_minutes`, `total_night_calls`, `total_intl_minutes`, `total_intl_calls`, and `customer_service_calls`.

## 2.4 Response Variable Distribution and Prior Probabilities

**Possible Exam Question:** *"Calculate the prior probabilities of the response variable classes. How does this affect model evaluation?"*

**Exam-Ready Written Reasoning:** In classification, the prior probability is the baseline proportion of each class in the dataset. If there is a severe class imbalance, a model can achieve a low overall error rate simply by assigning every observation to the majority class (the Accuracy Paradox). Consequently, overall accuracy is a misleading metric for imbalanced data. This structural reality dictates that we must evaluate subsequent models using the Confusion Matrix, specifically focusing on Sensitivity (the True Positive Rate).

```{r t2_prior_probabilities}
# Calculate prior probabilities (proportions) of the response classes
prop.table(table(df_clean$churn))

# Visualize the response distribution
ggplot(df_clean, aes(x = churn, fill = churn)) +
  geom_bar(color = "black") +
  theme_minimal() +
  labs(title = "Prior Probabilities of the Response Variable (Churn)")
```

**Written analysis of findings/output:** The prior probabilities reveal a severe class imbalance: 85.4% of the observations belong to the negative class (`FALSE`), and 14.6% belong to the positive class (`TRUE`). Because our objective is to identify customers who leave (the 14.6%), any model we build must be strictly evaluated on its Sensitivity. A baseline model predicting `FALSE` universally would yield 85.4% accuracy but zero business utility.

## 2.5 Bivariate Analysis: Predictors vs. Response

**Possible Exam Question:** *"Perform a bivariate analysis visualizing the relationship between the predictors and the response. Explain your choice of visualization for continuous predictors."*

**Exam-Ready Written Reasoning:** To identify features with high discriminatory power, we conduct a bivariate analysis. While scatter plots are common in regression, plotting a continuous predictor against a binary response causes severe overplotting. As recommended in ISLR, we use Density Estimation to visualize the distribution of continuous predictors conditional on the response class. If the density curve for the `TRUE` class is significantly shifted from the `FALSE` class, the feature strongly influences the probability of the response. (Note: Logistic regression does not assume predictors are normally distributed; it assumes a linear relationship between predictors and the log-odds of the response).

```{r t2_bivariate}
# DENSITY ESTIMATION FOR CONTINUOUS PREDICTORS
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = churn)) +
  geom_density(alpha = 0.5) + 
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Conditional Density Estimation of Continuous Predictors")

# STANDARDIZED BAR CHARTS FOR CATEGORICAL PREDICTORS
ggplot(df_clean, aes(x = international_plan, fill = churn)) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Response Probability Conditional on International Plan", y = "Proportion")
```

**Written analysis of findings/output:** The density plots indicate that `total_day_minutes` and `customer_service_calls` exhibit distinct distributional shifts conditional on the response, marking them as strong potential predictors. Conversely, variables like `total_day_calls` show perfectly overlapping conditional densities, suggesting they possess no discriminatory power. Furthermore, `number_vmail_messages` displays severe zero-inflation, implying a non-linear relationship. The bar chart indicates the probability of churn is significantly higher given the presence of an `international_plan`.

## 2.6 Feature Evaluation and Summary Statistics

**Possible Exam Question:** *"Summarize the relationship between your predictors and the response quantitatively. Which variables do you expect to be statistically significant in a logistic regression?"*

**Exam-Ready Written Reasoning:** Visual bivariate analysis must be substantiated with summary statistics. By computing the conditional means of the continuous predictors grouped by the response class, we establish mathematical confirmation of discriminatory power. Predictors with large differences in conditional means are expected to yield statistically significant coefficients (low p-values) in a logistic regression model.

```{r t2_conditional_means}
# Calculate conditional means for continuous predictors
df_clean %>%
  group_by(churn) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  t() 
```

**Written analysis of findings/output:** The conditional means confirm our visual analysis. `total_day_minutes` increases from a mean of 175.1 for non-churners to 205.2 for churners, and `customer_service_calls` increases from 1.45 to 2.2. These features provide strong separation. In contrast, `total_day_calls` and `total_night_calls` show virtually identical conditional means (\~100 calls), confirming they provide negligible information for classification. We expect these non-discriminatory variables to be statistically insignificant in our baseline models.


