---
title: "ExamPrep2"
format: html
editor: visual
---

**General Setup and Package Loading**

Before tackling any specific task, we load the standard suite of BAN404 packages. We use `tidyverse` for data manipulation and visualization, `janitor` for safe column naming, and the core statistical learning packages (`glmnet`, `gam`, `tree`, `randomForest`, `gbm`) so we don't have to interrupt our workflow later.

```{r setup_and_load}
# 0. Load necessary libraries
library(tidyverse)
library(janitor)
library(glmnet)       # Regularization (Ridge/LASSO)
library(gam)          # Generalized Additive Models (Non-linearities)
library(tree)         # Decision Trees
library(randomForest) # Bagging and Random Forests
library(gbm)          # Boosting

# 1. Load the datasets safely
# show_col_types = FALSE keeps the exam output clean. 
# clean_names() ensures no weird spaces or characters in column names.
df_task1 <- read_csv("task1_data.csv", show_col_types = FALSE) %>% clean_names()
df_customer <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()
```

# Task 2: Binary Classification (Predicting Churn)

## Historical Exam Context & Strategy

Before touching the code, we must recognize what the professor typically asks for in Classification EDA. Past exams (2023-2025) repeatedly ask to: \* *"Remove variables that cannot plausibly be assumed to be available..."* \* *"If necessary, recode categorical variables as factors. Motivate your choices..."* \* *"Use descriptive statistics and graphs in order to find potentially useful predictors."*

**Our Strategy:** We will systematically verify data types, hunt for "illegal" variables, clean the dataset, evaluate class imbalance, and iteratively visualize the predictors to find the best approach. We follow a strict `Before ➔ Code ➔ After` structure.

## Initial Import and Raw Data Inspection

**Possible Exam Question:** *"Load the dataset. Inspect its dimensions and data types. Check for missing values. Create a visual summary (e.g., boxplots) of all continuous variables in their raw state to identify differences in scale."*

**Exam-Ready Written Reasoning:** The very first step in any machine learning pipeline is to import the data safely and inspect its raw structure. Classification algorithms in R require the target variable to be a "factor". We must also check for missing values (`NAs`), as algorithms will either crash or drop rows, distorting our results. Finally, checking the raw scale of continuous variables via boxplots tells us if we need to standardize the data later for distance-based models (like KNN) or regularized models (like LASSO). We keep our original dataset untouched (`df_customer_raw`) as a reliable fallback.

```{r t2_raw_inspect}
# 1. LOAD DATA SAFELY
# read_csv is robust; clean_names() standardizes column names.
df_customer_raw <- read_csv("customer_data.csv", show_col_types = FALSE) %>% clean_names()

# 2. DIMENSIONS AND TYPES
# glimpse() allows us to quickly see rows/cols and data types.
glimpse(df_customer_raw)

# 3. MISSING VALUES CHECK
# colSums(is.na()) counts missing values per column.
colSums(is.na(df_customer_raw))

# 4. RAW DESCRIPTIVE BOXPLOTS
# We want to see the scale and outliers of all numeric variables BEFORE cleaning.
df_customer_raw %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightgray", color = "black") +
  coord_flip() + # Turns the plot sideways so variable names are readable
  theme_minimal() +
  labs(title = "Raw Scale and Outliers of All Numeric Variables")
```

**Written analysis of findings/output:** The `glimpse()` output reveals 2,666 observations and 20 columns, and `colSums` confirms there are exactly 0 missing values (meaning no imputation is required). However, the raw boxplots reveal three critical insights that will dictate our preprocessing: 1. **Scale Disparity:** Variables are on vastly different mathematical scales (e.g., `total_day_minutes` spans up to \~350, while `customer_service_calls` is bounded near 10). If we use distance-based models (KNN) or regularized models (LASSO) later, we must ensure these features are standardized so large numbers do not artificially dominate the model. 2. **Presence of Outliers:** There are severe right-tail outliers in almost all usage metrics (the black dots beyond the whiskers). Standard logistic regression can be sensitive to outliers, hinting that robust tree-based models might be safer. 3. **The Categorical Trap:** The boxplot for `area_code` looks highly unnatural, clustered entirely around the 400-500 range. This visually flags that `area_code` is not a continuous measurement, but rather a discrete geographic label (e.g., 408, 415, 510) that must be converted to a categorical factor in our next step.

## Hunting for Traps: Checking Business Logic

**Possible Exam Question:** *"Analyze the variables in the raw dataset. Test for high cardinality, disguised categories, and perfect multicollinearity. Explain your findings."*

**Exam-Ready Written Reasoning:** Before modifying the dataset, we must run diagnostic tests to find "illegal" or problematic variables. 1. **Disguised Categories:** We suspect `area_code` is a geographic label, not a math variable. We will check how many unique values it has. 2. **High Cardinality:** We suspect `state` has too many unique text values, which would create an explosion of dummy variables and risk severe overfitting. We will count them. 3. **Perfect Multicollinearity:** We suspect telecom billing uses a flat rate (Minutes \* Rate = Charge). We will test the correlation; if it is near 1.0, it will make our design matrix singular and crash a logistic regression.

```{r t2_business_tests}
# 1. TEST AREA CODE (Disguised Category Trap)
unique(df_customer_raw$area_code)

# 2. TEST STATE (High Cardinality Trap)
length(unique(df_customer_raw$state))

# 3. TEST MULTICOLLINEARITY (Minutes vs Charge Trap)
cor(df_customer_raw$total_day_minutes, df_customer_raw$total_day_charge)
```

**Written analysis of findings/output:** The tests confirm our three suspicions. First, `area_code` outputs only `415, 408, 510`, proving it is a discrete category, not a continuous number. Second, `state` outputs `51`; leaving it in would force the model to create 50 dummy variables, almost guaranteeing overfitting. Third, the correlation between day minutes and day charge is exactly `1`. Including both would cause perfect multicollinearity.

## Executing the Cleaning Pipeline

**Possible Exam Question:** *"Based on your diagnostic tests, clean the dataset. Remove redundant or harmful variables and ensure all categorical variables are correctly formatted as factors for classification."*

**Exam-Ready Written Reasoning:** Because our diagnostic tests proved the presence of high cardinality and perfect multicollinearity, we must drop the `state` variable and all four `charge` columns. Furthermore, R's classification algorithms (like logistic regression and random forests) strictly require the outcome variable and categorical predictors to be stored as `factors`. We will execute these changes to create our final, safe modeling dataset.

```{r t2_clean_execution}
# EXECUTE THE CLEANING PIPELINE based strictly on prior findings
df_clean <- df_customer_raw %>%
  # Drop state (overfitting) and charge columns (perfect multicollinearity)
  select(-state, -total_day_charge, -total_eve_charge, -total_night_charge, -total_intl_charge) %>%
  # Convert target and categorical variables to factors
  mutate(
    churn = as.factor(churn),
    area_code = as.factor(area_code),
    international_plan = as.factor(international_plan),
    voice_mail_plan = as.factor(voice_mail_plan)
  )

# Verify the final clean dimensions
dim(df_clean)
```

**Written analysis of findings/output:** The cleaning pipeline successfully executed. The output `2666 15` confirms that our row count remains intact (no data was lost), but our column count has been successfully reduced from 20 to 15 by stripping out the 5 harmful variables. `df_clean` is now mathematically safe for modeling.

## Target Distribution (Class Imbalance)

**Possible Exam Question:** *"Analyze the distribution of the target variable. Explain why this distribution dictates how you will evaluate your classification models later."*

**Exam-Ready Written Reasoning:** In binary classification, verifying the balance of the target classes is the most critical pre-modeling step. If a vast majority of the dataset belongs to one class, a naive model can achieve high overall accuracy simply by guessing the majority class every time (the "Accuracy Paradox"). Identifying severe imbalance forces us to abandon raw accuracy as our primary metric and instead rely on the Confusion Matrix, Sensitivity, and Specificity.

```{r t2_imbalance}
# Calculate exact proportions for the target variable
prop.table(table(df_clean$churn))

# Create a clean bar plot to visually display the imbalance
ggplot(df_clean, aes(x = churn, fill = churn)) +
  geom_bar(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Customer Churn", subtitle = "Checking for Class Imbalance")
```

**Written analysis of findings/output:** The output reveals "The Big Catch": severe class imbalance. Approximately 85.4% of the customers belong to the negative class (`FALSE`), while only 14.6% belong to the positive class (`TRUE`). Because our operational goal is to identify churners, any model we build must be evaluated on its ability to correctly catch that 14.6% (Sensitivity). A model that just predicts "FALSE" for everyone achieves 85.4% accuracy but is entirely useless for the business.

## 2.4 Visualizing Predictors: The Learning Process (Scatter vs. Density)

**Possible Exam Question:** *"Use graphs to find potentially useful predictors for churn. Discuss the limitations of scatter plots for binary outcomes and provide a superior alternative. Are we dealing with normal distributions, and does it matter?"*

**Exam-Ready Written Reasoning:** To find useful predictors, we must visualize them against the target. A naive attempt is a jittered scatter plot. However, plotting continuous variables against a binary target results in severe overplotting, and class imbalance heavily distorts the visual. The mathematically superior approach is using **Overlapping Density Plots**, which abstract away point counts and display the normalized *shape* of the distribution. If a variable is a strong predictor, its density curve for the "TRUE" class will be visibly shifted compared to the "FALSE" class. Note: This is a classic exam trap; Logistic regression DOES NOT require predictor variables (X) to be normally distributed, it only assumes the log-odds relate linearly to X.

```{r t2_visual_learning}
# ATTEMPT 1: JITTERED SCATTER PLOTS (The problematic approach)
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = churn, y = value, color = churn)) +
  geom_jitter(alpha = 0.3, width = 0.2) + # Jitter adds horizontal noise, but overplotting is still severe
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Attempt 1: Jittered Scatter Plots (Suffers from Overplotting)")

# ATTEMPT 2: OVERLAPPING DENSITY PLOTS (The optimal approach for continuous variables)
df_clean %>%
  select(where(is.numeric), churn) %>%
  pivot_longer(cols = -churn, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = churn)) +
  geom_density(alpha = 0.5) + # alpha=0.5 makes curves transparent to see overlap
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Attempt 2: Density Plots (Clear visualization of discriminatory power)")

# ATTEMPT 3: CATEGORICAL PREDICTORS (Stacked Bar Charts)
# For categorical predictors, we use position = "fill" to see standardized proportions
ggplot(df_clean, aes(x = international_plan, fill = churn)) +
  geom_bar(position = "fill", color = "black") +
  theme_minimal() +
  labs(title = "Churn Rate by International Plan", y = "Proportion of Customers")
```

**Written analysis of findings/output:** The scatter plot proved ineffective due to overplotting. The density plots, however, revealed clear patterns. Variables like `total_day_minutes` show a distinct rightward shift for the blue curve (`TRUE`), indicating that higher daily minutes correlate with higher churn. Conversely, `total_day_calls` shows perfectly overlapping curves, providing no mathematical separation between classes. Furthermore, the stacked bar chart reveals that the proportion of churn is drastically higher for customers who possess an `international_plan`.

## 2.5 Tabular Summary and EDA Master Conclusion

**Possible Exam Question:** *"Present your descriptive findings as a table. Based on all your EDA, summarize which variables have the highest discriminatory power and which appear useless. How might non-linear variables affect model choice?"*

**Exam-Ready Written Reasoning:** Visual intuition must always be backed up by hard mathematical numbers in an exam. By grouping the dataset by churn status and calculating the mean for all numeric columns, we establish a tabular baseline to definitively prove which variables separate the classes.

```{r t2_summary_table}
# CREATE A GROUPED SUMMARY TABLE
df_clean %>%
  group_by(churn) %>%
  # Calculate the mean for every numeric column
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  t() # Transpose for readability in the console
```

**Written analysis of findings/output (Master EDA Conclusion):** The grouped means table mathematically confirms our visual findings: 1. **Star Predictors:** The mean for `total_day_minutes` jumps substantially from 175.1 (FALSE) to 205.2 (TRUE), and `customer_service_calls` jumps from 1.45 to 2.2. These will be the primary drivers in our models. 2. **Useless Predictors:** `total_day_calls`, `total_eve_calls`, and `total_night_calls` have nearly identical means across both classes. We expect these to have high p-values and near-zero coefficients in a baseline logistic regression. 3. **Non-Linearity:** `number_vmail_messages` is severely zero-inflated (massive spike at 0 in the density plot), and `customer_service_calls` likely has a threshold effect (churn spikes exponentially after a few calls). Because standard logistic regression assumes a strictly linear relationship between predictors and log-odds, this non-linear structure strongly suggests that a flexible, tree-based model (like a Random Forest) might ultimately extract more signal from this dataset.
