---
title: "BAN404 Prospect Snippets"
subtitle: "Broader snippet ideas from old exam themes and compendiums"
author: "Exam preparation"
date: "2026"
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    fontsize: 10pt
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
  error: true
---

# Prospect snippets (not yet merged into ExamPrep naming convention)

These are **candidate snippet types** found from recurring old-exam patterns in `Previous Exam/` compendiums/transcripts.

```{r}
# Shared setup so all prospect chunks run in one file
suppressPackageStartupMessages({
  library(tidyverse)
})
```

## 1) Local linear regression helper (K-nearest local fit)

Use when exam gives a custom function `f(x0, x, y, K)` or asks to explain/implement local regression beyond plain KNN.

**How to adapt:**
- Change `X_VEC`, `Y_VEC`, `X0`, `K_VAL`.
- Keep `K_VAL < length(X_VEC)`.

```{r}
# Prospect: local linear regression at a query point x0
X_VEC <- c(DATA_HERE)
Y_VEC <- c(DATA_HERE)
X0 <- X0_VALUE
K_VAL <- K_VALUE

local_lin_pred <- function(x0, x, y, k) {
  # Find indices of k closest points to x0
  nn_idx <- order(abs(x - x0))[1:k]
  # Fit local linear model on neighbors only
  fit <- lm(y ~ x, data = tibble(x = x[nn_idx], y = y[nn_idx]))
  # Predict at x0
  as.numeric(predict(fit, newdata = tibble(x = x0)))
}

local_lin_pred(X0, X_VEC, Y_VEC, K_VAL)
```

Interpretation template: The local prediction at `x0=[X0]` is `[PRED]` using `K=[K_VAL]`; this reflects a **local** trend rather than one global slope.

## 2) Generic LOOCV loop for tuning K/lambda

Use when exam asks LOOCV for hyperparameter selection (seen repeatedly for K and lambda questions).

**How to adapt:**
- Replace `FIT_FN` and `PRED_FN` logic for your model type.
- Edit `GRID_VALUES` to tested candidates.

```{r}
# Prospect: generic LOOCV tuner
DF <- tibble(x = X_VAR, y = Y_VAR) # replace with your data frame/columns
GRID_VALUES <- c(1, 3, 5, 7, 9)

loo_mse <- map_dbl(GRID_VALUES, function(g) {
  preds <- map_dbl(seq_len(nrow(DF)), function(i) {
    train <- DF[-i, ]
    test  <- DF[i, ]
    # Example model: local polynomial degree = g placeholder
    fit <- lm(y ~ poly(x, degree = g, raw = TRUE), data = train)
    as.numeric(predict(fit, newdata = test))
  })
  mean((DF$y - preds)^2)
})

tibble(param = GRID_VALUES, loo_mse = loo_mse) %>% arrange(loo_mse)
```

Interpretation template: LOOCV is minimized at `[BEST_PARAM]`, so that is the preferred complexity/tuning value.

## 3) Threshold tuning by CV for logistic predictions

Use when exam says “0.5 is not optimal; find better threshold with cross-validation”.

**How to adapt:**
- Change `TARGET_BIN`, `FORMULA_HERE`, `THRESH_GRID`.
- Replace objective metric if exam prioritizes sensitivity/specificity/cost.

```{r}
# Prospect: threshold CV for logistic model
set.seed(123)
DF <- DATAFRAME_HERE
TARGET_BIN <- "TARGET_VAR"
FORMULA_HERE <- as.formula("TARGET_VAR ~ X1 + X2 + X3")
THRESH_GRID <- seq(0.1, 0.9, by = 0.05)
K <- 5

fold_id <- sample(rep(1:K, length.out = nrow(DF)))

score_threshold <- function(thr) {
  fold_scores <- map_dbl(1:K, function(k) {
    tr <- DF[fold_id != k, ]
    va <- DF[fold_id == k, ]

    fit <- glm(FORMULA_HERE, data = tr, family = binomial())
    p <- predict(fit, newdata = va, type = "response")
    yhat <- if_else(p >= thr, 1, 0)
    y <- as.integer(va[[TARGET_BIN]])

    # Balanced accuracy objective (good for imbalance)
    sens <- mean(yhat[y == 1] == 1)
    spec <- mean(yhat[y == 0] == 0)
    mean(c(sens, spec), na.rm = TRUE)
  })
  mean(fold_scores)
}

res <- tibble(threshold = THRESH_GRID, score = map_dbl(THRESH_GRID, score_threshold))
res %>% arrange(desc(score))
```

Interpretation template: The best threshold is `[THR]` because it maximizes `[METRIC]` instead of defaulting to 0.5.

## 4) Full classification metric table from confusion matrix

Use when exam asks for richer evaluation than plain accuracy.

**How to adapt:**
- Set positive class level in `POS_LABEL`.
- Use your predicted labels/probabilities.

```{r}
# Prospect: confusion-matrix derived metrics
ACTUAL <- factor(ACTUAL_HERE, levels = c("NO", "YES"))
PRED   <- factor(PRED_HERE, levels = c("NO", "YES"))
POS_LABEL <- "YES"

cm <- table(Pred = PRED, Actual = ACTUAL)
TP <- cm[POS_LABEL, POS_LABEL]
TN <- cm["NO", "NO"]
FP <- cm[POS_LABEL, "NO"]
FN <- cm["NO", POS_LABEL]

metrics <- tibble(
  accuracy = (TP + TN) / sum(cm),
  sensitivity = TP / (TP + FN),
  specificity = TN / (TN + FP),
  precision = TP / (TP + FP),
  recall = TP / (TP + FN),
  f1 = 2 * (precision * recall) / (precision + recall),
  balanced_accuracy = (sensitivity + specificity) / 2
)
metrics
```

Interpretation template: Model performance is `[GOOD/WEAK]` on `[METRIC]`; trade-off shows `[HIGH_SENS_LOW_SPEC or vice versa]`.

## 5) ROC + AUC + Youden threshold chooser

Use when exam asks for threshold-independent evaluation and optimal operating point.

**How to adapt:**
- Replace `P_PROB` and `Y_TRUE`.
- If package unavailable, keep confusion-grid fallback.

```{r}
# Prospect: ROC/AUC + best threshold (Youden J)
library(pROC)

Y_TRUE <- as.integer(Y_TRUE_HERE) # 0/1
P_PROB <- PRED_PROB_HERE

roc_obj <- roc(response = Y_TRUE, predictor = P_PROB)
auc_val <- auc(roc_obj)

best_thr <- coords(roc_obj, x = "best", best.method = "youden", ret = "threshold")
plot(roc_obj)
auc_val
best_thr
```

Interpretation template: AUC is `[AUC]`, indicating `[DISCRIMINATION_QUALITY]`; Youden-optimal threshold is `[THR]`.

## 6) Cost-sensitive threshold search

Use when false negatives and false positives have different business costs.

**How to adapt:**
- Set `COST_FN` and `COST_FP` from exam text.
- Adjust threshold grid granularity.

```{r}
# Prospect: minimize expected classification cost
Y_TRUE <- as.integer(Y_TRUE_HERE)
P_PROB <- PRED_PROB_HERE
COST_FN <- 5
COST_FP <- 1
THRESH <- seq(0.05, 0.95, by = 0.01)

cost_tbl <- tibble(threshold = THRESH) %>%
  mutate(
    pred = map(threshold, ~ if_else(P_PROB >= .x, 1L, 0L)),
    fn = map_dbl(pred, ~ sum(.x == 0 & Y_TRUE == 1)),
    fp = map_dbl(pred, ~ sum(.x == 1 & Y_TRUE == 0)),
    total_cost = COST_FN * fn + COST_FP * fp
  ) %>%
  arrange(total_cost)

cost_tbl %>% select(threshold, fn, fp, total_cost) %>% head(10)
```

Interpretation template: With `Cost(FN)=[COST_FN]` and `Cost(FP)=[COST_FP]`, threshold `[THR]` minimizes expected total error cost.

## 7) Ridge (manual matrix form) + lambda comparison

Use when exam asks conceptual/explicit ridge implementation without high-level wrappers.

**How to adapt:**
- Change `LAMBDA` values and matrix objects.
- Keep centering if intercept should be unpenalized.

```{r}
# Prospect: manual ridge using matrix algebra
X <- as.matrix(X_HERE)
y <- as.numeric(Y_HERE)
LAMBDA <- 100

# Center X and y so intercept is handled separately
Xc <- scale(X, center = TRUE, scale = FALSE)
yc <- y - mean(y)

p <- ncol(Xc)
I_p <- diag(p)

beta_ridge <- solve(t(Xc) %*% Xc + LAMBDA * I_p, t(Xc) %*% yc)
beta_ridge
```

Interpretation template: Increasing lambda shrinks coefficient magnitudes; this lowers variance but may increase bias.

## 8) Regression tree pruning with CV

Use when exam asks “fit tree, then improve/prune and justify complexity”.

**How to adapt:**
- Set formula and response.
- Choose best tree size using CV deviance.

```{r}
# Prospect: regression tree + pruning
library(tree)

DF <- DATAFRAME_HERE
TREE_FORMULA <- as.formula("Y_VAR ~ .")

tree_fit <- tree(TREE_FORMULA, data = DF)
cv_fit <- cv.tree(tree_fit)

best_size <- cv_fit$size[which.min(cv_fit$dev)]
pruned <- prune.tree(tree_fit, best = best_size)

plot(pruned)
text(pruned, pretty = 0)
best_size
```

Interpretation template: CV suggests tree size `[SIZE]`; pruning improves generalization by removing weak splits.

## 9) Random forest tuning grid (mtry, ntree)

Use when exam asks “can RF be improved” instead of single default model.

**How to adapt:**
- Adjust `MTRY_GRID`, `NTREE_GRID`.
- Replace metric for classification/regression target.

```{r}
# Prospect: RF tuning mini-grid
library(randomForest)

TRAIN <- TRAIN_DF
TEST  <- TEST_DF
TARGET <- "Y_VAR"

MTRY_GRID <- c(2, 3, 4)
NTREE_GRID <- c(200, 500)

grid <- crossing(mtry = MTRY_GRID, ntree = NTREE_GRID)

rf_res <- grid %>%
  mutate(
    fit = map2(mtry, ntree, ~ randomForest(as.formula(paste(TARGET, "~ .")),
                                           data = TRAIN, mtry = .x, ntree = .y, importance = TRUE)),
    pred = map(fit, ~ predict(.x, newdata = TEST)),
    mse = map_dbl(pred, ~ mean((TEST[[TARGET]] - .x)^2))
  ) %>%
  arrange(mse)

rf_res %>% select(mtry, ntree, mse)
```

Interpretation template: Best RF setting is `mtry=[MTRY], ntree=[NTREE]` with test error `[ERR]`.

## 10) Interaction-term check in logistic/linear model

Use when exam asks whether effect of one predictor depends on another.

**How to adapt:**
- Replace interacting variables and family.
- Compare nested models (with vs without interaction).

```{r}
# Prospect: interaction test via nested models
DF <- DATAFRAME_HERE

mod_main <- glm(TARGET_VAR ~ X1 + X2 + X3, data = DF, family = binomial())
mod_int  <- glm(TARGET_VAR ~ X1 * X2 + X3, data = DF, family = binomial())

anova(mod_main, mod_int, test = "Chisq")
summary(mod_int)
```

Interpretation template: Interaction term `[X1:X2]` is `[SIGNIFICANT / NOT SIGNIFICANT]`, so dependency between effects is `[SUPPORTED / NOT SUPPORTED]`.

## 11) Stratified train/test split for imbalanced classification

Use when class imbalance makes random split unstable.

**How to adapt:**
- Set target factor and train fraction.
- Keep class proportions matched across splits.

```{r}
# Prospect: stratified split (tidyverse only)
set.seed(123)
DF <- DATAFRAME_HERE
TARGET <- "TARGET_VAR"
TRAIN_FRAC <- 0.8

train_idx <- DF %>%
  mutate(.row_id = row_number()) %>%
  group_by(.data[[TARGET]]) %>%
  slice_sample(prop = TRAIN_FRAC) %>%
  pull(.row_id)

train_df <- DF %>% slice(train_idx)
test_df  <- DF %>% slice(setdiff(seq_len(nrow(DF)), train_idx))

prop.table(table(train_df[[TARGET]]))
prop.table(table(test_df[[TARGET]]))
```

Interpretation template: Stratification preserves class mix in train/test, making metric comparison more reliable under imbalance.

## 12) Bootstrap CI for a proportion (binary rate)

Use when exam asks CI for churn/claim probability using bootstrap and normal comparison.

**How to adapt:**
- Replace binary vector and `B` iterations.
- Change confidence level via quantiles.

```{r}
# Prospect: bootstrap CI for p = P(Y=1)
set.seed(123)
Y_BIN <- as.integer(BINARY_VECTOR_HERE) # 0/1
B <- 2000

boot_p <- replicate(B, {
  yb <- sample(Y_BIN, size = length(Y_BIN), replace = TRUE)
  mean(yb)
})

ci_pct <- quantile(boot_p, probs = c(0.025, 0.975))
p_hat <- mean(Y_BIN)
se_hat <- sd(boot_p)
ci_norm <- p_hat + c(-1.96, 1.96) * se_hat

ci_pct
ci_norm
```

Interpretation template: Bootstrap percentile CI is `[L,U]`; compared with normal CI `[L2,U2]`, they are `[SIMILAR/DIFFERENT]`, indicating `[APPROX_NORMALITY / SKEWNESS]` in sampling distribution.

## 13) Confidence vs prediction interval helper (quick exam text)

Use when exam asks both interval types for one profile and conceptual distinction.

**How to adapt:**
- Build `new_obs` with exact model predictors.
- Change `level` to exam requirement.

```{r}
# Prospect: CI vs PI from OLS
fit <- lm(Y_VAR ~ X1 + X2 + X3, data = TRAIN_DF)
new_obs <- tibble(X1 = X1_VALUE, X2 = X2_VALUE, X3 = X3_VALUE)

ci <- predict(fit, newdata = new_obs, interval = "confidence", level = 0.95)
pi <- predict(fit, newdata = new_obs, interval = "prediction", level = 0.95)

ci
pi
```

Interpretation template: CI estimates the mean response for this profile, while PI estimates one new individual outcome and is therefore wider.

## 14) Operational-model filter (availability-at-prediction-time)

Use when exam asks to remove variables “not plausibly available” at prediction time.

**How to adapt:**
- Set leakage/late variables in `DROP_VARS`.
- Keep this before all modeling chunks.

```{r}
# Prospect: operational predictor filtering
DF <- DATAFRAME_HERE
DROP_VARS <- c("LEAKAGE_VAR1", "POST_EVENT_VAR2")

df_operational <- DF %>% select(-any_of(DROP_VARS))

names(df_operational)
```

Interpretation template: We removed `[DROP_VARS]` because they are unavailable or leakage-prone at scoring time, improving realism of model evaluation.

## 15) Model leaderboard with one unified table

Use when exam asks to compare many models and recommend one.

**How to adapt:**
- Fill with your computed metrics.
- Add business metric column if required.

```{r}
# Prospect: unified model comparison table
leaderboard <- tibble(
  model = c("OLS", "LASSO", "Tree", "RF", "Logistic", "ClassTree"),
  task = c("reg", "reg", "reg", "reg", "class", "class"),
  metric_primary = c(MSE_OLS, MSE_LASSO, MSE_TREE, MSE_RF, BALACC_LOGIT, BALACC_TREE)
) %>%
  arrange(desc(metric_primary))

leaderboard
```

Interpretation template: Based on `[PRIMARY_METRIC]`, `[BEST_MODEL]` is preferred; final choice should also consider interpretability and operational constraints.

# Notes on why these prospects were selected

These themes were repeatedly seen in:
- `/home/runner/work/BAN404/BAN404/Previous Exam/BAN404_exam_2021.md` (bootstrap + regression coefficient bootstrap)
- `/home/runner/work/BAN404/BAN404/Previous Exam/BAN404_exam_2022.md` (threshold CV, logistic vs tree)
- `/home/runner/work/BAN404/BAN404/Previous Exam/BAN404_exam_2024.md` (LOOCV, logistic/tree/RF comparisons)
- `/home/runner/work/BAN404/BAN404/Previous Exam/BAN404_exam_2025.md` (bootstrap variance CI, threshold tuning)
- `/home/runner/work/BAN404/BAN404/Previous Exam/R_exam_comp_2024.qmd`
- `/home/runner/work/BAN404/BAN404/Previous Exam/2025/R_exam_comp_2025.qmd`
- `/home/runner/work/BAN404/BAN404/Previous Exam/ExamQuestion_thematic_analysis/Mapping_PreviousQuestions.csv`
